### 2022-12-21 C3.tex
Add files via upload

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------

\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
%\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
%\usepackage{bm}
\usepackage{graphicx}
\usepackage{tabularx}
%\usepackage{calligra}
\usepackage{array}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
%\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\usepackage{color}
\usepackage{ulem}
\usepackage{url}
\usepackage{yfonts}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}
\newcommand{\SVI}[0]{$\bf{S^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\CIII}[0]{$\bf{C^{3}}$}
\newcommand{\DVII}[0]{$\bf{D^{7}}$}
\newcommand{\VVII}[0]{$\bf{V^{7}}$}

\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\Imaginary}[0]{$ \textfrak{I} $}
\newcommand{\Real}[0]{$ \textfrak{R} $}
\newcommand{\Exchange}[0]{$\textfrak{X}$}

\newcommand{\nounderline}[3]{\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineab}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineac}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
\newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}

\newcommand{\scalar}[1]{$#1$}

\newcommand{\scalarsub}[2]{$#1_#2$}

%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
\journalcode{A}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation
\makeatletter
\font\dummyft@=dummy \relax
\makeatother

\begin{document}                  % DO NOT DELETE THIS LINE

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).
	{\LARGE \emph{\today}} \\
	\title{Delone lattice studies in \CIII, the space of three complex variables}
	%\title{Note on the transformation of three-space basis vectors to  corresponding matrix for Delaunay scalars}
	\shorttitle{properties of C3}

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
	\author[b]{Herbert J.}{Bernstein}

	\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
	\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}

	% Use \shortauthor to indicate an abbreviated author list for use in
	% running heads (you will need to uncomment it).

	\shortauthor{Andrews and Bernstein}

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	\keyword{Delaunay}
	\keyword{Delone}
	\keyword{Selling}
	\keyword{\CIII}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	\maketitle                        % DO NOT DELETE THIS LINE

	\begin{synopsis}
		Selling reduction and Delone reduction are
		considered in a space of complex variables.
	\end{synopsis}
	\newcommand{\si}[0]{$s_1$}
	\newcommand{\sii}[0]{$s_2$}
	\newcommand{\siii}[0]{$s_3$}
	\newcommand{\siv}[0]{$s_4$}
	\newcommand{\sv}[0]{$s_5$}
	\newcommand{\svi}[0]{$s_6$}
	\newcommand{\Svec} [0] {\{\si, \sii, \siii, \siv, \sv, \svi \}}
	\newcommand{\SvecA} [0] {\{-\si, -\si+\sii, \si+\siii, \si+\sv, \si+\siv, \si+\svi \}}

	\newcommand{\OPES}[0]{$E^3toS^6$}
	\newcommand{\OPESS}[0]{$$E^3toS^6$$}
	\newcommand{\MSVI}[0]{$M_{S^{6}}$}
	\newcommand{\MEIII}[0]{$M_{E^{3}}$}
	\newcommand{\Plus}[0]{$\textfrak{P}$}	
	\newcommand{\Minus}[0]{$\textfrak{M}$}

	\newcommand{\ci}[0]{$c_1$}
	\newcommand{\cii}[0]{$c_2$}
	\newcommand{\ciii}[0]{$c_3$}

	\begin{abstract}

		The Delone scalars are studied in \CIII, the space of three complex variables.

		{\bf Note:}  In his later publications, Boris Delaunay used the Russian version of his surname, Delone.\\

	\end{abstract}

	%-------------------------------------------------------------------------
	% The main body of the paper
	%-------------------------------------------------------------------------
	% Now enter the text of the document in multiple \section's, \subsection's
	% and \subsubsection's as required.

	% Appendices appear after the main body of the text. They are prefixed by
	% a single \appendix declaration, and are then structured just like the
	% body text.

	\section{Introduction}

	The scalars used by \citeasnoun{Delaunay1932} in his formulation of Selling reduction \cite{Selling1874}
	are (in the conventional order) $b \cdot c$, $a \cdot c$, $a \cdot b$, $a \cdot d$, 
	$b \cdot d$, $c \cdot d$, where $d = -a-b-c$. (As a mnemonic device, observe that the first three terms use
	$\alpha$, $\beta$, and $\gamma$, in that order, and the following terms use $a$, $b$, $c$, in that order.)

	\citeasnoun{andrews2019b} chose to represent the scalars in the space \SVI{},
	\Svec{} (defined in the order above), as a way to create a metric space
	for the measurement of the distance between lattices. They also consider the representation of this space as the
	space of three complex dimensions, \CIII{} or {$\{c_1$}, {$c_2$}, {$c_3\}$}. 	
	In \CIII{}, in terms of the Delone scalars, a vector is \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}, where the real and imaginary parts
	of each are the ``opposite'' scalars according to the definition of \citeasnoun{Delaunay1932} (see \citeasnoun{andrews2019a}).
	As a memonic device, note that the complex components involve ($\alpha,a$), ($\beta, b$), and ($\gamma,c$).

		\citeasnoun{andrews2019b} considered the matrix representations of the reflections in \CIII{} (and in \SVI{}). This paper will describe the boundary transformations at the edges of the fundamental
	unit of \CIII{}. In \SVI{}, the fundamental unit is the all negative orthant, which contains only and
	all of the reduced cells. In \SVI{} and \CIII{}, the boundaries are found where any scalar (or correspondingly the real 
	or imaginary part) is equal to zero. The rationale for this work was that it might lend insights into the
	topology of the space of lattices.

	\section{Notation}

	Complex numbers will be represented in Cartesian format $(x,y)$, where $x$ is the real part and
	$y$ is the imaginary part.

	We will represent a vector in \CIII{} by \{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\} as an alternative to  \{(\si,\siv ), (\sii,\sv),(\sii,\svi)\}.

	Next we define the operators in \CIII{} that will be used in the matrix descriptions of 
	the transformations at the boundaries of the fundamental unit.\\

	\begin{table}	
		\begin{tabular}{|l|l|l|l|}
			\hline
			Operator&Usage&Result&Name \\ \hline
			$\Minus{}_r $&  $\Minus{}_r(c_j)$ & $(-x_j, -x_j+y_j)$ & Minus real \\ \hline
			$\Minus{}_i $ & $\Minus{}_i(c_j)$ & $(x_j-y_j,-y_j) $ & Minus imag\\ \hline
			$\textfrak{P}_r$ & $ \textfrak{P}_r(c_j)$ & $(x_j, x_j)$ & Plus real\\ \hline
			$\textfrak{P}_i$ &  $\textfrak{P}_i(c_j)$  & $(y_j, y_j)$ & Plus imag\\ \hline
			$\Re$ &  $\Re(c_j)$ & $x_j$ & Real \\ \hline
			$\Im$  &$ \Im(c_j)$ & $y_j$ & Imaginary \\ \hline

		\end{tabular}

	\end{table}

	\section{Matrices of boundary transformations}

	For the boundary at $s_1$: (the real component of $c_1$).

	$\begin{bmatrix}
	\Minus{}_r & 0 & 0 \\
	\Plus{}_r &  \textrm{i}\Re{} & \Re{} \\
	\textfrak{P}_r & \textrm{i}\Im{} & \Im{}
	\end{bmatrix}$
	%
	$\begin{bmatrix}
	\Minus{}_r & 0 & 0 \\
	\Plus{}_r &\textrm{i}\Im{} & \Im{} \\
	\Plus{}_r & \textrm{i}\Re{} & \Re{} \\
	\end{bmatrix}$ 
	%
	$\begin{bmatrix}
	\Minus{}_r & 0 & 0 \\
	\Plus{}_r &  \Re{} & \textrm{i}\Re{} \\
	\Plus{}_r & \Im{} &\textrm{i}\Im{}
	\end{bmatrix}	$
	%
	$\begin{bmatrix}
	\Minus{}_r & 0 & 0 \\
	\Plus{}_r & \Im{} &\textrm{i}\Im{} \\
	\Plus{}_r &  \Re{} &\textrm{i}\Re{} \\
	\end{bmatrix}$ \\

	For the boundary at $s_4$: (the imaginary component of $c_1$).

	$\begin{bmatrix}
	\Minus{}_i & 0 & 0 \\
	\Plus{}_i &  \textrm{i}\Re{} & \Re{} \\
	\Plus{}_i & \textrm{i}\Im{} & \Im{}
	\end{bmatrix}$
	%
	$\begin{bmatrix}
	\Minus{}_i & 0 & 0 \\
	\Plus{}_i & \textrm{i}\Im{} & \Im{} \\
	\Plus{}_i &  \textrm{i}\Re{} & \Re{} \\
	\end{bmatrix}$ 
	%
	$\begin{bmatrix}
	\Minus{}_i & 0 & 0 \\
	\Plus{}_i &  \Re{} & \textrm{i}\Re{} \\
	\Plus{}_i & \Im{} & \textrm{i}\Im{}
	\end{bmatrix}	$
	%
	$\begin{bmatrix}
	\Minus{}_i & 0 & 0 \\
	\Plus{}_i & \Im{} & \textrm{i}\Im{} \\
	\Plus{}_i &  \Re{}& \textrm{i}\Re{} \\
	\end{bmatrix}$ \\

	For the boundary at \sii{} (the real component of \cii{}):

	$\begin{bmatrix}

	\textrm{i}\Re{} & \Plus{}_r & \Re{} \\
	0 & \Minus{}_r & 0 \\
	\textrm{i}\Im{} & \Plus{}_r & \Im{}
\end{bmatrix}$
%
$\begin{bmatrix}
	\textrm{i}\Im{} & \Plus{}_r &  \Im{} \\
	0 & \Minus{}_r & 0 \\
	\textrm{i}\Re{} & \Plus{}_r & \Re{} \\
\end{bmatrix}$ 
%
$\begin{bmatrix}
	\Re{} &\Plus{}_r &   \textrm{i}\Re{} \\
	0 & \Minus{}_r &  0 \\
	\Im{} &\Plus{}_r & \textrm{i}\Im{}
\end{bmatrix}	$ 
%
$\begin{bmatrix}
	\Im{} & \Plus{}_r & \textrm{i}\Im{} \\
	0 &\Minus{}_r &  0 \\
	\Re{} & \Plus{}_r &  \textrm{i}\Re{} \\
\end{bmatrix}$ \\

For the boundary at $s_5$: (the imaginary component of \cii{}).

$\begin{bmatrix}
\textrm{i}\Re{} & \Plus{}_i & \Re{} \\
0 & \Minus{}_i & 0 \\
\textrm{i}\Im{} & \Plus{}_i & \Im{}
\end{bmatrix}$
%
$\begin{bmatrix}
\textrm{i}\Im{} & \Plus{}_i &  \Im{} \\
0 & \Minus{}_i & 0 \\
\textrm{i}\Re{} & \Plus{}_ri& \Re{} \\
\end{bmatrix}$ 
%
$\begin{bmatrix}
\Re{} &\Plus{}_i &   \textrm{i}\Re{} \\
0 & \Minus{}_i &  0 \\
\Im{} &\Plus{}_i & \textrm{i}\Im{}
\end{bmatrix}	$ 
%
$\begin{bmatrix}
\Im{} & \Plus{}_i & \textrm{i}\Im{} \\
0 &\Minus{}_i &  0 \\
\Re{} & \Plus{}_i &  \textrm{i}\Re{} \\
\end{bmatrix}$ \\

For the boundary at \siii{} (the real component of \ciii{}):

$\begin{bmatrix}

\textrm{i}\Re{} & \Re{} & \Plus{}_r \\
\textrm{i}\Im{} & \Im{}& \Plus{}_r  \\
0 & 0& \Minus{}_r  \\
\end{bmatrix}$
%
$\begin{bmatrix}
\textrm{i}\Im{} &  \Im{} & \Plus{}_r \\
\textrm{i}\Re{} & \Re{} & \Plus{}_r \\
0 & 0 & \Minus{}_r \\
\end{bmatrix}$ 
%
$\begin{bmatrix}
\Re{} &   \textrm{i}\Re{} &\Plus{}_r \\
\Im{} & \textrm{i}\Im{}&\Plus{}_r \\
0  &  0 & \Minus{}_r \\
\end{bmatrix}	$ 
%
$\begin{bmatrix}
\Im{} & \textrm{i}\Im{} & \Plus{}_r \\
\Re{} &  \textrm{i}\Re{} & \Plus{}_r \\
0 &  0 &\Minus{}_r \\
\end{bmatrix}$ \\

For the boundary at \svi{} (the imaginary component of \ciii{}):

$\begin{bmatrix}

\textrm{i}\Re{} & \Re{} & \Plus{}_i \\
\textrm{i}\Im{} & \Im{}& \Plus{}_i \\
0 & 0 & \Minus{}_i 
\end{bmatrix}$
%
$\begin{bmatrix}
\textrm{i}\Im{} &  \Im{} & \Plus{}_i \\
\textrm{i}\Re{} & \Re{} & \Plus{}_i \\
0 & 0 & \Minus{}_i \\
\end{bmatrix}$ 
%
$\begin{bmatrix}
\Re{} &   \textrm{i}\Re{}  &\Plus{}_i \\
\Im{} & \textrm{i}\Im{} &\Plus{}_i \\
0 &  0 & \Minus{}_i \\
\end{bmatrix}	$ 
%
$\begin{bmatrix}
\Im{} & \textrm{i}\Im{} & \Plus{}_i \\
\Re{} &  \textrm{i}\Re{} & \Plus{}_i \\
0 &  0 &\Minus{}_i 
\end{bmatrix}$ \\

\section{Basics}

The standard representation of the identity operation is

	$\bf{c'}$  = 
	$\begin{bmatrix}
	$1$ & $0$ &  $0$ \\
	$0$ & $1$ & $0$ \\
	$0$ & $0$ & $1$ \\
	\end{bmatrix}$ 
	$\bf{c}$ \\

The identity in \CIII{} can also be written:

	$\bf{c'}$  = 
	$\begin{bmatrix}
	$1$ & $0$ &  $0$ \\
	$0$ & \Re{} +\textrm{i}\Im{} & $0$ \\
	$0$ & $0$ & \Re{} +\textrm{i}\Im{} \\
	\end{bmatrix}$ 
	$\bf{c}$
\\

\citeasnoun{Delaunay1932} does not consider the boundary transformations 
in detail. However, he uses them to define the process of Selling reduction. 
For example in \SVI{}, he lists the following as one of the possible results 
for a transformation on \si{}.
\SvecA{}
The third matrix above (the third boundary transform for $s_1$) implements this operation and interchanges the
real part of \ciii{} and the imaginary part of \cii{}.
~\\
	$\begin{bmatrix}
\Minus{}_r & 0 & 0 \\
\Plus{}_r &  \Re{} & \textrm{i}\Re{} \\
\Plus{}_r & \Im{} &\textrm{i}\Im{}
\end{bmatrix}	$
\\

In terms of \CIII, Delone's alternate transformation for the \si{} boundary would
exchange the real of \cii{} with the imaginary part of \ciii{}. That is the fourth
matrix in the list for \si{}. The other two transformations for \si{} can 
be generated from the two we have just displayed by the "exchange operation" \cite{andrews2019b} applied to the second and third \CIII{} elements. Delone 
did not describe the latter two transformation, perhaps because even a
single transformation was adequate to implement reduction. He already listed
two.

\section{Summary}

The considerable regularity of the transformation matrices (above) highlights one
of the important aspects of Selling reduction as used by \citeasnoun{Delaunay1932}
when compared to Niggli reduction \cite{Niggli1928}. Where all the 
boundaries of the fundamental orthant in \CIII ~and \SVI{} are basically the same (and indeed are 
related by reflections), the boundaries formed in Niggli reduction
are of multiple types, and the fundamental unit of the space
representing Niggli reduced cells (\GVI{}, see \citeasnoun{Andrews2014}) is non-convex.

The matrices display one of the properties of \CIII{} that make it a
useful representation. For the scalar being transformed, it and its 
opposite scalar create a unique row in each matrix. Those rows contain
only the minus operator and zeros, highlighting the unique relationship
of that pair of scalars.

Several aspects of C3 are obvious from inspecting the matrices above. 
First, we see that each boundary has four possible 
transformations that
can be applied. Because the each of the transformations at boundaries are 
self-inverse, they are the same transformations that would 
be used in the process of cell (lattice) reduction. 
\citeasnoun{Delaunay1932} and  \citeasnoun{Delone1975} give 
only two choices. Presumably for simplicity they did not list
the transformations that used the "exchange" operator 
(see \cite{andrews2019b}).

\section{Availability of code}

The $C^{++}$ ~code for \CIII{} is available in github.com, in
\url{https://github.com/duck10/LatticeRepLib.git}.

%\appendix

%\section{blah blah blah -- Supplementary Material}
\ack{{\bf Acknowledgements}}

Careful copy-editing and corrections by Frances C. Bernstein are 
gratefully acknowledged.	Our thanks to Jean Jakoncic and Alexei Soares for 
helpful conversations and access to data and facilities at 
Brookhaven National Laboratory.

\ack{{\bf Funding information}}

Funding for this research was provided in part by:  
US Department of Energy Offices of Biological and 
Environmental Research and of Basic Energy Sciences 
(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
U.S. National Institutes of Health (grant No. P41RR012408; 
grant No. P41GM103473; grant No. P41GM111244; 
grant No. R01GM117126,
grant No. 1R21GM129570); Dectris, Ltd.

\bibliography{reduced}

\bibliographystyle{iucr}

%-------------------------------------------------------------------------
% TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
%-------------------------------------------------------------------------

% Simple tables should use the tabular environment according to this
% model

% Postscript figures can be included with multiple figure blocks

\end{document}                    % DO NOT DELETE THIS LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2022-12-22 C3/C3.tex
trying to use github for LaTeX
git testing
reduced.bib - remove duplicate reference C3.tex change formatting of matrices per https://github.com/dspinellis/latex-advice

--- 2022-12-21 C3.tex
@@
 \usepackage{tabularx}
+\usepackage{booktabs}
 %\usepackage{calligra}
@@
 \newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
+
+%xxxxxx just for testing github % 2
 
@@
 	are (in the conventional order) $b \cdot c$, $a \cdot c$, $a \cdot b$, $a \cdot d$, 
-	$b \cdot d$, $c \cdot d$, where $d = -a-b-c$. (As a mnemonic device, observe that the first three terms use
-	$\alpha$, $\beta$, and $\gamma$, in that order, and the following terms use $a$, $b$, $c$, in that order.)
+	$b \cdot d$, $c \cdot d$, where $d = -a-b-c$. 
+	(As a mnemonic device, 
+	observe that the first three terms use
+	$\alpha$, $\beta$, and $\gamma$, 
+	in that order, 
+	and the following terms use $a$, $b$, $c$, in that order.)
 
-	\citeasnoun{andrews2019b} chose to represent the scalars in the space \SVI{},
-	\Svec{} (defined in the order above), as a way to create a metric space
-	for the measurement of the distance between lattices. They also consider the representation of this space as the
-	space of three complex dimensions, \CIII{} or {$\{c_1$}, {$c_2$}, {$c_3\}$}. 	
-	In \CIII{}, in terms of the Delone scalars, a vector is \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}, where the real and imaginary parts
-	of each are the ``opposite'' scalars according to the definition of \citeasnoun{Delaunay1932} (see \citeasnoun{andrews2019a}).
-	As a memonic device, note that the complex components involve ($\alpha,a$), ($\beta, b$), and ($\gamma,c$).
+	\citeasnoun{andrews2019b} chose to 
+	represent the scalars in the space \SVI{},
+	\Svec{} (defined in the order above), 
+	as a way to create a metric space
+	for the measurement of the distance between lattices. T
+	hey also consider the representation of this space as the
+	space of three complex dimensions, \CIII{} or 
+	{$\{c_1$}, {$c_2$}, {$c_3\}$}.
 
-		\citeasnoun{andrews2019b} considered the matrix representations of the reflections in \CIII{} (and in \SVI{}). This paper will describe the boundary transformations at the edges of the fundamental
-	unit of \CIII{}. In \SVI{}, the fundamental unit is the all negative orthant, which contains only and
-	all of the reduced cells. In \SVI{} and \CIII{}, the boundaries are found where any scalar (or correspondingly the real 
-	or imaginary part) is equal to zero. The rationale for this work was that it might lend insights into the
-	topology of the space of lattices.
+	In \CIII{}, in terms of the Delone scalars, 
+	a vector is \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}, 
+	where the real and imaginary parts
+	of each are the ``opposite'' scalars 
+	according to the definition of \citeasnoun{Delaunay1932} (see \citeasnoun{andrews2019a}).
+	As a memonic device, 
+	note that the complex components involve ($\alpha,a$), ($\beta, b$), and ($\gamma,c$).
+
+		\citeasnoun{andrews2019b} considered the matrix representations of the reflections in \CIII{} (and in \SVI{}). 
+		This paper will describe the boundary transformations 
+		at the edges of the fundamental	unit of \CIII{}. 
+		In \SVI{}, the fundamental unit is the all negative orthant, 
+		which contains only and	all of the reduced cells. 
+		In \SVI{} and \CIII{}, 
+		the boundaries are found where any scalar 
+		(or correspondingly the real or imaginary part) 
+		is equal to zero. The rationale for this work was 
+		that it might lend insights into the topology of
+		 the space of lattices.
 
@@
 
-	Complex numbers will be represented in Cartesian format $(x,y)$, where $x$ is the real part and
+	Complex numbers will be represented 
+	in Cartesian format $(x,y)$, 
+	where $x$ is the real part and
 	$y$ is the imaginary part.
 
-	We will represent a vector in \CIII{} by \{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\} as an alternative to  \{(\si,\siv ), (\sii,\sv),(\sii,\svi)\}.
+	We will represent a vector 
+	in \CIII{} by \{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\} 
+	as an alternative to  \{(\si,\siv ), (\sii,\sv),(\sii,\svi)\}.
 
-	Next we define the operators in \CIII{} that will be used in the matrix descriptions of 
+	Next we define the operators in \CIII{} t
+	hat will be used in the matrix descriptions of 
 	the transformations at the boundaries of the fundamental unit.\\
@@
 	\begin{tabular}{|l|l|l|l|}
-			\hline
-			Operator&Usage&Result&Name \\ \hline
-			$\Minus{}_r $&  $\Minus{}_r(c_j)$ & $(-x_j, -x_j+y_j)$ & Minus real \\ \hline
-			$\Minus{}_i $ & $\Minus{}_i(c_j)$ & $(x_j-y_j,-y_j) $ & Minus imag\\ \hline
-			$\textfrak{P}_r$ & $ \textfrak{P}_r(c_j)$ & $(x_j, x_j)$ & Plus real\\ \hline
-			$\textfrak{P}_i$ &  $\textfrak{P}_i(c_j)$  & $(y_j, y_j)$ & Plus imag\\ \hline
-			$\Re$ &  $\Re(c_j)$ & $x_j$ & Real \\ \hline
-			$\Im$  &$ \Im(c_j)$ & $y_j$ & Imaginary \\ \hline
-
+\toprule
+	Operator
+						&Usage	
+													&Result	
+																			&Name\\
+\midrule
+	$\Minus{}_r $		&  $\Minus{}_r(c_j)$ 		& $(-x_j, -x_j+y_j)$ 	& Minus real \\ 
+	$\Minus{}_i $ 		& $\Minus{}_i(c_j)$ 		& $(x_j-y_j,-y_j) $ 	& Minus imag\\ 
+	$\textfrak{P}_r$ 	& $ \textfrak{P}_r(c_j)$ 	& $(x_j, x_j)$ 			& Plus real\\ 
+	$\textfrak{P}_i$ 	&  $\textfrak{P}_i(c_j)$  	& $(y_j, y_j)$ 			& Plus imag\\ 
+	$\Real$ 			&  $\Real(c_j)$ 			& $x_j$ 				& Real \\ 
+	$\Imaginary$ 		&  $\Imaginary(c_j)$ 		& $y_j$ 				& Imaginary\\
+\bottomrule
 	\end{tabular}
@@
 	0				& \Minus{}_i	& 0 \\
-\textrm{i}\Re{} & \Plus{}_ri& \Re{} \\
+	\textrm{i}\Re{}	& \Plus{}_ri	& \Re{} \\
 \end{bmatrix}$ 
@@
 	\textrm{i}\Re{}	& \Re{}	& \Plus{}_r \\
-\textrm{i}\Im{} & \Im{}& \Plus{}_r  \\
-0 & 0& \Minus{}_r  \\
+	\textrm{i}\Im{}	& \Im{}	& \Plus{}_r  \\
+	0				& 0		& \Minus{}_r  \\
 \end{bmatrix}$
@@
 	\Re{}	&   \textrm{i}\Re{}	&\Plus{}_r \\
-\Im{} & \textrm{i}\Im{}&\Plus{}_r \\
+	\Im{}	& \textrm{i}\Im{}	&\Plus{}_r \\
 	0 		&  0				& \Minus{}_r \\
@@
 	\textrm{i}\Re{}	& \Re{}	& \Plus{}_i \\
-\textrm{i}\Im{} & \Im{}& \Plus{}_i \\
+	\textrm{i}\Im{}	& \Im{}	& \Plus{}_i \\
 	0				& 0		& \Minus{}_i 
@@
 real part of \ciii{} and the imaginary part of \cii{}.
-~\\
+
+\begin{center}
 		$\begin{bmatrix}
@@
 \end{bmatrix}	$
-\\
+\end{center}
 

### 2022-12-22 C3/C3.tex
fixing matrix to use booktabs
playing with a n extra line in the table. Goes away if the table environment is removed. Goes away in preprint mode.

@@
 
-\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
+\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
 %\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
@@
 	\end{abstract}
-
-	%-------------------------------------------------------------------------
-	% The main body of the paper
-	%-------------------------------------------------------------------------
-	% Now enter the text of the document in multiple \section's, \subsection's
-	% and \subsubsection's as required.
-
 	% Appendices appear after the main body of the text. They are prefixed by
@@
 	the transformations at the boundaries of the fundamental unit.\\
+	See Table \ref{tab:operators}.
 
 \begin{table}	
-	\begin{tabular}{|l|l|l|l|}
+	\label{tab:operators}
+	\begin{tabular}{c l l l }
 \toprule
@@
 	$\Imaginary$ 		&  $\Imaginary(c_j)$ 		& $y_j$ 				& Imaginary\\
+
 \bottomrule
 	\end{tabular}		
-
 \end{table}

### 2022-12-22 C3/C3.tex
Just messing with the table
just formatting of the .tex file

@@
 
-	The scalars used by \citeasnoun{Delaunay1932} in his formulation of Selling reduction \cite{Selling1874}
+	The scalars used by~\citeasnoun{Delaunay1932} in his formulation of Selling reduction ~\cite{Selling1874}
 	are (in the conventional order) $b \cdot c$, $a \cdot c$, $a \cdot b$, $a \cdot d$, 
@@
 
-	\citeasnoun{andrews2019b} chose to 
+	~\citeasnoun{andrews2019b} chose to 
 	represent the scalars in the space \SVI{},
@@
 	of each are the ``opposite'' scalars 
-	according to the definition of \citeasnoun{Delaunay1932} (see \citeasnoun{andrews2019a}).
+	according to the definition of~\citeasnoun{Delaunay1932} (see~\citeasnoun{andrews2019a}).
 	As a memonic device, 
@@
 
-		\citeasnoun{andrews2019b} considered the matrix representations of the reflections in \CIII{} (and in \SVI{}). 
+	~\citeasnoun{andrews2019b} considered the matrix representations 
+	of the reflections in \CIII{} (and in \SVI{}). 
 	This paper will describe the boundary transformations 
@@
 	the transformations at the boundaries of the fundamental unit.\\
-	See Table \ref{tab:operators}.
+	See Table ~\ref{tab:operators}.
 
@@
 			\toprule
-	Operator
-						&Usage	
-													&Result	
-																			&Name\\
+			Operator			&Usage						&Result					&Name\\
 			\midrule
@@
 			$\Imaginary$ 		&  $\Imaginary(c_j)$ 		& $y_j$ 				& Imaginary\\
-
 			\bottomrule
@@
 
-\citeasnoun{Delaunay1932} does not consider the boundary transformations 
-in detail. However, he uses them to define the process of Selling reduction. 
-For example in \SVI{}, he lists the following as one of the possible results 
-for a transformation on \si{}.
-\SvecA{}
-The third matrix above (the third boundary transform for $s_1$) implements this operation and interchanges the
+	~\citeasnoun{Delaunay1932} does not 
+	consider the boundary transformations 
+	in detail. 
+	However, he uses them to 
+	define the process of Selling reduction. 
+	For example in \SVI{}, 
+	he lists the following as one of the possible results 
+	for a transformation on \si{}:
+	\SvecA{}.
+	The third matrix above 
+	(the third boundary transform for $s_1$) 
+	implements this operation and interchanges the
 	real part of \ciii{} and the imaginary part of \cii{}.
@@
 
-In terms of \CIII, Delone's alternate transformation for the \si{} boundary would
-exchange the real of \cii{} with the imaginary part of \ciii{}. That is the fourth
-matrix in the list for \si{}. The other two transformations for \si{} can 
-be generated from the two we have just displayed by the "exchange operation" \cite{andrews2019b} applied to the second and third \CIII{} elements. Delone 
-did not describe the latter two transformation, perhaps because even a
-single transformation was adequate to implement reduction. He already listed
-two.
+	If considered in \CIII, Delone's alternate transformation 
+	for the \si{} boundary would
+	exchange the real of \cii{} 
+	with the imaginary part of \ciii{}. That is the fourth
+	matrix in the list for \si{}. 
+	The other two transformations for \si{} can 
+	be generated from the two we have 
+	just displayed by the "exchange operation"~\cite{andrews2019b} 
+	applied to the second and third \CIII{} elements. Delone 
+	did not describe the latter two transformation, 
+	perhaps because even a single transformation 
+	was adequate to implement reduction. He already listed two.
 
@@
 
-The considerable regularity of the transformation matrices (above) highlights one
-of the important aspects of Selling reduction as used by \citeasnoun{Delaunay1932}
-when compared to Niggli reduction \cite{Niggli1928}. Where all the 
-boundaries of the fundamental orthant in \CIII ~and \SVI{} are basically the same (and indeed are 
-related by reflections), the boundaries formed in Niggli reduction
-are of multiple types, and the fundamental unit of the space
-representing Niggli reduced cells (\GVI{}, see \citeasnoun{Andrews2014}) is non-convex.
+	The considerable regularity 
+	of the transformation matrices (above) highlights one
+	of the important aspects of Selling reduction 
+	as used by~\citeasnoun{Delaunay1932}
+	when compared to Niggli reduction~\cite{Niggli1928}. 
+	Where all the 	boundaries of the fundamental orthant 
+	in \CIII ~and \SVI{} are basically the same 
+	(and indeed are related by reflections), 
+	the boundaries formed in Niggli reduction are of multiple types, 
+	and the fundamental unit of the space representing 
+	Niggli reduced cells (\GVI{}, see~\citeasnoun{Andrews2014}) is non-convex.
 
-The matrices display one of the properties of \CIII{} that make it a
-useful representation. For the scalar being transformed, it and its 
-opposite scalar create a unique row in each matrix. Those rows contain
-only the minus operator and zeros, highlighting the unique relationship
-of that pair of scalars.
+	The matrices display one of the properties of \CIII{} 
+	that make it a useful representation. 
+	For the scalar being transformed, it and its 
+	opposite scalar create a unique row in each matrix. 
+	Those rows contain
+	only the minus operator and zeros, 
+	highlighting the unique relationship of that pair of scalars.
 
@@
 	First, we see that each boundary has four possible 
-transformations that
-can be applied. Because the each of the transformations at boundaries are 
-self-inverse, they are the same transformations that would 
+	transformations that can be applied. 
+	Because the each of the transformations at boundaries are self-inverse, 
+	they are the same transformations that would 
 	be used in the process of cell (lattice) reduction. 
-\citeasnoun{Delaunay1932} and  \citeasnoun{Delone1975} give 
+	~\citeasnoun{Delaunay1932} and~\citeasnoun{Delone1975} give 
 	only two choices. Presumably for simplicity they did not list
 	the transformations that used the "exchange" operator 
-(see \cite{andrews2019b}).
+	(see~\cite{andrews2019b}).
 
@@
 	Careful copy-editing and corrections by Frances C. Bernstein are 
-gratefully acknowledged.	Our thanks to Jean Jakoncic and Alexei Soares for 
+	gratefully acknowledged.
+	Our thanks to Jean Jakoncic and Alexei Soares for 
 	helpful conversations and access to data and facilities at 

### 2023-01-06 ProgramSummary/ProgramSummary..tex
continuing work on command line program writeup

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------

%xxxxxx just for testing github % 2
%xxxxxx just for testing github % 3

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
%\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
%\usepackage{xtabular]
%\usepackage{bm}
\usepackage{longtable}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
%\usepackage{calligra}
\usepackage{array}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
%\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\usepackage{color}
\usepackage{ulem}
\usepackage{url}
\usepackage{yfonts}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}
\newcommand{\BIV}[0]{$\bf{B^{4}}$}
\newcommand{\SVI}[0]{$\bf{S^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\CIII}[0]{$\bf{C^{3}}$}
\newcommand{\DVII}[0]{$\bf{D^{7}}$}
\newcommand{\DXIII}[0]{$\bf{D^{13}}$}
\newcommand{\VVII}[0]{$\bf{V^{7}}$}

\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\Imaginary}[0]{$ \textfrak{I} $}
\newcommand{\Real}[0]{$ \textfrak{R} $}
\newcommand{\Exchange}[0]{$\textfrak{X}$}

\newcommand{\nounderline}[3]{\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineab}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineac}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
\newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}

\newcommand{\scalar}[1]{$#1$}

\newcommand{\scalarsub}[2]{$#1_#2$}

%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
%\journalcode{A}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation
\makeatletter
\font\dummyft@=dummy \relax
\makeatother

\begin{document}                  % DO NOT DELETE THIS LINE

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).
	{\LARGE \emph{\today}} \\
	\title{Command line programs from LatticeRepLib}
	\shorttitle{Command line programs}

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
	\author[b]{Herbert J.}{Bernstein}

	\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
	\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}

	% Use \shortauthor to indicate an abbreviated author list for use in
	% running heads (you will need to uncomment it).

	\shortauthor{Andrews and Bernstein}

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	%	\keyword{Delaunay}
	%	\keyword{Delone}
	%	\keyword{Selling}
	%	\keyword{\CIII}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}
	\maketitle                        % DO NOT DELETE THIS LINE

	\begin{synopsis}
		Selling reduction and Delone reduction are
		considered in a space of complex variables.
	\end{synopsis}
	\newcommand{\si}[0]{$s_1$}
	\newcommand{\sii}[0]{$s_2$}
	\newcommand{\siii}[0]{$s_3$}
	\newcommand{\siv}[0]{$s_4$}
	\newcommand{\sv}[0]{$s_5$}
	\newcommand{\svi}[0]{$s_6$}
	\newcommand{\Svec} [0] {\{\si, \sii, \siii, \siv, \sv, \svi \}}
	\newcommand{\SvecA} [0] {\{-\si, -\si+\sii, \si+\siii, \si+\sv, \si+\siv, \si+\svi \}}

	\newcommand{\OPES}[0]{$E^3toS^6$}
	\newcommand{\OPESS}[0]{$$E^3toS^6$$}
	\newcommand{\MSVI}[0]{$M_{S^{6}}$}
	\newcommand{\MEIII}[0]{$M_{E^{3}}$}
	\newcommand{\Plus}[0]{$\textfrak{P}$}	
	\newcommand{\Minus}[0]{$\textfrak{M}$}

	\newcommand{\ci}[0]{$c_1$}
	\newcommand{\cii}[0]{$c_2$}
	\newcommand{\ciii}[0]{$c_3$}

	\section{Introduction}

	In the study of various spaces for representing lattices, a
	number of software tools have been prepared. Some of those
	are available as simple commandline programs. Most of them
	have flexible input and output their processed results in 
	a form that another can use that for processing. A few
	are terminal programs that produce analysis. A small
	number take no input and generate files for other uses.

	% Appendices appear after the main body of the text. They are prefixed by
	% a single \appendix declaration, and are then structured just like the
	% body text.

	\section{Data Inputs:}

	\begin{table}
		\caption{	All these are case-insensitive. If a particular
			input lattice is invalid, it is rejected with a message.}
		\begin{tabular}{l p{.2\textwidth} l l l}
			\toprule

			\begin{tabular}{l l l l}
				\underline{\textbf{Vector Input:}}&
				\parbox[t]{0.6\textwidth}{\textbf{g} 
					(or \textbf{v} or \textbf{g6}) for \GVI{} vectors\\
					\textbf{d} (or \textbf{d7}) for \DVII{} vectors\\
					\textbf{s} (or s6) for \SVI{}, Delone/Selling scalars\\
					\textbf{C3} for \CIII{} input (without parentheses or commas, \\
					~~~“C” would be interpreted as a C-centered unit cell)}
			\end{tabular}	\\[3pt]

			\underline{\textbf{RANDOM:}} Random (valid) unit cell generated\\

			\textbf{Crystal Lattice Input:} 
			``\textbf{A}'', ``\textbf{B}'', ``\textbf{C}'', ``\textbf{P}'', ``\textbf{R}'', ``\textbf{F}'', ``\textbf``{I}''	\\
			~~~~~followed by three axis lengths and three angles (in degrees)\\
			\underline{\textbf{semicolon}}: 
			lines beginning with a semicolon are treated as comments\\
			\underline{\textbf{END:}} ends the data input section\\
			\bottomrule
		\end{tabular}

	\end{table}	
	\subsection{Examples of unit cell inputs}
	\noindent
	P 10 20 30 90 111 90\\
	G 100 400 900  0 -215.02 0\\
	S6   0 -107.51  0   7.51 -400 -792.49\\
	; this is a comment

	\section{Programs}

	\begin{table}
		\caption{Programs for convert lattice representations}
		\begin{tabular}{l l l p{.2\textwidth} p{.5\textwidth} l}
			\toprule
			Name&	in&	out&	param& Output	\\
			\midrule
			CmdToB4&	y&	\BIV{}&&		For each input, 
			it produces the 4 vectors as a,b,c,d as E3 vectors, 
			and also the lengths of each of those vectors.\\[.9pt]
			CmdToC3&	y&	\CIII{}&&		
			For each input, produces the \CIII{} representation in the form 
			(\#,\#) (\#,\#) (\#,\#), 
			which does not conform to the format for input to other programs.\\[.9pt]
			CmdToCell&	y&	a,b,c,$\alpha$,$\beta$,$\gamma$&&		
			Converts to the conventional unit cell representation.\\[.9pt]
			CmdToD13&	y&	\DXIII{}&&		
			Outputs the lengths of the 13 unique vectors describing the Dirichlet cell.\\[.9pt]
			CmdToDC&	y&	y		\\[.9pt]
			CmdToG6&	y&	\GVI{}&&		Converts the input to \GVI{}\\[.9pt]
			CmdToS6&	y&	\SVI{}&&		Converts the input to \SVI{}\\[.9pt]
			\bottomrule

		\end{tabular}
	\end{table}

		\begin{table}
		\caption{Summary of programs manipulating data}
		\begin{tabular}{l l l p{.2\textwidth} p{.5\textwidth} l}
			\toprule
			Name&	in&	out&	param& Output	\\
			\midrule
			CmdCmplx&	NA&	y&&Currently, this just outputs some 
			programmed examples.		\\[.9pt]
			CmdDelone&	y&	\SVI{}&&Converts input to\SVI{}		\\[.9pt]
			CmdGen&	NA&	\GVI{}&	ngen& - number of examples to generate (optional)\\
			{}&&&type - &lattice type to generate, many options (optional) 
			c,t,h,o,r,h,m,a, 
			cP, cF,cI, tP, tI, hP, hR, oP, oF, oI, oS, mP, mC, mS, aP, 
			numeric Niggli types - 1-44, 
			Delone types - C1,C3,C5,~\dots 
			If no parameters are present, one each of the 44 Niggli types. 
			If only a number is present, that is how many of each of the 44 Niggli types to generate.
			For usage examples, see Table \ref{CMDGENexamples}\\[.9pt]
					CmdLM&	y&	cy&&		Lattice Matching. 
			The first input cell is used as the reference. 
			Succeeding cells are matched 
			as well as possible to the reference cell.\\[.9pt]
			CmdNiggli&	y&	\GVI{}&&		Produces the Niggli reduced cell for each input\\[.9pt]
			CmdPath&	?&	? && Clearly not working the way it should		\\[.9pt]
			CmdPerturb&	y&	y&	n-number of perturbations, parts per thousand to perturb&
			n cells perturbed normal to the input \SVI{} vector
			If a lattice centering (including P) is input,
			the output is in the same centering. Otherwise, the output is in \SVI{}.	\\[.9pt]
			CmdS6Refl&	y&	\SVI{}&	&	For each input, 
			all 24 \SVI{} reflections are produced.\\[.9pt]
			CmdSella&	y&	y		\\[.9pt]
			CmdSort&	y&	y&	"C3" or "seq'&	\\[.9pt]
			CmdVolume&	y&	mod&&		Outputs the unit cell and the volume of each input cell.\\[.9pt]
			MultiMetricDists&	y& mod&"Y/y"(optional)&	Parameter Y/y for maxima style output		\\[.9pt]
			PlotC3&	y&	mod& 
			1,2,or 3 for which \CIII{} coordinate to plot. 
			"seq" or "C3"& Plot one of the 3 \CIII{} coordinates, 
			sorted by minimizing point to point distance 
			for \SVI{}{ reflections or 
				sorting the \CIII{} components to put them 
				in the asymmetric unit.}	\\[.9pt]
			Radial&	y&	as radial	&&	
			Radial computes the polar distances in Angstroms 
			from the first input cell. 
			That is {a,$\alpha$}, {b,$\beta$}, and {c,$\gamma$} 
			as coordinates in complex space.\\[.9pt]
			SELLA&	y&	mod&&	
			Outputs the distances from each of the Bravais lattice types 
			(in \svi{}).	\\[.9pt]
			SVD&	y&	mod&& Outputs the singular value decomposition vectors
			and eigenvalues for the input lattice. This is done in \SVI{}.		\\[.9pt]			
			\bottomrule

		\end{tabular}
	\end{table}

	\begin{itemize}
	\item{CmdGen}

CmdGen is a program for creating examples of various types of lattice types.

There are two optional input parameters. The first is the count
of how many samples of each type to generate. The second (if present) is
the type of lattices to generate.  There is no input data
other than the command line parameters.

For examples of using CmdGen, see table \ref{CMDGENexamples}

The is considerable flexibility in the input types. However, they
are case-sensitive.

\begin{itemize}
	\item Niggli types by number 1-44 (see Table \ref{NiggliFormsI}).
	\item Delone types by Delone's designation - 
	C1,C3,C5, T1, T2, T5, R1, R3,\\ 
	O1A, O2, O3, O4, O5, O1B, M1A, M2A, M3, M4, M1B, M2B, 
	A1, A2, A3, H4 (see Table \ref{s6table}).
	\item Crystal system - c,h,t,o,m,a
\end{itemize}

\item SVD

On input to a number of lattices, 
SVD calculates a Singular Value Decomposition.
The output consists of the eigenvalues and 
the six eigenvectors.
It is important to check that the form of
each cell is the same; for instance, monoclinic
cell should have the unique interaxial angle
in the same place (\textit{e.g.} $\beta$)

As an example, consider using SVD to compute the information
for one of the

\textbf{Command line:}\\
CmdGen 5 33 | CmdSort seq | SVD\\
; SVD\\
S6 from input cell   0.000 -31.196   0.000 -71.963 -109.653 -131.880\\
S6 from input cell   0.000 -23.060   0.000 -77.367 -110.398 -125.599\\
S6 from input cell   0.000 -28.898   0.000 -54.985 -106.336 -82.229\\
S6 from input cell   0.000 -56.133   0.000 -103.736 -177.343 -123.204\\
S6 from input cell   0.000 -10.111   0.000 -50.618 -91.57100 -98.102\\

\textbf{eigenvalue     0}         vector         1 0 0 0 0 0\\
eigenvalue 413.012       vector         0 -0.1728 00 -0.4007 -0.6616 -0.6097\\
eigenvalue 48.3869       vector         0 -0.4009 0 -0.1129 -0.5164 0.7482\\
eigenvalue 14.0785       vector         0 -0.7060 0 -0.4581 0.5344 -0.0786\\
eigenvalue 8.53435       vector         0 0.5576 0 -0.7853 0.1001 0.2494\\
\textbf{eigenvalue     0}         vector         0 0 -1 0 0 0

The two zero eigenvalues correspond to the invariant
interaxial angles ($\alpha$ and $\gamma$, in this case).

\end{itemize}

\vspace{2pt}

\begin{table}

		\label{CMDGENexamples}
		\caption{CmdGen examples}

		\begin{tabular}{l l l l}
			\midrule
			CmdGen & generates a single example of each of the 44 Niggli types\\[2pt]
			CmdGen 2 & generates two examples of each of the 44 Niggli types\\[2pt]
			CmdGen 4 aP & generates four examples of each of the two anorthic Niggli types\\[2pt]
			CmdGen 2 17& generates 2 examples of Niggli 17, which is mC\\
			{}&\textbf{Output}:\\
			{}&; Niggli lattice type requested\\
			{}&; lattice type = 17\\
			{}&G6    90.428    90.428   142.222    83.992    83.992    22.888   IT\# = 17  mC\\
			{}&G6    79.033    79.033   111.127   -61.421   -73.872   -22.772   IT\# = 17  mC\\
			CmdGen 2 C5& generate two examples of Delone type "C5", which is
			primitive cubic.\\
			{}&\textbf{Output}:\\
			{}&; Delone lattice type input\\
			{}&; lattice type = C5\\
			{}&G6   181.275   181.275   181.275     0.000     0.000     0.000   IT\# = C5  cP\\
			{}&G6    74.526    74.526    74.526     0.000     0.000     0.000   IT\# = C5  cP\\[2pt]

			CmdGen 1 h& Generate a single example of each of the hexagonal (and rhombohedral) Bravais lattice (per Niggli).\\
			{}&\textbf{Output}:\\
			{}&; Niggli lattice type input\\
			{}&; lattice type = 2\\
			{}&G6   119.828   119.828   119.828   -45.302   -45.302   -45.302   IT\# = 2  hR\\
			{}&;\\
			{}&; lattice type = 4\\
			{}&G6    86.837    86.837    86.837    19.813    19.813    19.813   IT\# = 4  hR\\
			{}&;\\
			{}&; lattice type = 9\\
			{}&G6     3.688     3.688   145.128     3.688     3.688     3.688   IT\# = 9  hR\\
			{}&;\\
			{}&; lattice type = 12\\
			{}&G6    55.273    55.273   136.854     0.000     0.000   -55.273   IT\# = 12  hP\\
			{}&;\\
			{}&; lattice type = 22\\
			{}&G6    73.578    73.578   149.339     0.000     0.000   -73.578   IT\# = 22  hP\\
			{}&;\\
			{}&; lattice type = 24\\
			{}&G6   101.390   125.771   125.771   -91.974   -67.593   -67.593   IT\# = 24  hR\\
			{}&;\\
			\bottomrule
		\end{tabular}	
	\end{table}

		\begin{table}
		\caption{The table of \citeasnoun{Delaunay1932} describing the 24
			Bravais types in \SVI{}. It has been redone removing the images of the 
			Dirichlet cells, the not-reduced cells, and adding the ''lattice character", which
			describes the linear manifold of each type. The crystal family types have been
			renamed to modern usage: Q changed to T for tetrahedral, K changed
			to C for cubic, and T changed to A for anorthic. Where Delone
			in some places included two types in one table cell, they have been
			split into two (for example: ''M1" becomes ''M1A" and ''M1B").

			Note that five types (O3, M3, M2B, A2, and A3) are not normal 
			crystallographic types. They are boundary types, and they have
			fewer free parameters than the generic type requires. For instance,
			O3 (character: rs0 rs0) has only two free parameters (r and s), whereas
			an ordinary orthorhombic type requires three variables.}
		\includegraphics[width=0.75\textwidth]{DeloneTypesGrid_3}\\
		\label{s6table}		
	\end{table}

\begin{table}
	\caption{Roof/Niggli symbol, International Tables (IT) lattice character, Bravais lattice type, unsorted ${\mathbf{DC}^{\mathbf{7}}}$ subspace, boundary polytope.
		Note that the variables r, s and t are non-negative, and u, v and w may be
		positive, negative or zero as constrained below.}
	\begin{center}
		\begin{tabular}{|c|c|c|c|c|}
			\hline
			{\bf Roof/}     &{\bf IT}&{\bf Bravais}&{\bf Unsorted ${\mathbf{DC}^{\mathbf{7}}}$}&{\bf Bound-}\\
			{\bf Niggli}&{\bf Lattice}&{\bf Lattice}&{\bf Subspace}&{\bf ary}\\
			{\bf Symbol}&{\bf Char}&{\bf Type}&                             & {\bf Polytope} \\
			\hline
			44A&3&${\bf cP}$&$(r,r,r,2r,2r,2r,3r)$&$12345\!=\!12\hat{3}\!=\!12\hat{4}\!=\!12\hat{5}$\\
			\hline
			44C&1&$cF$&$(r,r,r,r,r,r,2r)$&12679ACD\\
			\hline
			44B&5&$cI$&$(r,r,r,4 r/3,4 r/3, 4 r/3,r)$&$\text{12F2}^{\prime} \text{F}^{\prime} = 1\hat{2}\hat{\text{F}}$\\
			\hline
			45A&11&${\bf tP}$&$(r,r,t,r+t,r+t,2r,2r+t)$&$1345 = 1\hat{3} = 1\hat{4} = 1\hat{5}$\\
			45B&21&${\bf tP}$&$(r,s,s,2s,r+s,r+s,r+2s$&$2345 = 2\hat{3} = 2\hat{4} = 2\hat{5}$\\
			\hline
			45D&6&$tI$&$(r,r,r,r-w/2,r-w/2,2r+w,r),$&\\
			&&&$-r \leq w \leq 0$&$\text{12FF}^{\prime} = 12\hat{\text{F}}$\\
			45D&7&$tI$&$[r,r,r,2*r+u,r-u/2,r-u/2,r],$&\\
			&&&$-r \leq u \leq 0$&$\text{12F2}^{\prime} = 1{\hat{2}}\text{F}$\\
			45C&15&$tI$&$(r,r,t,t,t,2r,t)$&158BF\\
			45E&18&$tI$&$(r,s,s,-r/2+2s,s,s,-r/2+2s)$&$\text{2ADA}^{\prime}  = 2{\hat{A}}\text{D}$\\
			\hline
			48A&12&${\bf hP}$&$(r,r,t,r+t,r+t,r,r+t])$&134E\\
			48B&22&${\bf hP}$&$(r,s,s,s,r+s,r+s,r+s)$&2458\\
			\hline
			49C&2&$hR$&$(r,r,r,2r-u,2r-u,2r-u,3r-u),$&\\
			&&&$0 < u \leq r$&$121^\prime 2^\prime = \hat{1}\hat{2}$\\
			49D&4&$hR$&$(r,r,r,2r+u,2r+u,2r+u,3r+3u),$&\\
			&&&$-r \leq u \leq 0$&$121^\prime 2^\prime = \hat{1}\hat{2}$\\
			49B&9&$hR$&$(r,r,t,t,t,r,r+t)$&1679ACD\\
			49E&24&$hR$&$(r,s,s,s+r/3,s+r/3,s+r/3,s)$&$2F2^\prime \text{F}^\prime = \hat{2}\hat{\text{F}}$\\
			\hline
			50C&32&${\bf oP}$&$(r,s,t,s+t,r+t,r+s,r+s+t)$&$345 = \hat{3} = \hat{4} = \hat{5}$\\
			\hline
			50D&13&$oC$&$(r,r,t,r+t,r+t,2r+w,2r+t+w),$&\\
			&&&$-r \leq w \leq 0$&134\\
			50E&23&$oC$&$(r,s,s,u+2*s,s+r,s+r,u+2*s+r),$&\\
			&&&$-s \leq u \leq 0$&245\\
			50A&36&$oC$&$(r,s,t,s+t,t,r+s,s+t)$&35B\\
			50B&38&$oC$&$(r,s,t,s+t,r+t,s,s+t)$&34E\\
			50F&40&$oC$&$[r,s,t,t,r+t,r+s,r+t)$&458\\
			\hline
			51A&16&$oF$&$(r,r,s,r+s+u,r+s+u,-2u,s),$&\\
			&&&$-r \leq u \leq 0$&$\text{1F1}^{\prime} = \hat{1}\text{F}$\\
			51B&26&$oF$&$(r,s,t,-r/2+s+t,t,s,-r/2+s+t)$&$\text{ADA}^{\prime}  = \hat{A}\text{D}$\\
			\hline
			52A&8&$oI$&$(r,r,r,2r+u,2r+v,-u-v,r),$&\\
			&&&$-r \leq u \leq 0, -r \leq v \leq 0$&12F\\
			52B&19&$oI$&$ (r,s,s,2s-u,s,s,-r+2s+u),$&\\
			&&&$0 < u \leq r $&\\
			&&&$ (r,s,s,2s-u,s,s,r+2s-u),$&\\
			&&&$r < u \leq s $&29C = 2AD\\
			52C&42&$oI$&$r,s,t,t,t,r+s,t$&58BF\\
			\hline
		\end{tabular}
	\end{center}
	\label{NiggliFormsI}
\end{table}%

	\section{Availability of code}

	The $C^{++}$ ~code for C++ is available in github.com, in
	\url{https://github.com/duck10/LatticeRepLib.git}.

	\ack{{\bf Acknowledgements}}

	\ack{{\bf Funding information}}

	Funding for this research was provided in part by:  
	US Department of Energy Offices of Biological and 
	Environmental Research and of Basic Energy Sciences 
	(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
	U.S. National Institutes of Health (grant No. P41RR012408; 
	grant No. P41GM103473; grant No. P41GM111244; 
	grant No. R01GM117126,
	grant No. 1R21GM129570); Dectris, Ltd.

	%	\bibliography{reduced}
	%	
	%	\bibliographystyle{iucr}

	%-------------------------------------------------------------------------
	% TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
	%-------------------------------------------------------------------------

	% Simple tables should use the tabular environment according to this
	% model

	% Postscript figures can be included with multiple figure blocks

\end{document}                    % DO NOT DELETE THIS LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-01-09 C3/C3.tex
testing and update .gitignore
test
just testing
continuing work on command line program writeup
improving formatting mostl

@@
 
-\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
+%xxxxxx just for testing github % 2
+%xxxxxx just for testing github % 3
+
+\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
 %\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
@@
 \newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
-
-%xxxxxx just for testing github % 2
 
@@
 	according to the definition of~\citeasnoun{Delaunay1932} (see~\citeasnoun{andrews2019a}).
-	As a memonic device, 
+	As a mnemonic device, 
 	note that the complex components involve ($\alpha,a$), ($\beta, b$), and ($\gamma,c$).

### 2023-01-09 ProgramSummary/ProgramSummary..tex
improving formatting mostl

@@
 \usepackage{array}
+ \usepackage{setspace}
  \DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
@@
 
+\DeclareMathOperator{\mr}{\textfrak{P}_i}
 \begin{document}                  % DO NOT DELETE THIS LINE
+	\singlespacing
+	$\mr$
 
@@
 	% for use in running heads (you will need to uncomment it).
-	{\LARGE \emph{\today}} \\
+	{\Large PDF generated: \emph{\today}} \\
 	\title{Command line programs from LatticeRepLib}
@@
 	\newcommand{\ciii}[0]{$c_3$}
+	\begin{abstract}
+	\end{abstract}
 
@@
 	\section{Data Inputs:}
+	\subsection{Table of input types}
+
+	In general, there are 5 types of input lines, see Table \ref{inputtypes}.
+	Except for "END", they can be combined in any order.
 
 	\begin{table}
+		\label{inputtypes}
 		\caption{	All these are case-insensitive. If a particular
@@
 
-			\textbf{Crystal Lattice Input:} 
+			\underline{\textbf{Crystal lattice input:}} 
 			``\textbf{A}'', ``\textbf{B}'', ``\textbf{C}'', ``\textbf{P}'', ``\textbf{R}'', ``\textbf{F}'', ``\textbf``{I}''	\\
@@
 
+	\subsection{Filters - change lattice representation}
+
 	\begin{table}
-		\caption{Programs for convert lattice representations}
-		\begin{tabular}{l l l p{.2\textwidth} p{.5\textwidth} l}
+		\renewcommand{\arraystretch}{1.2}
+		\caption{Programs to convert lattice representations}
+		\begin{tabular}{l l l p{.7\textwidth}  l}
 			\toprule
-			Name&	in&	out&	param& Output	\\
+			Name		&	in	&	out				& Output	\\
 			\midrule
-			CmdToB4&	y&	\BIV{}&&		For each input, 
+			CmdToB4		&	y	&	\BIV{}		&		For each input, 
 			it produces the 4 vectors as a,b,c,d as E3 vectors, 
 			and also the lengths of each of those vectors.\\[.9pt]
-			CmdToC3&	y&	\CIII{}&&		
+			CmdToC	3	&	y	&	\CIII{	}	&		
 			For each input, produces the \CIII{} representation in the form 
@@
 			which does not conform to the format for input to other programs.\\[.9pt]
-			CmdToCell&	y&	a,b,c,$\alpha$,$\beta$,$\gamma$&&		
+			CmdToCell	&	y	&	
+			\parbox[t]{0.07\textwidth}{${	a,b,c,}$ \\ ${\alpha,\beta,\gamma}$}		&		
 			Converts to the conventional unit cell representation.\\[.9pt]
-			CmdToD13&	y&	\DXIII{}&&		
-			Outputs the lengths of the 13 unique vectors describing the Dirichlet cell.\\[.9pt]
-			CmdToDC&	y&	y		\\[.9pt]
-			CmdToG6&	y&	\GVI{}&&		Converts the input to \GVI{}\\[.9pt]
-			CmdToS6&	y&	\SVI{}&&		Converts the input to \SVI{}\\[.9pt]
+			CmdToDC		&	y	&	\DXIII{}		&	Outputs the lengths of the 13 unique vectors describing the Dirichlet cell.\\[.9pt]			
+			CmdToG6		&	y	&	\GVI{}		&		Converts the input to \GVI{}\\[.9pt]
+			CmdToS6		&	y	&	\SVI{}		&		Converts the input to \SVI{}\\[.9pt]
+			Radial		&	y	&	Polar	&
+				CSomputes the polar distances in Angstroms 
+				from the first input cell. 
+				That is {(a,$\alpha$)}, {(b,$\beta$)}, and {(c,$\gamma$)} 
+				as coordinates in complex space.\\[.9pt]
 			\bottomrule
@@
 
-		\begin{table}
+\subsection{Data processing programs}
+
+		\begin{longtable}{l l l p{.2\textwidth} p{.5\textwidth} l}
 			\caption{Summary of programs manipulating data}
-		\begin{tabular}{l l l p{.2\textwidth} p{.5\textwidth} l}
+			\label{processingprogs}\\ %NOTE: that lineskip is necessary to prevent toprule from crashing
 			\toprule
-			Name&	in&	out&	param& Output	\\
+			Name		&	in	&	out		&	\begin{parbox}[t]{2cm}{
+				command line\\
+				param{s}}
+			\end{parbox}	& Output	\\
 			\midrule
-			CmdCmplx&	NA&	y&&Currently, this just outputs some 
-			programmed examples.		\\[.9pt]
-			CmdDelone&	y&	\SVI{}&&Converts input to\SVI{}		\\[.9pt]
-			CmdGen&	NA&	\GVI{}&	ngen& - number of examples to generate (optional)\\
-			{}&&&type - &lattice type to generate, many options (optional) 
+			CmdCmplx	&	NA	&	y		&\hrulefill	&Currently, 
+				this just outputs some programmed examples.		\\[.9pt]
+			CmdDelone	&	y	&	\SVI{}	&\hrulefill	&Converts input to \SVI{}		\\[.9pt]
+			CmdGen		&	NA	&	\GVI{}	&	ngen - number of examples to generate (optional)\\
+				{}		&		&			&type		& - lattice type to 
+				generate, many options (optional) 
 				c,t,h,o,r,h,m,a, 
@@
 				For usage examples, see Table \ref{CMDGENexamples}\\[.9pt]
-					CmdLM&	y&	cy&&		Lattice Matching. 
+			CmdLM		&	y	&	y		&	\hrulefill		&		Lattice Matching. 
 				The first input cell is used as the reference. 
@@
 				as well as possible to the reference cell.\\[.9pt]
-			CmdNiggli&	y&	\GVI{}&&		Produces the Niggli reduced cell for each input\\[.9pt]
-			CmdPath&	?&	? && Clearly not working the way it should		\\[.9pt]
-			CmdPerturb&	y&	y&	n-number of perturbations, parts per thousand to perturb&
+			CmdNiggli	&	y	&	\GVI{}	&\hrulefill	&		
+				Produces the Niggli reduced cell for 	each input\\[.9pt]
+			CmdPath		&	?	& ?			&\hrulefill	& Clearly not working the way it should		\\[.9pt]
+			CmdPerturb	&	y	& y			&	n-number of perturbations,
+
+				parts per thousand to perturb			&
 				n cells perturbed normal to the input \SVI{} vector
@@
 				the output is in the same centering. Otherwise, the output is in \SVI{}.	\\[.9pt]
-			CmdS6Refl&	y&	\SVI{}&	&	For each input, 
+			CmdS6Refl	&	y	&	\SVI{}	&\hrulefill	&	For each input, 
 				all 24 \SVI{} reflections are produced.\\[.9pt]
-			CmdSella&	y&	y		\\[.9pt]
-			CmdSort&	y&	y&	"C3" or "seq'&	\\[.9pt]
-			CmdVolume&	y&	mod&&		Outputs the unit cell and the volume of each input cell.\\[.9pt]
-			MultiMetricDists&	y& mod&"Y/y"(optional)&	Parameter Y/y for maxima style output		\\[.9pt]
-			PlotC3&	y&	mod& 
-			1,2,or 3 for which \CIII{} coordinate to plot. 
-			"seq" or "C3"& Plot one of the 3 \CIII{} coordinates, 
-			sorted by minimizing point to point distance 
-			for \SVI{}{ reflections or 
-				sorting the \CIII{} components to put them 
-				in the asymmetric unit.}	\\[.9pt]
-			Radial&	y&	as radial	&&	
-			Radial computes the polar distances in Angstroms 
-			from the first input cell. 
-			That is {a,$\alpha$}, {b,$\beta$}, and {c,$\gamma$} 
-			as coordinates in complex space.\\[.9pt]
-			SELLA&	y&	mod&&	
+			CmdSella	&	y	&	y		&\hrulefill	&
+				SELLA produces 2 output: std::out contains the match of
+				each of the 24 Delone types; SVG for the Grimmer
+				diagram is written to a file with a name such as 
+					\verb#SEL__2023-01-07.08_56_42.svg# with date/time stamp.\\
+			CmdSort		&	y	&	y		&"C3" or "seq'&	\\[.9pt]
+			CmdVolume	&	y	&	mod		&\hrulefill	&		Outputs the unit cell 
+				and the volume of each input cell.\\[.9pt]
+			MultiMetricDists&y	& mod		&"Y/y"(optional)&	
+				Parameter Y/y for maxima style output		\\[.9pt]
+			PlotC3		&	y	&	mod		& 			& Plot 
+				the three \CIII{} coordinates\\[.9pt]
+			SELLA		&	y	&	mod		&\hrulefill	&	
 				Outputs the distances from each of the Bravais lattice types 
 				(in \svi{}).	\\[.9pt]
-			SVD&	y&	mod&& Outputs the singular value decomposition vectors
+			SVD			&	y	&	mod		&\hrulefill	& Outputs the singular value decomposition vectors
 				and eigenvalues for the input lattice. This is done in \SVI{}.		\\[.9pt]			
@@
 
-		\end{tabular}
-	\end{table}
+		\end{longtable}
+%	\end{table}
 
@@
 	\item{CmdGen}
+
+	\subsection{Individual program details}
 
@@
 	\item Niggli types by number 1-44 (see Table \ref{NiggliFormsI}).
+\begin{singlespace}
 		\item Delone types by Delone's designation - 
@@
 	A1, A2, A3, H4 (see Table \ref{s6table}).
+\end{singlespace}
 	\item Crystal system - c,h,t,o,m,a
-\end{itemize}
-
-\item SVD
-
-On input to a number of lattices, 
-SVD calculates a Singular Value Decomposition.
-The output consists of the eigenvalues and 
-the six eigenvectors.
-It is important to check that the form of
-each cell is the same; for instance, monoclinic
-cell should have the unique interaxial angle
-in the same place (\textit{e.g.} $\beta$)
-
-As an example, consider using SVD to compute the information
-for one of the
-
-\textbf{Command line:}\\
-CmdGen 5 33 | CmdSort seq | SVD\\
-; SVD\\
-S6 from input cell   0.000 -31.196   0.000 -71.963 -109.653 -131.880\\
-S6 from input cell   0.000 -23.060   0.000 -77.367 -110.398 -125.599\\
-S6 from input cell   0.000 -28.898   0.000 -54.985 -106.336 -82.229\\
-S6 from input cell   0.000 -56.133   0.000 -103.736 -177.343 -123.204\\
-S6 from input cell   0.000 -10.111   0.000 -50.618 -91.57100 -98.102\\
-
-\textbf{eigenvalue     0}         vector         1 0 0 0 0 0\\
-eigenvalue 413.012       vector         0 -0.1728 00 -0.4007 -0.6616 -0.6097\\
-eigenvalue 48.3869       vector         0 -0.4009 0 -0.1129 -0.5164 0.7482\\
-eigenvalue 14.0785       vector         0 -0.7060 0 -0.4581 0.5344 -0.0786\\
-eigenvalue 8.53435       vector         0 0.5576 0 -0.7853 0.1001 0.2494\\
-\textbf{eigenvalue     0}         vector         0 0 -1 0 0 0
-
-The two zero eigenvalues correspond to the invariant
-interaxial angles ($\alpha$ and $\gamma$, in this case).
-
 \end{itemize}
@@
 	\end{table}
+
+	\item SVD
+
+	On input to a number of lattices, 
+	SVD calculates a Singular Value Decomposition.
+	The output consists of the eigenvalues and 
+	the six eigenvectors.
+	It is important to check that the form of
+	each cell is the same; for instance, monoclinic
+	cell should have the unique interaxial angle
+	in the same place (\textit{e.g.} $\beta$)\\
+
+	As an example, consider using SVD to compute the information
+	for one of the
+
+	\begin{samepage}
+		{\nopagebreak{
+				\textbf{Command line:}\\
+				CmdGen 5 33 | CmdSort seq | SVD\\
+				; SVD\\
+				S6 from input cell   0.000 -31.196   0.000 -71.963 -109.653 -131.880\\
+				S6 from input cell   0.000 -23.060   0.000 -77.367 -110.398 -125.599\\
+				S6 from input cell   0.000 -28.898   0.000 -54.985 -106.336 -82.229\\
+				S6 from input cell   0.000 -56.133   0.000 -103.736 -177.343 -123.204\\
+				S6 from input cell   0.000 -10.111   0.000 -50.618 -91.57100 -98.102\\
+			}
+		}
+	\end{samepage}
+
+	\begin{samepage}
+		\nopagebreak{
+			\textbf{Output}:\\
+			\textbf{eigenvalue     0};         vector         1 0 0 0 0 0\\
+			eigenvalue 413.012;       vector         0 -0.1728 00 -0.4007 -0.6616 -0.6097\\
+			eigenvalue 48.3869;       vector         0 -0.4009 0 -0.1129 -0.5164 0.7482\\
+			eigenvalue 14.0785;       vector         0 -0.7060 0 -0.4581 0.5344 -0.0786\\
+			eigenvalue 8.53435;       vector         0 0.5576 0 -0.7853 0.1001 0.2494\\
+			\textbf{eigenvalue     0};         vector         0 0 -1 0 0 0
+		}
+	\end{samepage}
+
+	The two zero eigenvalues correspond to the invariant
+	interaxial angles ($\alpha$ and $\gamma$, in this case).
+
+\end{itemize}
 
@@
 
-	The $C^{++}$ ~code for C++ is available in github.com, in
+	The $C^{++}$ ~code is available in github.com, in
 	\url{https://github.com/duck10/LatticeRepLib.git}.

### 2023-01-10 ProgramSummary/ProgramSummary..tex
changes for HJB, mostly. And a couple of errors fixed

@@
 \newcommand{\DVII}[0]{$\bf{D^{7}}$}
-\newcommand{\DXIII}[0]{$\bf{D^{13}}$}
+\newcommand{\DXIII}[0]{$\bf{DC^{13}}$}
 \newcommand{\VVII}[0]{$\bf{V^{7}}$}
@@
 				\parbox[t]{0.6\textwidth}{\textbf{g} 
-					(or \textbf{v} or \textbf{g6}) for \GVI{} vectors\\
-					\textbf{d} (or \textbf{d7}) for \DVII{} vectors\\
-					\textbf{s} (or s6) for \SVI{}, Delone/Selling scalars\\
+					 for \GVI{} vectors\\
+					\textbf{s} for \SVI{}, Delone/Selling scalars\\
 					\textbf{C3} for \CIII{} input (without parentheses or commas, \\
-					~~~“C” would be interpreted as a C-centered unit cell)}
+					~~~“C” would be interpreted as a C-centered unit cell)\\
+					\textbf{U} for DC7 unsorted}\\
 			\end{tabular}	\\[3pt]
@@
 			\underline{\textbf{Crystal lattice input:}} 
-			``\textbf{A}'', ``\textbf{B}'', ``\textbf{C}'', ``\textbf{P}'', ``\textbf{R}'', ``\textbf{F}'', ``\textbf``{I}''	\\
+			``\textbf{A}'', ``\textbf{B}'', ``\textbf{C}'', ``\textbf{P}'', ``\textbf{R}'', ``\textbf{H}'',``\textbf{F}'', ``\textbf{I}''	\\
 			~~~~~followed by three axis lengths and three angles (in degrees)\\
@@
 				Parameter Y/y for maxima style output		\\[.9pt]
-			PlotC3		&	y	&	mod		& 			& Plot 
-				the three \CIII{} coordinates\\[.9pt]
+			PlotC3		&	y	&	mod		& \hrulefill	& Plot 
+				the three \CIII{} coordinates. See Figure~\ref{plotc3}.\\[.9pt]
 			SELLA		&	y	&	mod		&\hrulefill	&	
@@
 
+	\begin{figure}
+		\includegraphics[width=\textwidth]{PlotC3}
+		\label{plotc3}
+		\caption{Output of PlotC3}
+	\end{figure}
+
 	\section{Availability of code}

### 2023-01-12 ProgramSummary/ProgramSummary..tex
Changed the data for CmdPath to match the current code. The earlier code  was to do PointDistanceFolloer type options, which are not needed. THis program needs more work.

@@
 				Produces the Niggli reduced cell for 	each input\\[.9pt]
-			CmdPath		&	?	& ?			&\hrulefill	& Clearly not working the way it should		\\[.9pt]
+			CmdPath		&	y	& \SVI{}			&
+			no. of points	& If the number of points is 
+			not used on the command line, the default number to generate is 20.		\\[.9pt]
 			CmdPerturb	&	y	& y			&	n-number of perturbations,

### 2023-01-12 ProgramSummary/ProgramSummary..tex
updating CmdPath writeup

@@
 			no. of points	& If the number of points is 
-			not used on the command line, the default number to generate is 20.		\\[.9pt]
+			not on the command line, the default number 
+			to generate is 20.	If there is only one point, 
+			then the second point is generated as the 
+			Niggli reduced cell. If more than two points are input, 
+			each successive pair will
+			generate a list of output points; so if
+			3 cells are input, the output will be 
+			2 times the number of points requested.	\\[.9pt]
 			CmdPerturb	&	y	& y			&	n-number of perturbations,

### 2023-02-05 ProgramSummary/ProgramSummary..tex
Descriptions of the command line programs for lattice studies

@@
 	\singlespacing
-	$\mr$
 
@@
 	\newcommand{\ciii}[0]{$c_3$}
-	\begin{abstract}
-	\end{abstract}
+%	\begin{abstract}
+%	\end{abstract}
 
@@
 		\renewcommand{\arraystretch}{1.2}
-		\caption{Programs to convert lattice representations}
+		\caption{Programs to convert lattice representations.
+		NOTE: although all can take all of the input types, \BIV{}. \DXIII{}, 
+	Polar write output that cannot be used as input.}
 		\begin{tabular}{l l l p{.7\textwidth}  l}
@@
 			CmdToCell	&	y	&	
-			\parbox[t]{0.07\textwidth}{${	a,b,c,}$ \\ ${\alpha,\beta,\gamma}$}		&		
-			Converts to the conventional unit cell representation.\\[.9pt]
+			\parbox[t]{0.07\textwidth}{${	a,b,c,}$ \\ ${\alpha,\beta,\gamma}$}	&					Converts to the conventional unit cell representation
+			If the input is already unit cell parameters, then the output
+			will be in the same lattice centering as the input.\\[.9pt]
 			CmdToDC		&	y	&	\DXIII{}		&	Outputs the lengths of the 13 unique vectors describing the Dirichlet cell.\\[.9pt]			
@@
 				param{s}}
-			\end{parbox}	& Output	\\
+			\end{parbox}	& Output	\\[.9pt]
 			\midrule
@@
 			CmdDelone	&	y	&	\SVI{}	&\hrulefill	&Converts input to \SVI{}		\\[.9pt]
-			CmdGen		&	NA	&	\GVI{}	&	ngen - number of examples to generate (optional)\\
+			CmdDists	&	y	& mod		&\hrulefill	&For on input list of n
+			cells, n-1 distances will be output.\\[.9pt]
+			CmdGen		&	NA	&	\GVI{}	&	ngen	&- number of examples to generate (optional)\\
 				{}		&		&			&type		& - lattice type to 
-				generate, many options (optional) 
+				generate (optional); many options 
 				c,t,h,o,r,h,m,a, 
@@
 			CmdNiggli	&	y	&	\GVI{}	&\hrulefill	&		
-				Produces the Niggli reduced cell for 	each input\\[.9pt]
+				Produces the Niggli reduced cell for 	each input.\\[.9pt]
 			CmdPath		&	y	& \SVI{}			&
@@
 					\verb#SEL__2023-01-07.08_56_42.svg# with date/time stamp.\\
-			CmdSort		&	y	&	y		&"C3" or "seq'&	\\[.9pt]
+			CmdSort		&	y	&	y		&"C3" or "seq'&	CmdSort has 2 possible actions. If "seq" is specified
+			as the sort type, then in the incoming list of 
+			lattices, in each pair, the second one is multipled by the 24 \SVI{} reflections, and the one that
+			is closest to the first is chosen for output.\\[.9pt]
 			CmdVolume	&	y	&	mod		&\hrulefill	&		Outputs the unit cell 
 				and the volume of each input cell.\\[.9pt]
-			MultiMetricDists&y	& mod		&"Y/y"(optional)&	
-				Parameter Y/y for maxima style output		\\[.9pt]
+%			MultiMetricDists&y	& mod		&"Y/y"(optional)&	
+%				Parameter Y/y for maxima style output		\\[.9pt]
 			PlotC3		&	y	&	mod		& \hrulefill	& Plot 
-				the three \CIII{} coordinates. See Figure~\ref{plotc3}.\\[.9pt]
+				the three \CIII{} coordinates. See Figure~\ref{plotc3}. The gray lines
+				connect successive pairs of points 
+				in the input. The output graphics file
+				is an SVG file whose name is in the output
+				in standard output.\\[.9pt]
 			SELLA		&	y	&	mod		&\hrulefill	&	
@@
 	\begin{itemize}
-	\item{CmdGen}
+		  \setlength\itemsep{3em}
+	\item{\textbf{CmdGen}}
 
@@
 
-	\item SVD
+	\item \textbf{SVD}
 
@@
 				\textbf{Command line:}\\
-				CmdGen 5 33 | CmdSort seq | SVD\\
+				CmdGen 5 33 \textbar~ CmdSort seq \textbar~ SVD\\
 				; SVD\\
@@
 
+	\item \textbf{CmdS6Refl}
+
+	CmdS6Refl generates the 24 reflections (see \cite{Andrews2019b})
+	of each input lattice. Duplicate outputs of each input are not removed.
+
+	\item \textbf{CmdPath}
+
+	CmdPath outputs points along a line between pairs of lattice.
+
+	The number of points per segment is a commandline parameter.
+	If no parameter is input, the default is 20. This is not
+	the number of steps; it is the number of points that will be
+	output for a segment (pair of input lattices).
+
+\begin{enumerate}
+	\item Only a single lattice is input. A second point will
+	be generated as the Niggli reduced lattice, and series of points
+	will be output between them.
+
+	\item Two lattices are input. Then the points will be
+	created between them.
+
+	\item More than two lattices are input. Then each point
+	and its succeeding point will generate a segment. Each segment
+	will be output as the number of points specified. Note: the
+	final point of one segment and the first point of the next
+	segment will be duplicates.
+\end{enumerate}
+
+	The line between points is calculated in \SVI{}. The other
+	likely space would be \GVI{}, but the results seem to be the same.
+	As an example, in the case of 3 items input and 5 requested
+	output lattices, the output will consist of 10 lines, 5
+	for each successive pair of lattices, and the 5th and
+	6th points will be the same.\\
+
+	\textbf{Command Line:}\\
+	 CmdPath 5 \\
+	; Path generator\\
+	f 10 10 10 90 90 90 90\\
+	end\\
+
+	\textbf{Output:}\\
+	; no. of points output is 5\\
+	S    0.00000   0.00000   0.00000 -100.00000 -100.00000 -100.00000    1\\
+	S    6.25000   6.25000   6.25000 -100.00000 -100.00000 -100.00000    2\\
+	S   12.50000  12.50000  12.50000 -100.00000 -100.00000 -100.00000    3\\
+	S   18.75000  18.75000  18.75000 -100.00000 -100.00000 -100.00000    4\\
+	S   25.00000  25.00000  25.00000 -100.00000 -100.00000 -100.00000    5\\
+
+	\textbf{Command Line:}\\
+	 CmdPath 5  \textbar~ CmdToCell\\
+	; To Cell\\
+	f 10 10 10  90 90 90\\
+	end\\
+
+	\textbf{Output:}\\
+	; no. of points output is 5\\
+	P  10.00000  10.00000  10.00000  90.00000  90.00000  90.00000\\
+	P   9.35414   9.35414   9.35414  85.90396  85.90396  85.90396\\
+	P   8.66025   8.66025   8.66025  80.40593  80.40593  80.40593\\
+	P   7.90569   7.90569   7.90569  72.54240  72.54240  72.54240\\
+	P   7.07107   7.07107   7.07107  60.00000  60.00000  60.00000\\
+
 \end{itemize}
@@
 			fewer free parameters than the generic type requires. For instance,
-			O3 (character: rs0 rs0) has only two free parameters (r and s), whereas
+			O3 (character: \textit{rs0 rs0}) has only two free parameters (r and s), whereas
 			an ordinary orthorhombic type requires three variables.}
-		\includegraphics[width=0.75\textwidth]{DeloneTypesGrid_3}\\
+		\includegraphics[width=0.85\textwidth]{DeloneTypesGrid_3}\\
 		\label{s6table}		
@@
 
-	\ack{{\bf Acknowledgements}}
+	\ack{\bf Acknowledgements}
 
-	\ack{{\bf Funding information}}
+	\ack{\bf Funding information}
 
@@
 
-	%	\bibliography{reduced}
-	%	
-	%	\bibliographystyle{iucr}
+		\bibliography{reduced}
+
+		\bibliographystyle{iucr}
 

### 2023-02-05 LatticeMatching/LM.tex
LatticeMatching  1st return from referees

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
%\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{graphicx}
\usepackage{calligra}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\usepackage{color}
%\usepackage{ulem}
\newcommand{\NKS}[1]{{\color{red}~\textsf{#1}}}
\usepackage{url}
\usepackage{booktabs}
%\usepackage{multirow}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}
%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
\journalcode{J}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation
\makeatletter
\font\dummyft@=dummy \relax
\makeatother

\begin{document}                  % DO NOT DELETE THIS LINE

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	% HJB IUCr id IUCr6484
	% NKS IUCr ID: IUCr7572
	% lca ORCID  0000-0002-4451-1641
	% HJB ORCID 0000-0002-0517-8532
	% NKS ORCID 0000-0003-2786-6552
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).
	{\LARGE \emph{\today}} \\
	\title{Approximating Lattice Similarity}
	%\title{Note on the transformation of three-space basis vectors to  corresponding matrix for Delaunay scalars}
	\shorttitle{Lattice Matching}

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
	\author[b]{Herbert J.}{Bernstein}
	\author[c]{Nicholas K.}{Sauter}

	\aff[a]{Ronin Institute, 9515 NE 137th St., Kirkland, WA, 98034-1820 \country{USA}}
	\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973-5000 \country{USA}}
	\aff[c]{Lawrence Berkeley National Laboratory, 1 Cyclotron Rd., Berkeley, CA, 94720 \country{USA}}

	% Use \shortauthor to indicate an abbreviated author list for use in
	% running heads (you will need to uncomment it).

	\shortauthor{Andrews, Bernstein, and Sauter}

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	\keyword{Lattice Matching}
	\keyword{Delaunay}
	\keyword{Delone}
	\keyword{Niggli}
	\keyword{Selling}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	\maketitle                        % DO NOT DELETE THIS LINE

	\begin{synopsis}
		A method is proposed for transforming unit cells for a group of crystals so that they all appear
		as similar as possible to a selected cell.
	\end{synopsis}

	\newcommand{\OPES}[0]{$E^3toS^6$}
	\newcommand{\OPESS}[0]{$$E^3toS^6$$}
	\newcommand{\SVI}[0]{$\bf{S^{6}}$}
	\newcommand{\GVI}[0]{$\bf{G^{6}}$}
	\newcommand{\CIII}[0]{$\bf{C^{3}}$}
	\newcommand{\EIII}[0]{$\bf{E^{3}}$}
	\newcommand{\RIII}[0]{$\bf{R^{3}}$}
	\newcommand{\DVII}[0]{$\bf{D^{7}}$}
	\newcommand{\VVII}[0]{$\bf{V^{7}}$}
	\newcommand{\MSVI}[0]{$M_{S^{6}}$}
	\newcommand{\MEIII}[0]{$M_{E^{3}}$}
	\newcommand{\EXIII}[0]{$\bf{E^{3 \times 3}}$}
	\newcommand{\scalarsub}[2]{$#1_#2$}
	\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}

	\begin{abstract}

		A method is proposed for choosing unit cells for a group of crystals so that they all appear
		as nearly similar as possible to a selected cell. Related unit cells with varying cell parameters 
		or indexed with different lattice centering can be accommodated.

	\end{abstract}

	%-------------------------------------------------------------------------
	% The main body of the paper
	%-------------------------------------------------------------------------
	% Now enter the text of the document in multiple \section's, \subsection's
	% and \subsubsection's as required.

	% Appendices appear after the main body of the text. They are prefixed by
	% a single \appendix declaration, and are then structured just like the
	% body text.

	\section{Introduction}

	A common problem in crystallography is to provide a list of the unit cells of several (or many) 
	crystals so that they can be visually compared, making it easier to identify meaningful clusters
	of crystals of related morphology. Collections of experimental unit cell parameters have 
	been created based on similarity of morphology (for example, see \citeasnoun{Donnay1963}), and,
	in recent years, the clustering of unit cells from the
	myriad of images in serial crystallography has become increasingly important \cite{keable2021room}. We have created a method
	to group unit cells to serve these needs and have addressed this problem in the space \SVI{} \cite{Andrews2019b}.

	\section{Background and notation}

	\subsection{The space \SVI{}}
	\citeasnoun{Andrews2019b} introduced the space \SVI{} as an alternative
	representation of crystallographic lattices. The space is defined in terms of the
	``Selling scalars'' used in Selling reduction \cite{Selling1874} and by \citeasnoun{Delaunay1932}%
		\footnote{In his later publications, Boris Delaunay used the Russian version of his surname, Delone.}
	for the classification of lattices. A point $\bf{s}$ in \SVI{} is defined by

	$\bf{s}$  = [\vdotv{b}{c}, \vdotv{a}{c}, \vdotv{a}{b}, \vdotv{a}{d}, \vdotv{b}{d}, \vdotv{c}{d}]
	where
	${\mathbf d} = -{\mathbf a} - {\mathbf b} - {\mathbf c}$.
	As a mnemonic to remember the order: the terms involve, in order, $ \alpha, \beta,
	\gamma, \mathbf{a}, \mathbf{b}, \mathbf{c}$.

	\subsection{The space \GVI{}}
	\citeasnoun{Andrews2014} introduced the space \GVI{} based on the metric tensor
	and operated on by Niggli reductions \cite{Niggli1928}. A point in \GVI{}
	is defined by \\
	$\bf{g}$ =   [$\mathit{a^2}$, $\mathit{b^2}$, $\mathit{c^2}$, 2\vdotv{b}{c}, 2\vdotv{a}{c}, 2\vdotv{a}{b}]

	\subsection{Similarity}

In Euclidean geometry, two objects are described as ``similar'' if they are identical except for a scale factor, see Euclid, in \cite{heath1956thirteen}.
	In crystallography, we can say that all face-centered cubic unit cells are similar (assuming
	that they are in the same presentation). On the other hand, not all primitive orthorhombic 
	unit cells are similar.  In a metric space, we
	refer to two objects as ``approximately similar'' if the distance between them after scaling
	to the same size is, in some sense, small, {\it e.g}. commensurate with the experimental errors
	in determination of the unit cells.  The algorithm below attempts to find the representation of one 
	cell that is nearest-to-similar to some other cell. For a given reference cell, the probe cell
	will be transformed to other choices of unit cell that would generate the probe's lattice, and the
	closest match to the reference will be chosen for the result. Finally, the lattice centering
	of the reference cell will be restored (if necessary).

	\section{Algorithm}

	We start with a collection of experimental unit cells. From among them, we select or create the
	``reference'' cell; that is the one to which all the rest will be matched as closely as possible.

	We transform the reference cell by many operations in the course of exploring alternative
	lattice representations. For each newly
	generated lattice representation, we accumulate the transformations needed to convert  back to the 
	original reference cell.  All of these operations are
	performed in \SVI{}. (The alternative space, \GVI{}, is less convenient
	because the $ \mathbf{G^6}$ fundamental unit is non-convex.) To avoid duplication, for each step 
	we only accumulate transformations
	that have not already been found.

	To begin, each input cell is transformed to the \SVI{} representation, and then 
	Selling-reduced (see \citeasnoun{Delaunay1932} and \citeasnoun{Andrews2019a}). As
	there is a need to be able to reverse the reduction, the reduction transformation 
	is saved for use in later stages.

	The following transformations of the reference will be done in three stages.

	First, the 24 \SVI{} reflections are
	applied \cite{Andrews2019b} and the results stored. The store of \SVI{} vectors and their generating matrices holds 24 entries each at that point.

	Next the boundary (reduction) transformations \cite{Andrews2019b} are applied to
	the results of the previous step. Then, the 24 reflections are applied again. In
	each step, only newly found results are stored. These latter two steps are repeated 
	at least once in order to gain better coverage of possibly useful transformations. 
	The counts of entries for each iteration are 24, 1566, 45876, and finally 1016726.
	Three iterations, {\it i.e.} 45876 entries, have been sufficient in test cases to date.

	Next, all the accumulated, transformed representations of the reference cell must be rescaled and the saved
	transformations inverted.
	The \SVI{} vectors are all scaled to the same length (see Section~\ref{section:modified}), and the transformation matrix
	attached to each vector is inverted, thereby yielding the operation to return a lattice to the vicinity of the
	original reference cell. For more efficient searching in the final step, it is helpful to use
	a nearest neighbor search function such as NearTree \cite{andrews2016}.

	\section{Why must the \SVI{} vectors be scaled?} 	\label{section:modified}

	All similar lattices lie on lines that go through the origin of \SVI{}. Fig. \ref{similarity} shows
	the distinction between the cases where the transformed points are scaled to all be at the same
	distance from the origin as the reference point ( Fig. \ref{similarity}\textbf{a}) and the case where they are not ( Fig. \ref{similarity}\textbf{b}).
	Fig. \ref{zone} illustrates the way in which scaling all the reference points to the same
	6-spherical surface defines the zones of approximate similarity.
	Any non-zero scale factor will produce the same correct result. In \SVI{}
	the reflections maintain the distance from the origin, but the boundary transformations may not.
	Repeating: the only way to guarantee that the
	separation line for two regions goes through the origin is to have all the points at
	the same radius.

	\begin{figure}
		\label{similarity}
		\includegraphics[height=3.8cm]{similarity}
		\caption{Frame \textbf{a} shows the case where the transformed point \textbf{T} has been scaled to be at the same distance 
			from the origin as the reference point \textbf{R}.  In Frame \textbf{b}, point \textbf{T} has not been scaled, and some areas
			are incorrectly assigned to point R.  In each frame, the straight line between points \textbf{R} and \textbf{T} separates the 
			regions closer to each of the points.}
	\end{figure}

	\begin{figure}
		\label{zone}
		\includegraphics[height=5cm]{zone2a}
		\caption{A two-dimensional example of the determining similarity. Each transformed copy of the reference cell
			is normalized to a constant length in the chosen space (here, \SVI{}). Each transformed and normalized cell then 
			defines a zone in which every point in that zone is closer to the transformed and normalized cell defining that zone than it is
			to the transformed and normalized cell defining any other zone.  In this example, each point within the textured zone (which extends
			to infinity) is closer to the
			gray-centered point than it is to any of the black points.}
	\end{figure}

\section{Angular measure of fit}

Because the measure of similarity is independent of scale, projecting
points onto a spherical surface does not modify the similarity. The
angle between a probe point and the reference point is a meaningful measure of how
similar the two points are.

	\section{Generating the approximation}

	The following operations are performed for each of the probe lattices in 
	the original list.  For a given probe lattice, the closest approximation among all of the 
	transformed reference points is found. If there are multiple representations of the 
	reference point that are equally close, then all should be examined. For the case of multiples, a
	method must be used to find the preferred one. For our purposes, we have found
	it convenient to choose the one for which the unreduced \GVI{} distances to the transformed
	reference is the smallest. Other choices might be useful for other purposes.

	Once the preferred result has been found, the corresponding inverted transformation is used
	to place the vector in the region of the original reduced reference cell. Finally, the inverse of the reduction
	operation that was performed on the reference cell is used to create the best match
	to the original reference. If it is desirable to restore lattice centering, then that operation must also  be
	performed.

	\section{Examples}
	\subsection{A rhombohedral example}
	\citeasnoun{LeTrong2007} cite several structures for phospholipase A2 (krait neurotoxin) that were reported as 
	different structures but were actually all the same structure \cite{bernstein2020}. Expanding their search using the program SAUC 
	\cite{mcgill2014},
	we find a total of six structures, four of which are identical in 
	two pairs. Table \ref{table:LT1} lists the unit cells as reported in the Protein
	Data Bank \cite{Bernstein1977} \cite{Berman2000}. In Tables \ref{table:LT2}, \ref{table:LT3}, and \ref{table:LT4}, the first
	entry in each table is used as the reference, and the following five entries are matched as 
	closely as possible to the presentation of the reference cell. In Table \ref{table:LT2}, a
	rhombohedral presentation 1DPY was chosen as the reference. In Table \ref{table:LT3},
	C-centered cell 1G2X was chosen as the reference. In Table \ref{table:LT4}, the hexagonal cell 1U4J
	was chosen as the reference. In each case, the probe cells
	were return in the same presentation, including lattice 
	centering as the reference cell. So the resulting centerings
	were hR, mC, and hP, respectively, for each matched cell,
	regardless of the input centering, which had been determined
	by crystallographic analysis.
	~~\\

	\begin{table}
		\begin{center}
			\caption{Unit cells of phospholipase A2 from the Protein Data Bank.}		
			\vspace{3mm}
			\begin{tabular}{lccccccccccc}
				\toprule
				PDB &   & a &b &c&$\alpha$&$\beta$&$\gamma$ \\
				\midrule
				1DPY &R& 57.98& 57.98 &57.98 &92.02  &92.02 &92.02 \\ 
				1FE5 &R& 57.98& 57.98 &57.98 &92.02  &92.02 &92.02 \\ 
				1G0Z &H& 80.36& 80.36 &99.44 &90     &90    &120   \\ 
				1G2X &C& 80.95& 80.57 &57.1  &90     &90.35 &90    \\ 
				1U4J &H& 80.36& 80.36 &99.44 &90     &90    &120   \\ 
				2OSN &R& 57.10 & 57.10  &57.10  &89.75  &89.75 &89.75  \\ 
				\bottomrule
			\end{tabular}
			\label{table:LT1}
		\end{center}
	\end{table}

	\begin{table}
		\begin{center}
			\caption{The data of Table \ref{table:LT1} matching a rhombohedral reference. The 
				reference cell is marked in boldface.}
			\vspace{3mm}
			\begin{tabular}{lccccccccccc}
				\toprule
				PDB &   a & b & c & $\alpha$ & $\beta$ & $\gamma$ & fit(degrees) \\ \midrule
\textbf{1DPY} & \textbf{57.98}  & \textbf{57.98}  & \textbf{57.98}  &
\textbf{92.02}  & \textbf{92.02}  & \textbf{92.02} & 0 \\ 
1FE5 &57.980& 57.980 & 57.980 & 92.020 & 92.020 & 92.020 & 0    \\
1G0Z &57.020& 57.020 & 57.020 & 90.395 & 90.395 & 89.605 & 2.11 \\
1G2X &57.106& 57.106 & 57.100 & 89.752 & 90.248 & 90.270 & 2.11 \\
1U4J &57.020& 57.020 & 57.020 & 90.395 & 90.395 & 89.605 & 2.11 \\
2OSN &57.100& 57.100 & 57.100 & 90.250 & 90.250 & 89.750 & 2.11 \\

				\bottomrule
			\end{tabular}		
			\label{table:LT2}
		\end{center}
	\end{table}

	\begin{table}
		\begin{center}
			\caption{The data of Table \ref{table:LT1} matching a monoclinic reference. The 
				reference cell is marked in boldface.}
			\vspace{3mm}
			\begin{tabular}{lccccccccccc}
				\toprule[1pt]
								PDB &   a & b & c & $\alpha$ & $\beta$ & $\gamma$ & fit(degrees) \\ \midrule
\textbf{1G2X} & \textbf{80.95}  & \textbf{80.57}0  & \textbf{57.10}  & \textbf{90}  & \textbf{90.35}  & \textbf{90} & \textbf{0}   \\
1DPY & 83.42999  & 80.53835  & 57.98115  & 87.0908  & 89.9992  & 90.00114 & 3.14          \\
1FE5 & 83.42999 &  80.53835  & 57.98115  & 87.0908  & 89.9992  & 90.00114 & 3.14          \\
1G0Z & 80.91861  & 80.35937  & 57.02143  & 89.9996  & 90.5644  & 90.00144 & 0.09        \\
1U4J & 80.91861  & 80.35937  & 57.02143  & 89.9996  & 90.5603  & 90.00144 & 0.09       \\
2OSN & 80.92799  & 80.57842  & 57.10254  & 90.0002  & 90.3502  & 89.99745 & 0.01      \\

				\bottomrule[1pt]
			\end{tabular}
			\label{table:LT3}
		\end{center}
	\end{table}

	\begin{table}
		\begin{center}
			\caption{The data of Table \ref{table:LT1} matching a hexagonal reference. The reference cell is marked in boldface.}
			\vspace{3mm}
			\begin{tabular}{lccccccccccc}
				\toprule
PDB &   a & b & c & $\alpha$ & $\beta$ & $\gamma$ & fit(degrees) \\ \midrule         
\textbf{1U4J} & \textbf{80.36} & \textbf{80.36} & \textbf{99.44} & \textbf{90} & \textbf{90} & \textbf{120} & \textbf{0} \\
1DPY & 83.4287 & 80.5380 & 101.5974 & 91.6597 & 90 & 121.195 & 4.30   \\
1FE5 & 83.4287 & 80.5380 & 101.5974 & 91.6597 & 90 & 121.195 & 4.30   \\
1G0Z & 80.3600 & 80.3600 & 99.4400  & 90 & 90 & 120 & 0          \\
1G2X & 80.5809 & 80.5809 & 99.3468  & 90.0138 & 89.986 & 120.009 & 0.09  \\
2OSN & 80.5752 & 80.5752 & 99.3307  & 90 & 90 & 120 & 0.08  \\

				\bottomrule
			\end{tabular}		
			\label{table:LT4}
		\end{center}
	\end{table}

	\subsection{Adenosine receptor A2A}
	Unit cells were determined automatically from frames from serial crystallography data collection for adenosine receptor A2A, PDB ID: 5NLX, \cite{weinert2017}.

	Three example unit cells were chosen from several hundred indexed data frames.
	Two are C-centered and one is primitive. 
	Table \ref{table:signal0} is the reported data, and tables
	\ref{table:signal1}, \ref{table:signal2}, and \ref{table:signal3}
	are the approximate similarity matches.
	\begin{table}
		\begin{center}
			\caption{Adenosine receptor A2A: 5NLX.  }
~~\\

			\begin{tabular}{clccccccccccc}
				\toprule
				serial\\
				\midrule
				1&C&39.741&183.767&140.649&90&90&90.\\
				2&P&40.160&142.899&92.417&90&102.480&90.\\
				3&C&180.613&40.156&142.737&90&90.017&90.\\
				\bottomrule 
			\end{tabular}
			\label{table:signal0}
		\end{center}
	\end{table}

	\begin{table}
		\begin{center}
			\caption{Adenosine receptor A2A:  Approximating a C-centered cell.
				The reference cell is marked in boldface.
				 Centering in parentheses indicates the lattice centering before matching.}
			\vspace{3mm}
			\begin{tabular}{clccccccccccc}
				\toprule
serial&center &a & b & c & $\alpha$ & $\beta$ & $\gamma$ & fit(degrees) \\ \midrule         
1&\textbf{C}    & \textbf{39.741} & \textbf{183.767} & \textbf{140.649}  & \textbf{90}  &\textbf{90}  &\textbf{90} &  \textbf{0}\\
2&(P)  & 40.160 & 180.467 & 142.899  & 90  & 90  & 89.931 & 1.37981\\
3&C    & 40.156 & 180.613 & 142.737  & 89.983  & 90  & 90 &  1.30423\\
				\bottomrule
			\end{tabular}
		\end{center}
	\label{table:signal1}
	\end{table}

	\begin{table}
		\begin{center}
			\caption{Adenosine receptor: A2A. Approximating a primitive cell.   The reference cell	is marked in boldface.  Centering in parentheses indicates the lattice centering before matching.}
			~~\\c

			\begin{tabular}{lccccccccccc}
				\toprule
serial&center& a & b & c & $\alpha$ & $\beta$ & $\gamma$ & fit(degrees) 	\\			
\toprule
				2&\textbf{P} & \textbf{40.160} & \textbf{142.899} & \textbf{92.417} & \textbf{90} & \textbf{102.480} & \textbf{90}&\textbf{0}	\\  
1&(C)   & 39.741 & 140.649  & 94.008  & 90 &  102.21  & 90 &  1.37981 \\
3&(C)   & 40.156 & 142.73  & 92.512  & 89.983 & 102.535 & 90 &  0.08374 \\				\bottomrule
			\end{tabular}
		\end{center}
	\label{table:signal2}
	\end{table}

%	\begin{table}
%		\vspace{3mm}
%		\begin{center}
%			\begin{tabular}{lccccccccccc}
%				\toprule
%				P &	\textbf{40.160}& \textbf{142.899} & \textbf{92.417} &
%				\textbf{90.} &\textbf{102.480} & \textbf{90.} \\
%				C  & 39.741 & 140.649 & 94.008 & 90. & 102.203 & 90.   \\
%				C  & 40.156 & 142.737 & 92.512 & 89.983 & 102.535 & 90.   \\ 
%				\bottomrule
%			\end{tabular}			
%			\label{table:signal2}
%		\end{center}
%	\end{table}

%	\begin{table}
%		\begin{center}
%			\caption{Adenosine receptor: A2A. Approximating a C-centered alternate presentation.   
%				The reference cell is marked in boldface.}
%			\vspace{3mm}
%			\begin{tabular}{lccccccccccc}
%				\toprule
%				C & 180.613 & 40.156 & 142.737 & 90. & 90.017 & 90. \\ 
%				C &  39.741& 183.767& 140.649 & 90. & 90. & 90 \\  
%				P &  40.160 &142.899 & 92.417 & 90. &102.480 & 90. \\ 
%				\bottomrule
%			\end{tabular}			
%		\end{center}
%	\end{table}

	\begin{table}
		\begin{center}
			\caption{Adenosine receptor: A2A. Approximating a C-centered cell.   The reference cell	is marked in boldface.  Centering in parentheses indicates the lattice centering before matching.}
			~~\\

			\begin{tabular}{clccccccccccc}
				\toprule
serial&center& a & b & c & $\alpha$ & $\beta$ & $\gamma$ & fit(degrees) \\
\midrule
				3&	\textbf{C} &  \textbf{180.613} &  \textbf{40.156} & \textbf{142.737} &  \textbf{90} &   \textbf{90.017} &  \textbf{90} &\textbf{0}  \\
			1&	C &  183.767 &  39.741 & 140.649 &  90 &  90 &  90 &1.31 \\
			2&	(P) &  180.467 &  40.160 & 142.899 &  90 &  90 &  89.930 &0.07\\
				\bottomrule
			\end{tabular}
			\label{table:signal3}
		\end{center}
	\end{table}

	\subsection{Points along a line in \SVI}
	Tables \ref{table:AD1} and \ref{table:AD2} present two views of artificial data. 
	A line of points in \SVI{} was created from the C-centered [80.95, 80.95, 57.10, 90, 90.35, 90] 
	to the A-centered [57.10, 80.95, 80.95, 90, 90, 90.35] representation of the same cell of phospholipase A2. The 
	series of intervening points  interpolated in S6 and shown in Table \ref{table:AD1} (each as the reduced unit cell except for the endpoints), 
	and the lattice-matched results are shown
	in Table \ref{table:AD2}.

	In Table \ref{table:AD2},
	the first line is the reference cell. In both cases that is the C-centered cell above.
	The final cell is the same cell but in the A-centered presentation. The points between
	are equally spaced points in \SVI{} between those two centered points. Table \ref{table:AD1}
	presents the list of points as generated, and Table \ref{table:AD2} lists the same cells in the lattice-matching
	presentation. Because the initial cell was C-centered, the following cells are also
	in that presentation, although the intermediate cells are not C-centered.

	\begin{table}
		\begin{center}
			\caption{A line of unit cells generated by interpolating 
				between the first and last points in \SVI.}
			~~\\

			\begin{tabular}{lccccccccc}
				\toprule
				&a&b&c&$\alpha$&$\beta$&$\gamma$ \\ \midrule
				C & 80.95    &  80.95  & 57.1    &  90      &  90.35   & 90    \\  
				P & 57.24 &57.24 &80.68 &129.50 &94.21   &90     \\  
				P & 57.24 &57.24 &80.68 &124.46 &98.26  & 90    \\  
				P & 57.24 &57.24 &80.68 &119.70 &102.36 &90     \\  
				P & 57.24 &57.24 &80.68 &115.16 &106.52 &90     \\  
				P & 57.24 &57.24 &80.68 &110.78 &110.78 &90     \\  
				P & 57.24 &57.24 &80.68 &106.53 &115.16 &90     \\  
				P & 57.24 &57.24 &80.68 &102.36 &119.70 &90     \\ 
				P & 57.24 &57.24 &80.68 &98.26  &124.46 &90     \\  
				P & 57.24 &57.2& 80.68 &94.209   &129.50 &90  \\ 
				A &  57.10 & 80.95 & 80.95 & 90 & 90 & 90.35		\\	 
				\bottomrule
			\end{tabular}			
			\label{table:AD1}
		\end{center}
	\end{table}

	\begin{table}
		\begin{center}
			\caption{Table 13 lists the same cells in the same order after transformation to approximately match the reference cell.
				The reference cell is marked in boldface.}
			\vspace{3mm}
			\begin{tabular}{ccccccccc}
				\toprule
 a & b & c & $\alpha$ & $\beta$ & $\gamma$ & fit(degrees) \\ \midrule         
\textbf{80.950} & \textbf{80.950}  &  \textbf{57.100}  & \textbf{90}  & \textbf{90.350}    & \textbf{90} & \textbf{0}  \\   
80.680 & 88.685  &  57.240  & 86.171  & 94.210    & 95.086 & 5.687  \\
80.680 & 95.7216  & 57.240  & 83.0450  & 98.260    & 99.564 &  12.17  \\
80.680 & 102.287 &  57.240 &  80.280 &  102.360  & 103.547 &  19.35  \\
80.680 & 108.450 &  57.240 &  77.787 &  106.520  & 107.167 & 27.08  \\
80.677 & 114.284 &  57.240 &  75.499 &  110.775  & 110.520 & 35.10  \\
80.680 & 108.450 &  57.240 &  77.780 &  106.530  & 107.167 & 27.09  \\
80.680 & 102.287 &  57.240 &  80.280 &  102.360  & 103.547 &  19.35  \\
80.680 & 95.722  &  57.240  & 83.045  & 98.260    & 99.564 &  12.17  \\
80.680 & 88.685  &  57.200  & 86.172  & 94.209    & 95.086 & 5.70  \\
80.950 & 80.950  &  57.100  & 90  & 90.350    & 90 & 0  \\
				\bottomrule
			\end{tabular}			
			\label{table:AD2}
		\end{center}
	\end{table}

	\subsection{Examples from the Protein Data Bank}

	The program SAUC \cite{mcgill2014} was used to query the
	Protein Data Bank (PDB). The search started from the
	C-centered unit cell of PDB entry 1RGX (resistin) requesting the nearest 50 cells;
	26 unique cells resulted. Because there was no limit on
	how far the points could be from the probe, some cells
	differ significantly from the search cell. The results are listed
	in Table \ref{table:PDB1} in their published representation. Table \ref{tablePDB2} lists the same cells
	in the same order.

	\begin{table}
		\begin{center}
			\caption{Unit cells from the Protein Data Bank. Cells listed are nearest the C-centered cell of 1RGX and 
				keeping only one representative of each protein type. The search was performed 
				using the program SAUC.}		
			\vspace{3mm}
			\begin{tabular}{lcccccccclcccc} \toprule
				\rotatebox{0}{PDB}&& a&b&c&$\alpha$&$\beta$&$\gamma$ \\ \midrule
				1RGX  &  C &   49.021  &  52.475 &   96.609  &  90 &   96.53 &   90 \\
				1R8M &  P  & 33.429  & 95.775  & 33.665   & 90     & 101.67~  & 90   \\
				2FXO  & P  & 40.157  & 41.867  & 97.795   & 91.11  & 92.73   & 107.18 \\
				4RNE  & C  & 37.656  & 54.197  & 95.677   & 90     & 90      & 90    \\
				3MGD  & C  & 57.933  & 56.341  & 99.721   & 90     & 98.86   & 90   \\
				5YO3  & C  & 40.328  & 50.126  & 94.237   & 90     & 90      & 90   \\
				4GZN  & C  & 40.218  & 60.641  & 96.119   & 90     & 90      & 90   \\
				3VVW  & C  & 195.72  & 37.420  & 40.280   & 90     & 94.66   & 90   \\
				4BHV  & P  & 33.078  & 33.621  & 99.138   & 90     & 96.75   & 90   \\
				3IHU  & C  & 54.646  & 79.135  & 103.244~  & 90     & 102.08~  & 90   \\
				5WOU  & C  & 36.429  & 53.884  & 94.219   & 90     & 90      & 90   \\
				3NHM  & C  & 56.616  & 40.408  & 99.617   & 90     & 102.28~  & 90   \\
				5K2L  & P  & 29.130  & 29.130  & 94.257   & 90     & 90      & 90   \\
				3T47  & P  & 26.152  & 94.356  & 29.196   & 90     & 97.19   & 90   \\
				4RUV  & P  & 31.376  & 31.376  & 94.804   & 90     & 90      & 90   \\
				5ED9  & C  & 86.371  & 34.743  & 99.839   & 90     & 101.49~  & 90   \\
				1SIP  & C  & 32.180  & 62.520  & 95.760   & 90     & 90      & 90   \\
				2SAM  & C  & 62.700  & 32.200  & 96.100   & 90     & 90      & 90   \\
				1YTJ  & C  & 62.300  & 32.100  & 96.300   & 90     & 90      & 90   \\
				4HHX  & C  & 34.790  & 73.610  & 95.900   & 90     & 90      & 90   \\
				3W92  & P  & 31.760  & 33.552  & 94.998   & 90     & 90      & 90   \\
				167D  & P  & 33.200  & 33.200  & 96.040   & 90     & 90      & 120  \\
				4QEG  & P  & 31.237  & 31.237  & 93.848   & 90     & 90      & 90   \\
				2YGG  & C  & 200.700~ & 38.350  & 34.100   & 90     & 91.35   & 90   \\
				1OZ7  & P  & 37.966  & 95.258  & 42.611   & 90     & 112.58~  & 90   \\
				6NFS  & P  & 30.584  & 34.753  & 94.679   & 90     & 90      & 90   \\ \bottomrule
			\end{tabular}			
			\label{table:PDB1}
		\end{center}
	\end{table}

	\begin{table}
		\begin{center}
			\caption{Data in Table \ref{table:PDB1} best-matched to PDB entry 1RGX.  The 
				reference cell is marked in boldface.}
			\vspace{3mm}
			\begin{tabular}{lccccccccccc} \toprule
				\rotatebox{0}{PDB}&  a&b&c&$\alpha$&$\beta$&$\gamma$ & fit(degrees)\\ \midrule

				\textbf{1RGX}  &\textbf{49.021} &\textbf{52.475}
				&\textbf{96.609}   &\textbf{90.}  &\textbf{96.53}   &\textbf{90.}   &\textbf{0} \\

%1RGX & 49.021 & 52.475 & 96.609 & 90 & 96.53 & 90 & 0         \\
1R8M & 42.374 & 52.020 & 95.775 & 90 & 90 & 90.41 & 2.67 \\
2FXO & 48.706 & 66.020 & 97.795 & 89.04 & 93.21 & 87.50 & 3.07 \\
4RNE & 37.656 & 54.197 & 95.677 & 90 & 90 & 90 & 3.80 \\
3MGD & 57.933 & 56.341 & 99.721 & 90 & 98.86 & 90 & 2.91 \\
5YO3 & 40.328 & 50.126 & 94.237 & 90 & 90 & 90 & 2.84 \\
4GZN & 40.218 & 60.641 & 96.119 & 90 & 90 & 90 & 3.86 \\
3VVW & 54.979 & 54.979 & 99.633 & 86.02 & 100.74 & 85.78 & 3.90  \\
4BHV & 47.165 & 47.165 & 99.138 & 85.27 & 94.73 & 89.07 & 2.71 \\
3IHU & 54.646 & 79.135 & 103.244 & 90 & 102.08 & 90 & 5.07   \\
5WOU & 36.429 & 53.884 & 94.219 & 90 & 90 & 90 & 4.01 \\
3NHM & 56.616 & 40.408 & 99.617 & 90 & 102.28 & 90 & 4.39  \\
5K2L & 41.196 & 41.196 & 94.257 & 90 & 90 & 90 & 2.65 \\
3T47 & 41.563 & 36.677 & 94.356 & 90 & 90 & 83.65 & 2.91 \\
4RUV & 44.372 & 44.372 & 94.804 & 90 & 90 & 90 & 2.54 \\
5ED9 & 46.548 & 67.682 & 99.839 & 82.70 & 100.65  & 107.73 & 5.82   \\
1SIP & 35.158 & 57.508 & 95.760 & 90 & 90 & 84.31 & 4.83 \\
2SAM & 35.242 & 57.582 & 96.10 & 90 & 90 & 84.20 & 4.83 \\
1YTJ & 35.042 & 57.348 & 96.30 & 90 & 90 & 84.36 & 4.86 \\
4HHX & 40.709 & 63.858 & 95.90 & 90 & 90 & 80.10 & 4.63 \\
3W92 & 46.200 & 46.200 & 94.998 & 90 & 90 & 86.86 & 2.75 \\
167D & 33.20 & 57.504 & 96.040 & 90 & 90 & 90 & 5.26 \\
4QEG & 44.176 & 44.176 & 93.848 & 90 & 90 & 90 & 2.54 \\
2YGG & 51.318 & 51.318 & 102.166 & 82.83 & 98.954 & 96.71 & 3.40  \\
1OZ7 & 44.886 & 67.078 & 95.258 & 90 & 90 & 82.86 & 4.45 \\
6NFS & 46.294 & 46.294 & 94.679 & 90 & 90 & 82.70 & 2.99 \\
				\bottomrule
			\end{tabular}
			\label{tablePDB2}
		\end{center}
	\end{table}

	\section{Summary}

	A method is proposed for transforming unit cells for a group of crystals so that they all appear
	as similar as possible to a selected cell. The search for
	cells similar to the reference cell
	 in done using the reduced cell comparing to 
	other possible unit cells nearby in the space \SVI{}. 
	At the end, the lattice centering of the reference cell
	is restored.

	\section{Availability of code}

	The C++ code for lattice matching in S6 is available in github.com, in\\
	\url{https://github.com/duck10/LatticeRepLib.git}

	%\appendix

	%\section{blah blah blah -- Supplementary Material}
	\ack{{\bf Acknowledgements}}

	Careful copy-editing and corrections by Frances C. Bernstein are 
	gratefully acknowledged. Our thanks to Jean Jakoncic and Alexei Soares for 
	helpful conversations and access to data and facilities at 
	Brookhaven National Laboratory. We thank Ronald Stenkamp for pointing
	us to \citeasnoun{LeTrong2007}. We gratefully acknowledge
	Jörg Standfuss for permission to use the Adenosine Receptor A2A data.
	Richard Gildea helped in securing data for examples.

	\ack{{\bf Funding information}}

	Funding for this research was provided in part by:  
	US Department of Energy Offices of Biological and 
	Environmental Research and of Basic Energy Sciences 
	(grant No. KP1607011, grant No. DE-SC0012704 and in earlier years grant No. DE-AC02-98CH10886); 
	U.S. National Institutes of Health
	(grant No. P30GM133893, and in earlier years,
	grant No. P41RR012408; 
	grant No. P41GM103473; grant No. P41GM111244; 
	grant No. R01GM117126,
	grant No. 1R21GM129570); and in earlier years Dectris, Ltd.

	\bibliography{Reduced}

	\bibliographystyle{iucr}

	%-------------------------------------------------------------------------
	% TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
	%-------------------------------------------------------------------------

	% Simple tables should use the tabular environment according to this
	% model

	% Postscript figures can be included with multiple figure blocks

\end{document}                    % DO NOT DELETE THIS LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-02-06 LatticeMatching/LM.tex
update in reply to referees 1,2,3, but not completely the "Report"

@@
 
-	\subsection{The space \GVI{}}
-	\citeasnoun{Andrews2014} introduced the space \GVI{} based on the metric tensor
-	and operated on by Niggli reductions \cite{Niggli1928}. A point in \GVI{}
-	is defined by \\
-	$\bf{g}$ =   [$\mathit{a^2}$, $\mathit{b^2}$, $\mathit{c^2}$, 2\vdotv{b}{c}, 2\vdotv{a}{c}, 2\vdotv{a}{b}]
+%	\subsection{The space \GVI{}}
+%	\citeasnoun{Andrews2014} introduced the space \GVI{} based on the metric tensor
+%	and operated on by Niggli reductions \cite{Niggli1928}. A point in \GVI{}
+%	is defined by \\
+%	$\bf{g}$ =   [$\mathit{a^2}$, $\mathit{b^2}$, $\mathit{c^2}$, 2\vdotv{b}{c}, 2\vdotv{a}{c}, 2\vdotv{a}{b}]
 
@@
 
+	Because the 24 operations defining reflection are unitary and
+	in \SVI{} they are simply perturbations of the six values,
+	they retain the values and signs of he six values, 
+	simply rearranging the six scalars.
+
 	Next the boundary (reduction) transformations \cite{Andrews2019b} are applied to
@@
 	Three iterations, {\it i.e.} 45876 entries, have been sufficient in test cases to date.
+
+	Although the six scalars are all negative for Selling
+	reduced unit cells, the boundary transformations are
+	not unitary, and so not retain the six negative values.
 
@@
 	to the original reference. If it is desirable to restore lattice centering, then that operation must also  be
-	performed.
+	performed; the search returns a primitive representation
+	of the unit cell.
 
@@
 		\begin{center}
-			\caption{Adenosine receptor A2A: 5NLX.  }
+			\caption{Adenosine receptor A2A: 5NLX, unit cells as reported.  }
 ~~\\
@@
 	in Table \ref{table:PDB1} in their published representation. Table \ref{tablePDB2} lists the same cells
-	in the same order.
+	in the same order as in Table \ref{table:PDB1}, but with 
+	the same lattice centering as 1RGX, which is the first, 
+	reference entry.
 
@@
 	cells similar to the reference cell
-	 in done using the reduced cell comparing to 
+	 in done using the reduced cell and comparing to 
 	other possible unit cells nearby in the space \SVI{}. 

### 2023-02-07 LatticeMatching/LM.tex
fix Euclid reference instead of only Wikipedia

@@
 
-In Euclidean geometry, two objects are described as ``similar'' if they are identical except for a scale factor, see Euclid, in \cite{heath1956thirteen}.
+In Euclidean geometry, two objects are described as ``similar'' if they are identical except for a scale factor, see Euclid, in \cite{heath1956thirteen} and a longer description in Wikipedia  \cite{enwiki:1097100366}.
 	In crystallography, we can say that all face-centered cubic unit cells are similar (assuming

### 2023-02-08 MeasuringLattices_uv5016/MeasuringLattices.tex
Matching Lattices (ACTA paper uv5016)  creation

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
%\usepackage{bm}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
%\usepackage{calligra}
\usepackage{array}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
%\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\DeclareRobustCommand{\captionpar}{\par}
\usepackage{color}
\usepackage{ulem}
\usepackage{url}
\usepackage{yfonts}
\usepackage{relsize} 
\usepackage{wrapfig}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}

\newcommand{\SVI}[0]{$\bf{S^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\CIII}[0]{$\bf{C^{3}}$}
\newcommand{\DVII}[0]{$\bf{D^{7}}$}
\newcommand{\VVII}[0]{$\bf{V^{7}}$}

\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\Imaginary}[0]{$ \textfrak{I} $}
\newcommand{\Real}[0]{$ \textfrak{R} $}
\newcommand{\Exchange}[0]{$\textfrak{X}$}

\newcommand{\nounderline}[3]{\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineab}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineac}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
\newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}

\newcommand{\scalar}[1]{$#1$}

\newcommand{\scalarsub}[2]{$#1_#2$}

%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
\journalcode{A}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation
\makeatletter
\font\dummyft@=dummy \relax
\makeatother

\begin{document}                  % DO NOT DELETE THIS LINE
	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).
	\raggedbottom
	\setlength{\parskip}{0pt}
	{\LARGE \emph{\today}} \\
	\title{Measuring Lattices}
	%\title{Note on the transformation of three-space basis vectors to  corresponding matrix for Delaunay scalars}
	\shorttitle{Measuring Lattices}

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
	\author[b]{Herbert J.}{Bernstein}

	\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
	\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973-5000 \country{USA}}

	% Use \shortauthor to indicate an abbreviated author list for use in
	% running heads (you will need to uncomment it).

	\shortauthor{Andrews and Bernstein}

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	\keyword{Delaunay}
	\keyword{Delone}
	\keyword{Selling}
	\keyword{lattices}
	\keyword{unit cell}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	\maketitle                        % DO NOT DELETE THIS LINE
	\begin{synopsis}
        Unit cells are used to represent crystallographic lattices.
		Calculations measuring the differences between unit
		cells are used to provide metrics for measuring meaningful
		distances between three-dimensional crystallographic
		lattices.  This is a surprisingly complex and
		computationally demanding problem.  We present a
		review of the current best practice using 
		Delaunay-reduced unit cells
		in the six-dimensional real space of Selling
		scalar cells \SVI{} and the equivalent three-dimensional complex space \CIII{}.
	\end{synopsis}
	\newcommand{\si}[0]{$s_1$}
	\newcommand{\sii}[0]{$s_2$}
	\newcommand{\siii}[0]{$s_3$}
	\newcommand{\siv}[0]{$s_4$}
	\newcommand{\sv}[0]{$s_5$}
	\newcommand{\svi}[0]{$s_6$}
	\newcommand{\Svec} [0] {\{\si, \sii, \siii, \siv, \sv, \svi \}}
	\newcommand{\SvecA} [0] {\{-\si, -\si+\sii, \si+\siii, \si+\sv, \si+\siv, \si+\svi \}}

	\newcommand{\OPES}[0]{$E^3toS^6$}
	\newcommand{\OPESS}[0]{$$E^3toS^6$$}
	\newcommand{\MSVI}[0]{$M_{S^{6}}$}
	\newcommand{\MEIII}[0]{$M_{E^{3}}$}
	\newcommand{\Plus}[0]{$\textfrak{P}$}	
	\newcommand{\Minus}[0]{$\textfrak{M}$}

	\newcommand{\ci}[0]{$c_1$}
	\newcommand{\cii}[0]{$c_2$}
	\newcommand{\ciii}[0]{$c_3$}

	\begin{abstract}

		Unit cells are used to represent crystallographic lattices.
		Calculations measuring the differences between unit
		cells are used to provide metrics for measuring meaningful
		distances between three-dimensional crystallographic
		lattices.  This is a surprisingly complex and
		computationally demanding problem.  We present a
		review of the current best practice using 
		Delaunay-reduced unit cells
		in the six-dimensional real space of Selling
		scalar cells \SVI{} and the equivalent three-dimensional complex space \CIII{}.
		The process is a simplified version of the process
		needed when working with the more complex
		six-dimensional real space of Niggli-reduced unit
		cells \GVI{}.  Obtaining a distance begins with identification of the fundamental region in the space, continues with
		conversion to primitive cells and reduction, analysis of
		distances to the boundaries of the fundamental unit,
		and is completed by a comparison of direct paths to
		boundary-interrupted paths, looking for a path of
		minimal length.

	\end{abstract}

	%-------------------------------------------------------------------------
	% The main body of the paper
	%-------------------------------------------------------------------------
	% Now enter the text of the document in multiple \section's, \subsection's
	% and \subsubsection's as required.

	\section{History}

	Human fascination with crystals has a long history. 105,000 years ago,
	someone had a collection of calcite crystals 
	(Iceland spar) \cite{wilkins2021innovative}. 
	Theophrastus (ca. 372-287 B.C.), a student of Plato and successor to Aristotle, wrote the first known treatise on gems (``On Stones'')  \cite{enwiki:1114534722}.

	Figure \ref{timeline} notes a few key events in cataloging 
	crystal properties. We start with \citeasnoun{Kepler1611} (translated in \citeasnoun{kepler1966}) and Steno  (see \citeasnoun{authier2013early}) who
	conjectured on the structures of crystals.  \citeasnoun{hauy1800addition} created the first catalog of minerals.

\begin{figure}
	\includegraphics[width=\textwidth]{TimeLine_rev}
	\label{timeline}
	\caption{Some key dates in the history of modern crystallography}
\end{figure}

	With the creation of the 1978 version of the Chemical Information System (CIS)  \cite{bernstein1979nih},
	building on the earlier CIS \cite{feldmann1972application} which lacked cell search capabilities,
	online computerized retrieval systems become
	available for searching on ranges of unit cell parameters from
	the Cambridge Structural Database (CSD) \cite{kennard1977computer}. From 
	a few tens of thousands in 1979, the community has progressed to 
	collection of several hundreds of thousands of sets of
	unit cell parameters, with approximately one million in the
	CSD by 2019 \cite{taylor2019million} alone.  When placed in
	a common SAUC database (see below), the number of cells in the 
	Protein Data Bank and the Crystallographic Open Database total
	more than 600,000 entries as of summer 2022.   Below, we discuss algorithms for 
	robust searching on ranges of unit cell parameters. The need for
	ranges can be due to uncertainty of measurements, misidentified
	symmetry classes, variable composition, and searches for
	related materials.

	There have been multiple databases that allow searching
	on the values of unit cell parameters, including:
		\begin{itemize}
			\label{databases}

		\item Chemical Information System (CIS) (1978 version),
		 transferred to private management in 1984 \cite{kadec1985transfer}.

		\item Protein Data Bank, Rutgers Univ. \cite{Bernstein1977} \cite{Berman2000}.

		\item Cambridge Structural Database.  The WebCSD 
		\cite{hayward2019cambridge}	cell search uses the
		\cite{hayward2019cambridge} version of \GVI{}.

		\item American Mineralogist Crystal Structure Database \cite{downs2003american}.

		\item Crystal Lattice Structures. U.S. Naval Research Laboratory, now in the
		AFLOW library of crytallographic prototypes \cite{mehl2017aflow}.

		\item Crystallographic Database for Minerals and Their Structural Analogues, 
		Russian Foundation of Basic Research
		\cite{chichagov2001mincryst}.

		\item ICSD Web: the Inorganic Crystal Structure Database, FIZ Karlsruhe
		\cite{ruhl2019inorganic}.

		\item Crystallography Open Database, University of Cambridge \cite{gravzulis2012crystallography}

		\item  SAUC (Search of Alternate Unit Cells),  \cite{McGill2014}
	\end{itemize}

Various paradigms are used 
by different databases, sometimes only simple ranges on unit cell
edge length. SAUC uses the methods described herein.

\section{Introduction}

%	\cite{andrewsa2019faster}
Crystallographers, in general, seem convinced that unit cells
are simple; after all, we learn about them on the first day of
studying crystallography.

Likewise, lattices are considered to be simple. Of course
there are complications, such as space groups, but the concept
of the simple lattice is felt to be almost as simple as
unit cells.

Stumbling blocks arise when the need to compare various unit
cells is included in the tasks. For example, if the variations are due
to temperature changes or site substitutions, it is not
uncommon for the variations to be largely confined to a single
parameter, such as one unit cell axial length.

A different issue arises with the need to compare the unit
cells of an \textbf{a}rbitrary group of materials. For instance, one
might want to search in a database of all known materials. In
those cases, more sophisticated tools are needed.

What modern problems require comparing arbitrary groups of cells?
\begin{itemize}
	\item database searches (approximately a million unit cells are known)
	\item clustering for serial crystallography (thousands to hundreds of thousands of diffraction images in a single study)
	\item finding candidate materials for solving protein crystal structures 
	\cite{nanao2022id23}
	\item searches for minerals or alloys (both of which may have large compositional variation)
	\item Bravais lattice determination
	\item Studies in epitaxy \cite{yang2014unit}
\end{itemize}

\section{Techniques used to measure lattices}

Several techniques other than common linear algebra methods will be used in measuring distances between unit cells.
\begin{itemize}
	\item 	Metric vector spaces, see Section \ref{spaces}
	\item 	Simple topology, see Section \ref{topology}
	\item   ``Fundamental region'' (also called ``fundamental unit''),
	see Section \ref{fundamental}
	\item 	Unit cell reduction, see Section \ref{reduction}
	\item 	Projectors, see Appendix \ref{projectors}

\end{itemize}

\section{Unit cell parameters}

\label{unitcells}
If we are to measure the distance between two lattices, we will need
to have a space where the measure can be performed. The conventional 
representation of a unit cell is the ``unit cell parameters'', 
a, b, c, $\alpha$, $\beta$, $\gamma$. Mathematically, this is a
point in $\bf{R^3 \times R^3}$. That is the product of two different spaces; a
point in one is not necessarily a point in the other; worse in this
case is the fact that a point in the length space means nothing
in the angle space. It is still possible to define a distance, but 
that would require inventing a unit that mixes the units of the length
space and the angle space.

\section{Spaces}
\label{spaces}

In order to measure a meaningful distance between two objects, the objects need to
be placed in a ”metric space”. For crystallographic unit cells, this will often be a
vector space amenable to the tools and techniques of
linear algebra, at least locally. Because six free parameters are needed to define a unit
cell, the space will need to be at least six-dimensional. Depending on
other criteria, more dimensions may be needed; gluing the edges of
a manifold generates the need for more dimensions.

\section{A digression on topology}
\label{topology}

One aspect of topology is the study of objects and the relationships
of their edges, which can be ``glued'' together to make new objects
and disclose new relationships among the parts. Here we look at a simple 
example that demonstrates some of the issues that arise
in measuring distances between lattices.

\subsection{Create a simple, 1D unit cell}

\begin{figure}
	\includegraphics[width=\textwidth]{line_1}
	\label{line_1}
	\caption{This is just an empty single unit cell.}
\end{figure}

\subsection{Put two atoms into the unit cell}

	Here we put atoms at the ends of the cell so that they
	will be close to their counterparts in the adjacent cell.
\begin{figure}
	\includegraphics[width=0.4\textwidth]{line_2a}
	\label{line_2}
	\caption{Put an atom at or near each of two boundaries of the unit cell}
\end{figure}

\subsection{Connect the unit cell with its neighbors}

	Here we see that the ``correct'' distance between A and B is
	quite small compared to the distance within the 
	fundamental cell.
\begin{figure}
	\includegraphics[width=\textwidth]{line_3}
	\label{line_3}
	\caption{Here we show just the nearest neighbor cells}
\end{figure}

\subsection{Another choice: wrap around into a loop}
	A technique based in topology is to wrap one edge of a
	cell through a higher dimensional space to join
	another edge of the cell to make a loop. Doing this can make the 
	problem seem simpler, since there is only one cell
	to be considered. However, the problem of distance calculation has
	not been simplified. Now the distance has to be measured
	in both directions around the loop, and the shorter distance
	reported. This process of joining is termed ``gluing''. In
	addition, the distance must be measured within the curved
	surface.

     Gluing two edges of the one-dimensional cell
	turns the cell into a circle, making the problem
	two dimensional. 	If 
	we had started with a two-dimensional cell,
	then gluing opposite pairs of edges could be used to
	make a three-dimensional torus. In either case, the dimensionality of the system
	would be increased by one. 
\begin{figure}
	\includegraphics[width=\textwidth]{loop}		
	\label{loop}
	\caption{}
\end{figure}

\subsection{Or you could twist the loop to make a M{\"o}bius strip}

Consider the above case where the cell was on a 2-D surface.
Instead of simply joining the ends, we can twist the loop before
we glue the ends together, making a M{\"o}bius strip.

%The issue of dimensionality arises again here. The twist angle  adds yet another dimension, so the 3-D loop becomes an object that requires a fourth dimension to specify a point fully.

\begin{figure}
	\includegraphics[height=0.45\textwidth]{mobius_1}
	\label{mobius_1}
	\caption{Twisting before joining makes a M{\"o}bius strip}
\end{figure}

\subsection{Summary of the properties on joining}

One aspect of gluing is that the dimensionality increases. The issue of measuring distances also 
become more complex. With added dimensions, the directions in which to look for shortest paths may increase. Similarly
the glued boundaries themselves add to the cases to
be considered in looking for shortest paths.

\section{Spaces, continued}

In this paper, we use several spaces to represent unit cells.
Other choices of spaces are possible, but this set will suffice 
for the current topics.

\begin{itemize}
	\item 	Lengths in \r{A}ngstroms and angles in degrees, a 
	space of $R^3 \times R^3$).
Although the presentation of unit cell parameters as edge lengths and
interaxial angles is familiar, it does not present an obvious metric (see Section \ref{unitcells}). A
possible alternative would be to treat each length and angle pair as
a point in polar coordinates. This unexplored variation makes use
of the association of each axis with its related angle,\textit{ e.g.} \textbf{a}
and $\alpha$ (see appendix \ref{polar}).
	\item 	\GVI{} (Niggli space)
	(see appendix \ref{G6})
	\item 	\SVI{} (Delone space)
	(see appendix \ref{S6})
	\item 	\CIII{} 
	(see appendix \ref{C3})
%	\item 
\end{itemize}

For future reference, we are in the process of extending 
this list by consideration of a seven-dimensional space of Wigner-Seitz cells \cite{Bernstein2023}.

	% Appendices appear after the main body of the text. They are prefixed by
	% a single \appendix declaration, and are then structured just like the
	% body text.

	\section{Fundamental region}
	\label{fundamentaltext}

	In working with a space in which there are 
	multiple symmetry-related occurrences of items of interest (atoms, molecules, or,
	in this case, unit cells), it is conventional to choose a fundamental
	region \cite{enwiki:1028832752}. In a two-dimensional graph, it would often be chosen as the all-plus
	quadrant. In crystallography, it might be a unit cell, a
	particular asymmetric unit of a crystal, or even a molecule.

	The conventional fundamental unit for Niggli space is the region
	of \GVI{} that contains the Niggli-reduced unit cells. This choice
	has the disadvantage that the fundamental region is non-convex,
	and the boundaries are complex \cite{Andrews2014}.

	For Delone space, the fundamental unit is conventionally chosen
	as the all-negative orthant of \SVI{}, which contains all 
	reduced cells. In this orthant, all non-boundary 
	points (that is, those having no zero scalars) correspond to
	a computable unit cell. All points with one or two zero scalars also
	correspond to valid unit cells. Boundaries that have
	four, five, or six zeros have no valid cells. For three zeros, there are
	valid cells except for those points with a prototype of
	$s_1, s_2, s_3, 0, 0, 0$ \cite{Andrews2019b}.

	The all-negative orthant (excluding the boundaries) is 
	an ``open'' manifold in topological terms; all points
	there correspond to exactly one lattice. The fact that
	certain boundaries do not correspond to valid unit
	cells complicates certain aspects of its topology.

	\section{Measuring}

	Measuring the distance between two lattices
	comes down to finding a geodesic between them, {\it i.e.}
	a minimal length path between them obeying the rules for points
	in the space. We first give a simple example of points in
	a 2-D manifold, which demonstrates some of the complexity of treating lattices.

	\subsection{Step 1, fundamental unit}
	\label{fundamental}

	The obvious step to find the shortest distance between two
	lattices is to simply compute the Euclidean distance between
	them in the fundamental unit. That distance might not be the
	shortest that can be found, but no geodesic could 
	be longer. It makes a useful starting point. Call 
	this distance $d_0$.

	\subsection{Boundaries}

	The next step is to consider the boundaries of the fundamental
	unit. If there is no boundary with a shorter distance to
	 \textit{either} of	the lattices than $d_0$, then $d_0$ is the required minimum.

	If either lattice does have a boundary closer to it than
	$d_0$, call the shortest distance from the first lattice
	to any boundary $d_1$ and the shortest distance
	from the second lattice to any boundary $d_2$.  If 
	$d_1 +d_2 > d_0$ then $d_0$ is still the required
	minimum and we are done.   It is only when both lattices
	are closer than this to some boundaries that we need to
	engage in a more complex search for geodesics.

	As a simple example of treating boundaries, consider a plane with a single
	symmetry element, two-dimensional point group 4.

	\includegraphics[width = 0.5\textwidth ]{4_1}

	We choose the lower left quadrant to be the fundamental unit.
    This is an arbitrary choice.

	\includegraphics[width = 0.5\textwidth ]{4_1_1}

	Now put a single atom, A, in the plane, and apply the
	4-fold operation.

		\includegraphics[width = 0.5\textwidth ]{4_2}

	Now let us consider what happens if we place three atoms, A, B,  and C in the fundamental unit.

\begin{figure}
		\includegraphics[width = 0.5\textwidth ]{4_3}
		\label{4_3}
	\caption{}
\end{figure}	
	It is clear that within the fundamental unit, ignoring all symmetry
	transforms and considering B and C, B is the atom closer to
	A. Draw a circle centered on A that touches B. We see that the
	circle crosses the boundary and a portion is outside of the
	fundamental unit. Any atom that is within that region outside the fundamental unit must be examined to see if it contains
	an atom.

	\includegraphics[width = 0.5\textwidth ]{4_5}

	Applying the 4-fold operation to the shaded segment, we
	see that the atom C (in the fundamental unit) is closer to A
	than B is to A.

	\includegraphics[width = 0.5\textwidth ]{4_5_1}

	The next part of the problem is how to compute the actual 
	distance from A to C. The obvious answer is to apply the 
	4-fold symmetry operation either to A or to C and measure
	the Euclidean distance. We will see in the next section that
	such simplistic solutions are not sufficient for lattices.

	\subsection{Spaces for measuring}

	From this point on, only two spaces will be used for
	specific examples, \SVI{}, the
	space of Selling scalars as six real vector components,
	$( P,Q,R,S,T,U )$ = 
	({$\mathbf{b} \cdot \mathbf{c}$},
	{$\mathbf{a} \cdot \mathbf{c}$},
	{$\mathbf{a} \cdot \mathbf{b}$},
	{$\mathbf{a} \cdot \mathbf{d}$},
	{$\mathbf{b} \cdot \mathbf{d}$},
	{$\mathbf{c} \cdot \mathbf{d}$}) where {$\mathbf{d}=-\mathbf{a}-\mathbf{b}-\mathbf{c}$}, 
	 and \CIII{}, the same scalars as \SVI{}, but used as the real and imaginary parts
	of three complex vector components, $(P+iS, Q+iT, R+iU)$, with each complex number written as a column two-vector.  See sections \ref{S6} and
	\ref{C3} in the appendices. \CIII{} is simply an alter ego of \SVI{}, but 
	some operations are simpler to describe in \CIII{}. The
	reason for limiting our consideration to these two spaces is their simplicity
	compared to other spaces. As stated above, the fundamental
	unit of \SVI{} is a simply-connected, convex manifold, with
	only a few excluded regions (multiple zeros).
	%	\textbf{might need to add the section here about measuring only with the metric of the fundamental unit.}

	For any given lattice there are infinitely many choices of cells from that lattice that can be used to represent it.  The point of reduction is to unambiguously choose a single one of those choices of cells.  In the absence of
	experimental error that is what, for example, Niggli reduction does.  In the presence of experimental error things are not
	so simple and any of several equally valid,
	but very different, unit cells may end up being chosen
	to represent the same lattice \cite{Gruber1973} \cite{McGill2014}.

	\subsection{Reflections, a simple example}

	Reflections in \SVI{} were defined in \citeasnoun{Andrews2019b}. They are
	the operations that create isometric copies of a lattice, always within the
	fundamental unit. When comparing two lattices, it is necessary to 
	examine the distances from a reduced cell in one lattice to each of the reflections of another reduced cell in the other lattice.

	Returning to Figure \ref{4_3} (where the distance to objects
	outside the fundamental unit were considered), consider instead the
	 two-dimensional point group 4mm. Figure
	\ref{4mm_3} shows a subset of the reflections of the basic set of three points.
	Immediately, it is clear that in this case, the 
	distance between two copies of
	point A is the shortest.

\begin{centering}
\begin{figure}
	\includegraphics[width=.8\textwidth]{4mm_3}
	\label{4mm_3}
	\caption{Recreating the above example, but with point group 4mm. }
\end{figure}
\end{centering}

\subsection{Reflections in \SVI{} and \CIII{}}

``Reflections'' were defined by\citeasnoun{Andrews2019b} as
the 24 operations that create different, reduced versions of
a reduced cell in \SVI{} and \CIII{} (isometric copies).
By this definition, the 
4-fold operation and the mirrors in point group 4mm (see Figure \ref{4mm_3})
are reflections.

	An important difference between Niggli space and Delone space
	is that Niggli space has the vector scalars sorted by 
	somewhat complex rules. \SVI{} (that is, Delone space)
	being unsorted results in there being multiple points in the
	fundamental unit for each lattice. In general, there are 24 
	reflections, but equal scalars and zero scalars can lead to degeneracies.

	The reflections are simpler to describe in \CIII{}, but,
	of course, the same ones exist in \SVI{}. Here we use the \CIII{} vectors: $c_1$, $c_2$, and $c_3$, see Appendix \ref{C3}.

First there are the three interchanges of scalars: $c_1\Longleftrightarrow c_2$, $c_1\Longleftrightarrow c_3$, 
and $c_2\Longleftrightarrow c_3$.

There are also reflections that interchange elements within 
scalars.
\begin{figure}
		\includegraphics[width=0.5\textwidth]{C3_ex}
%$\begin{bmatrix}$
%\begin{tabular}{ c c l c }
%		$\begin{pmatrix}
%		s_1 \\
%		s_4 
%	\end{pmatrix}$ 
%	$\begin{pmatrix}
%		s_2 \\
%		s_5
%	\end{pmatrix}$  &
%%\begin{centering}
%			\includegraphics[width =0.5cm ]{arrows3}
%%\end{centering}
%&
%
% 
%$\begin{pmatrix}
%		s_3\\
%		s_6
%	\end{pmatrix}$
%\end{tabular}
%$\end{bmatrix}$

	\caption{Example of interchanging real and imaginary scalars
	for the case of reduction of scalar $s_1$.}

\end{figure}

And finally each pair of the reflections needs to be combined
giving a total of 24.

\subsection{Reduction}
\label{reduction}

The objective of reduction in \SVI{} (and in \CIII{}) is to
make each Selling scalar less than or equal to zero.
If any scalar is greater than zero, there are two stages
to a reduction step. First,
a vector is composed and added to the vector being reduced.
Then two scalars are interchanged as shown in Figure \ref{S6_ops_B}.

While any scalar can be the one to be reduced, this figure shows
the operations for $s_1$.

\begin{figure}
	\includegraphics[height=0.65\textwidth]{S6_ops_C}
	\label{S6_ops_B}
	\caption{Steps for reduction in \CIII{}}
\end{figure}

	\subsection{Boundary transformations}

	The transformations for crossing boundaries of the 
	fundamental unit are the reduction operations.

		The section on fundamental regions (see Section \ref{fundamentaltext}) shows the problem of points outside
	the fundamental unit and how boundary transformations were applied
	there. The boundary transformations for lattices are more complex
	than those of space groups.

\begin{enumerate}
	\item  The boundaries must be identified.
	\item  The transformations at the boundaries in \GVI{} and
	 	\SVI{} are not isometric, meaning that they do not preserve
	 	distance measure.
	\item  For a particular boundary, there may be more than one
	 	allowed transformation available.
	\item  Applying a transformation to a point near a boundary may generate
	 	a new point that is near another boundary, requiring further
	 	searching.
	 \item A point that is in the fundamental unit but
	 close to a boundary might need to be transformed in order to
	 find other nearby points.
	 \item All measurements must be done in the metric of the
	 fundamental unit. A corollary is that when two points are
	 being treated, one must be held fixed and within the 
	 fundamental unit while the possible boundary transformations 
	 and reflections of the other are treated.
	 \item If a point is near more than one boundary,
	 then there are multiple transformations and reflections
	  to consider.
\end{enumerate}

	The fundamental unit in \SVI{} is the all-negative orthant. The
	boundaries are all where one (or more) of the scalars equals zero.
	Two points that are close together and have a corresponding scalar that is positive 
	in one and negative in the other may have representations in 
	the fundamental unit that are far apart.

	In Figure \ref{DUAL}a,
	the two points with $\gamma=89.8^{\circ}$  and $\gamma=90^{\circ}$ are far apart.

	Figure \ref{DUAL}b demonstrates a further complexity. Because the angles $\alpha$
	and $\beta$ were chosen to be $90^{\circ}$, all points in Figure \ref{DUAL}a sit
	on two boundaries ($beta$=$90^{\circ}$ and  $\gamma$=$90^{\circ}$, 
	$s_4$=$0$ and $s_5$=$0$). Perturbing those boundary points may cause them to be within
	the fundamental unit, but, on the other hand,  may cause them to be outside and thus be far from
	the neighboring points (after reduction). All the gray lines
	correspond to members of lattice pairs that are close together
	but are far apart in the fundamental unit (that is, Delone-reduced).

\begin{figure}
		\includegraphics[width =0.8\textwidth ]{DUAL}
		\label{DUAL}

	\caption{
	\captionpar{}In both figures (a) and (b), a series of points has been generated in \SVI{}. The basic cell is a=10, b=10, c=10, $\alpha=90^{\circ}$, $\beta=90^{\circ}$, and $\gamma$ ranges sequentially from $10^{\circ}$ to $170^{\circ}$, in $0.2^{\circ}$ increments. Each cell was Delone-reduced. Each was then placed in the \CIII{} asymmetric unit \cite{Andrews2019b}. Then, in sequence from the first cell, for the succeeding cell, all 24 reflections of the succeeding cell were examined, and the reflection that is closest to the first cell chosen. Calculation then proceeds to the next pair (where the new first is the second chosen in the previous step). A line (shown in gray) was drawn between each successive pair of points. The individual points are shown as circles.
	\captionpar{}~~
	\captionpar{}This figure shows the points converted to \protect\CIII, and the c3 coordinate is displayed.  In figure (a) on the left the points are generated as described. The gray connection lines, except for the one connection $\gamma=90^{\circ}$ to $\gamma=89.8^{\circ}$, are hidden by the circles. That single long line represents a case when distance measured in the fundamental unit is quite long, although the originally generated points are quite close together. In figure (b) on the right, initially, three copies of each point were generated, and all points were perturbed by 1\% in a random direction orthogonal to the generated vector (point). Then (in the order of generation), the above sequencing was performed.
    \captionpar{}~~
    \captionpar{}Because the basic structure of the point generation requires that  all points be on multiple boundaries, there are many cases where the distances in the fundamental unit are long. Every gray line is a case where the simple Euclidean distance in the fundamental unit is long, but the generated points are close together.
	}
\end{figure}

As seen in Figure \ref{DUAL}, points that are a small distance apart may
be far apart in the fundamental unit. As an example, we choose a primitive
cubic unit cell, \SVI{} vector $\begin{bmatrix}{ 0, 0, 0, 10, 10, 10}\end{bmatrix}$. In Figure \ref{perturbed},
1000 copies have all been perturbed by 1\% in a direction orthogonal to that vector.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{perturbed}\label{perturbed}

	\caption{1000 copies of \SVI{} vector $\begin{bmatrix}{ 0, 0, 0, 10, 10, 10}\end{bmatrix}$ were 
	perturbed by 1\% orthogonal to the vector. As in Figure \ref{DUAL}, for
	each pair of points, the second point was chosen to be the reflection that
	is closest to the first point. The choice of reflection pairs was done by
	treating the list of generated points as a sequence. Long lines indicate
	pairs that are far apart within the fundamental unit. }
\end{figure}
\section{Boundaries (and reduction) in \SVI{} and \CIII{}}

	\subsection{Multiple boundaries}

	If a point is near two boundaries, all the cases for each boundary need to be considered (recall
	that in \SVI{} there are four possible boundary transformations at
	each of the six boundaries).
	In addition, applying the transformations of one of the boundaries
	and then those of the second is necessary. That means that
	there are 16 possible results. Further still, the 
	case of applying them in the other order must be considered (resulting
	in another 16 cases).

	For the case of being near three boundaries, similar complexity to
	the case for two boundaries must be considered: all single 
	boundaries, all boundary pairs (in both orders) and all 
	triples (in all possible orders).

	\subsection{Repeated boundaries}

	In some cases, applying a boundary transformation results
	in one of the modified scalars being near a boundary
	that it was not near before. Such a case requires that the
	new neighbor boundary also be considered.

	\subsection{Iteration}

	Much of the complexity of determining the shortest distance between two
	lattices is due to the need to repeat steps of reduction and reflection.
	Among the places in the process where this becomes necessary is when a
	reduction step creates a new point that is near a boundary.

	\subsection{Measuring lattice distances in \SVI{}}

\begin{figure}
\includegraphics[height=0.6\textwidth]{S6_1}
\caption{In space \SVI{}, choose two axes. The operations shown
will be a cartoon of the operations in six-space.}
\label{S6_1}
\end{figure}

\begin{figure}

\includegraphics[height=0.6\textwidth]{S6_2}
\caption{Define a fundamental unit (all non-positive region).}
\label{S6_2}
\end{figure}

\begin{figure}
\includegraphics[height=0.6\textwidth]{S6_3}
\caption{Put in a unit cell (a point in \SVI{}).}
\label{S6_3}
\end{figure}

\begin{figure}
\includegraphics[height=0.6\textwidth]{S6_4}
\caption{Define a search radius. In this case, it
	crosses a boundary and a region is outside the
	fundamental unit. Cells there are \textit{not} reduced.}
\label{S6_4}
\end{figure}

\begin{figure}
\includegraphics[height=0.6\textwidth]{S6_5}
\caption{The non-reduced region is marked in crosshatch.}
\label{S6_5}
\end{figure}

\begin{figure}
	\includegraphics[height=0.6\textwidth]{S6_6}
	\caption{	Unlike in the 4-fold example, there are multiple 
		(up to 4) 
		places to go. A further complication is that the transformations
		at the boundaries do \textbf{not} preserve distances. Distances
		must be measured in the frame of the fundamental unit.}
	\label{S6_6}
\end{figure}

\begin{figure}

\includegraphics[height=0.6\textwidth]{S6_7}
\caption{Sometimes, in the process, another boundary is crossed. Then further iterations may be required.}
\label{S6_7}
\end{figure}

The distance between two lattices is determined by first computing $d_0$, the minimum of all the Euclidean distances between reflections of those two points within the fundamental unit.  If either lattice is closer to any boundary than $d_0$, then we need to iterate through all possible paths that may cross a boundary.

If the second point is near a boundary, then the four transforms for that boundary are applied. We need to compute the tunneled distances. If the generated points are not near another boundary, then we are done. If they are near a new boundary, then we need to iteratively apply the boundary transforms.

\begin{enumerate}
	\item Reduce both unit cells.
	\item Choose one point (``first point'') to be the fixed point, and the other (``second point'') to be the one to which transformations are applied.
	\item Compute $d_0$, the minimum of all the Euclidean distances between the first point and all reflections of the second point (the reflected points are all within the fundamental unit).
	\item Compute $d_1$, the minimum of all the Euclidean distances between the first point and all boundaries.
	\item Compute $d_2$, the minimum of all the Euclidean distances between the second point and all boundaries.
	\item If $d_1+d_2 < d_0$, then there may be a path between
	some reflection of the first point and  some reflection of
	the second point that is shorter than $d_0$ that crosses a boundary.  It is sufficient to explore possibly shorter
	paths by leaving the
	first point fixed and only considering reflections of the
	second point.  For each boundary to be considered there are
	at most four possible paths from whichever of the two points
	is close enough to that boundary to satisfy $d_1+d_2 < d_0$.
	For such a case we need to iterate through all four possible paths that may cross the boundary.
	\item If the second point is near two or three boundaries, then each boundary needs to be treated by itself. Then all pairs (and all triples) in all permuted orders must be transformed through the boundaries by all four paths of each boundary.  For a given \SVI{} boundary there are four possible boundary transforms to consider which may take a reflection
	of the second point outside of the fundamental unit (see Figure \ref{S6_6}).
	\item If the points produced in the above steps are near new boundaries, then the above steps need to be iterated using the new points. 
	\item N.B., the distance must be measured in the metric
	of the fundamental unit; see section \ref{measure}.
\end{enumerate}

In the above discussion of gluing in topology, mention was made of the
increase in dimensions as edges are glued and more if there are twists.
Here, the boundary transformation operations interchange two scalars, and
that is a twist operation. Counting up the gluings and twists leads to
the conclusion that the required dimension of the space in which to embed the glued manifold is quite high; immediately,
we understand that there are potentially going to be many paths to examine.

\subsection{Measuring with the metric of the fundamental unit}
\label{measure}
The problem that arises in computing distances between lattices
is that it is necessary to transform points to be outside the 
fundamental unit. The boundary transformations are not
isometric; distance measure is not preserved on crossing out
of the fundamental unit. Here we describe how to create a measure
for those points that is isometric.

If the second point of a measurement is near a boundary, then project the point
onto the boundary (the dashed arrow in Figure \ref{TMB}) and
extend the line segment from the second point to the boundary
to twice the length creating a transformed point creating
an artificial mirror point. The ``Tunneled Mirrored Boundaries'' distance is found by locating the crossing point on the boundary where
the line from the first point to the artificial mirror point
crosses the boundary and then finding the four boundary transforms
of the crossing point.
Applying a boundary transform to
the crossing point provides an alternate boundary point
to be used for a broken path (first point to crossing point
point -- equivalent to boundary-mapped crossing point, followed
by mapped crossing point to second point).  This broken
path is sometimes shorter than the direct $d_0$ path and
is entirely in the fundamental unit.  If two boundaries
are close, then similar actions on both lattice points are 
needed.

\begin{figure}
	\includegraphics[width=.9\textwidth]{TMB}
\label{TMB}
	\caption{Tunneled Mirrored Boundary}
\end{figure}

\subsection{Algorithm}

	For finding the shortest distance between two reduced lattices, the first step
	is to compute the simple Euclidean distance between them.
	Each lattice is considered as a point in a space of six or
	more dimensions with a metric derived from the cell reduction
	rules for that space.

	If the original second point is near one (or more) boundaries, then the 
	the boundary searches (above) are performed.

	If new near-boundary-contacts are detected, then the same kind of searches
	is performed.

	Distances must be measured in the metric of the fundamental unit 
	\textit{not in the metric of any region outside the fundamental unit}.

	\subsection{Nearest neighbors, the ``Post Office Problem''}

	Up to this point, the discussion has targeted finding the shortest 
	distance between two lattices. Sometimes, the problem is to find the collection of points
	that are near a particular probe. One important use is for
	creating and for searching within
	databases of crystals. Known as the nearest neighbor problem, it is also
	known as the post-office problem \cite{Andrews2001}, and \cite{Andrews2016}. (see Appendix \ref{nearest}.)

\section{Determining Bravais lattice types}

\subsection{Notation}

\begin{figure}
		\includegraphics[width=0.3\textwidth]{PQRSTU} 
	\label{PQRSTU}
	\caption{Bravais organized the scalars into a ``tetrahedron''\cite{Delaunay1932}.}
\end{figure}
   The International
		Tables for Crystallography define this assignment of the scalars (PQRSTU) in
		the tetrahedron \cite{donnay1953international}. Other authors used alternative assignments.

		As \SVI{}: \textbf{s} = $\begin{pmatrix}{P,Q,R, S,T,U}\end{pmatrix}$\\ = $\begin{pmatrix}{s_1, s_2, s_3, s_4, s_5, s_6}\end{pmatrix}$.\\

		As \CIII{}:
		$\begin{bmatrix}
			\begin{pmatrix}
				P   \\
				S
			\end{pmatrix}
			\begin{pmatrix}
				Q   \\
				T
			\end{pmatrix}
			\begin{pmatrix}
				R   \\
				U
			\end{pmatrix}
		\end{bmatrix}$

	\subsection{Bravais lattice types in \SVI{}}

	\citeasnoun{Delaunay1932} and \citeasnoun{Delone1975} produced
	a table classifying the reduced forms of the Bravais types as 24 cases.
	Figure \ref{s6table} shows the same classification (the 
	letter codes for crystal systems have been changed to modern notation. For some types, revised names have been given
	 to Delone's types where a cell included two types (for example,
	Delone's ``M1'' is split into ``M1A'' and ``M1B''). In Figure \ref{s6table}, the 
	Bravais type is listed, and the restrictions on \SVI{} scalars is shown
	graphically as the Bravais tetrahedron and listed as the lattice character.

	Figure \ref{Btype} describes the contents of each table entry
	in Figure \ref{s6table}.

	\begin{figure}
		\includegraphics[width=0.75\textwidth]{DeloneTypesGrid_3}
		\caption{The table of \citeasnoun{Delaunay1932} describing the 24
			Bravais types in \SVI{}. It has been redone removing the images of the 
			Dirichlet cells, the non-reduced cells, and adding the ``lattice character'', which
			describes the linear manifold of each type. The crystal family types have been
			renamed to modern usage: Q changed to T for tetrahedral, K changed
			to C for cubic, and T changed to A for anorthic. Where Delone
			in some places included two types in one table cell, they have been
			split into two (for example: ``M1'' is split into ``M1A'' and ``M1B''.  Note that five types (O3, M3, M2B, A2, and A3) are not normal 
			crystallographic types. They are boundary types, and they have
			fewer free parameters than the generic type requires. For instance,
			O3 (character: rs0 rs0) has only two free parameters (r and s), whereas
			an ordinary orthorhombic type requires three variables.}
		\label{s6table}		
	\end{figure}

	\begin{figure}
		\begin{center}
			\includegraphics[width=0.4\textwidth]{C1_square}
			\includegraphics[width=0.4\textwidth]{C5_square}
			\label{Btype}
			\caption{Explaining the information in Figure \ref{s6table}; it includes the
				name of the Delone type (\textit{e.g.}: C1),the Bravais type
				(\textit{e.g.}: cI for body-centered cubic), and the lattice character that describes
				the linear manifold of each Delone type (\textit{e.g.}: 
				\mbox{000 rrr}).
				The markings on the tetrahedra describe the restriction on the \SVI{}
				scalars; they encode the same information that is in the lattice
				character. A zero (circle) means that the value of that scalar is
				zero, several with the same lines means they are constrained to be
				equal; in Figure \ref{Btype}, they are a pair of double
				line hatches.}

		\end{center}

	\end{figure}

\subsection{How to determine Bravais lattice types}

			First, the projectors for each type are necessary.  There are two ways to determine the Bravais lattice type.
			\begin{enumerate}
				\item The program BGAOL uses methods essentially the same as those used for distance calculations. 
				For a given input unit cell, the reflections and boundary transformations are applied. The 
				original cell and each of the transformed cells, is projected onto each of
				the manifolds of the lattice types (24 in the cases of \SVI{}). The distance from
				the probe cell to the projected cell is the agreement factor. For each lattice type,
				the cell with the shortest distance is reported.
				\item The program SELLA \cite{andrews2022unnecessarily} starts from vectors representing each of the lattice
				types and transforms them using the reflections and first layer boundary transformations.
				(In the case of multiple boundaries, all combinations must be used as in distance calculations).
				Then the distance from the reduced probe to each of the transformed manifolds of each type
				is computed. The shortest distance to one of the transformed manifolds is reported. If
				needed, the inverse reduction of the input (unreduced) unit cell can be applied to the 
				best projected cell to get the best approximation of the input.

%				\item For each type, transform the character by the reflections and by the boundary transformations. In cases where there are two zeros in the character, the products of the boundary transformations (in both orders) are needed. For the cases of three zeros, all pairs and all triples in all orders are required.
%				\item For a given unit cell: for each type compute the minimum distance to the list of different representations of the type. Choose the minimum.
%				\item For a given unit cell, for each reflection and for each boundary transformation, compute the minimum distance to each type. In some cases, this may require some iterations of products of reflections and boundary transformations. For cases where two or three boundaries are ``near'', products of the transforms and the reflections are necessary.
			\end{enumerate}

	\subsection{What do you need to compute distances?}

\begin{itemize}
	\item 		Reflections
	 		\begin{itemize}
		 			\item 	\GVI{} \cite{Andrews2014}
		 			\item \SVI{} \cite{Andrews2019b}
		 		\end{itemize}
	\item	Boundary transforms
	 		\begin{itemize}
		 			\item 	\GVI{} \cite{Andrews2014}
		 			\item 	\SVI{} \cite{Andrews2019a} (only two, following Delone, but reflections generate a total of 4)
		 		\end{itemize}
	\item 		Bravais lattice projectors
	 		\begin{itemize}
		 			\item 	\GVI{} \cite{paciorek1992projection}
		 			\item 	\SVI{} \cite{andrews2022generating}
		 		\end{itemize}
	\item 		Lattice characters
	 		\begin{itemize}
		 			\item \GVI{} \cite{Andrews2014}
		 			\item \SVI{} see Section \ref{characters}
		 	\end{itemize}
\end{itemize}

\subsection{\SVI{} Bravais lattice characters}
		\label{characters}

Each Bravais type defined by \citeasnoun{Delaunay1932} corresponds
to a linear manifold in \SVI{}. The definitions of each type
were described by markings on the elements of the tetrahedra
for each type in Figure \ref{s6table}. For implementation
in software, it is
more convenient to use the ``lattice characters'' that describe
the restrictions.

\begin{center}
	$\begin{tabular}  {lll lll lll lll}

		\begin{tabular}{l l}
			type & Bravais type \\
&\& character \\
			C1&cI (rrr rrr) \\ 
			C3&cF (rr0 rr0) \\ 
			C5&cP (000 rrr) \\ 
			T1&tI (rrs rrs) \\ 
			T2&tI (rr0 rrs) \\ 
			T5&tP (000 rrs) \\ 
			R1&hR (rrr sss) \\ 
			R3&hR (rr0 sr0) \\ 
			H4&hP (00r rrs) 
		\end{tabular}&&

		\begin{tabular}{l l}
			type & Bravais type \\
			&\& character \\
			O1A& oF (rrs rrt)               	  \\
			O1B& oI (rst rst)              	  \\
			O2&oI (rs0 srt)               	  \\
			\textit{O3} &\textit{oI} \textit{(rs0 rs0)} \textit{*}	\\
			O4&oS (00r sst)               	  \\
			O5&oP (000 rst)               	  \\
			O4&oS (00r sst)               	  \\
			O5&oP (000 rst)               	  \\
			&\\
		\end{tabular}&&

		$\begin{tabular}  {ll}       
			type & Bravais type \\
&\& character \\
			M1B &mC (rst rsu) \\
			M2A &mC (rs0 stu) \\
			\textit{M2B}& \textit{mC}  \textit{(rs0 rst)} {  *} \\
			\textit{M3}& \textit{mC}  \textit{(rs0 ts0)} {  *} \\
			M4 &mP  (00r stu) \\
			A1 &aP  (rst uvw) \\
			\textit{A2}& \textit{aP}   \textit{(rs0 tuv)} {  *} \\
			A3 &aP (rs0 tu0) {  *} \\
			M1A& mC (rrs ttu) 
		\end{tabular} $

	\end{tabular} $\\
{ *} cases of non-crystallographic \SVI{} scalars ($90^{\circ}$ angles)
\end{center}

\subsection{Lattice Matching}

A common problem in crystallography is to provide a list of the
unit cells of several (or many) crystals so that they can be visually
compared, making it easier to identify meaningful clusters
of crystals of related morphology. Data collections of
unit cells have been created based on similarity of morphology
(for example, see \citeasnoun{Donnay1963}.  In recent years,
the clustering of unit cells from the myriad of images in serial
crystallography has been increasingly important.

For each transformed vector, one must store the transformation that will bring it back to the original presentation.

Start with a collection of experimental unit cells. From
among them, we select or create the “reference” cell; that is the
one to which all the rest will be matched as closely as possible.

We transform the reference cell by many operations in the
course of exploring alternative lattice representations. For each
newly generated lattice representation,we accumulate the transformations
needed to convert back to the original reference cell.
All of these operations are be performed in \SVI{}.

Only accumulate transformations
that have not already been found (to avoid duplication).

Next all the accumulated, transformed representations of the
reference cell must be scaled  to a common length and the saved transformations
inverted which allows one to return a vector to
its original presentation.  For improved searching
in the final step, it is convenient to store the  the vectors in a nearest neighbor search
application such as NearTree \cite{Andrews2016}. This ends the 
preprocessing of the reference lattice.

To match a probe lattice to the reference cell, first the 
probe lattice should be
Selling reduced. Then the nearest point of the
collection of transformed lattice points to the reduced probe
is found. The stored transform of the nearest reference is
applied to the reduced probe to generate the best approximation.

As an example, consider the case of krait toxin phospholipase A2
\cite{LeTrong2007}. Table \ref{table:LT1} shows 6 reported forms
from the Protein Data Bank. The lattice type and unit cell parameters
are as initially reported.

Table \ref{table:LT2} shows the same data in the same order after 
lattice matching to the first entry (1DPY). The immediate conclusion
is that the crystals are likely the same crystal form.

\begin{table}
	\begin{center}
		\caption{Unit cells of phospholipase A2 from the Protein Data Bank.}		
		\vspace{3mm}
		\begin{tabular}{lccccccccccc}
			\toprule
			PDB &   & a &b &c&$\alpha$&$\beta$&$\gamma$ \\
			\midrule
			1DPY &R& 57.98& 57.98 &57.98 &92.02  &92.02 &92.02 \\ 
			1FE5 &R& 57.98& 57.98 &57.98 &92.02  &92.02 &92.02 \\ 
			1G0Z &H& 80.36& 80.36 &99.44 &90     &90    &120   \\ 
			1G2X &C& 80.95& 80.57 &57.10 &90     &90.35 &90    \\ 
			1U4J &H& 80.36& 80.36 &99.44 &90     &90    &120   \\ 
			2OSN &R& 57.10 & 57.10  &57.10  &89.75  &89.75 &89.75  \\ 
			\bottomrule
		\end{tabular}
		\label{table:LT1}
	\end{center}
\end{table}

\begin{table}
	\begin{center}
		\caption{The data of Table \ref{table:LT1} matching a rhombohedral reference. The 
			reference cell is marked in boldface.}
		\vspace{3mm}
		\begin{tabular}{lccccccccccc}
			\toprule
			PDB &   a & b & c & $\alpha$ & $\beta$ & $\gamma$ \\ \midrule
			\textbf{1DPY} & \ \textbf{57.98}  & \textbf{57.98}  & \textbf{57.98}  &
			\textbf{92.02}  & \textbf{92.02}  & \textbf{92.02}  \\ 
			1FE5 &57.98  &57.98  &57.98  &92.02 & 92.02  & 92.02  \\ 
			1G0Z &57.02  &57.02  &57.02  &90.39  &90.39 &89.61  \\
			1G2X &57.10  &57.11  &57.11  &89.25  &90.25  &90.25  \\
			1U4J &57.02  &57.02  &57.02  &90.39  &90.39  &89.61  \\ 
			2OSN &57.10  &57.10  &57.10  &90.25  &90.25  &89.75  \\ 
			\bottomrule
		\end{tabular}		
		\label{table:LT2}
	\end{center}
\end{table}

%\begin{tabular}{l}
%	\begin{tabular}{l c c c c c c c }
%	c 10 15 20 90 110 90\\
%	r 15 15 15   95 95 95\\
%	r 15 15 15   85  85 85
%	\caption{An example of finding a C-centered representation of two rhombohedral lattices}
%\end{tabular}\\
%
%\begin{tabular}{l c c c c c c c }
%	C   10.000  15.000  20.000  90.000 110.000  90.000    REFERENCE\\
%	C   20.268  22.118  26.725  90.00 146.18010  90.00000\\
%	C   15.000  34.690  20.268  73.015 132.500 120.515
%	\caption{}
%\end{tabular}
%\label{LM}
%\caption{An example of lattice matching}
%\end{tabular}\\
\section{Various available programs}

	\itemize{
		\item SAUC: Search for related unit cells
		\item BGAOL: Determine likely cells of higher symmetry
		\item Follower: Compute distances between cells along a path
		\item LatticeMatching: Try to match a reference cell
		\item SELLA: Determine likely cells of higher symmetry
		\item various cmd programs\textendash
	}
\subsection{SAUC (Search of Alternative Unit Cells)}

	\textbf{SAUC} is software plus an online database of the Protein Data Bank unit cells and
the Crystallographic Open Database unit cells. \\
\url{https://iterate.sourceforge.net/sauc/}\\
SAUC provides searching for several alternate distance measures,
including \SVI{} and \GVI{}. Searching by nearest neighbor
allows specifying various criteria, such as percent range, 
nearest neighbors, or within a sphere of specified radius.\\

A useful example is that of krait toxin phospholipase A2
\cite{LeTrong2007}.

Starting from PDB entry 1u4j with cell parameters: (80.36, 80.36, 99.44, 90, 90, 120)	in space group R3 with a distance cutoff
of 3.5\AA  ~using the \GVI{} metric, four structures of phospholipase A2 are found.
\vspace{.2cm}

$\begin{tabular}{l l l}
	PDB code & G6 distance & protein name \\
	1u4j &  0.0 & {Phospholipase~A2~isoform 2} \\
	1g0z & 0.0 & Phospholipase~A2 \\
	1g2x & 0.9 &  Phospholipase~A2 \\
	2osn  & 0.9  & Phospholipase~A2~isoform 3 
	\label{phospholipase}
	\caption{}
\end{tabular}$

\vspace{.2cm}

\subsection{BGAOL - Bravais General Analysis of Lattices}

	\textbf{BGAOL} is an online program for seeking possible Bravais lattice types for a given  unit cell.

	\url{https://iterate.sourceforge.net/bgaol/} \\

  Consider 1,8-terpin \cite{Mighell2001} with unit cell 
10.912(2), 22.79(4), 10.705(2), 90, 120.64, 90.

The BGAOL results (only the best 4, which are also 1,8-terpins) are:

\begingroup
\setlength{\tabcolsep}{1.6pt}
\renewcommand{\arraystretch}{0.925}
\begin{tabular}{llcrlcccccc}
% && &&unit \\
 lattice &  distance &&unit& cell\\
 oF  &  \bf{0.025}&18.419 &22.790 &10.912& 90& 90 & 90 \\
mC  &  \bf{0.025}& 10.912&18.419  &12.634 &90&115.585 &90\\
mI&\bf{0.000}&10.704 &22.790&10.705&90&118.713& 90\\
 mI  &  \bf{0.018} &14.652 &10.912& 14.651&90&102.109 &90\\
	\caption{}
	\label{bgaol}
\end{tabular}
\endgroup

The unit cells reported in Table \ref{bgaol} are then
matched to the lattice type in column 1. If the 
discovered lattice type is monoclinic (in this case mI),
then the unit cell is projected into the manifold of the 
discovered type.

A web site on which searches may be done  is located at \url{http://iterate.sourceforge.net/bgaol/}.

\section{Summary}

Unit cells are used to represent crystallographic lattices.
Calculations measuring the differences between unit
cells are used to provide metrics for measuring meaningful
distances between three-dimensional crystallographic
lattices.  This is a surprisingly complex and
computationally demanding problem.  We present a
review of the current best practice using 
Delaunay-reduced unit cells
in the six-dimensional real space of Selling
scalar cells \SVI{} and the equivalent three-dimensional complex space \CIII{}.
The process is a simplified version of the process
needed when working with the more complex
six-dimensional real space of Niggli-reduced unit
cells \GVI{}.  Obtaining a distance begins with identification of the fundamental region in the space, continues with
conversion to primitive cells and reduction, analysis of
distances to the boundaries of the fundamental unit,
and is completed by a comparison of direct paths to
boundary-interrupted paths, looking for a path of
minimal length.

\appendix
\section{Availability of code}
    \label{code}

Code supporting this paper is available on github
at \url{http://github.com/duck20/LatticeRepLib.git}
and \url{http://github.com/yayahjb/ncdist.git}.

\section{Projectors}
	\label{projectors}

Orthogonal projectors onto linear subspaces can be calculated from 
any algebraic descriptor of that subspace that can provide a matrix $X$ with vectors that at least span that subspace.  Redundant vectors can help to improve the accuracy of the calculation.  If $A$ is the columnspace of $X$, then
$P=A(A^TA)^{-1}A^T$ is a projector onto that linear subspace,
and $PP=P$ and $P^T=P$.

The application of a projector yields the least-squares best fit.

\begin{figure}
				\includegraphics[width=0.8\textwidth]{Orthogonal_projection}

	\caption{Example of projectors: they are the operators that do orthogonal projection.}
\end{figure}
\section{Nearest neighbor search}
\label{nearest}

Wikipedia has a useful description of nearest neighbor searching \cite{enwiki:1103897119}.

Code for nearest neighbor searching using NearTree is available on GitHub
at \url{http://github.com/yayahjb/neartree.git}. NearTree uses the algorithm of
\citeasnoun{kalantari1983data} with enhancements for other kinds of searches. 
%\textbf{HJB, do we need to have more than the above short mention? Mention NearTree here?}

\section{Polar coordinates}
\label{polar}

Expressing unit cell parameters as edge lengths and angles does 
not lead to a easily interpreted metric: lengths and angles are
not commensurate. An unexplored alternative is to express them in
polar coordinates, which is equivalent to expressing them in complex
coordinates.

Delone pointed
out that the ``opposite'' scalars form special pairs. The Bravais
tetrahedron is the origin of his use of the term ``opposite''. Another way
to view this relationship is to organize the six scalars into
three pairs, treating the first member of the each pair as
the real component of a complex number and the second member of
each pair as the imaginary component.
Write the definition of an \SVI{} vector $(s_1, s_2, s_3, s_4, s_5, s_6)$ in \CIII{}:\\

$\begin{bmatrix}$
	\begin{tabular}{ c c l c }
		$\begin{pmatrix}
			s_1 \\
			s_4 
		\end{pmatrix}$

		$\begin{pmatrix}
			s_2 \\
			s_5
		\end{pmatrix}$

		$\begin{pmatrix}
			s_3\\
			s_6
		\end{pmatrix}$
	\end{tabular}
	$\end{bmatrix}$ \\ \\

Rewriting the definitions of the \SVI{} scalars, we
see that for each \CIII{} scalar, the angle in the real
position, is related to the edge vector used in the 
imaginary position. For instance for $c_1$, alpha is 
associated with $a$. For reflections, these associations
are not changed.\\

$\begin{bmatrix}$
	\begin{tabular}{ c c l c }
	$\begin{pmatrix}
		\bf{b \cos{\alpha}~ c} \\
		\bf{a \cdot d} 
	\end{pmatrix}$

	$\begin{pmatrix}
		\bf{a \cos{\beta} ~c} \\
		\bf{b \cdot d}
	\end{pmatrix}$

	$\begin{pmatrix}
		\bf{a \cos{\gamma}~ b}\\
		\bf{c \cdot d}
	\end{pmatrix}$
\end{tabular}
	$\end{bmatrix}$ \\

The stability of the angle/edge length association implies
that a potentially  useful new coordinate system for describing unit cells 
might be $\begin{pmatrix} a,\alpha \end{pmatrix}$, 
 $\begin{pmatrix} b,\beta \end{pmatrix}$,  
 $\begin{pmatrix} c,\gamma \end{pmatrix}$.

\section{The linear space \GVI ~(Niggli space)}
\label{G6}
%---------------------------------------------------------------------------

			Label  the cell axes ${\textbf{a}, \textbf{b}, \textbf{c}}$. The scalars are	\vdotv{\textbf{a}}{\textbf{a}},~ \vdotv{\textbf{b}}{\textbf{b}},~ \vdotv{\textbf{c}}{\textbf{c}},
			~ 2\vdotv{\textbf{b}}{\textbf{c}},~ 2\vdotv{\textbf{a}}{\textbf{c}},~ 2\vdotv{\textbf{b}}{\textbf{c}} where,
			{\it e.g.},
			\vdotv{\textbf{b}}{\textbf{c}} represents the dot product of the ${\textbf{b}}$  and ${\textbf{c}}$  axes. In order to organize these six quantities as a vector space in which one
			can compute simple Euclidean distances, we describe this set of scalars 
			as a vector, ${\textbf{g}}$, with components, $ {g}_1, {g}_2, {g}_3, {g}_4, {g}_5, {g}_6 $. \cite{Andrews2014}
		It should be noted that the first three scalars of \GVI{} are the
		 diagonal elements of the metric tensor. The second three are the 
		 off-diagonal elements, doubled. One reason for doubling is that those
		 elements occur twice within the metric tensor.

\section{The linear space \SVI{}~ (Delone space)}
\label{S6}

		\SVI{}: \citeasnoun{Andrews2019b} 
			presented a simple and the fastest currently
			known representation of lattices as the six Selling 
			scalars obtained from the dot products of the unit cell axes in addition to the negative of their sum (a body diagonal). Labeling 
			the cell axes ${\textbf{a}, \textbf{b}, \textbf{c}}$ , and ${\textbf{d}}$ (${\textbf{d}} =$ $-{\textbf{a}}-\!{\textbf{b}}-\!{\textbf{c}}$), 
			The scalars are	\vdotv{\textbf{b}}{\textbf{c}},~ \vdotv{\textbf{a}}{\textbf{c}},~ \vdotv{\textbf{a}}{\textbf{b}},
			~ \vdotv{\textbf{a}}{\textbf{d}},~ \vdotv{\textbf{b}}{\textbf{d}},~ \vdotv{\textbf{c}}{\textbf{d}} where,
			{\it e.g.},
			\vdotv{\textbf{b}}{\textbf{c}} represents the dot product of the ${\textbf{b}}$  and ${\textbf{c}}$  axes. For the purpose of organizing these six quantities as a vector space in which one can compute simple Euclidean distances, we describe this set of scalars 
			as a vector, ${\textbf{s}}$, with components, $ {s}_1, {s}_2, {s}_3, {s}_4, {s}_5, {s}_6 $.\\
		 		{ A cell is Selling-reduced if all six components are negative or zero \cite{Delaunay1932}.  
			\item Reversing the  observation that a Buerger-reduced cell is a good stepping stone to a Selling-reduced cell \cite{Allmann1968},
			a Selling-reduced cell can be a very efficient stepping stone to a Buerger-reduced cell from which a Niggli-reduced cell can quickly be derived.
			\item {Selling showed that the all-obtuse (Selling parameters all-negative) case is unique.
			 Every \SVI{} vector with all scalars non-positive corresponds to a valid unit cell.}

\section{The linear space \CIII{}}
\label{C3}

	\CIII{}~ is a space created from the scalars of \SVI{}. \\
		\vspace{.1cm}

\begin{center}
		$\begin{bmatrix}
		\begin{pmatrix}
			s_1 \\
			s_4
		\end{pmatrix}
		\begin{pmatrix}
			s_2 \\
			s_5
		\end{pmatrix}
		\begin{pmatrix}
			s_3 \\
			s_6
		\end{pmatrix}
	\end{bmatrix}$\\
\end{center}	
    ~~\\
	This associates each axis with its associated angle \cite{Andrews2019b}. 
	~~\\

	\begin{center}
		$\begin{bmatrix}
			\begin{pmatrix}
				\alpha   \\
				\textbf{a}
			\end{pmatrix}
			\begin{pmatrix}
				\beta   \\
				\textbf{b}
			\end{pmatrix}
			\begin{pmatrix}
				\gamma   \\
				\textbf{c}
			\end{pmatrix}
		\end{bmatrix}$
	\end{center}

\section{Geodesics}
\label{geodesic}

A geodesic is a curve representing in some sense the shortest path between two points in a surface \cite{enwiki:1116672577}.

\ack{{\bf Acknowledgements}}

Careful copy-editing and corrections by Frances C. Bernstein are 
gratefully acknowledged. Elizabeth Kincaid is to be thanked
for artwork.

\ack{{\bf Funding information}}

Funding for this research was provided in part by:  
US Department of Energy Offices of Biological and 
Environmental Research and of Basic Energy Sciences 
(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
U.S. National Institutes of Health (grant No. P41RR012408; 
grant No. P41GM103473; grant No. P41GM111244; 
grant No. R01GM117126,
grant No. 1R21GM129570); Dectris, Ltd.

\bibliography{Reduced}

\bibliographystyle{iucr}

%-------------------------------------------------------------------------
% TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
%-------------------------------------------------------------------------

% Simple tables should use the tabular environment according to this
% model

% Postscript figures can be included with multiple figure blocks

\end{document}                    % DO NOT DELETE THIS LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-02-16 C3/C3.tex
Updating text of article C3 uv5016, "Measuring Lattices: adding file for replies to referees.

@@
 
-	\keyword{Delaunay}
+	\keyword{lattice}
+	\keyword{reduction}
 	\keyword{Delone}
@@
 	note that the complex components involve ($\alpha,a$), ($\beta, b$), and ($\gamma,c$).
+	Also consider that each complex term uses all 
+	four 3-space vectors; for example, $c_1$ is ($b \cdot c$, $a \cdot d$)
 
@@
 		\label{tab:operators}
+		\caption{The Result column values give the products of each operator as applied to $(x_j, y_j)$}.
 		\begin{tabular}{c l l l }
@@
 	$\begin{bmatrix}
+		\label{thirdmatrix}
 		\Minus{}_r	& 0			& 0 \\
@@
 	\end{bmatrix}$ 
-	$\bf{c}$ \\
+	$\bf{c}$. \\
 
@@
 	\end{bmatrix}$ 
-	$\bf{c}$
+	$\bf{c}$.
 	\\
@@
 	\SvecA{}.
-	The third matrix above 
-	(the third boundary transform for $s_1$) 
+	The third boundary transform for $s_1$
 	implements this operation and interchanges the
-	real part of \ciii{} and the imaginary part of \cii{}.
+	real part of \ciii{} and the imaginary part of \cii{}:
 
@@
 	Where all the 	boundaries of the fundamental orthant 
-	in \CIII ~and \SVI{} are basically the same 
+	\SVI{} (and also \CIII{}) are basically the same 
 	(and indeed are related by reflections), 
@@
 	The matrices display one of the properties of \CIII{} 
-	that make it a useful representation. 
+	that make it a useful conceptual representation. 
 	For the scalar being transformed, it and its 
@@
 
-	Several aspects of C3 are obvious from inspecting the matrices above. 
+	Several aspects of \CIII{} are obvious from inspecting the matrices above. 
 	First, we see that each boundary has four possible 

### 2023-02-28 MeasuringLattices_uv5016/MeasuringLattices.tex
Finished response to reviews and updated the tex file to show the changes

@@
 
+In crystallography, a unit cell represents an entire lattice, displaying all the symmetries of a crystal except for the translations.
+The entire lattice is generated by assembling all the translational repetitions of the unit cell.
+
 Stumbling blocks arise when the need to compare various unit
@@
 a manifold generates the need for more dimensions.
+The dimensionality of a unit cell itself should be distinguished from the higher dimensions of the spaces in which it may be useful to embed the cell.
 
@@
 	in this case, unit cells), it is conventional to choose a fundamental
-	region \cite{enwiki:1028832752}. In a two-dimensional graph, it would often be chosen as the all-plus
+	region.
+``A fundamental domain or fundamental region is a subset of the space which contains exactly one point from each of these orbits.
+It serves as a geometric realization for the abstract set of representatives of the orbits.'' \cite{enwiki:1028832752}.
+In a two-dimensional graph, it would often be chosen as the all-plus
 	quadrant. In crystallography, it might be a unit cell, a
@@
 
-	The obvious step to find the shortest distance between two
+	The obvious first step in finding the shortest distance between two
 	lattices is to simply compute the Euclidean distance between
@@
 
-\begin{figure}
+%\begin{figure}
 		\includegraphics[width = 0.5\textwidth ]{4_3}
 		\label{4_3}
-	\caption{}
-\end{figure}	
+%	\caption{}
+%\end{figure}	
 	It is clear that within the fundamental unit, ignoring all symmetry

### 2023-03-06 C3/C3.tex
FInish cover letter to mois for Measuing Lattices C3, start on adding section on graphics
C3.tex -- updating

@@
 
-\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
-%\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
+\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
 \usepackage{amssymb}
@@
 	~\citeasnoun{andrews2019b} chose to 
-	represent the scalars in the space \SVI{},
+	represent the Selling scalars in the space \SVI{},
 	\Svec{} (defined in the order above), 
 	as a way to create a metric space
-	for the measurement of the distance between lattices. T
-	hey also consider the representation of this space as the
+	for the measurement of the distance between lattices. 
+	They also consider the representation of this space as the
 	space of three complex dimensions, \CIII{} or 
@@
 
-	In \CIII{}, in terms of the Delone scalars, 
-	a vector is \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}, 
+	In \CIII{}, in terms of the Selling scalars, 
+	a vector is defined as \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}, 
 	where the real and imaginary parts
@@
 	note that the complex components involve ($\alpha,a$), ($\beta, b$), and ($\gamma,c$).
-	Also consider that each complex term uses all 
-	four 3-space vectors; for example, $c_1$ is ($b \cdot c$, $a \cdot d$)
+	Additionally, each complex term uses all 
+	four 3-space vectors; for example, $c_1$ is ($b \cdot c$, $a \cdot d$).
 
@@
 	of the reflections in \CIII{} (and in \SVI{}). 
-	This paper will describe the boundary transformations 
+	This paper describes the boundary transformations 
 	at the edges of the fundamental	unit of \CIII{}. 
@@
 	In \SVI{} and \CIII{}, 
-	the boundaries are found where any scalar 
-	(or correspondingly the real or imaginary part) 
-	is equal to zero. The rationale for this work was 
+	the boundaries located where any \SVI{} scalar 
+	(or correspondingly in\CIII{}, the real or imaginary part) 
+	equals to zero. The rationale for this work was 
 	that it might lend insights into the topology of
@@
 
-	If considered in \CIII, Delone's alternate transformation 
+	Considered in \CIII, Delone's alternate transformation 
 	for the \si{} boundary would
@@
 	just displayed by the "exchange operation"~\cite{andrews2019b} 
-	applied to the second and third \CIII{} elements. Delone 
+	applied to the second and third \CIII{} coordinates. Delone 
 	did not describe the latter two transformation, 
@@
 
+	\section{Graphical display of projections}
+
+%	Although printing the image of a 6-dimensional object
+%	is impossible, there is some utility to creating the
+%	2-dimensional projections of \CIII{} onto the three
+%	coordinates.
+
 	\section{Summary}
 
-	The considerable regularity 
-	of the transformation matrices (above) highlights one
-	of the important aspects of Selling reduction 
-	as used by~\citeasnoun{Delaunay1932}
-	when compared to Niggli reduction~\cite{Niggli1928}. 
-	Where all the 	boundaries of the fundamental orthant 
-	\SVI{} (and also \CIII{}) are basically the same 
-	(and indeed are related by reflections), 
-	the boundaries formed in Niggli reduction are of multiple types, 
-	and the fundamental unit of the space representing 
-	Niggli reduced cells (\GVI{}, see~\citeasnoun{Andrews2014}) is non-convex.
+The transformation matrices shown above demonstrate the 
+considerable regularity of Selling reduction, as used by 
+Delaunay (1932), in comparison to Niggli 
+reduction \cite{Niggli1928}. While all the boundaries of the non-positive 
+orthant of \SVI{} are essentially the same (and 
+related by reflections), the boundaries formed in 
+Niggli reduction are of multiple types, and the fundamental unit 
+of the space representing Niggli reduced cells (\GVI{}, see \citeasnoun{Andrews2014}) is non-convex.
 
-	The matrices display one of the properties of \CIII{} 
-	that make it a useful conceptual representation. 
-	For the scalar being transformed, it and its 
-	opposite scalar create a unique row in each matrix. 
-	Those rows contain
-	only the minus operator and zeros, 
-	highlighting the unique relationship of that pair of scalars.
+The matrices also display one of the unique properties of \CIII{}, 
+which makes it a useful conceptual representation. 
+For the scalar being transformed, it and its opposite scalar
+create a unique row in each transformation matrix. These rows contain only
+the minus operator and zeros, highlighting the 
+unique relationship of that pair of scalars.
 
-	Several aspects of \CIII{} are obvious from inspecting the matrices above. 
-	First, we see that each boundary has four possible 
-	transformations that can be applied. 
-	Because the each of the transformations at boundaries are self-inverse, 
-	they are the same transformations that would 
-	be used in the process of cell (lattice) reduction. 
-	~\citeasnoun{Delaunay1932} and~\citeasnoun{Delone1975} give 
-	only two choices. Presumably for simplicity they did not list
-	the transformations that used the "exchange" operator 
-	(see~\cite{andrews2019b}).
+Several aspects of \CIII{} are evident from inspecting the matrices. 
+First, each boundary has four possible 
+transformations that can be applied. Since each of 
+the transformations at boundaries are self-inverse, they are the 
+same transformations that would be used in the process of 
+cell (lattice) reduction. \citeasnoun{Delaunay1932} and \citeasnoun{Delone1975} give only two choices, presumably for simplicity, 
+omitting transformations that use the "exchange" operator 
+(see \citeasnoun{andrews2019b}).
 

### 2023-03-08 C3/C3.tex
C3 -- comment out section  title for graphics
C3 added graphics, sent to HJB

@@
 	\begin{synopsis}
-		Selling reduction and Delone reduction are
-		considered in a space of complex variables.
+		The space \CIII{} is explained in more detail than
+		in the original description. Boundary transformations
+		of the fundamental unit are described in detail. 
+		A graphical presentation of the basic coordinates
+		is described and illustrated.
 	\end{synopsis}
@@
 
-		The Delone scalars are studied in \CIII, the space of three complex variables.
+		The Delone (Selling) scalars, which are used in 
+		unit cell reduction and in lattice type determination,
+		are studied in \CIII, the space of 3 complex variables.
+		The 3 complex coordinate planes are composed of the 6
+		Delone scalars.
 
@@
 
-%	Although printing the image of a 6-dimensional object
-%	is impossible, there is some utility to creating the
-%	2-dimensional projections of \CIII{} onto the three
-%	coordinates.
+	The 2-dimensional nature of the three coordinates of \CIII{} 
+	suggests their use for graphical display.
+
+	As an example, we use Phospholipase A2 (retrieved from 
+	the Protein Data Bank \cite{Bernstein1977}), which has had several similar
+	or identical structures determined \cite{LeTrong2007}. \citeasnoun{andrews2019b}
+	found additional cases (see Table \ref{PLA2})
+
+\begin{table}
+	\begin{tabular}{l c c c c c c c c c}
+		\toprule
+		PDB id & Centering & a&b&c&$\alpha$&$\beta$&$\gamma$\\
+		\midrule
+	1DPY & R& 57.98& 57.98& 57.98& 92.02& 92.02& 92.02\\
+	1FE5 & R& 57.98& 57.98& 57.98& 92.02& 92.02& 92.02\\
+	1G0Z & H& 80.36& 80.36& 99.44& 90&    90&    120\\
+	1G2X & C& 80.95& 80.57& 57.1 & 90&    90.35&  90\\
+	1U4J & H& 80.36& 80.36& 99.44& 90&    90&    120\\
+	2OSN & R& 57.10& 57.10& 57.10& 89.75& 89.75&  89.75\\
+	\bottomrule
+\end{tabular}	\\
+\caption{Phospholipase A2 unit cells}
+\label{PLA2}
+\end{table}
+
+Below, Figure \ref{fig1} shows the unit cells as 
+reported (the centering of lattices has not been removed).
+The following figures show various transformations and
+embellishments of the reported cells.
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{1_35}
+	\caption{Phospholipase A2 unit cells as reported. }
+	\label{fig1}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{2_06}
+	\caption{The unit cells Niggli reduced. The similarity 
+	of the 3 projections is indicative of the exact or nearly
+exact rhombohedral symmetry.}
+	\label{fig2}
+\end{figure}
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{3_03}
+	\caption{The unit cells Delone reduced. }
+	\label{fig3}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{4_13}
+	\caption{The unit cells, Niggli and 5 copies were perturbed 2\%
+	orthogonally to \SVI{} vector. }
+	\label{fig4}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{5_21}
+	\caption{The unit cells Niggli, and 5 copies were perturbed 10\%
+		orthogonally to \SVI{} vector. }
+	\label{fig5}
+\end{figure}
 

### 2023-03-09 C3/C3.tex
C3 -- add the generating command to generate the graphics
updating to (I hope) final version

@@
 		unit cell reduction and in lattice type determination,
-		are studied in \CIII, the space of 3 complex variables.
-		The 3 complex coordinate planes are composed of the 6
-		Delone scalars.
+		are studied in \CIII, the space of three complex variables.
+		The three complex coordinate planes are composed of the six
+		Delone scalars. The transformations at boundaries of the
+		Selling reduced orthant are described as matrices of operators.
+		A graphical representation as the projections onto the
+		three coordinates is described.
 
@@
 	for the measurement of the distance between lattices. 
-	They also consider the representation of this space as the
+	~\citeasnoun{andrews2019b} also considered the representation 
+	of this space as the
 	space of three complex dimensions, \CIII{} or 
@@
 	of each are the ``opposite'' scalars 
-	according to the definition of~\citeasnoun{Delaunay1932} (see~\citeasnoun{andrews2019a}).
+	according to the definition of~\citeasnoun{Delaunay1932}
+	(opposite in terms of the Bravais tetrahedron of scalars) (see~\citeasnoun{andrews2019a}).
 	As a mnemonic device, 
@@
 	This paper describes the boundary transformations 
-	at the edges of the fundamental	unit of \CIII{}. 
+	at the edges of the fundamental	unit of \CIII{}
+	and also a graphical presentation. 
 	In \SVI{}, the fundamental unit is the all negative orthant, 
-	which contains only and	all of the reduced cells. 
+	which contains only, and all, of the reduced cells. 
 	In \SVI{} and \CIII{}, 
 	the boundaries located where any \SVI{} scalar 
-	(or correspondingly in\CIII{}, the real or imaginary part) 
-	equals to zero. The rationale for this work was 
-	that it might lend insights into the topology of
-	the space of lattices.
+	(or correspondingly in \CIII{}, a real or imaginary part) 
+	equals to zero.
 
@@
 
-	Next we define the operators in \CIII{} t
-	hat will be used in the matrix descriptions of 
-	the transformations at the boundaries of the fundamental unit.\\
-	See Table ~\ref{tab:operators}.
+	Next we define the operators in \CIII{} that 
+	will be used in the matrix descriptions of 
+	the transformations at the boundaries of the fundamental unit,
+	see Table ~\ref{tab:operators}.
 
@@
 	\section{Matrices of boundary transformations}
+	~~\\
 
@@
 	he lists the following as one of the possible results 
-	for a transformation on \si{}:
+	for a transformation on \si{}, translated to \SVI{}:
 	\SvecA{}.
-	The third boundary transform for $s_1$
+	The third boundary transform for $s_1$ above
 	implements this operation and interchanges the
@@
 	with the imaginary part of \ciii{}. That is the fourth
-	matrix in the list for \si{}. 
+	matrix above in the list for \si{}. 
 	The other two transformations for \si{} can 
 	be generated from the two we have 
-	just displayed by the "exchange operation"~\cite{andrews2019b} 
+	just mentioned by the ``exchange operation''~\cite{andrews2019b} 
 	applied to the second and third \CIII{} coordinates. Delone 
-	did not describe the latter two transformation, 
+	did not describe the latter two transformations, 
 	perhaps because even a single transformation 
-	was adequate to implement reduction. He already listed two.
+	was adequate to implement reduction. He had already listed two.
 
@@
 
-	The 2-dimensional nature of the three coordinates of \CIII{} 
+	The two-dimensional nature of the three coordinates of \CIII{} 
 	suggests their use for graphical display.
 
-	As an example, we use Phospholipase A2 (retrieved from 
+	As an example, we use Phospholipase A2 (\textbf{PLA2}) (retrieved from 
 	the Protein Data Bank \cite{Bernstein1977}), which has had several similar
@@
 reported (the centering of lattices has not been removed).
-The following figures show various transformations and
+The following figures (Figures \ref{fig2} to \ref{fig5}) show various transformations and
 embellishments of the reported cells.
@@
 	\includegraphics[width=1.0\textwidth]{2_06}
-	\caption{The unit cells Niggli reduced. The similarity 
+	\caption{The PLA2 unit cells Niggli-reduced. The similarity 
 	of the 3 projections is indicative of the exact or nearly
@@
 	\includegraphics[width=1.0\textwidth]{3_03}
-	\caption{The unit cells Delone reduced. }
+	\caption{The PLA2 unit cells Delone-reduced. }
 	\label{fig3}
@@
 	\includegraphics[width=1.0\textwidth]{4_13}
-	\caption{The unit cells, Niggli and 5 copies were perturbed 2\%
+	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 2\%
 	orthogonally to \SVI{} vector. }
@@
 	\includegraphics[width=1.0\textwidth]{5_21}
-	\caption{The unit cells Niggli, and 5 copies were perturbed 10\%
+	\caption{The PLA2 unit cells Niggli-reduced, and five copies were perturbed 10\%
 		orthogonally to \SVI{} vector. }
@@
 considerable regularity of Selling reduction, as used by 
-Delaunay (1932), in comparison to Niggli 
-reduction \cite{Niggli1928}. While all the boundaries of the non-positive 
+Delaunay (1932); The reduction operations of Niggli 
+reduction \cite{Niggli1928} are more complex. 
+While all the boundaries of the non-positive 
 orthant of \SVI{} are essentially the same (and 
@@
 Niggli reduction are of multiple types, and the fundamental unit 
-of the space representing Niggli reduced cells (\GVI{}, see \citeasnoun{Andrews2014}) is non-convex.
-
-The matrices also display one of the unique properties of \CIII{}, 
-which makes it a useful conceptual representation. 
-For the scalar being transformed, it and its opposite scalar
-create a unique row in each transformation matrix. These rows contain only
-the minus operator and zeros, highlighting the 
-unique relationship of that pair of scalars.
+of the space representing Niggli-reduced cells (\GVI{}, see \citeasnoun{Andrews2014}) is non-convex.
 
@@
 cell (lattice) reduction. \citeasnoun{Delaunay1932} and \citeasnoun{Delone1975} give only two choices, presumably for simplicity, 
-omitting transformations that use the "exchange" operator 
+omitting transformations that use the ``exchange'' operator 
 (see \citeasnoun{andrews2019b}).
@@
 
-	The $C^{++}$ ~code for \CIII{} is available in github.com, in
+	The $C^{++}$ ~code for \CIII{} and related 
+	software tools is available in github.com, in
 	\url{https://github.com/duck10/LatticeRepLib.git}.
@@
 
-	\bibliography{reduced}
+	\bibliography{Reduced}
 
@@
 
+%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>plotc3
+%; Graphical output SVG file =PLT__2023-03-07.13_43_35.svg
+%
+%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmdniggli | plotc3
+%; Graphical output SVG file =PLT__2023-03-07.13_44_06.svg
+%
+%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmddelone | plotc3
+%; Graphical output SVG file =PLT__2023-03-08.07_11_03.svg
+%
+%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmdniggli | cmdperturb 5 20 | plotc3
+%; Graphical output SVG file =PLT__2023-03-08.09_00_13.svg
+%
+%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmdniggli | cmdperturb 5 100 | plotc3
+%; Graphical output SVG file =PLT__2023-03-08.09_00_21.svg
+
 \end{document}                    % DO NOT DELETE THIS LINE

### 2023-03-13 C3/C3.tex
changes from HJB & FCB

@@
 	for the measurement of the distance between lattices. 
-	~\citeasnoun{andrews2019b} also considered the representation 
+	They also considered the representation 
 	of this space as the
@@
 	Additionally, each complex term uses all 
-	four 3-space vectors; for example, $c_1$ is ($b \cdot c$, $a \cdot d$).
+	four three-space vectors; for example, $c_1$ is ($b \cdot c$, $a \cdot d$).
 
@@
 	In \SVI{}, the fundamental unit is the all negative orthant, 
-	which contains only, and all, of the reduced cells. 
+	which contains only, and all of, the reduced cells. 
 	In \SVI{} and \CIII{}, 
@@
 	(or correspondingly in \CIII{}, a real or imaginary part) 
-	equals to zero.
+	equals zero.
 
@@
 
-	Complex numbers will be represented 
+	Complex numbers are represented 
 	in Cartesian format $(x,y)$, 
@@
 
-	We will represent a vector 
+	We represent a vector 
 	in \CIII{} by \{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\} 
@@
 
-	The standard representation of the identity operation is
+	The standard representation of the identity operation is\\
 
@@
 
-	The identity in \CIII{} can also be written:
+	The identity in \CIII{} can also be written:\\
 
@@
 	\includegraphics[width=1.0\textwidth]{5_21}
-	\caption{The PLA2 unit cells Niggli-reduced, and five copies were perturbed 10\%
+	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 10\%
 		orthogonally to \SVI{} vector. }
@@
 considerable regularity of Selling reduction, as used by 
-Delaunay (1932); The reduction operations of Niggli 
+Delaunay (1932). The reduction operations of Niggli 
 reduction \cite{Niggli1928} are more complex. 
@@
 
-	\bibliographystyle{iucr}
+	\bibliographystyle{iucrit}
 

### 2023-03-28 MeasuringLattices_uv5016/MeasuringLattices.tex
updating mostly Polar

@@
 
+In the following discussion we will follow the common practice of viewing the circle as a two-dimensional object and the torus as a three-dimensional object.
+Formally, as a manifold, a circle is one-dimensional and a torus is a two-dimensional.”
+
 \subsection{Create a simple, 1D unit cell}
@@
 	region.
-``A fundamental domain or fundamental region is a subset of the space which contains exactly one point from each of these orbits.
-It serves as a geometric realization for the abstract set of representatives of the orbits.'' \cite{enwiki:1028832752}.
+``Given a topological space and a group acting on it, the images of a single point under the group action form an orbit of the action.
+A fundamental domain or fundamental region is a subset of the space which contains exactly one point from each of these orbits.'' \cite{enwiki:1028832752}.
 In a two-dimensional graph, it would often be chosen as the all-plus
@@
 			a Selling-reduced cell can be a very efficient stepping stone to a Buerger-reduced cell from which a Niggli-reduced cell can quickly be derived.
-			\item {Selling showed that the all-obtuse (Selling parameters all-negative) case is unique.
+			\item {Selling showed that the all-obtuse (Selling parameters all-negative) case is unique \cite{Selling1874}.
 			 Every \SVI{} vector with all scalars non-positive corresponds to a valid unit cell.}

### 2023-05-09 MeasuringLattices_uv5016/MeasuringLattices.tex
changes for Mois

@@
 
+	\section{Forward}
+
+	"Measuring Lattices" is intended to be a summary of an invited talk
+	for satellite meeting, "Mathematics and Computer Science for Materials Innovation: Crystal Lattice Classifications" of August, 2022, ECM33
+	meeting (https://www.ecm33.fr/). The intent is describe in brief
+	the basics of implementing a metric space that creates
+	a measure of the "distances" between lattices.
+
 	\section{History}
@@
 \end{itemize}
+
+Crystallographic software delivered with diffractometers, used for processing
+beamline data, etc. has produced more than one and quarter million structures. The structures have had internally consistent lattice and
+symmetry identification, but there have been serious omissions in 
+providing globally consistent lattice and symmetry identification
+because of structures being done in the incorrect space group.
+\citeasnoun{LeTrong2007} published the case of phospholipase
+A2, and more were found in \citeasnoun{McGill2014}.
+Not having failed to be observed is quite different from always working. Two regions of the space of lattice determination come into play here. 
+First, how much error do we want to consider? For protein work, especially for serial crystallography, the errors can be large. Second, the simple methods commonly used in 
+diffractometer software are likely to fail for 
+some of the less common cases: centered lattices commonly cause interesting issues; lattices that are needle- or pancake-shaped are often challenging;
+the five-fold ambiguity cited by \citeasnoun{Gruber1973} also causes confusion. The cited five-fold ambiguity is only for the particular
+case that Gruber listed, but interchanges of the axes to produce
+other, Buerger-reduced cells.
 

### 2023-05-22 C3/C3.tex
changes to make latex work :wq
Signed-off-by: larry andrews <larry6640995@gmail.com>
updating mostly Polar
uv5016 -- fix quote marks C3 -- replying to reviewers

@@
 
-\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
+\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
 \usepackage{amssymb}
@@
 \newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
-\newcommand{\Imaginary}[0]{$ \textfrak{I} $}
-\newcommand{\Real}[0]{$ \textfrak{R} $}
-\newcommand{\Exchange}[0]{$\textfrak{X}$}
+\newcommand{\Imaginary}[0]{\mathcal{I}}
+\newcommand{\Real}[0]{\mathcal{R}}
+\newcommand{\Exchange}[0]{$\mathcal{X}$}
 
@@
 	\newcommand{\MEIII}[0]{$M_{E^{3}}$}
-	\newcommand{\Plus}[0]{$\textfrak{P}$}	
-	\newcommand{\Minus}[0]{$\textfrak{M}$}
+	\newcommand{\Plus}[0]{\mathcal{P}}	
+	\newcommand{\Minus}[0]{\mathcal{M}}
 
@@
 	\begin{abstract}
-
 		The Delone (Selling) scalars, which are used in 
@@
 	in \CIII{} by \{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\} 
-	as an alternative to  \{(\si,\siv ), (\sii,\sv),(\sii,\svi)\}.
+	as an alternative to  \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}.
 
@@
 			$\Minus{}_i $ 		& $\Minus{}_i(c_j)$ 		& $(x_j-y_j,-y_j) $ 	& Minus imag\\ 
-			$\textfrak{P}_r$ 	& $ \textfrak{P}_r(c_j)$ 	& $(x_j, x_j)$ 			& Plus real\\ 
-			$\textfrak{P}_i$ 	&  $\textfrak{P}_i(c_j)$  	& $(y_j, y_j)$ 			& Plus imag\\ 
-			$\Real$ 			&  $\Real(c_j)$ 			& $x_j$ 				& Real \\ 
-			$\Imaginary$ 		&  $\Imaginary(c_j)$ 		& $y_j$ 				& Imaginary\\
+			$\mathcal{P}_r$ 	& $ \mathcal{P}_r(c_j)$ 	& $(x_j, x_j)$ 		& Plus real\\ 
+			$\mathcal{P}_i$ 	&  $\mathcal{P}_i(c_j)$  	& $(y_j, y_j)$ 		& Plus imag\\ 
+			$\Real$ 			&  $\Real(c_j)$ 			& $x_j$ or $(x_j,0)$		& Real \\ 
+			$\Imaginary$ 		&  $\Imaginary(c_j)$ 		& $y_j$ pr $(y_j,0)$ 		& Imaginary\\
 			\bottomrule
@@
 
+	\citeasnoun{andrews2019b} listed the boundary transformations
+	for each of the six \SVI{} boundaries. For each,
+	they only listed 2 transformations, following \citeasnoun{Delaunay1932}
+	and \citeasnoun{Delone1975}. That list is incomplete in the sense
+	that applying the 24 reflections gives 24 boundary transformation,
+	four for each boundary, rather than 12.
+
 	For the boundary at $s_1$: (the real component of $c_1$).
@@
 		\Minus{}_r		& 0					& 0 \\
-		\Plus{}_r		&  \textrm{i}\Re{}	& \Re{} \\
-		\textfrak{P}_r	& \textrm{i}\Im{}	& \Im{}
+		\Plus{}_r		&  \textrm{i}\Real{}	& \Real{} \\
+		\mathcal{P}_r	& \textrm{i}\Imaginary{}	& \Imaginary{}
 	\end{bmatrix}$
@@
 		\Minus{}_r	& 0 				& 0 \\
-		\Plus{}_r	&\textrm{i}\Im{}	& \Im{} \\
-		\Plus{}_r	& \textrm{i}\Re{}	& \Re{} \\
+		\Plus{}_r	&\textrm{i}\Imaginary{}	& \Imaginary{} \\
+		\Plus{}_r	& \textrm{i}\Real{}	& \Real{} \\
 	\end{bmatrix}$ 
@@
 		\Minus{}_r	& 0			& 0 \\
-		\Plus{}_r	&  \Re{}	& \textrm{i}\Re{} \\
-		\Plus{}_r	& \Im{}		&\textrm{i}\Im{}
+		\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
+		\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
 	\end{bmatrix}	$
@@
 		\Minus{}_r	& 0			& 0 \\
-		\Plus{}_r	& \Im{}		&\textrm{i}\Im{} \\
-		\Plus{}_r	&  \Re{}	&\textrm{i}\Re{} \\
+		\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{} \\
+		\Plus{}_r	&  \Real{}	&\textrm{i}\Real{} \\
 	\end{bmatrix}$ \\
@@
 		\Minus{}_i	& 0					& 0 \\
-		\Plus{}_i	&  \textrm{i}\Re{}	& \Re{} \\
-		\Plus{}_i	& \textrm{i}\Im{}	& \Im{}
+		\Plus{}_i	&  \textrm{i}\Real{}	& \Real{} \\
+		\Plus{}_i	& \textrm{i}\Imaginary{}	& \Imaginary{}
 	\end{bmatrix}$
@@
 		\Minus{}_i	& 0					& 0 \\
-		\Plus{}_i	& \textrm{i}\Im{}	& \Im{} \\
-		\Plus{}_i	&  \textrm{i}\Re{}	& \Re{} \\
+		\Plus{}_i	& \textrm{i}\Imaginary{}	& \Imaginary{} \\
+		\Plus{}_i	&  \textrm{i}\Real{}	& \Imaginary{} \\
 	\end{bmatrix}$ 
@@
 		\Minus{}_i	& 0			& 0 \\
-		\Plus{}_i	&  \Re{}	& \textrm{i}\Re{} \\
-		\Plus{}_i	& \Im{}		& \textrm{i}\Im{}
+		\Plus{}_i	&  \Real{}	& \textrm{i}\Real{} \\
+		\Plus{}_i	& \Imaginary{}		& \textrm{i}\Imaginary{}
 	\end{bmatrix}	$
@@
 		\Minus{}_i	& 0		& 0 \\
-		\Plus{}_i	& \Im{}	& \textrm{i}\Im{} \\
-		\Plus{}_i	&  \Re{}& \textrm{i}\Re{} \\
+		\Plus{}_i	& \Imaginary{}	& \textrm{i}\Imaginary{} \\
+		\Plus{}_i	&  \Real{}& \textrm{i}\Imaginary{} \\
 	\end{bmatrix}$ \\
@@
 
-		\textrm{i}\Re{}	& \Plus{}_r		& \Re{} \\
+		\textrm{i}\Real{}	& \Plus{}_r		& \Real{} \\
 		0				& \Minus{}_r	& 0 \\
-		\textrm{i}\Im{}	& \Plus{}_r		& \Im{}
+		\textrm{i}\Imaginary{}	& \Plus{}_r		& \Imaginary{}
 	\end{bmatrix}$
@@
 	$\begin{bmatrix}
-		\textrm{i}\Im{}	& \Plus{}_r		&  \Im{} \\
+		\textrm{i}\Imaginary{}	& \Plus{}_r		&  \Imaginary{} \\
 		0				& \Minus{}_r	& 0 \\
-		\textrm{i}\Re{}	& \Plus{}_r		& \Re{} \\
+		\textrm{i}\Real{}	& \Plus{}_r		& \Real{} \\
+	\end{bmatrix}$ e
+	%
+	$\begin{bmatrix}
+		\Real{}	&\Plus{}_r		&   \textrm{i}\Real{} \\
+		0		& \Minus{}_r	&  0 \\
+		\Imaginary{}	&\Plus{}_r		& \textrm{i}\Imaginary{}
+	\end{bmatrix}	$ 
+	%
+	$\begin{bmatrix}
+		\Imaginary{}	& \Plus{}_r	& \textrm{i}\Imaginary{} \\
+		0		&\Minus{}_r	&  0 \\
+		\Real{}	& \Plus{}_r	&  \textrm{i}\Real{} \\
+	\end{bmatrix}$ \\
+
+	For the boundary at \sv: (the imaginary component of \cii{}).
+
+	$\begin{bmatrix}
+		\textrm{i}\Real{}	& \Plus{}_i		& \Real{} \\
+		0				& \Minus{}_i	& 0 \\
+		\textrm{i}\Imaginary{}	& \Plus{}_i		& \Imaginary{}
 	\end{bmatrix}$
@@
 	$\begin{bmatrix}
-		\Re{}	&\Plus{}_r		&   \textrm{i}\Re{} \\
-		0		& \Minus{}_r	&  0 \\
-		\Im{}	&\Plus{}_r		& \textrm{i}\Im{}
+		\textrm{i}\Imaginary{}	& \Plus{}_i		&  \Imaginary{} \\
+		0				& \Minus{}_i	& 0 \\
+		\textrm{i}\Real{}	& \Plus{}_ri	& \Real{} \\
+	\end{bmatrix}$ 
+	%
+	$\begin{bmatrix}
+		\Real{}	&\Plus{}_i		&   \textrm{i}\Real{} \\
+		0		& \Minus{}_i	&  0 \\
+		\Imaginary{}	&\Plus{}_i		& \textrm{i}\Imaginary{}
 	\end{bmatrix}	$ 
@@
 	$\begin{bmatrix}
-		\Im{}	& \Plus{}_r	& \textrm{i}\Im{} \\
-		0		&\Minus{}_r	&  0 \\
-		\Re{}	& \Plus{}_r	&  \textrm{i}\Re{} \\
-	\end{bmatrix}$ \\
-
-	For the boundary at $s_5$: (the imaginary component of \cii{}).
-
-	$\begin{bmatrix}
-		\textrm{i}\Re{}	& \Plus{}_i		& \Re{} \\
-		0				& \Minus{}_i	& 0 \\
-		\textrm{i}\Im{}	& \Plus{}_i		& \Im{}
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\textrm{i}\Im{}	& \Plus{}_i		&  \Im{} \\
-		0				& \Minus{}_i	& 0 \\
-		\textrm{i}\Re{}	& \Plus{}_ri	& \Re{} \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\Re{}	&\Plus{}_i		&   \textrm{i}\Re{} \\
-		0		& \Minus{}_i	&  0 \\
-		\Im{}	&\Plus{}_i		& \textrm{i}\Im{}
-	\end{bmatrix}	$ 
-	%
-	$\begin{bmatrix}
-		\Im{}	& \Plus{}_i	& \textrm{i}\Im{} \\
+		\Imaginary{}	& \Plus{}_i	& \textrm{i}\Imaginary{} \\
 		0		&\Minus{}_i	&  0 \\
-		\Re{}	& \Plus{}_i	&  \textrm{i}\Re{} \\
+		\Real{}	& \Plus{}_i	&  \textrm{i}\Real{} \\
 	\end{bmatrix}$ \\
@@
 
-		\textrm{i}\Re{}	& \Re{}	& \Plus{}_r \\
-		\textrm{i}\Im{}	& \Im{}	& \Plus{}_r  \\
+		\textrm{i}\Real{}	& \Real{}	& \Plus{}_r \\
+		\textrm{i}\Imaginary{}	& \Imaginary{}	& \Plus{}_r  \\
 		0				& 0		& \Minus{}_r  \\
@@
 	$\begin{bmatrix}
-		\textrm{i}\Im{}	&  \Im{}	& \Plus{}_r \\
-		\textrm{i}\Re{}	& \Re{}		& \Plus{}_r \\
+		\textrm{i}\Imaginary{}	&  \Imaginary{}	& \Plus{}_r \\
+		\textrm{i}\Real{}	& \Real{}		& \Plus{}_r \\
 		0				& 0			& \Minus{}_r \\
@@
 	$\begin{bmatrix}
-		\Re{}	&   \textrm{i}\Re{}	&\Plus{}_r \\
-		\Im{}	& \textrm{i}\Im{}	&\Plus{}_r \\
+		\Real{}	&   \textrm{i}\Real{}	&\Plus{}_r \\
+		\Imaginary{}	& \textrm{i}\Imaginary{}	&\Plus{}_r \\
 		0 		&  0				& \Minus{}_r \\
@@
 	$\begin{bmatrix}
-		\Im{}	& \textrm{i}\Im{}	& \Plus{}_r \\
-		\Re{}	&  \textrm{i}\Re{}	& \Plus{}_r \\
+		\Imaginary{}	& \textrm{i}\Imaginary{}	& \Plus{}_r \\
+		\Real{}	&  \textrm{i}\Real{}	& \Plus{}_r \\
 		0		&  0				&\Minus{}_r \\
@@
 
-		\textrm{i}\Re{}	& \Re{}	& \Plus{}_i \\
-		\textrm{i}\Im{}	& \Im{}	& \Plus{}_i \\
+		\textrm{i}\Real{}	& \Real{}	& \Plus{}_i \\
+		\textrm{i}\Imaginary{}	& \Imaginary{}	& \Plus{}_i \\
 		0				& 0		& \Minus{}_i 
@@
 	$\begin{bmatrix}
-		\textrm{i}\Im{}	&  \Im{}	& \Plus{}_i \\
-		\textrm{i}\Re{}	& \Re{}		& \Plus{}_i \\
+		\textrm{i}\Imaginary{}	&  \Imaginary{}	& \Plus{}_i \\
+		\textrm{i}\Real{}	& \Real{}		& \Plus{}_i \\
 		0				& 0			& \Minus{}_i \\
@@
 	$\begin{bmatrix}
-		\Re{}	&   \textrm{i}\Re{} &\Plus{}_i \\
-		\Im{}	& \textrm{i}\Im{}	&\Plus{}_i \\
+		\Real{}	&   \textrm{i}\Real{} &\Plus{}_i \\
+		\Imaginary{}	& \textrm{i}\Imaginary{}	&\Plus{}_i \\
 		0		&  0				& \Minus{}_i \\
@@
 	$\begin{bmatrix}
-		\Im{}	& \textrm{i}\Im{}	& \Plus{}_i \\
-		\Re{}	&  \textrm{i}\Re{}	& \Plus{}_i \\
+		\Imaginary{}	& \textrm{i}\Imaginary{}	& \Plus{}_i \\
+		\Real{}	&  \textrm{i}\Real{}	& \Plus{}_i \\
 		0		&  0				&\Minus{}_i 
 	\end{bmatrix}$ \\
+
+\subsection{Derivation}
+
+Consider the \SVI{} vector 	
+	\{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\};
+in \CIII{} that vector is:  \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}.
+Stated in \CIII{}, the problem is to find a matrix that converts a \CIII{}
+vector to a more reduced one. Take the case of \si positive and the
+other scalars negative.\\
+
+ \{(-\si,~\siv-\si ),~(\siii+\si,~\sii+\si), (\svi+\si,~\sv+\si)\}\\
+ =
+\begin{tabular}{ccc}
+	 $\begin{bmatrix}
+ 	~\\~\\~\\
+ \end{bmatrix}$
+
+$ \begin{Bmatrix}
+ $(\si,\siv )$\\ $(\sii,\sv)$\\$(\siii,\svi)$
+\end{Bmatrix}$
+\end{tabular}\\
+
+Here, one of the four possible result vectors has been chosen 
+(arbitrarily).
+
+Element 1 of result vector is (-\si,~\siv-\si ), and
+it is created by multiplying the beginning \CIII{} vector by the
+first row of the transformation matrix. In this case, we see
+that \ci of the result only contains scalars from \ci of
+the input, so the first row must contain some operator to
+be multiplied times input \ci, and the second and third
+elements of the matrix row must be zero. We create an operator
+($Minus~real$) to transform the input \ci to the output \ci; so
+the first row of the reduction matrix is: $\Minus{}_i	 0		 0$.
+
+Element 2 of the result vector is (\siii+\si,~\sii+\si),
+and
+it is created by multiplying the beginning \CIII{} vector by the
+second row of the transformation matrix. The result \cii
+contains \si in both the real and imaginary parts. So
+the first matrix element in that row must insert \si, the
+real part of the input \ci, into both the real and imaginary
+parts. An operator to produce that is created 
+(Plus{}~real). Next we see that the result contains \sii and \siii.
+\siii appears in the real part of the result element and in the
+third part of the input vector. So the third element of the 
+second matrix row must extract the \siii, the real part of 
+\ciii of the input. So the second row now contains the Plus real
+operator and the element to put the  real part of \ciii of the input.
+The final matrix element of the second row (the second element)
+must extract the real part of input \cii and place it in the
+imaginary part of the output vector. The resulting second
+row of the matrix is $\Plus_r~\Real~i\Real$.
+
+Using similar logic, the third row of the matrix is:
+$\Plus_r~\Imaginary~i\Imaginary$.
+
+Having derived the first matrix, the other 23 can
+be derived by the same process. However because all the
+\SVI boundaries are related by the 24 reflection, the
+other 23 matrices can be generated simply
+by multiplying by each of the reflection matrices.
 
@@
 		$1$	& $0$						&  $0$ \\
-		$0$	& \Re{} +\textrm{i}\Im{}	& $0$ \\
-		$0$	& $0$						& \Re{} +\textrm{i}\Im{} \\
+		$0$	& \Real{} +\textrm{i}\Imaginary{}	& $0$ \\
+		$0$	& $0$						& \Real{} +\textrm{i}\Imaginary{} \\
 	\end{bmatrix}$ 
@@
 			\Minus{}_r	& 0			& 0 \\
-			\Plus{}_r	&  \Re{}	& \textrm{i}\Re{} \\
-			\Plus{}_r	& \Im{}		&\textrm{i}\Im{}
+			\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
+			\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
 		\end{bmatrix}	$
@@
 	\begin{table}
-	\begin{tabular}{l c c c c c c c c c}
+		\begin{tabular}{m{0.08\textwidth} c c c c c c c c c}
 			\toprule
@@
 			\midrule
-	1DPY & R& 57.98& 57.98& 57.98& 92.02& 92.02& 92.02\\
-	1FE5 & R& 57.98& 57.98& 57.98& 92.02& 92.02& 92.02\\
-	1G0Z & H& 80.36& 80.36& 99.44& 90&    90&    120\\
-	1G2X & C& 80.95& 80.57& 57.1 & 90&    90.35&  90\\
-	1U4J & H& 80.36& 80.36& 99.44& 90&    90&    120\\
-	2OSN & R& 57.10& 57.10& 57.10& 89.75& 89.75&  89.75\\
+			1DPY & hR& 57.98& 57.98& 57.98& 92.02& 92.02& 92.02\\
+			~~\& 1FE5 \\
+			1G0Z  & hP& 80.36& 80.36& 99.44& 90&    90&    120\\
+			~~\& 1U4J\\
+			1G2X & mC& 80.95& 80.57& 57.1 & 90&    90.35&  90\\
+			2OSN & hR& 57.10& 57.10& 57.10& 89.75& 89.75&  89.75\\
 			\bottomrule
@@
 	\end{figure}
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{5_21}
+	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 10\%
+		orthogonally to \SVI{} vector. }
+	\label{fig5}	
+\end{figure}
+
 	\begin{figure}
@@
 	\begin{figure}
-	\includegraphics[width=1.0\textwidth]{4_13}
-	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 2\%
-	orthogonally to \SVI{} vector. }
-	\label{fig4}
+		\includegraphics[width=1.0\textwidth]{5_38}
+		\label{5_38}
+		\caption{The PLA2 cells Delone reduced and
+		5 copies perturbed by 10\%.}
 	\end{figure}
@@
 	\begin{figure}
-	\includegraphics[width=1.0\textwidth]{5_21}
-	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 10\%
-		orthogonally to \SVI{} vector. }
-	\label{fig5}
+		\includegraphics[width=1.0\textwidth]{6_53}
+		\label{6_53}
+		\caption{The same cells as plotted in Figure \ref{5_38} 
+			and Delone reduced.}
+	\end{figure}
+
+\begin{figure}
+	\includegraphics[width=1.0\textwidth]{7_14}
+	\label{7_14}
+		\caption{The same cells as plotted in Figure \ref{5_38} 
+	and Delone reduced, and then each multiplied by the 24
+	\SVI{} reflections.}
 \end{figure}
@@
 	\url{https://github.com/duck10/LatticeRepLib.git}.
+	The program CmdToC3 uses the required files.
 
@@
 
-	\bibliographystyle{iucrit}
+	\bibliographystyle{iucr}
 

### 2023-05-22 MeasuringLattices_uv5016/MeasuringLattices.tex
uv5016 -- fix quote marks C3 -- replying to reviewers

@@
 
-	"Measuring Lattices" is intended to be a summary of an invited talk
-	for satellite meeting, "Mathematics and Computer Science for Materials Innovation: Crystal Lattice Classifications" of August, 2022, ECM33
+	``Measuring Lattices'' is intended to be a summary of an invited talk
+	for satellite meeting, ``Mathematics and Computer Science for Materials Innovation: Crystal Lattice Classifications'' of August, 2022, ECM33
 	meeting (https://www.ecm33.fr/). The intent is describe in brief
 	the basics of implementing a metric space that creates
-	a measure of the "distances" between lattices.
+	a measure of the ``distances'' between lattices.
 
@@
 the five-fold ambiguity cited by \citeasnoun{Gruber1973} also causes confusion. The cited five-fold ambiguity is only for the particular
-case that Gruber listed, but interchanges of the axes to produce
+case that Gruber listed, but interchanges of the axes produce
 other, Buerger-reduced cells.

### 2023-05-31 C3/C3.tex

@@
 \begin{tabular}{ccc}
-	 $\begin{bmatrix}
- 	~\\~\\~\\
- \end{bmatrix}$
+
+ 	RedMat
 
@@
 
+We need to solve for RedMat.
 Here, one of the four possible result vectors has been chosen 
@@
 by multiplying by each of the reflection matrices.
+
+\subsection{Example of use}
+
+To demonstrate the application of the above matrices,
+example C++ code was written to create matrices of operators;
+see the supplementary material for the code or the complete
+library reference in the Availability of Code below.
+
+Create a \CIII{} vector c3 with
+the values \{(1,2),(3,4),(5,6)\} (which does not
+correspond to a valid unit cell). Multiply it by the matrix
+of operators derived above.
+
+The matrix multiply can be implemented as follows"
+
+\begin{verbatim}
+C3 C3::c3_s1(const C3& a) {
+    C3 c;
+    c[0] = c3_Minus_Real(a[0]);
+    c[1] = c3_Plus_Real(a[0]) + c3_Real(a[1]) +
+        c3_I_Times_Real(a[2]);
+    c[2] = c3_Plus_Real(a[0]) + c3_Imag(a[1]) +
+        c3_I_Times_Imag(a[2]);
+    return c;
+}
+\end{verbatim}
+
+Therefore, RedMat = \\
+\begin{center}
+	$\begin{bmatrix}
+	\Minus{}_r	& 0			& 0 \\
+	\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
+	\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
+\end{bmatrix}	$
+\end{center}
+
+For example:
+
+\begin{center}
+\{(-1.000,1.000) (4.000,6.000) (5.000,7.000)\} =\\
+$\begin{bmatrix}
+	\Minus{}_r	& 0			& 0 \\
+	\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
+	\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
+\end{bmatrix}	$
+$\begin{Bmatrix}
+	(1,2)\\
+	(3,4)\\
+	(5,6)
+\end{Bmatrix}$\\
+
+\end{center}
+
+which can be verified by hand.
 
@@
 \begin{figure}
+	\includegraphics[width=1.0\textwidth]{19_57}
+	label{19\_57}
+	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 10\%
+		orthogonally to \SVI{} vector, followed by Niggli reduction.}
+\end{figure}
+
+	\begin{figure}
 		\includegraphics[width=1.0\textwidth]{3_03}

### 2023-06-14 C3/C3.tex
responses to the reviewers' comment 2023-06-12

@@
 
-\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
+\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
 \usepackage{amssymb}
@@
 
+	The C3 vector transformations are represented as multiplication by a matrix of the operators defined in Table 1.
+When a complex vector component is 'multiplied' by an operator, the result is as defined in the table."
+
 	For the boundary at $s_1$: (the real component of $c_1$).
@@
 		\Plus{}_i	& \textrm{i}\Imaginary{}	& \Imaginary{} \\
-		\Plus{}_i	&  \textrm{i}\Real{}	& \Imaginary{} \\
+		\Plus{}_i	&  \textrm{i}\Real{}	& \Real{} \\
 	\end{bmatrix}$ 
@@
 		\Plus{}_i	& \Imaginary{}	& \textrm{i}\Imaginary{} \\
-		\Plus{}_i	&  \Real{}& \textrm{i}\Imaginary{} \\
+		\Plus{}_i	&  \Real{}& \textrm{i}\Real{} \\
 	\end{bmatrix}$ \\
@@
 		\textrm{i}\Real{}	& \Plus{}_r		& \Real{} \\
-	\end{bmatrix}$ e
+	\end{bmatrix}$
 	%
@@
 		0				& \Minus{}_i	& 0 \\
-		\textrm{i}\Real{}	& \Plus{}_ri	& \Real{} \\
+		\textrm{i}\Real{}	& \Plus{}_i	& \Real{} \\
 	\end{bmatrix}$ 
@@
 
+\subsection{Example C++ implementation}
+
+\begin{verbatim}
+static std::complex<double> c3_Minus_Real(std::complex<double>& a) {
+   const std::complex<double> c = { -a.real(), -a.real() + a.imag() };
+   return c;
+}
+
+C3 C3::c3_s1(const C3& a) {
+   C3 c;
+   c[0] = c3_Minus_Real(a[0]);
+   c[1] = c3_Plus_Real(a[0]) + c3_Real(a[1]) +
+   c3_I_Times_Real(a[2]);
+   c[2] = c3_Plus_Real(a[0]) + c3_Imag(a[1]) +
+   c3_I_Times_Imag(a[2]);
+   return c;
+}
+\end{verbatim}
+
 \subsection{Example of use}
@@
 
-The matrix multiply can be implemented as follows"
-
-\begin{verbatim}
-C3 C3::c3_s1(const C3& a) {
-    C3 c;
-    c[0] = c3_Minus_Real(a[0]);
-    c[1] = c3_Plus_Real(a[0]) + c3_Real(a[1]) +
-        c3_I_Times_Real(a[2]);
-    c[2] = c3_Plus_Real(a[0]) + c3_Imag(a[1]) +
-        c3_I_Times_Imag(a[2]);
-    return c;
-}
-\end{verbatim}
+The matrix multiplication can be implemented as follows (including
+only one example of the functions for operators (c3\_Minus\_Real))"
 
@@
 		\Minus{}_r	& 0			& 0 \\
-	\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-	\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
+		\Plus{}_r	&  \textrm{i}\Real{}	& \Real{} \\
+		\Plus{}_r	& \textrm{i}\Imaginary{}		&\Imaginary{}
 	\end{bmatrix}	$
@@
 		\Minus{}_r	& 0			& 0 \\
-	\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-	\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
+\Plus{}_r	&  \textrm{i}\Real{}	& \Real{} \\
+\Plus{}_r	& \textrm{i}\Imaginary{}		&\Imaginary{}
 	\end{bmatrix}	$
@@
 
-	The identity in \CIII{} can also be written:\\
+	The identity in \CIII{} could also be written:\\
 
@@
 	\\
+	which is simply replacement of the value one with an
+	equivalent representation as operators.
 
@@
 \begin{figure}
-	\includegraphics[width=1.0\textwidth]{19_57}
-	label{19\_57}
+	\includegraphics[width=1.0\textwidth]{19_52}
+	\label{19_57}
 	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 10\%

### 2023-07-14 C3/C3.tex
uv5017,uv5018,uv5019 fixing reviewer comments and page proofs

@@
 			$\Real$ 			&  $\Real(c_j)$ 			& $x_j$ or $(x_j,0)$		& Real \\ 
-			$\Imaginary$ 		&  $\Imaginary(c_j)$ 		& $y_j$ pr $(y_j,0)$ 		& Imaginary\\
+			$\Imaginary$ 		&  $\Imaginary(c_j)$ 		& $y_j$ or $(y_j,0)$ 		& Imaginary\\
 			\bottomrule
@@
 ($Minus~real$) to transform the input \ci to the output \ci; so
-the first row of the reduction matrix is: $\Minus{}_i	 0		 0$.
+the first row of the reduction matrix is: $\Minus{}_r	 0		 0$.
 
@@
 imaginary part of the output vector. The resulting second
-row of the matrix is $\Plus_r~\Real~i\Real$.
+row of the matrix is $\Plus_r~i\Real~\Real$.
 
 Using similar logic, the third row of the matrix is:
-$\Plus_r~\Imaginary~i\Imaginary$.
+$\Plus_r~i\Imaginary~\Imaginary$.
 

### 2023-07-21 SELLA/Sella.tex
adding SELLA paper

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
%\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage{bm}
\usepackage[fleqn]{amsmath}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `basename #1 .tif`.png}
%\usepackage{amsbsy}
\usepackage{float}
\usepackage{url}
%\usepackage{yfonts}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[outdir=./]{epstopdf}
\usepackage{epsfig}
\usepackage{booktabs}

\newcommand{\SVI}[0]{$\bf{S^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\scalarsub}[2]{$#1_#2$}
\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}

     %-------------------------------------------------------------------------
     % Information about journal to which submitted
     %-------------------------------------------------------------------------
     \journalcode{A}              % Indicate the journal to which submitted
                                  %   A - Acta Crystallographica Section A
                                  %   B - Acta Crystallographica Section B
                                  %   C - Acta Crystallographica Section C
                                  %   D - Acta Crystallographica Section D
                                  %   E - Acta Crystallographica Section E
                                  %   F - Acta Crystallographica Section F
                                  %   J - Journal of Applied Crystallography
                                  %   M - IUCrJ
                                  %   S - Journal of Synchrotron Radiation

\begin{document}                  % DO NOT DELETE THIS LINE

     %-------------------------------------------------------------------------
     % The introductory (header) part of the paper
     %-------------------------------------------------------------------------

     % The title of the paper. Use \shorttitle to indicate an abbreviated title
     % for use in running heads (you will need to uncomment it).

	{\LARGE \emph{\today}} \\
\title{SELLA - A Program for Determining Bravais Lattice Types}
%\shorttitle{SELLA  Determining Bravais Lattice Types}

     % Authors' names and addresses. Use \cauthor for the main (contact) author.
     % Use \author for all other authors. Use \aff for authors' affiliations.
     % Use lower-case letters in square brackets to link authors to their
     % affiliations; if there is only one affiliation address, remove the [a].

\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
\author[b]{Herbert J.}{Bernstein}
\author[c]{Nicholas K.}{Sauter}

% ORCID for LCA - 0000-0002-4451-1641

\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}, Rochester Institute of Technology, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
\aff[c]{Lawrence Berkeley National Laboratory, 1 Cyclotron Rd., Berkeley, CA, 94720 \country{USA}}

% Use \shortauthor to indicate an abbreviated author list for use in
% running heads (you will need to uncomment it).

\shortauthor{Andrews, Bernstein, and Sauter}

     % Use \vita if required to give biographical details (for authors of
     % invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
%\vita{Author's biography}

     % Keywords (required for Journal of Synchrotron Radiation only)
     % Use the \keyword macro for each word or phrase, e.g. 
     % \keyword{X-ray diffraction}\keyword{muscle}

%\keyword{keyword}
\keyword{Unit cell reduction}
\keyword{Delaunay}
\keyword{Delone}
\keyword{Niggli}
\keyword{Selling}
\keyword{Bravais Lattice}

     % PDB and NDB reference codes for structures referenced in the article and
     % deposited with the Protein Data Bank and Nucleic Acids Database (Acta
     % Crystallographica Section D). Repeat for each separate structure e.g
     % \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

%\PDBref[optional name]{refcode}
%\NDBref[optional name]{refcode}

\maketitle                        % DO NOT DELETE THIS LINE

\begin{synopsis}
A method for determining likely Bravais lattice types based on Selling (Delone) 
reduction. It is a complete, closed solution.
\end{synopsis}

\begin{abstract}
We introduce a new Bravais lattice determination algorithm. SELLA is a 
straight-forward algorithm and a program for determining Bravais lattice type 
based on Selling (Delone) reduction. It is a complete, closed solution, 
and it provides a clear metric of fit to each type.
\end{abstract}

{\bf Note:}  Boris Delaunay in his later publications used the Russian version 
of his surname: Delone. We will follow that choice.\\

     %-------------------------------------------------------------------------
     % The main body of the paper
     %-------------------------------------------------------------------------
     % Now enter the text of the document in multiple \section's, \subsection's
     % and \subsubsection's as required.

\section{Introduction}

We introduce a new Bravais lattice determination algorithm. The Bravais 
lattice types were created by \citeasnoun{Bravais1850}. \citeasnoun{Delaunay1932} 
and \citeasnoun{Niggli1928} developed methods for the identification 
of the Bravais lattice type of a crystal using the measured unit cell 
dimensions; their methods were exact only if the cell parameters 
exactly corresponded to the actual type \cite{Patterson1957}. 
Delone discussed the issues of scalars that have nearly zero values, 
but the broader issues of other measurement errors were not 
discussed. \citeasnoun{andrews2014} review the literature on 
the efforts to create methods to utilize data that contain unavoidable 
measurement error.

SELLA is a algorithm and program for determining Bravais lattice 
type based on Selling (Delone) reduction. It is a complete, closed solution, and it provides a clear metric of fit to each type.

\section{Terminology}

\citeasnoun{Delaunay1932} found that the 14 Bravais lattice types 
could be allocated among 24 types that are determined by the 
type of Dirichlet unit cell (termed the Voronoi region \cite{Voronoi1908} or 
Dirichlet region \cite{dirichlet1850} by 
Delone). Delone used the convention of \citeasnoun{Bravais1850}
with the sides of the tetrahedron labeled by the labels of the six
Selling scalars. Figure \ref{fig:PQRSTU} shows the labeling used in
the International Tables for Crystallography \cite{Henry1952}; 
Table \ref{table:ScalarConcordance} shows other choices that have been used. The
 contents for each type are described in Figure \ref{fig:terminology}.

CHARACTER

\citeasnoun{Burzlaff1985} revised the labeling of the types.
Here we have returned to the numbering of the types to that 
of \citeasnoun{Delaunay1932}.
In some cases, Delone included two forms within a single 
type. \citeasnoun{Burzlaff1985}
renumbered by ordinals. Here we have labeled the Delone 
types that have multiple items as "A" and "B". Table \ref{table:DeloneTypeConcordance}
shows the various labelings. Figure \ref{fig:DeloneTypes} 
recapitulates the figure of \citeasnoun{Delaunay1932} using 
more modern notation. Table \ref{table:ScalarConcordance} 
lists the various labelings that have been 
used for the sides of the Bravais tetrahedron. 
Figure \ref{fig:basictriangle} shows the arrangement of the 
symbols as used in the International Tables for Crystallography.

\begin{table}
		\begin{tabular}{cccccccccccc} 
			\toprule

	\rotatebox{80}{\citeasnoun{Delaunay1932}}
		  & \rotatebox{80}{\citeasnoun{Henry1952}  }
		   & \rotatebox{80}{\citeasnoun{Patterson1957} }
		 & \rotatebox{80}{\citeasnoun{Burzlaff1985}}
		   &\rotatebox{80}{\citeasnoun{andrews2019} } \\	
			\midrule
		g&p&P&h$_{23}$&\scalarsub{s}{1}\\		
		k&r&R&h$_{12}$&\scalarsub{s}{3}\\		
		l&s&S&h$_{14}$&\scalarsub{s}{4}\\		
		m&t&T&h$_{24}$&\scalarsub{s}{5}\\		
		h&q&Q&h$_{13}$&\scalarsub{s}{2}\\		
		n&u&U&h$_{34}$&\scalarsub{s}{6}\\		
		\bottomrule
	\end{tabular}

	\caption{The various terms used to describe the sides of 
		the Bravais tetrahedron. (In the figures in \citeasnoun{Delaunay1932} 
		the tetrahedron has been rotated by -120 degrees)}
	\label{table:ScalarConcordance}
\end{table}

\begin{table}
	\begin{tabular}{ccccccccc}
		\toprule
		 \cite{Delaunay1932} & \cite{Burzlaff1985}   &this  \\
		&  & paper \\
		\midrule
		K$_I$ & K1 & C1 \\		
		K$_{III}$ & K2 & C3 \\		
		K$_V$ & K3 & C5 \\		
		Q$_I$ & Q1 & T1 \\		
		Q$_{II}$ & Q2 & T2 \\		
		Q$_V$ & Q3 & T5\\		
		R$_I$ & R1 & R1 \\		
		R$_{III}$ & R2 & R3 \\		
		O$_I$ & O1 & O1A  \\		
		O$_I$ & O2 & O1B  \\		
		O$_{II}$ & O3 & O2   \\		
		O$_{III}$ & O4 & O3   \\		
		O$_{IV}$ & O5 & O4   \\		
		O$_V$ & O6 & O5   \\		
		M$_I$ & M1 & M1A  \\		
		M$_I$ & M2 & M1B  \\		
		M$_{II_2}$  & M3 & M2A  \\		
		M$_{II}$ & M4 & M2B  \\		
		M$_{III}$ & M5 & M3  \\		
		M$_{IV}$ & M6 & M4  \\		
		T$_I$ & T1 & A1  \\ 		
		T$_{II}$ & T2 & A2 \\		
		T$_{III}$ & T3 & A3 \\		
		H$_{IV}$  & H1 & H4 \\ 
		\bottomrule
	\end{tabular}
~\\ ~\\
	\caption{The 24 Delone types. \cite{Delaunay1932}}
	\label{table:DeloneTypeConcordance}
\end{table}

\subsection{The space \SVI{}}
\cite{andrews2019} recast the Selling parameters to define a metric space, \SVI{}.
We use the base vectors, $\bf{a}$, $\bf{b}$, $\bf{c}$ of the unit cell to define $\bf{d}=-(\bf{a}+\bf{b}+\bf{c})$.
A point ${s}$ in \SVI{} is defined as:\\
${s}$= [\scalarsub{s}{1}, \scalarsub{s}{2}, \scalarsub{s}{3},\scalarsub{s}{4}, \scalarsub{s}{5}, \scalarsub{s}{6}] \\
where \scalarsub{s}{1}=\vdotv{b}{c}, ~\scalarsub{s}{2}=\vdotv{a}{c}, ~\scalarsub{s}{3}=\vdotv{a}{b}, 
~\scalarsub{s}{4}=\vdotv{a}{d}, ~\scalarsub{s}{5}= \vdotv{b}{d}, and \scalarsub{s}{6}=\vdotv{c}{d}.

\subsection{Terminology of the Bravais lattice types}
The Bravais tetrahedron symbol. Figure \ref{fig:PQRSTU}  and Table \ref{table:DeloneTypeConcordance}\\
The different ways that the scalars have been labeled  Figure \ref{fig:terminology} \\
The table of the Delone types. Figure \ref{fig:DeloneTypes} \\

\section{Algorithms for creation of an exhaustive list of the polytopes of the Bravais types}

\subsection{Deriving the projectors and perps}

We will use Selling reduction; Niggli reduction has a fundamental 
unit that is non-convex, and some boundaries of the fundamental 
unit are partly closed and partly open \cite{andrews2014} leading 
to considerable complexity. Because Niggli reduction divides the 
fundamental unit into separate all obtuse and all acute regions, 
 the transitions between those are quite complex \cite{andrews2014}. 
In space $\bf{S^{6}}$ \cite{andrews2019b}, the fundamental 
unit is the simple all-negative orthant of a 6-dimensional 
Cartesian space. Only two kinds of operations in the space 
will need to be considered: the 6 boundary transform operations 
at the boundaries of the orthant (zeros of an axis), and 
24 reflections \cite{andrews2019}

Twenty-four lattice types were enumerated by \citeasnoun{Delaunay1932}. 
Three of those are triclinic, and we will ignore them because all crystals 
can be indexed as triclinic. We 
shall need all representations of all non-triclinic lattice 
types within the $\bf{S^{6}}$ fundamental unit (all Selling-reduced 
cells) or those generated beyond the boundaries by boundary transforms. 
Three forms (M3, M2B, and O3) are reduced dimension boundaries that do
not correspond to unique Bravais types; any lattice that falls in those
three is indexed by one of the other types.

 Seven of the 21 Delone types do not have a zero in their $\bf{S^{6}}$ 
 representation; the presence of zero indicates that a form is 
 on the boundary of the $\bf{S^{6}}$ fundamental unit. Since a 
 point on the boundary transforms to another boundary 
 point \cite{andrews2019b}, we can generate the Virtual 
 Cartesian Point (VCP) \cite{andrews2019b} appropriate to 
 each zero in the $\bf{S^{6}}$ vector. Because the reduction
  operations do not commute, in the cases where there is more 
  than one zero in the vector, all orderings of applying 
  reductions to all of the zeros must be generated. (For example, 
  for two zeros, there will be six vectors produced.) Again, 
  there may be many duplicate results.

 Finally, the projectors must be computed for each of the many 
 sample vectors of the 21 types. We will also need the 
 ``perps'', the operators that give the normal vector to each 
 such manifold. For a particular polytope, the projector applied to the probe gives
 the least-squares best fit point within that polytope. Perps are computed by subtracting the 
 corresponding projector from the unit matrix. The norm of a perp times
 a probe is the distance from that polytope.

 \subsubsection{Generating sample vectors from all Bravais types}

 Step 1: Generate any random vector that contains no zero 
 values and no duplicate values.  For a later operation, 
 it should only contain values large enough that adding or
 subtracting the smallest
 representable floating point number will not change the 
 value in computer arithmetic ($DBL\_MIN$ in the C language).

 Step 2; For each Delone type, multiply its projector by 
 the vector from Step 1. Store  the results in a list 
 with entries for each type (see Table \ref{table:DeloneTypes} 
 for the projectors).

 Step 3:

a) If the vector contains no zeros, generate the 24 reflections
 and store in the list by type.

\textit{\textbf{LCA--need to normalize the vectors, just like in lattice matching. ? Can we use the 
lattice matching code here?}}

b) If the vector contains one zero, apply the boundary transform operation for that boundary
to the vector, and
 store the products of the vector and its 24 reflections in the list. Note, 
 this is where the choice of $DBL\_MIN$ has effect; adding or
 subtracting such a small value will not change
 any value but zero.

 c) If the vector contains more than one zero and \underline{not} $DBL\_MIN$, 
 first set the zero to $DBL\_MIN$,  and then return it to 
 step b). Then set the second zero to $DBL\_MIN$ (with the
 first zero still zero), and send that result to step b). 
 If there is a third zero,  process it the same way as two zeros and return to c).

 Continue those steps until all the types have been processed. 
 There will be duplicates within each type,
 which will be eliminated later.

 \subsubsection{Generate projectors from sample vectors}

 Each vector in the list from the previous steps is a representative of one
 of the Delone types. Although a vector is a one-dimensional polytope,
 these vectors encode the information about the full polytopes of the
 Delone types in their zeros and their duplicate values.

 We begin by defining a function: \\ \relax
\underline{Fraction} computes the reciprocal of the 
number of elements that equal \\ \relax
 an input value or zero if there are no duplicates.

 The following steps will be repeated for each of the vectors in the
 list generated above. We generate a matrix \textit{m}.

 Step 1: Zero any elements that have been set to $DBL\_MIN$.

 Step 2: For each of the six scalars

 For i in 1-6
 	\textit{m}[i,i] = \{if \textit{Fraction}(s[i]) == 0.0 or abs(s[i]) == 0.0\} then 0.0 else \textit{Fraction}(s[i])

For i in 1-6
For j in 1-6
\textit{m}[i,j] 	= \{if s[i]==s[j] or abs(s[j]) $>$ 0\} then~\textit{Fraction}(s[j]) else 0.0

Add m to a list of the projectors.

To finish, the duplicate projectors are removed within each type, 
leaving 239 non-triclinic projectors. (If the triclinic cases are 
included, there are 10 more.) See Table \ref{table:TypeCounts}.

\subsection {Measuring the fit}
The distance from the given polytope is just
the norm of the perp for that polytope applied to the probe.

\section{Degenerate Delone types}

Three of the 21 non-triclinic Delone types are only boundaries of other types. In Figure 
\ref{fig:degenerate}, Delone types M2B and M3 are in the row for orthorhombic, and
O3 is in the row for rhombohedral and tetragonal. These types do not need to
be searched for in general surveys.

\section{Verification}

\subsection{Rationale}

Most of the proposals for Bravais lattice determination in the 
presence of experimental error do not have a clear proof that 
they are complete. For many years there were verbal complaints 
that available programs would not infrequently miss experimental 
cases (for example, TRACER \cite{lawton1965} and many anecdotal 
communications).  To our knowledge, only the method of 
\cite{tomiyasu2012} has clearly demonstrated completeness. 
The justification that Sella is complete has three parts.

First: We require that the cell being tested be reduced, and
all Bravais lattice types within the \SVI{} fundamental unit (all
\SVI{} scalars negative) be generated. Therefore,
 the 24 possible forms of a 
reduced cell will be close to one of the 24 copies of the types. 
For the reflections, we do not need to consider the case of 
points near a boundary that might be outside; the boundary 
transforms must be treated for that.

Second: In the case of the boundary transform operations, we take the case of a point that is infinitesimally distant from the polytope of a 
Bravais lattice type that has one zero in its scalars. Take 
the point in the polytope that is closest to the experimental 
point. We transform both by the corresponding boundary transform 
operation. We have generated a new pair of points in a 
different location, one of which is in some representation of 
the same Bravais lattice type. The points are still 
infinitesimally distant from each other.

Third: Assume we have a point that is outside the fundamental 
unit and is near an undiscovered polytope for a Bravais lattice 
type. If the point is near a boundary defined by a single 
zero, then we apply a boundary transform to the point and to 
the Bravais polytope, bringing them both into the fundamental 
unit. If all of the variations of all of the Bravais 
lattice types within and on the fundamental unit have 
been generated, then this putative undiscovered polytope 
has already been included in the list of possibilities. 
Similar arguments apply to the cases of two or three zeros.

If we apply all of the boundary transform operations to the 
representations of the Bravais lattice types, then 
there are no places that a point in the fundamental 
unit will fail to find the closest Bravais lattice 
(or lattices). As stated above, there are 239 
non-triclinic possible representations.

What might occur if some reduced point has an unreduced 
copy near the all-negative orthant?  Would it be 
possible for it to be close to one of the copies of one 
of the 21 Delone types? The reduced point is obviously in 
the all-negative orthant. Further, because all possible 
copies of all of the Delone type manifolds have been 
generated, the reduced point will be near a copy of the 
manifold it was near when it was unreduced.

\subsection{Testing against known cells}

89539 unit cells were extracted from the Protein Data Bank 
(\citeasnoun{bernstein1977} and \citeasnoun{berman2000}). 
When evaluated with Sella, all but 57 were found at zero 
distance from the described crystal class. The 57 were 
identified as incorrect unit cells for the assigned 
crystal class; however, the distances from the assigned 
classes were small because the deviations from the 
appropriate parameters were modest (see Table \ref{tab:PDBCells}).
\begin{table}
	\label{tab:PDBCells}

\begin{tabular}{ccccccc}
	\toprule
PDB ID & Sp.Gr. & Delone type&distance ({{\AA}}) \\ 
\midrule
		1NUI&P3121&H4&0.0439\\ 
		2NW8&P3121&H4&0.0401\\ 
		2J1L&P3121&H4&0.0307\\ 
		1VYI&P3121&H4&0.2504\\ 
		3FZ8&P32&H4&0.0575\\ 
		2J5W&P3221&H4&0.0543\\ 
		2VAM&P3221&H4&0.1166\\ 
		1Y8J&P3221&H4&0.0389\\ 
		1F4V&P3221&H4&0.0276\\ 
		2ZP9&P6&H4&0.0528\\ 
		1DDR&P61&H4&0.0726\\ 
		1DDS&P61&H4&0.0726\\ 
		1SGU&P61&H4&0.0295\\ 
		3CDC&P61&H4&0.0559\\ 
		2WAG&P6122&H4&0.0377\\ 
		1LBM&P6122&H4&0.0312\\ 
		2O9Z&P62&H4&0.0402\\ 
		1ELZ&P6322&H4&0.4035\\ 
		1SA0&P65&H4&0.0681\\ 
		3KQI&P6522&H4&0.0331\\ 
		2F9Y&P6522&H4&0.0470\\ 
		3E73&P6522&H4&0.0523\\ 
		3CAP&H3&H4&0.0586\\ 
		2UX6&H32&H4&0.0386\\ 
		1ODT&H32&H4&0.0816\\ 
		2FDI&P43&T5&0.0758\\ 
		3HQ7&P43212&T5&0.2869\\ 
		1E2Q&P43212&T5&0.5482\\ \\
\bottomrule
\end{tabular}
\begin{tabular}{ccccccc}
	\toprule
	PDB ID & Sp.Gr. & Delone type&distance ({{\AA}}) \\ 
	\midrule
			1SHN&P43212&T5&0.0491\\ 
			1C50&P43212&T5&1.9040\\ 
			2A1L&P43212&T5&0.0528\\ 
			1S2L&P43212&T5&0.0334\\ 
			1H9S&P41&T5&1.0149\\ 
			1H9R&P41&T5&1.0141\\ 
			2F0Q&P41&T5&0.0260\\ 
			3EYM&P41&T5&0.0476\\ 
			1WV5&P41212&T5&0.0250\\ 
			2DF3&P41212&T5&0.0387\\ 
			1L6O&P41212&T5&0.0346\\ 
			3A58&P41212&T5&0.2976\\ 
			2IU9&P41212&T5&0.0373\\ 
			4BTP&P41212&T5&0.0666\\ 
			1W54&P4212&T5&0.0420\\ 
			2PBE&P42212&T5&0.0548\\ 
			3BR5&P42212&T5&0.2203\\ 
			1GMD&P42212&T5&0.1400\\ 
			1GMC&P42212&T5&0.1402\\ 
			2CCN&P42212&T5&0.0447\\ 
			1CE1&P212121&O5&0.6472\\ 
			3DQ2&P212121&O5&0.1058\\ 
			1Z2A&P212121&O5&0.0670\\ 
			2HA3&P212121&O5&0.3070\\ 
			3BNJ&I4122&T2&0.9518\\ 
			4MGP&I-42d&T1&0.2128\\ 
			4ASL&C2221&O4&0.3383\\ 
			3K7M&P432&C5&0.0517\\ 
			3GBN&I213&C1&1.2661\\ 
	\bottomrule
\end{tabular}

	\caption{}
\end{table}
\subsection{Testing for continuity}

The polytopes of the Bravais lattice types are characterized by required zeros
of scalars and/or multiplets of equal values. For instance, all primitive 
monoclinic cells have two 90 degree angles and so two zero scalars. Consider R1 as
an example for multiplets. The character for R1 is $ \bf{rrr sss} $; two triplets.

For the case of M4, primitive monoclinic, the character is $ \bf{00r stu} $. 
For a particular value for r,s,t,u, the best monoclinic cell will not
depend on whatever values are substituted for the zeros. In the plane
corresponding to the two zero scalars, we can draw a circle, each point
on the circle corresponding to a lattice. Reducing each lattice and calculating
the distance of the reduced cell from $ \bf{00r stu} $ will give a curve that
should not have any discontinuities. See Figure \ref{fig:M4}.

Similarly, Bravais type O5 has a triplet of zeros, character $ \bf{000 rst} $. Here we
can generate a sphere that surrounds the orthorhombic point. See Figure \ref{fig:O5}.

Bravais type O4 has two zeros and a pair of required equal values, 
character $ \bf{00r sst} $. The zeros can be treated the same as above. 
The pair of equal values must 
have an average value that remains equal to $\bf{s}$ in the lattice character. 
That leaves 3 values to vary, and we can create a sphere around 
the point $ \bf{00r sst} $. See Figure \ref{fig:O4}.

\section{Comparisons}

Comparison to other distance measure; Data for other measures taken
from \citeasnoun{andrews2014}.

\begin{tabular}{ccccccc}
	\toprule
	Lattice character&G6 distance  &BGAOL   Z score&XDS QI &Scaled QI &Sella\\	
	\midrule
			mP               &     20.138  & 0.657    &   1.0  & 0.13  & 1.06\\	
			oC               &     125.958 &  3.560   &    23. & 8 3.09& 13.82\\	
			mC               &     125.150 &  4.085   &    23. & 4 3.03& 13.80\\	
	\bottomrule
\end{tabular}

~\\

     % Appendices appear after the main body of the text. They are prefixed by
     % a single \appendix declaration, and are then structured just like the
     % body text.

     %-------------------------------------------------------------------------
     % The back matter of the paper - acknowledgements and references
     %-------------------------------------------------------------------------

     % Acknowledgements come after the appendices

	\ack{{\bf Acknowledgements}}
	Careful copy-editing and corrections by Frances C. Bernstein are 
gratefully acknowledged.  	Our thanks to Jean Jakoncic and Alexei Soares for 
helpful conversations and access to data and facilties at 
Brookhaven National Laboratory.
     % References are at the end of the document, between \begin{references}
     % and \end{references} tags. Each reference is in a \reference entry.

	\ack{{\bf Funding information}}

Funding for this research was provided in part by: US Department of Energy Offices of Biological and Environmental Research and of Basic Energy Sciences (grant No. DE-AC02-98CH10886 ; grant No. E-SC0012704); U.S. National Institutes of Health (grant No. P41RR012408; grant No. P41GM103473; grant No. P41GM111244; grant No. R01GM117126, grant No. 1R21GM129570; grant No. P30GM133893); Dectris, Ltd.

\bibliographystyle{iucr}
\bibliography{Reduced}

     %-------------------------------------------------------------------------
     % TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
     %-------------------------------------------------------------------------

     % Simple tables should use the tabular environment according to this
     % model

\begin{table}
\caption{The 21 non-triclinic Delone types with the count of representations.
The asterisks indicate non-crystallographic types.}
\label{table:TypeCounts}
\begin{tabular}{lr}
Delone type & Count    \\
\hline
H4&12\\
C1&1\\
C3&3\\
C5&16\\
R1&4\\
R3&12\\
T1&3\\
T2&6\\
T5&48\\
O1A&3\\
O1B&1\\
O2&6\\
O3$^*$&9\\
O4&36\\
O5&16\\
M1A&6\\
M1B&3\\
M2A&12\\
M2B$^*$&12\\
M3$^*$&18\\
M4&12\\
\bottomrule
\end{tabular}
\end{table}

\fontseries{Courier}{
\begin{table}
	\caption{Projectors for the 24 Delone types}
	\label{table:DeloneTypes}
	\begin{tabular}{lcl}
		\toprule
C1  & (cI) & [1 1 1 1 1 1/  1 1 1 1 1 1/  1 1 1 1 1 1/  1 1 1 1 1 1/  1 1 1 1 1 1/  1 1 1 1 1 1] \\ 
C3  & (cF) & [1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 0/  1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 0] \\ 
C5  & (cP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 1 1/  0 0 0 1 1 1/  0 0 0 1 1 1] \\ \\

T1  & (tI) & [1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 1 0 0 1/  1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 1 0 0 1] \\ 
T2  & (tI) & [1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 0/  1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 1] \\ 
T5  & (tP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 1 0/  0 0 0 1 1 0/  0 0 0 0 0 1] \\ \\

R1  & (rP) & [1 1 1 0 0 0/  1 1 1 0 0 0/  1 1 1 0 0 0/  0 0 0 1 1 1/  0 0 0 1 1 1/  0 0 0 1 1 1] \\ 
R3  & (rP) & [1 1 0 0 1 0/  1 1 0 0 1 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  1 1 0 0 1 0/  0 0 0 0 0 0] \\ \\

O1A & (oF) & [1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 1 0 0 0/  1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 1] \\ 
O1B & (oI) & [1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 1 0 0 1/  1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 1 0 0 1] \\ 
O2  & (oI) & [1 0 0 0 1 0/  0 1 0 1 0 0/  0 0 0 0 0 0/  0 1 0 1 0 0/  1 0 0 0 1 0/  0 0 0 0 0 1] \\ 
O3  & (oI) & [1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0/  1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0] \\ 
O4  & (oS) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 1 0 0 0/  0 0 0 1 1 0/  0 0 0 1 1 0/  0 0 0 0 0 1] \\ 
O5  & (oP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ \\

M1A & (mC) & [1 1 0 0 0 0/  1 1 0 0 0 0/  0 0 1 0 0 0/  0 0 0 1 1 0/  0 0 0 1 1 0/  0 0 0 0 0 1] \\ 
M1B & (mC) & [1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 1 0 0 0/  1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 1] \\ 
M2A & (mC) & [1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0/  1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 1] \\ 
M2B & (mC) & [1 0 0 0 0 0/  0 1 0 1 0 0/  0 0 0 0 0 0/  0 1 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ 
M3  & (mC) & [1 0 0 0 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0] \\ 
M4  & (mP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 1 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ \\

A1  & (aP) & [1 0 0 0 0 0/  0 1 0 0 0 0/  0 0 1 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ 
A2  & (aP) & [1 0 0 0 0 0/  0 1 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ 
A3  & (aP) & [1 0 0 0 0 0/  0 1 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 0] \\ \\

H4  & (hP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 1 1 1 0/  0 0 1 1 1 0/  0 0 1 1 1 0/  0 0 0 0 0 1] \\ 
\bottomrule
	\end{tabular}
\end{table}
}
     % Postscript figures can be included with multiple figure blocks

\begin{figure}
\caption{Bravais triangle with labeling as used in the International Tables}
\includegraphics[width=3cm]{PQRSTU}
\label{fig:PQRSTU}
\end{figure}

\begin{figure}
	\caption{Delone type description}
	\label{fig:terminology}
	\includegraphics[width=12cm]{terminology}
\end{figure}

\begin{figure}
	\caption{The Delone types}
	\label{fig:DeloneTypes}
	\includegraphics[width=12cm]{DeloneTypesGrid_2}
\end{figure}

\begin{figure}
	\caption{}
	\label{fig:degenerate}
	\includegraphics[width=12cm]{NullSpaceDistribution}
\end{figure}

\begin{figure}
	\label{fig:M4}
	\caption{Delone type M4: there are two zeros. A circle of 100 points with
		radius 0.1 is centered at 0,0. The lattices are reduced and the points are
		plotted at the distance from M4.} The red points indicate the original
	circle. The blue points are for the reduced cells.
	\includegraphics[width=12cm]{M4-100-2-zeros}
\end{figure}

\begin{figure}
	\caption{Delone type O5: there are three zeros. A sphere 
		is centered at 0,0,0, with radius 0.1. The lattices are 
		reduced, and the point is 
		placed at a distance equal to the distance from type O5. 50,000
		points are plotted.}
	\label{fig:O5}
	\includegraphics[width=12cm]{O5-50K-3-zeros}
\end{figure}

\begin{figure}
	\caption{Delone type O4: There are two zeros and one pair.
	A sphere with radius 0.1 was placed around one point. At each
sphere point, the lattice was reduced and the point on the sphere
was assigned a color (0.01) to blue (0.0075). 100,000 points are plotted.}
	\label{fig:O4}
	\includegraphics[width=12cm]{O4_Cropped-2-zeros-1-pair}
\end{figure}

\end{document}                    % DO NOT DELETE THIS LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-10-04 DC7unsrt_uv5017/DC7unsrt.tex
udate and add References for Web Site

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
%\documentclass{iucr}              % DO NOT DELETE THIS LINE
\usepackage{graphicx}
\usepackage{amssymb}
%\usepackage{epstopdf}
\usepackage{url}
%\usepackage{bm}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\usepackage{bm}
\usepackage[fleqn]{amsmath}

\newcommand{\vdotv}[2]{${\mathbf{#1} \cdot \mathbf{#2}}$}
\newcommand{\Imaginary}[0]{$ \textfrak{I} $}
\newcommand{\Real}[0]{$ \textfrak{R} $}
\newcommand{\Exchange}[0]{$\textfrak{X}$}

\newcommand{\nounderline}[3]{\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineab}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineac}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
\newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
\newcommand{\SVI}[0]{$\mathbf{S}^{6}$}
\newcommand{\GVI}[0]{$\mathbf{G}^{6}$}
\newcommand{\CIII}[0]{$\mathbf{C}^{3}$}

     %-------------------------------------------------------------------------
     % Information about journal to which submitted
     %-------------------------------------------------------------------------
     \journalcode{A}              % Indicate the journal to which submitted
                                  %   A - Acta Crystallographica Section A
                                  %   B - Acta Crystallographica Section B
                                  %   C - Acta Crystallographica Section C
                                  %   D - Acta Crystallographica Section D
                                  %   E - Acta Crystallographica Section E
                                  %   F - Acta Crystallographica Section F
                                  %   J - Journal of Applied Crystallography
                                  %   M - IUCrJ
                                  %   S - Journal of Synchrotron Radiation

\begin{document}                  % DO NOT DELETE THIS LINE

     %-------------------------------------------------------------------------
     % The introductory (header) part of the paper
     %-------------------------------------------------------------------------

     % The title of the paper. Use \shorttitle to indicate an abbreviated title
     % for use in running heads (you will need to uncomment it).

\title{An Invertible Seven-Dimensional Dirichlet Cell Characterization of Lattices}
\shorttitle{DC7unsrt}

     % Authors' names and addresses. Use \cauthor for the main (contact) author.
     % Use \author for all other authors. Use \aff for authors' affiliations.
     % Use lower-case letters in square brackets to link authors to their
     % affiliations; if there is only one affiliation address, remove the [a].

\cauthor[a]{Herbert J.}{Bernstein}{yayahjb@gmail.com}{}
\author[b]{Lawrence C.}{Andrews}
\author[c,d]{Mario}{Xerri}

\aff[a]{Ronin Institute for Independent Scholarship, c/o NSLS-II, Brookhaven National Laboratory, Bldg 745, P.O. Box 5000, Upton, NY \country{USA}}
\aff[b]{Ronin Institute for Independent Scholarship, Kirkland, WA \country{USA} }
\aff[c]{College of Engineering, Cornell Univ., Ithaca, NY \country{USA} }
\aff[d]{Brookhaven National Laboratory, Bldg 745, P.O. Box 5000, Upton, NY \country{USA}}
     % Use \shortauthor to indicate an abbreviated author list for use in
     % running heads (you will need to uncomment it).

\shortauthor{Bernstein, Andrews and Xerri}

     % Use \vita if required to give biographical details (for authors of
     % invited review papers only). Uncomment it.

%\vita{Author's biography}

     % Keywords (required for Journal of Synchrotron Radiation only)
     % Use the \keyword macro for each word or phrase, e.g. 
     % \keyword{X-ray diffraction}\keyword{muscle}

%\keyword{keyword}
\keyword{Dirichlet Cell}
\keyword{Voronoi Cell}
\keyword{Wigner-Seitz Cell}
\keyword{Niggli Reduction}

     % PDB and NDB reference codes for structures referenced in the article and
     % deposited with the Protein Data Bank and Nucleic Acids Database (Acta
     % Crystallographica Section D). Repeat for each separate structure e.g
     % \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

%\PDBref[optional name]{refcode}
%\NDBref[optional name]{refcode}

\maketitle                        % DO NOT DELETE THIS LINE

\begin{synopsis}
Starting from a Niggli-reduced cell, a crystallographic lattice may be characterized by seven parameters describing the Dirichlet cell: the three 
shortest non-coplanar lattice vector lengths, the three shorter 
of each pair of face diagonal lengths and the shortest body diagonal length,
from which the Niggli-reduced cell may be recovered.
\end{synopsis}

\begin{abstract}
Characterization of crystallographic lattices is an important tool in
structure solution, crystallographic database searches and
clustering of diffraction images in serial crystallography.
Characterization of lattices by Niggli-reduced cells
(based on the three shortest non-coplanar lattice vectors)
or by Delaunay-reduced cells (based on four non-coplanar 
vectors summing to zero
and all meeting at obtuse or right angles) are commonly used.  The
Niggli cell derives from Minkowski reduction.  The Delaunay cell derives
from Selling reduction.  All are related to the Wigner-Seitz (or
Dirichlet, or Voronoi) cell of the lattice, which consists of the
points at least as close to a chosen lattice point as they are to any 
other lattice point.  We call the three non-coplanar lattice vectors
chosen, the Niggli-reduced cell edges.
Starting from a Niggli-reduced cell, the Dirichlet cell is
characterized by the planes determined by thirteen lattice half-edges:
the midpoints of the three Niggli cell edges, the six Niggli cell face-diagonals and the four body-diagonals, but seven of the
lengths are sufficient:  three edge lengths, the three shorter of
each pair of face-diagonal lengths, and the shortest body-diagonal length.
These seven are sufficient to recover the 
Niggli-reduced cell.

\end{abstract}

     %-------------------------------------------------------------------------
     % The main body of the paper
     %-------------------------------------------------------------------------
     % Now enter the text of the document in multiple \section's, \subsection's
     % and \subsubsection's as required.

\section{Introduction}

Algorithms for quantifying the differences among lattices are used for 
Bravais lattice determination, for database lookup of unit cells to select 
candidates for molecular replacement, and for clustering to 
group together images from serial crystallography. For crystallography,
there are many alternative representations to choose from as 
a basis for distance calculations.  \citeasnoun{Andrews1980} defined
$\mathbf{V}^7$, a perturbation-stable space in which, using real and 
reciprocal space Niggli
reduction, a lattice is represented by three cell-edge lengths, three reciprocal 
cell-edge lengths, and the cell volume.  While suitable for database
searches, $\mathbf{V}^7$ was found to have difficulties in some other uses. For lattice type determination, issues arose when working near right angles. 
 \citeasnoun{Andrews1988} then
defined $\mathbf{G}^6$ that uses a modified metric tensor and an iterative search through 25 alternative
reduction boundary transforms \cite{Gruber1973} to work in a satisfactory manner both for
database searches and lattice identification in the presence of experimental
error.  \citeasnoun{Andrews2014} discussed sewing together regions of the fundamental
region of $\mathbf{G}^6$ under Niggli reduction at fifteen boundaries.  \citeasnoun{Andrews2019} 
presented the simplest and  fastest currently known representation of lattices as the six Selling 
scalars obtained from the dot products of the unit cell axes in 
addition to the negative of their sum (a body diagonal). Labeling 
three linearly independent vectors $\mathbf{a},\, \mathbf{b}, \mathbf{c}$ spanning the lattice, and $\mathbf{d}$ ($\mathbf{d} = -\mathbf{a}-\!\mathbf{b}-\!\mathbf{c}$), 
the Selling scalars are

\begin{center}
	\vdotv{b}{c},~ \vdotv{a}{c},~ \vdotv{a}{b},
	~ \vdotv{a}{d},~ \vdotv{b}{d},~ \vdotv{c}{d}
\end{center}

\noindent{}(where, {\it e.g.}, \vdotv{b}{c}   represents the dot product of 
	the $\bf{b}$  and $\bf{c}$  axes). For the purpose of organizing 
	these six quantities as a vector space in which one can
compute simple Euclidean distances, we describe the set of scalars as a 
vector, $\bf{s}$, with components, $ s_1, s_2, s_3, ..., s_6 $.  
The cell is Selling-reduced if all six components are negative or zero \cite{Delone1933}.  
Reversing Allman's observation that a Buerger-reduced 
cell is a good stepping stone to a Selling-reduced cell \cite{Allmann1968},
a Selling-reduced cell can be an efficient stepping stone to
a  Buerger-reduced cell that quickly reduces to a Niggli-reduced cell.
Note that the negatives of the Selling scalars of a Selling-reduced cell
are non-negative so that the six square roots provide a convenient six-parameter 
characterization of a lattice \cite{kurlin2022complete}.

In this paper we consider lattice representation based on
the Wigner-Seitz \cite{wigner1933} (or Dirichlet, or Voronoi) cell of the lattice, which consists of the points at least as close to a chosen lattice point as they are to any other lattice point.  
Starting from a Niggli-reduced cell, the Dirichlet cell is
characterized by the planes determined by thirteen lattice half-edges:
the midpoints of the three Niggli-reduced cell edges, the six Niggli-reduced cell face-diagonals, and the four body-diagonals, but seven of the Niggli-reduced cell lattice vector
lengths are sufficient:  three Niggli-reduced cell-edge lengths, the three shorter of each pair of 
face-diagonal lengths, and the shortest body-diagonal length.
The Niggli-reduced cell may be recovered easily from these seven
quantities.

A Wigner-Seitz cell is a polyhedron of six, eight, ten, twelve or fourteen
faces.  The general fourteen-face case is a truncated octahedron.  See
Fig. \ref{fig:truncoctahedron}.

When one is creating a metric to compute distances between lattices they
are likely to be based directly or indirectly either on the components
of the Niggli-reduced cell or on the components of the Selling-reduced
cell.  One disadvantage in use of the former is that the space
of Niggli-reduced cells admits both all-acute and all-obtuse presentations,
dividing the space into two distinct components with a disruptive boundary
between them.  Use of Selling-reduction avoids this problem by restricting
our attention to the all-obtuse case, making it easier to find smooth
metrics.  The Wigner-Seitz cell is often presented entirely in the context
of Selling-reduction, but in this paper we are using the approach of \cite{Hart2019}
and working in terms of Niggli-reduction in order to understand the primary
static characteristics of the space, especially the invertibility of
this presentation.  The question of improving stability -- especially when
computing distances between latices in the $---$ and $+++$ components is left for a future
investigation.

\section{Background}

That crystals are built from some regular assembly of basic parts was clear already in ancient times.    In 1611 Kepler described this relationship (\citeasnoun{Kepler1611}, translated in \citeasnoun {kepler1966}). Steno was asked to prepare a catalog of a ''cabinet of curiosities"; this is considered the first database of crystals (minerals in this case) \cite{Steno1669}.  See Fig. \ref{fig:timeline} for a timeline
of lattice characterization from Steno onwards.  In the 19th century 
indices were published with interaxial angles of crystals, specifically 
for the identification of minerals.

Following the discovery of x-rays, 
catalogs of unit cell parameters started to be published \cite{wyckoff1931}.
Often, these were arranged by crystal system and then sorted by some of the cell parameters. However, related minerals with distortions or deformed into another crystal system or incorrectly attributed to another could be difficult to find. Clearly, a metric for relating unit cells was required.  See \cite{Andrews2023} for additional
background information on the problem.

\subsection{Reduced Cells}

Niggli \cite{Niggli1928} and Delaunay \cite{Delaunay1933} (aka Delone) devised ``reduced cells'', which allowed for a more standard presentation of some of the crystal data.
The Buerger-reduced cell is simpler than the Niggli-reduced cell, having fewer
constraints \cite{Buerger1957} \cite{Azaroff1958} \cite{Buerger1960}.  All Niggli-reduced cells are Buerger-reduced, but not all Buerger-reduced cells are Niggli-reduced.  Roof \cite{Roof1967} provided an updated and authoritative reference to the lattice characters of Niggli-reduced cells.

Finally, in the 1970s, the National Institute of Health (NIH) and the Environmental Protection Agency (EPA) joined to create the online searchable Chemical Information System (CIS) \cite{Heller1976} \cite{Bernstein1979}.
Along with physical measurements such as nuclear magnetic resonance and infrared, NIH/EPA wanted to include unit cell searching. At the time, there was no commonly accepted method to compute the ''distance" between two unit cells (equivalently, lattices).

There were two problems.

The first problem was that measured unit cell parameters (conventionally, [a, b, c, alpha, beta, gamma]  for the cell lengths and angles) always have experimental error in their determinations.  Further, closely related compounds of interest might have slightly different cell-edge parameters.  That means that the problem to be solved is ''the nearest neighbor problem” also known as ''the post office problem". An exact match is not sufficient.

The second issue is related to the problem of experimental error, but it manifests in a different way. It is well-known that for any given lattice, there is an infinity of unit cells that can be chosen.  The problem is that two unit cells from the same lattice  may not look the same.  The face-centered cells: [10, 10, 10, 90, 90, 90]  and [10, 10, 10, 120, 120, 90] are two different choices from a single crystal lattice.  Only the first of these is Niggli reduced, but lattice variation may make the second the reduced one.

\section{The Unsorted $\mathbf{DC}^7$ Cell, dc7unsrt}

We define the Wigner-Seitz cell as consisting of the points which
are no farther from a given lattice point than they are from any
other lattice point.   As \citeasnoun{Hart2019} has shown, the
Wigner-Seitz cell centered on a given lattice point is contained 
entirely within the convex envelope of the immediate 
neighbors of a given lattice point, {\it i.e.} in terms of twenty-six
Miller indices:
~~\\
\vspace{-5mm}
\[
(1,0,0),\,(0,1,0),\,(0,0,1),\,(-1,0,0),\,(0,-1,0),\,(0,0,-1),
\]
~~\\

\vspace{-30mm}
\[(0,1,1),\,(1,0,1),\,(1,1,0),\,(0,1,-1),\,(-1,0,1),\,(1,-1,0),
\]
~~\\

\vspace{-30mm}
\[
(0,-1,\,-1),\,(-1,0,-1),\,(-1,-1,0),\,(0,-1,1),\,(1,0,-1),\,(-1,1,0),
\]
~~\\

\vspace{-30mm}
\[
(1,1,1),(1,-1,-1),\,(-1,1,-1),\,(-1,-1,1),
\]
~~\\

\vspace{-30mm}
\[
(-1,-1,-1),(-1,1,1),(1,-1,1,),(1,1,-1)
\]

We organize the lattice in terms of a basis of the three shortest distances.
The Wigner-Seitz cell is symmetric around the given lattice point, so the thirteen Miller indices are sufficient:
~~\\

\vspace{-20mm}
\[
(1,0,0),\,(0,1,0),\,(0,0,1),
\]
~~\\

\vspace{-30mm}
\[
(0,1,1),\,(1,0,1),\,(1,1,0),\,(0,1,-1),\,(-1,0,1),\,(1,-1,0),
\]
~~\\

\vspace{-30mm}
\[
(1,1,1),\,(1,-1,-1),\,(-1,1,-1),\,(-1,-1,1),
\]

Formally, the definition of the Wigner-Seitz cell is:

Let $\mathbf{R}$ be the space of reals, $\mathbf{Z}$ be the
space of integers.  Let $L$ be an $\mathbf{R}^3$ lattice with Minkowski basis $\mathbf{a},\, \mathbf{b},\, \mathbf{c}\, \in \mathbf{R}^3$, i.e. such that $h\mathbf{a}+k\mathbf{b}+l\mathbf{c}, h \in \mathbf{Z}, k \in \mathbf{Z}, l \in \mathbf{Z}$ spans $L$ and $||\mathbf{a}||, ||\mathbf{b}||, ||\mathbf{c}||$ are minimal.  We define the Wigner-Seitz cell of $L$ as
\[
WS(L) = \{ w \in \mathbf{R}^3 \backepsilon \forall x \in L, x \neq 0, ||w|| \leq ||x-w|| \}
\]

If we translate this cell to each element of $L$, we tile the space and have a
Voronoi decomposition.

Niggli-reduction provides an unambiguous Minkowski reduction. Assume the cell formed by $\mathbf{a}, \mathbf{b}, \mathbf{c}$ is Niggli reduced, with  $||\mathbf{a}||\, \leq\, ||\mathbf{b}|| \leq ||\mathbf{c}||$.  Define the $\mathbf{G}^6$ vector of
the cell as
\[
\{r,\,s,\,t,\,u,\,v,\,w\}
\]
\[=\{ \mathbf{a} \cdot \mathbf{a},\, \mathbf{b} \cdot \mathbf{b},\, \mathbf{c} \cdot \mathbf{c},\,  2 \mathbf{b} \cdot \mathbf{c},\, 2 \mathbf{a} \cdot \mathbf{c},\, 2 \mathbf{a} \cdot \mathbf{b}\}
\]
As a Niggli-reduced cell, we either have all of $u,v,w \leq 0$, which we
annotate as the $---$ case, or all of $u,v,w > 0$, which we annotate
as the $+++$ case, and in both cases $|u| \leq s,\, |v| \leq r,\, |w| \leq r$.

The full set of conditions for a Niggli-reduced cell are
given in conditions \ref{eq:niggli_start} -- \ref{eq:niggli_end}:
\begin{equation}
\text{require } 0 < r \leq s \leq t \label{eq:niggli_start}
\end{equation}
\begin{equation}
\text{if } r = s; \text{ then require } |u| \leq |v|
\end{equation}
\begin{equation}
\text{if } s = t; \text{ then require } |v| \leq |w| 
\end{equation}
\begin{equation}
\text{require } u > 0 \text{ and } v > 0 \text{ and } w > 0 \text{ or}
\end{equation}
\begin{equation}
\text{~~~~require } u \leq 0 \text{ and } v \leq 0 \text{ and } w \leq 0 
\end{equation}
\begin{equation}
\text{require } |u| \leq s
\end{equation}
\begin{equation}
\text{require } |v| \leq r
\end{equation}
\begin{equation}
\text{require } |w| \leq r
\end{equation}
\begin{equation}
\text{require } t \leq r + s + t + u + v + w 
\end{equation}
\begin{equation}
\text{if } u = s \text{ then require }  w \leq 2\,v
\end{equation}
\begin{equation}
\text{if } v = r \text{ then require }  w \leq 2\,u
\end{equation}
\begin{equation}
\text{if } w = r \text{ then require }  v \leq 2\,u
\end{equation}
\begin{equation}
\text{if } u = -s \text{ then require }  w = 0
\end{equation}
\begin{equation}
\text{if } v = -r \text{ then require }  w = 0
\end{equation}
\begin{equation}
\text{if } w = -r \text{ then require }  v = 0
\end{equation}
\begin{equation}
\text{if } t = r + s + t + u + v + w 
\text{ then require } 2\,r + 2\, v + w \leq 0 \label{eq:niggli_end}
\end{equation}

Using the $\mathbf{G}^6$ vector component definitions, we define a  $\mathbf{DC}^{13}$ cell as the squares of the three edge lengths, the six face diagonal lengths and the
four body diagonal lengths, {\it i.e.}:
\[
dc_{13,raw}(r,s,t,u,v,w)=
\]
\[
[r,\,s,\,t,
\]
\[
s+t-u,\, s+t+u,\, r+t-v,\, r+t+v,\, r+s-w,\, r+s+w, 
\]
\[
r+s+t+u+v+w,\, r+s+t+u-v-w,
\]
\[
r+s+t-u+v-w,\, r+s+t-u-v+w]
\]

If we sort the elements of $\mathbf{DC}^{13}$ and only present the first
seven elements, we have $\mathbf{DC}^{7}$ as discussed in \cite{bernstein2021},
which is a smooth, but ambiguous, characterization of lattices.  $\mathbf{DC}^{7}$ 
is not invertible in some cases unless the symmetry is known 
{\it a priori}, or some elements after the seventh are retained.  Bright
\cite{bright2021a} has demonstrated the $\mathbf{DC}^{7}$ ambiguity with
the cells i: $ [ 2.8284,$ $ 3.162277$, $ 3.4641 $, $ 117.157$, $ 107.8295$, $116.5651 ] $
and ii: $ [ 2.8284$, $ 3.162277$, $ 3.4641$, $ 123.211$, $ 107.8295$, $ 109.59748 ] $
as $ [a,b,c,\alpha,\beta,\gamma] $, or
i: $ [8,\,10,\,12,\,-10,\,-6,\,-4]$ and ii: $ [8,\,10,\,12,\,-12,\,-6,\,-3 ] $
as the $\mathbf{G}^6$ vectors $[r,\,s,\,t,\,u,\,v,\,w]$
for which the sorted $\mathbf{DC}^{13}$ elements after Niggli reduction are
~~\\

\vspace{-20mm}
\[ \text{i:}\, 2.44949,\, 2.82843,\, 3.16228,\]
~~\\

\vspace{-30mm}
\[ 3.16228,\, 3.4641,\, 3.4641,\, 3.74166,\,\mathbf{4},\]
~~\\

\vspace{-30mm}
\[4.24264,\, 4.47214,\, 5.09902,\,  5.09902,\, 6.16441, \]
~~\\

\vspace{-20mm}
\noindent{}and 
~~\\

\vspace{-20mm}
\[ \text{ii:}\, 2.44949,\, 2.82843,\, 3.16228,\]
~~\\

\vspace{-30mm}
\[ 3.16228,\, 3.4641,\, 3.4641,\, 3.74166,\, \mathbf{4.24264},\]
~~\\

\vspace{-30mm}
\[4.24264,\, 4.89898,\, 4.89898,\, 5.2915,\, 5.65685, \]
~~\\

\vspace{-20mm}
\noindent{}respectively, which do not differ until the eighth element.

If all of $u,\,v,\,w \leq 0$, then $s+t+u \leq s+t-u$, 
$r+t+v \leq r+t-v$, $r+s+w \leq r+s-w$, {\it i.e.} the three summed
squares of the face diagonals are no larger than the corresponding differences,
and $r+s+t+u+v+w \leq $ \{ $r+s+t+u-v-w$, $r+s+t-u+v-w$, $r+s+t-u-v+w$ \}.
In this $---$ case, the main body diagonal is no larger than the three remaining body diagonals.

On the other hand, if all of $u,v,w > 0$, then $s+t+u > s+t-u$, 
$r+t+v > r+t-v$, $r+s+w > r+s-w$; {\it i.e.} the squares of the
lengths of longer face diagonals formed by the Niggli-reduced
cell-edge vector sums are strictly greater than the corresponding
shorter face diagonals formed by the Niggli-reduced
cell-edge vector differences.  Turning to the body diagonals,
the square of the length of the main body diagonal is
$r+s+t+u+v+w$.  The squares of the lengths of the other three
body diagonals are $ \{r+s+t+u-v-w$, $r+s+t-u+v-w$, $r+s+t-u-v+w\}$.
Thus the square of the length of smallest squared body diagonal is strictly less than $r+s+t+u+v+w$, and strictly greater than $r+s+t-|u|-|v|-|w|$.

Thus we can recover $r,s,t,u,v,w$ from the three cell-edge lengths,
the three shorter of each pair of face diagonal lengths, and the shortest squared body diagonal length
as explained in the following theorem and proof.

\subsection{Theorem}
\label{subsec::invert}

Let $g = {r,s,t,u,v,w}$ be a $\mathbf{G}^6$ Niggli-reduced
cell satisfying the Niggli-reduction conditions 
\ref{eq:niggli_start} -- \ref{eq:niggli_end}.  Let
\[ dc7_{unsrt}=[dc7_{unsrt,1},dc7_{unsrt,2},dc7_{unsrt,3},\]
\[ dc7_{unsrt,4},dc7_{unsrt,5},dc7_{unsrt,6},dc7_{unsrt,7} ] \]
\[ =[r,s,t,s+t-|u|,r+t-|v|,r+s-|w|,\] 
\[ min(r+s+t+u+v+w, r+s+t+u-v-w,\]
\[ r+s+t-u+v-w, r+s+t-u-v+w)]
\]
\[ =[r,s,t,s+t-|u|,r+t-|v|,r+s-|w|,\] 
\[ r+s+t-|u|-|v|-|w|+2\,min(max(0,u),max(0,v),max(0,w))]
\]
be the unsorted $\mathbf{DC}^{7}$ presentation of $g$.
Then the values of the components of $dc7_{unsrt}$
unambiguously determine the values of the components
of $g$.

\subsubsection{Proof of Theorem \ref{subsec::invert}}
~~\\

In the Niggli-reduced $---$ case we have
\begin{equation}
  dc7_{unsrt} = [r,s,t,s+t+u,r+t+v,r+s+w,r+s+t+u+v+w]
  \label{mmm}
\end{equation}
and in the Niggli-reduced $+++$ case we have
\begin{equation}
dc7_{unsrt} = [r,s,t,s+t-u,r+t-v,r+s-w,r+s+t-u-v-w+2\,min(u,v,w)]
  \label{ppp}
\end{equation}

If we subtract the face diagonal from the matching pairs of edges, we get the
absolute values of $u,v,w$
\[
dc7_{unsrt,2}+dc7_{unsrt,3}-dc7_{unsrt,4} 
\]
\[
= s+t-(s+t-|u|)=|u|
\]
\[
dc7_{unsrt,1}+dc7_{unsrt,3}-dc7_{unsrt,5}
\]
\[
= r+t-(r+t-|v|)=|v|
\]
\[
dc7_{unsrt,1}+dc7_{unsrt,2}-dc7_{unsrt,6}
\]
\[
= r+s-(r+s-|w|)=|w|
\]

\noindent{}from which we can compute an estimate, $\tau$, of the shortest
body diagonal that is exact for $---$ and a strict underestimate
for $+++$:
\[\tau\,= \,r+s+t-|u|-|v|-|w|\]
\[=\,dc7_{unsrt,1}\,+\,dc7_{unsrt,2}\,+\,dc7_{unsrt,3} \]
\[ -(dc7_{unsrt,2},+\,dc7_{unsrt,3},-\,dc7_{unsrt,4})\]
\[ -(dc7_{unsrt,1}\,+\,dc7_{unsrt,3}\,-\,dc7_{unsrt,5})\]
\[ -(dc7_{unsrt,1}\,+,dc7_{unsrt,2}\,-\,dc7_{unsrt,6})\]
\[=\,-\,dc7_{unsrt,1}\,-\,dc7_{unsrt,2}\,-\,dc7_{unsrt,3}\]
\[\,+\,dc7_{unsrt,4}\,+\,dc7_{unsrt,5}\,+ \,dc7_{unsrt,6}\]
  If $\tau=dc7_{unsrt,7}$,
we can be certain that all of $u,\,v,\,w\, \leq 0$.  If  $\tau \neq dc7_{unsrt,7}$ and the difference is larger than the possible experimental
or rounding errors, we can be certain that all of $u,v,w > 0$.  Thus the
Niggli cell can be recovered from $dc7_{unsrt}$.
Q.E.D.

\subsection{Recovering the components of the Bright example}

For example, the Niggli-reduced $\mathbf{G}^6$ versions of the cells in the
Bright example above are i: $ [ 6,\, 8,\, 10,\, 8,\, 4,\, 2] $ and ii: $ [ 6,\, 8,\, 10,\, -6,\, -2,\, -4 ]$.
Note that the former has all of $u,\, v,\, w$ positive and the latter negative.
Table \ref{table::recover} shows the unsorted $\mathbf{DC}^7$ vectors
and the process of recovery of the $\mathbf{G}^6$ vectors.

\subsection{Examples using Phospholipase A2}

As noted in \cite{McGill2013}, the structures in the PDB for Phospholipase
A2 include slightly different experimental determinations of presented
as different lattices, {\it e.g.} 1U4J in space group H 3 as 
$ [ 80.36,\,	80.36,\,	99.44,\,	90,\,	90,\,	120 ] $,
 1G2X in space group C 1 2 1 as

$ [ 80.949,\,	80.572,\,	57.098,\,	90,\,	90.35,\,	90] $ , and 1FE5 in space group R 3 2 as
$ [ 57.98,\,	57.98,\,	57.98,\,	92.02,\,	92.02,\,	92.02 ] $.
The primitive Niggli-reduced $\mathbf{G}^6$ $[r,s,t,u,v,w ]$ vectors are:
$$ \text{1U4J:  } [ 3251.278,\,	3251.278,\,	3251.27,\,	44.826,\,	44.826,\,	44.826  ] $$
$$ \text{1G2X:  } [ 3260.182,\,	3261.147,\,	3261.147,\,	30.447,\,	28.234,\,	28.234  ] $$
$$ \text{1FE5:  } [ 3361.68,\,	3361.68,\,	3361.68,\,	-236.987,\,	-236.987,\,	-236.987 ] $$
from which we compute the three $\mathbf{DC}^7$ vectors as
$$ [ r,s,t,s+t-|u|, r+t-|v|, r+s-|w|, r+s+t-|u|-|v|-|w|+2 min(|u|,|v|,|w|)] $$
in the first two $+++$ cases and as
$$ [ r,s,t,s+t-|u|, r+t-|v|, r+s-|w|, r+s+t-|u|-|v|-|w|] $$
in the final $---$ case:
$$ \text{1U4J:  } [ 3251.278,\,	3251.278,\,	3251.278,\,	6457.73,\,	6457.73,\,	6457.73,\,	9709.008 ] $$
$$ \text{1G2X:  } [ 3260.182,\,	3261.147,\,	3261.147,\,	6491.847,\,	6493.095,\,	6493.095,\,	9752.029 ] $$
$$ \text{1FE5:  } [ 3361.68,\,	3361.68,\,	3361.68,\,	6486.373,\,	6486.373,\,	6486.373,\,	9374.079 ] $$

In each case the original primitive Niggli-reduced $\mathbf{G}^6$ $[r,s,t,u,v,w ]$ vectors can be recovered by simply copying the first three elements of
the unsorted $\mathbf{DC}^7$ vector, $r,s,t$ then computing the three absolute values of $u,v,w$ as the differences $s+t$ minus the fourth element of the
unsorted $\mathbf{DC}^7$ vector, $r+t$ minus the fifth element of the
unsorted $\mathbf{DC}^7$ vector, and $r+s$ minus the sixth element.  
What remains is to compute $\tau = r+s+t-|u|-|v|-|w|$ and compare it to the
seventh element.  If they are the same to within rounding error, as they
are, indeed, for the 1FE5 case, then this is a $---$ case and the absolute
values of $u,v,w$ we have computed are the negatives of the actual values.
Otherwise this is a $+++$ case.  Completion of this example is left as an
exercise to the reader, but to help in checking your work, the three
values of $\tau$ are $9619.356$, $9695.561$, and $9374.079$.

\subsection{Example of the effect of Small Perturbations an F-centered Cell}

A well-know example occurs for face-centered cubic lattices. Choosing an initial face-centered unit cell of ($10 \sqrt(2)$, $10 \sqrt(2)$, $10 \sqrt(2)$, $90$, $90$, $90$), the primitive Niggli-reduced cell is {$10$, $10$, $10$, $60$, $60$, $60$). However, small changes in the original F-centered cell can change the 
primitive Niggli-reduced cell to ( $10....$, $10....$, $10....$, $90$, $120$, $120$ )
The cells  in Table \ref{table::fcent} are all derived from F-centered 
$ ( 14.14,\, 14.14,\, 14.14,\, 90,\, 120,\ 120 ) ] $ perturbing 
by 0.01\% normal to the vector.  The resulting unsorted $\mathbf{DC}^7$ vectors are given in Table \ref{table::fcentdc7}.

While these look very different, when the differences are smoothed
by permutations and gluing at boundaries, the distances among these cells are all small.

\section{The Boundaries of Unsorted $\mathbf{DC}^7$}

Whether we are working in seven dimensions with $\mathbf{DC}^7$ 
or in six dimensions with $\mathbf{G}^6$, the reduced cells
form a manifold for which it is useful to understand the
boundaries.   Inasmuch as seven-dimensional unsorted $\mathbf{DC}^7$ 
cells are invertibly derived from six-dimensional $\mathbf{G}^6$
cells, the six-dimensional boundary polytopes
of the manifold of valid Wigner-Seitz-reduced cells in $\mathbf{DC}^7$ can be derived directly
from the fifteen five-dimensional boundary polytopes of the manifold of valid Niggli-reduced
cells in $\mathbf{G}^6$ as described in \cite{Andrews2014} and then
applying equations (\ref{mmm}) and (\ref{ppp}) to the $\mathbf{G}^6$
descriptions of the boundaries.   The manifold
 of Wigner-Seitz-reduced cells in unsorted $\mathbf{DC}^7$
 is divided similarly to the way in which the manifold of Niggli-reduced cells is
 divided on the basis of whether $\tau < dc7_{usrt,7}$ for $+++$  or $\tau = dc7_{usrt,7}$
 for $---$.  The fifteen boundary polytopes are sufficient to then
 describe the primitive lattice characters, but, as with $\mathbf{G}^6$,
 seven additional special-position subspaces are needed to fully
 describe some of the centered cases.  See \cite{Andrews2014} for a
 discussion of the special-position subspaces.

 For consistency, we break from past tradition and use $u,v,w$ for
 both the all-positive and all-negative cases rather than using $-u,-v,-w$ for the
 $---$ case.

 Recall that in linear algebra an orthogonal projector $P$ into a subspace $S_{sub}$ of
 a space $S$ is a symmetric matrix that maps any element $s \in S$ into an element
 $s_{sub} \in S{sub}$ and satisfies $P^2=P$.  Orthogonal projectors are
 commonly just called "projectors".  Note than ${I-P}s)$ is perpendicular to
 $Ps$.  Projectors are often computed by {\it ad hoc} singular value decomposition (SVD).  In
 this case they were computed by the column-space operations of the
 symbolic algebra package maxima, the open source version of DOE macsyma \cite{bogen1974macsyma}

 \subsection{Equal-cell-edge case}

 Recall that $r \leq s \leq t$ for Niggli-reduced cells.  The first two boundaries are the equal-edge boundary cases.  For both of these cases, the boundaries subdivide
 into one $---$ subcase and three $+++$ subcases, one for each of $u$, $v$, or $w$
 being minimal.  Then at least one of $r+s+t+u-v-w$, $r+s+t-u+v-w$, or $r+s+t-u-v+w$ respectively
 is the minimal body diagonal.

 \begin{itemize}
     \item{Case 1. $r=s$:  The cells in this case may be either $+++$ or $---$.\\
     $dc7_{unsrt,1}=dc7_{unsrt,2}$.
     \[dc7_{unsrt}  \text{ subspace 1}=[r,r,t,r+t-|u|,r+t-|v|,2r-|w|,\]
     \[2r+t-|u|-|v|-|w|+min(max(0,u),max(0,v),max(0,w))]\]
         \[dc7_{unsrt} \text{ boundary 1 }---\text{ projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{5}{14}&\frac{5}{14}&\frac{-1}{7}&\frac{1}{7}
&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{5}{14}&\frac{5}{14}&\frac{-1}{7}&\frac{1}{7}
&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{-1}{7}&\frac{-1}{7}&\frac{6}{7}&\frac{1}{7}
&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{-1}{7}
&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{6}{7}
&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}
&\frac{6}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}
&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}
\end{pmatrix}
\end{equation*}
\end{center}

\[dc7_{unsrt} \text{ boundary 1 } +++ \text{ minimal }u \text{ projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{1}{2}&\frac{1}{2}&0&0&0&0&0\\[.25em]
\frac{1}{2}&\frac{1}{2}&0&0&0&0&0\\[.25em]
0&0&\frac{4}{5}&\frac{1}{5}&\frac{-1}{5}&\frac{-1}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{1}{5}&\frac{4}{5}&\frac{1}{5}&\frac{1}{5}&\frac{-1}{5}\\[.25em]
0&0&\frac{-1}{5}&\frac{1}{5}&\frac{4}{5}&\frac{-1}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{-1}{5}&\frac{1}{5}&\frac{-1}{5}&\frac{4}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{1}{5}&\frac{-1}{5}&\frac{1}{5}&\frac{1}{5}&\frac{4}{5}
\end{pmatrix}
\end{equation*}
\end{center}

\[dc7_{unsrt} \text{ boundary 1 } +++ \text{ minimal }v \text{ projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{1}{2}&\frac{1}{2}&0&0&0&0&0\\[.25em]
\frac{1}{2}&\frac{1}{2}&0&0&0&0&0\\[.25em]
0&0&\frac{4}{5}&\frac{-1}{5}&\frac{1}{5}&\frac{-1}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{-1}{5}&\frac{4}{5}&\frac{1}{5}&\frac{-1}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{1}{5}&\frac{1}{5}&\frac{4}{5}&\frac{1}{5}&\frac{-1}{5}\\[.25em]
0&0&\frac{-1}{5}&\frac{-1}{5}&\frac{1}{5}&\frac{4}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{1}{5}&\frac{1}{5}&\frac{-1}{5}&\frac{1}{5}&\frac{4}{5}
\end{pmatrix}
\end{equation*}
\end{center}

\[dc7_{unsrt} \text{ boundary 1 } +++ \text{ minimal }w \text{ projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{5}{14}&\frac{5}{14}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}\\[.25em]
\frac{5}{14}&\frac{5}{14}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}\\[.25em]
\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{6}{7}&\frac{1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{-1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{6}{7}
\end{pmatrix}
\end{equation*}
\end{center}

\[dc7_{unsrt} \text{ boundary 1 transforms}=\]
  \[ ---: [s,r,t,v+t+r,u+t+s,w+s+r,w+v+u+t+s+r] \]
  \[ +++: [s,r,t,-v+t+r,-u+t+s,-w+s+r,-w-v+2 min(u,v,w)-u+t+s+r]\]

}

     \item{Case 2. $s=t$:  The cells in this case may be either $+++$ or $---$.\\
     $dc7_{unsrt,2}=dc7_{unsrt,3}$.
     \[dc7_{unsrt}  \text{ subspace 2}=[r,s,s,2s-|u|,r+s-|v|,r+s-|w|,\]
     \[r+2s-|u|-|v|-|w|+min(max(0,u),max(0,v),max(0,w))]\]
     \[dc7_{unsrt} \text{ boundary 2 }---\text{ projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{6}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{-1}{7}&\frac{5}{14}&\frac{5}{14}&\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{-1}{7}&\frac{5}{14}&\frac{5}{14}&\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{6}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{6}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}
\end{pmatrix}
\end{equation*}
\end{center}

\[dc7_{unsrt} \text{ boundary 2 } +++ \text{ minimal }u \text{ projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{6}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{1}{7}&\frac{5}{14}&\frac{5}{14}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{5}{14}&\frac{5}{14}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{6}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}
\end{pmatrix}
\end{equation*}
\end{center}

\[dc7_{unsrt} \text{ boundary 2 } +++ \text{ minimal }v \text{ projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{4}{5}&0&0&\frac{-1}{5}&\frac{1}{5}&\frac{-1}{5}&\frac{1}{5}\\[.25em]
0&\frac{1}{2}&\frac{1}{2}&0&0&0&0\\[.25em]
0&\frac{1}{2}&\frac{1}{2}&0&0&0&0\\[.25em]
\frac{-1}{5}&0&0&\frac{4}{5}&\frac{1}{5}&\frac{-1}{5}&\frac{1}{5}\\[.25em]
\frac{1}{5}&0&0&\frac{1}{5}&\frac{4}{5}&\frac{1}{5}&\frac{-1}{5}\\[.25em]
\frac{-1}{5}&0&0&\frac{-1}{5}&\frac{1}{5}&\frac{4}{5}&\frac{1}{5}\\[.25em]
\frac{1}{5}&0&0&\frac{1}{5}&\frac{-1}{5}&\frac{1}{5}&\frac{4}{5}]
\end{pmatrix}
\end{equation*}
\end{center}

\[dc7_{unsrt} \text{ boundary 2 } +++ \text{ minimal }w \text{ projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{4}{5}&0&0&\frac{-1}{5}&\frac{-1}{5}&\frac{1}{5}&\frac{1}{5}\\[.25em]
0&\frac{1}{2}&\frac{1}{2}&0&0&0&0\\[.25em]
0&\frac{1}{2}&\frac{1}{2}&0&0&0&0\\[.25em]
\frac{-1}{5}&0&0&\frac{4}{5}&\frac{-1}{5}&\frac{1}{5}&\frac{1}{5}\\[.25em]
\frac{-1}{5}&0&0&\frac{-1}{5}&\frac{4}{5}&\frac{1}{5}&\frac{1}{5}\\[.25em]
\frac{1}{5}&0&0&\frac{1}{5}&\frac{1}{5}&\frac{4}{5}&\frac{-1}{5}\\[.25em]
\frac{1}{5}&0&0&\frac{1}{5}&\frac{1}{5}&\frac{-1}{5}&\frac{4}{5}
\end{pmatrix}
\end{equation*}
\end{center}

 \[dc7_{unsrt} \text{ boundary 2 transforms}=\]
 \[ ---: [r,t,s,u+t+s,w+s+r,v+t+r,w+v+u+t+s+r] \]
 \[ +++: [r,t,s,-u+t+s,-w+s+r,-v+t+r,-w-v+2 min(u,v,w)-u+t+s+r]\]
}
 \end{itemize}

 The special-position subspaces $\hat{1}$ and $\hat{2}$ are obtained by adding the constraints $1^\prime\!: \{u = v\}$ and
 $2^\prime\!:\{v=w\}$, respectively.

 \subsection{$90^\circ$ case}

 The $90^\circ$ case marks a possible transition between $---$ and $+++$.
 All the cells with a $90^\circ$ angle are in $---$.

 \begin{itemize}
     \item{Case 3. $u=0$:  The cells in this case must be $---$.\\
     $dc7_{unsrt,2}\,+\,dc7_{unsrt,3}\,-\,dc7_{unsrt,4}=0$.
     \[dc7_{unsrt}  \text{ subspace 3}=[r,s,t,s+t,r+t+v,r+s+w,r+s+t+v+w]\]
     \[dc7_{unsrt} \text{ boundary 3 projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{3}{4}&0&0&0&\frac{1}{4}&\frac{1}{4}&\frac{-1}{4}\\[.25em]
0&\frac{2}{3}&\frac{-1}{3}&\frac{1}{3}&0&0&0\\[.25em]
0&\frac{-1}{3}&\frac{2}{3}&\frac{1}{3}&0&0&0\\[.25em]
0&\frac{1}{3}&\frac{1}{3}&\frac{2}{3}&0&0&0\\[.25em]
\frac{1}{4}&0&0&0&\frac{3}{4}&\frac{-1}{4}&\frac{1}{4}\\[.25em]
\frac{1}{4}&0&0&0&\frac{-1}{4}&\frac{3}{4}&\frac{1}{4}\\[.25em]
\frac{-1}{4}&0&0&0&\frac{1}{4}&\frac{1}{4}&\frac{3}{4}
\end{pmatrix}
\end{equation*}
\end{center}

     \[dc7_{unsrt} \text{ boundary 3 transform}=\]
     \[ [r,s,t,-u+t+s,v+t+r,w+s+r,w+v+2 min(u,-v,-w)-u+t+s+r]\]
     }
     \item{Case 4. $v=0$:  The cells in this case must be $---$.\\
     $dc7_{unsrt,1}\,+\,dc7_{unsrt,3}\,-\,dc7_{unsrt,5}=0$.
     \[dc7_{unsrt} \text{ subspace 4}=[r,s,t,s+t+u,r+t,r+s+w, r+s+t+u+w]\]
     \[dc7_{unsrt} \text{ boundary 4 projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{2}{3}&0&\frac{-1}{3}&0&\frac{1}{3}&0&0\\[.25em]
0&\frac{3}{4}&0&\frac{1}{4}&0&\frac{1}{4}&\frac{-1}{4}\\[.25em]
\frac{-1}{3}&0&\frac{2}{3}&0&\frac{1}{3}&0&0\\[.25em]
0&\frac{1}{4}&0&\frac{3}{4}&0&\frac{-1}{4}&\frac{1}{4}\\[.25em]
\frac{1}{3}&0&\frac{1}{3}&0&\frac{2}{3}&0&0\\[.25em]
0&\frac{1}{4}&0&\frac{-1}{4}&0&\frac{3}{4}&\frac{1}{4}\\[.25em]
0&\frac{-1}{4}&0&\frac{1}{4}&0&\frac{1}{4}&\frac{3}{4}
\end{pmatrix}
\end{equation*}
\end{center}

      \[dc7_{unsrt} \text{ boundary 4 transform}=\]
      \[ [r,s,t,u+t+s,-v+t+r,w+s+r,w-v+u+2 min(-u,v,-w)+t+s+r]\]

     }
     \item{Case 5. $w=0$:  The cells in this case must be $---$.\\
     $dc7_{unsrt,1}\,+\,dc7_{unsrt,2}\,-\,dc7_{unsrt,6}=0$.
     \[dc7_{unsrt} \text{ subspace 5}=[r,s,t,s+t+u,r+t+v,r+s,r+s+t+u+v]\]
     \[dc7_{unsrt} \text{ boundary 5 projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{2}{3}&\frac{-1}{3}&0&0&0&\frac{1}{3}&0\\[.25em]
\frac{-1}{3}&\frac{2}{3}&0&0&0&\frac{1}{3}&0\\[.25em]
0&0&\frac{3}{4}&\frac{1}{4}&\frac{1}{4}&0&\frac{-1}{4}\\[.25em]
0&0&\frac{1}{4}&\frac{3}{4}&\frac{-1}{4}&0&\frac{1}{4}\\[.25em]
0&0&\frac{1}{4}&\frac{-1}{4}&\frac{3}{4}&0&\frac{1}{4}\\[.25em]
\frac{1}{3}&\frac{1}{3}&0&0&0&\frac{2}{3}&0\\[.25em]
0&0&\frac{-1}{4}&\frac{1}{4}&\frac{1}{4}&0&\frac{3}{4}
\end{pmatrix}
\end{equation*}
\end{center}

      \[dc7_{unsrt} \text{ boundary 5 transform}=\]
      \[[r,s,t,u+t+s,v+t+r,-w+s+r,-w+v+u+2 min(-u,-v,w)+t+s+r]\]
     }
 \end{itemize}

 In each ninety-degree case, the special-position subspace consists of $\hat{3}, \hat{4}, \hat{5}\!: \{u = v = w = 0\}$,
{\it i.e.} the primitive orthorhombic case, and we take
$3^\prime\!: \{v = w = 0\}$, $4^\prime\!: \{u = w = 0\}$, and $5^\prime\!: \{u = v = 0\}$.

\subsection{Face-diagonal  case}

 Recall that $|u| \leq s$, $|v| \leq r$, and $|w| \leq r$.  Equality marks
 the transition from edges being smaller than face diagonals to face diagonals
 possibly being smaller than the Niggli-reduced cell edges.

 \begin{itemize}
     \item{Case 6. $s=u, v \ge w$:  The cells in this case must be $+++$.\\
     $\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i}) < dc7_{unsrt,7}$\\
     $dc7_{unsrt,2}=dc7_{unsrt,2}\,+\,dc7_{unsrt,3}\,-\,dc7_{unsrt,4}$\\
          equivalent to
     $dc7_{unsrt,3}\,=\,dc7_{unsrt,4}$\\
     $dc7_{unsrt,3}\,-\,dc7_{unsrt,5} \ge
     dc7_{unsrt,2}\,-\,dc7_{unsrt,6}$.
     \[dc7_{unsrt} \text{ subspace 6}=[r,s,t,t,r+t-v,r+s-w, r+t-v+w],\, v \geq w >0 \]
     \[dc7_{unsrt} \text{ boundary 6 projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{4}{5}&\frac{-1}{5}&0&0&\frac{-1}{5}&\frac{1}{5}&\frac{1}{5}\\[.25em]
\frac{-1}{5}&\frac{4}{5}&0&0&\frac{-1}{5}&\frac{1}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{1}{2}&\frac{1}{2}&0&0&0\\[.25em]
0&0&\frac{1}{2}&\frac{1}{2}&0&0&0\\[.25em]
\frac{-1}{5}&\frac{-1}{5}&0&0&\frac{4}{5}&\frac{1}{5}&\frac{1}{5}\\[.25em]
\frac{1}{5}&\frac{1}{5}&0&0&\frac{1}{5}&\frac{4}{5}&\frac{-1}{5}\\[.25em]
\frac{1}{5}&\frac{1}{5}&0&0&\frac{1}{5}&\frac{-1}{5}&\frac{4}{5})
\end{pmatrix}
\end{equation*}
\end{center}

    \[dc7_{unsrt} \text{ boundary 6 transform}=\]
    \[ [r,s,-u+t+s,t,w-v-u+t+s+r,-w+s+r,-v+t+r] \]
     }

     \item{Case 7. $s=u, v < w$:  The cells in this case must be $+++$.\\
     $\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i}) < dc7_{unsrt,7}$\\
     $dc7_{unsrt,2}=dc7_{unsrt,2}\,+\,dc7_{unsrt,3}\,-\,dc7_{unsrt,4}$\\
          equivalent to
     $dc7_{unsrt,3}\,=\,dc7_{unsrt,4}$\\
     $dc7_{unsrt,3}\,-\,dc7_{unsrt,5} <
     dc7_{unsrt,2}\,-\,dc7_{unsrt,6}$.
     \[dc7_{unsrt} \text{ subspace 7}=[r,s,t,t,r+t-v,r+s-w, r+t+v-w],\, w > v > 0 \]
     \[dc7_{unsrt} \text{ boundary 7 projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{6}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{6}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{-1}{7}&\frac{1}{7}&\frac{5}{14}&\frac{5}{14}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{1}{7}&\frac{5}{14}&\frac{5}{14}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{-1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{6}{7}
\end{pmatrix}
\end{equation*}
\end{center}

 \[dc7_{unsrt} \text{ boundary 7 transform}=\]
 \[ [r,s,-u+t+s,t,-w+v-u+t+s+r,-w+s+r,-2*w+v+2*min(2*s-u,w,w-v)+t+r] \]
  }

     \item{Case 8. $s=-u$:  The cells in this case must be $---$.\\
    $\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i}) = dc7_{unsrt,7}$\\
     $dc7_{unsrt,2}=dc7_{unsrt,2}\,\,+dc7_{unsrt,3}\,-\,dc7_{unsrt,4}$\\
     equivalent to
     $dc7_{unsrt,3}\,=\,dc7_{unsrt,4}$.
     \[dc7_{unsrt} \text{ subspace 8}=[r,s,t,t,r+t+v,r+s+w,r+t+v+w],\, v, w \leq 0 \]
     \[dc7_{unsrt} \text{ boundary 8 projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{4}{5}&\frac{-1}{5}&0&0&\frac{1}{5}&\frac{1}{5}&\frac{-1}{5}\\[.25em]
\frac{-1}{5}&\frac{4}{5}&0&0&\frac{1}{5}&\frac{1}{5}&\frac{-1}{5}\\[.25em]
0&0&\frac{1}{2}&\frac{1}{2}&0&0&0\\[.25em]
0&0&\frac{1}{2}&\frac{1}{2}&0&0&0\\[.25em]
\frac{1}{5}&\frac{1}{5}&0&0&\frac{4}{5}&\frac{-1}{5}&\frac{1}{5}\\[.25em]
\frac{1}{5}&\frac{1}{5}&0&0&\frac{-1}{5}&\frac{4}{5}&\frac{1}{5}\\[.25em]
\frac{-1}{5}&\frac{-1}{5}&0&0&\frac{1}{5}&\frac{1}{5}&\frac{4}{5}
\end{pmatrix}
\end{equation*}
\end{center}

      \[dc7_{unsrt} \text{ boundary 8 transform}=\]
      \[ [r,s,u+t+s,t,w+v+u+t+s+r,w+s+r,2 w+v+2 min(u+2*s,-w-v,-w)+t+r]\]
     }

     \item{Case 9. $r=v, u \ge w$:  The cells in this case must be $+++$. \\
     $\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i}) < dc7_{unsrt,7}$\\
     $dc7_{unsrt,1}=dc7_{unsrt,1}\,+\,dc7_{unsrt,3}\,-\,dc7_{unsrt,5}$\\
          equivalent to
     $dc7_{unsrt,3}\,=\,dc7_{unsrt,5}$\\
     $dc7_{unsrt,3}\,-\,dc7_{unsrt,4} \ge
     dc7_{unsrt,1}\,-\,dc7_{unsrt,6}$.
     \[dc7_{unsrt} \text{ subspace 9}=[r,s,t,s+t-u,t,r+s-w,s+t-u+w],\, 0 < w \leq u \]
     \[dc7_{unsrt} \text{ boundary 9 projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{4}{5}&\frac{-1}{5}&0&\frac{-1}{5}&0&\frac{1}{5}&\frac{1}{5}\\[.25em]
\frac{-1}{5}&\frac{4}{5}&0&\frac{-1}{5}&0&\frac{1}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{1}{2}&0&\frac{1}{2}&0&0\\[.25em]
\frac{-1}{5}&\frac{-1}{5}&0&\frac{4}{5}&0&\frac{1}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{1}{2}&0&\frac{1}{2}&0&0\\[.25em]
\frac{1}{5}&\frac{1}{5}&0&\frac{1}{5}&0&\frac{4}{5}&\frac{-1}{5}\\[.25em]
\frac{1}{5}&\frac{1}{5}&0&\frac{1}{5}&0&\frac{-1}{5}&\frac{4}{5}
\end{pmatrix}
\end{equation*}
\end{center}

    \[dc7_{unsrt} \text{ boundary 9 transform}=\]
    \[ [r,s,-v+t+r,w-v-u+t+s+r,t,-w+s+r,-u+t+s]\]
     }

     \item{Case A. $r=v, u < w$:  The cells in this case must be $+++$.\\
     $\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i}) < dc7_{unsrt,7}$\\
     $dc7_{unsrt,1}=dc7_{unsrt,1}\,+\,dc7_{unsrt,3}\,-\,dc7_{unsrt,5}$\\
          equivalent to
     $dc7_{unsrt,3}\,=\,dc7_{unsrt,5}$\\
     $dc7_{unsrt,3}\,-\,dc7_{unsrt,4} <
     dc7_{unsrt,1}\,-\,dc7_{unsrt,6}$.
     \[dc7_{unsrt} \text{ subspace A}=[r,s,t,s+t-u,t,r+s-w,s+t+u-w],\, 0 < u < w \]
     \[dc7_{unsrt} \text{ boundary A projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{6}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{1}{7}&\frac{6}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{-1}{7}&\frac{5}{14}&\frac{1}{7}&\frac{5}{14}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{1}{7}&\frac{-1}{7}&\frac{5}{14}&\frac{1}{7}&\frac{5}{14}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{6}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}
\end{pmatrix}
\end{equation*}
\end{center}

\[dc7_{unsrt} \text{ boundary A transform}=\]
\[[r,s,-v+t+r,-w-v+u+t+s+r,t,-w+s+r,-2 w+2 min(2r-v,w,w-u)+u+t+s]\]
     }

     \item{Case B. $r=-v$:  The cells in this case must be $---$.\\
     $\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i}) = dc7_{unsrt,7}$\\
     $dc7_{unsrt,1}=dc7_{unsrt,1}\,+\,dc7_{unsrt,3}\,-\,dc7_{unsrt,5}$\\
          equivalent to
     $dc7_{unsrt,3}\,=\,dc7_{unsrt,5}$.
     \[dc7_{unsrt} \text{ subspace B}=[r,s,t,s+t+u,t,r+s+w,s+t+u+w],\, u, w \leq 0 \]
     \[dc7_{unsrt} \text{ boundary B projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{4}{5}&\frac{-1}{5}&0&\frac{1}{5}&0&\frac{1}{5}&\frac{-1}{5}\\[.25em]
\frac{-1}{5}&\frac{4}{5}&0&\frac{1}{5}&0&\frac{1}{5}&\frac{-1}{5}\\[.25em]
0&0&\frac{1}{2}&0&\frac{1}{2}0&0\\[.25em]
\frac{1}{5}&\frac{1}{5}&0&\frac{4}{5}&0&\frac{-1}{5}&\frac{1}{5}\\[.25em]
0&0&\frac{1}{2}&0&\frac{1}{2}&0&0\\[.25em]
\frac{1}{5}&\frac{1}{5}&0&\frac{-1}{5}&0&\frac{4}{5}&\frac{1}{5}\\[.25em]
\frac{-1}{5}&\frac{-1}{5}&0&\frac{1}{5}&0&\frac{1}{5}&\frac{4}{5}
\end{pmatrix}
\end{equation*}
\end{center}

    \[dc7_{unsrt} \text{ boundary B transform}=\]
    \[[r,s,v+t+r,w+v+u+t+s+r,t,w+s+r,2 w+2 min(v+2*r,-w-u,-w)+u+t+s]\]
     }

     \item{Case C. $r=w, u \ge v$:  The cells in this case must be $+++$.\\
     $\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i}) < dc7_{unsrt,7}$\\
     $dc7_{unsrt,1}=dc7_{unsrt,1}\,+\,dc7_{unsrt,2}\,-\,dc7_{unsrt,6}$\\
          equivalent to
     $dc7_{unsrt,2}\,=\,dc7_{unsrt,6}$\\
     $dc7_{unsrt,2}\,-\,dc7_{unsrt,4} \ge
     dc7_{unsrt,1}\,-\,dc7_{unsrt,5}$.
     \[dc7_{unsrt} \text{ subspace C}=[r,s,t,s+t-u,r+t-v,s,s+t-u+v],\, 0 <v \leq u \]
     \[dc7_{unsrt} \text{ boundary C projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{4}{5}&0&\frac{-1}{5}&\frac{-1}{5}&\frac{1}{5}&0&\frac{1}{5}\\[.25em]
0&\frac{1}{2}&0&0&0&\frac{1}{2}&0\\[.25em]
\frac{-1}{5}&0&\frac{4}{5}&\frac{-1}{5}&\frac{1}{5}&0&\frac{1}{5}\\[.25em]
\frac{-1}{5}&0&\frac{-1}{5}&\frac{4}{5}&\frac{1}{5}&0&\frac{1}{5}\\[.25em]
\frac{1}{5}&0&\frac{1}{5}&\frac{1}{5}&\frac{4}{5}&0&\frac{-1}{5}\\[.25em]
0&\frac{1}{2}&0&0&0&\frac{1}{2}&0\\[.25em]
\frac{1}{5}&0&\frac{1}{5}&\frac{1}{5}&\frac{-1}{5}&0&\frac{4}{5}
\end{pmatrix}
\end{equation*}
\end{center}

    \[dc7_{unsrt} \text{ boundary C transform}=\]
    \[[r,-w+s+r,t,-w+v-u+t+s+r,-v+t+r,s,-u+t+s]\]
     }
     \item{Case D. $r=w, u < v$:  The cells in this case must be $+++$.\\
     $\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i}) <
     dc7_{unsrt,7}$\\
     $dc7_{unsrt,1}=dc7_{unsrt,1}\,+\,dc7_{unsrt,2}\,-\,dc7_{unsrt,6}$\\
          equivalent to
     $dc7_{unsrt,2}\,=\,dc7_{unsrt,6}$\\
     $dc7_{unsrt,2}\,-\,dc7_{unsrt,4} <
     dc7_{unsrt,1}\,-\,dc7_{unsrt,5}$.
     \[dc7_{unsrt}\text{ subspace D} =[r,s,t,s+t-u,r+t-v,s,s+t+u-v],\, 0<u<v \]
     \[dc7_{unsrt} \text{ boundary D projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{6}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{1}{7}&\frac{5}{14}&\frac{-1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{5}{14}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{-1}{7}&\frac{6}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{5}{14}&\frac{-1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{5}{14}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}]
\end{pmatrix}
\end{equation*}
\end{center}

    \[dc7_{unsrt} \text{ boundary D transform}=\]
    \[[r,-w+s+r,t,-w-v+u+t+s+r,-v+t+r,s,2min(v,v-u,2r-w)-2v+u+t+s]\]

 }
     \item{Case E. $r=-w$:  The cells in this case must be $---$.\\
     $\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i}) = dc7_{unsrt,7}$\\
     $dc7_{unsrt,1}=dc7_{unsrt,1}\,+\,dc7_{unsrt,2}\,-\,dc7_{unsrt,6}$\\
          equivalent to
     $dc7_{unsrt,2}\,=\,dc7_{unsrt,6}$.
     \[dc7_{unsrt} \text{ subspace E } =[r,s,t,s+t+u,r+t+v,s,s+t+u+v], u, v \leq 0\]
     \[dc7_{unsrt} \text{ boundary E projector } =\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{4}{5}&0&\frac{-1}{5}&\frac{1}{5}&\frac{1}{5}&0&\frac{-1}{5}\\[.25em]
0&\frac{1}{2}&0&0&0&\frac{1}{2}&0\\[.25em]
\frac{-1}{5}&0&\frac{4}{5}&\frac{1}{5}&\frac{1}{5}&0&\frac{-1}{5}\\[.25em]
\frac{1}{5}&0&\frac{1}{5}&\frac{4}{5}&\frac{-1}{5}&0&\frac{1}{5}\\[.25em]
\frac{1}{5}&0&\frac{1}{5}&\frac{-1}{5}&\frac{4}{5}&0&\frac{1}{5}\\[.25em]
0&\frac{1}{2}&0&0&0&\frac{1}{2}&0\\[.25em]
\frac{-1}{5}&0&\frac{-1}{5}&\frac{1}{5}&\frac{1}{5}&0&\frac{4}{5}
\end{pmatrix}
\end{equation*}
\end{center}

    \[dc7_{unsrt} \text{ boundary E transform}=\]
    \[[r,w+s+r,t,w+v+u+t+s+r,v+t+r,s,2v+2min((-v)-u,-v,w+2*r)+u+t+s]\]

     }
 \end{itemize}

 The special-position subspaces of the face-diagonal boundary polytopes 6, 8, 9, B, C and E are empty because such a
special position would require a common point in the all acute $+ + +$ and all obtuse $- - -$ cases, but they only meet
at the axial planes of the $u,v,w$ subspace, which are excluded from the all acute $+ + +$ cases.  For cases
7, A and D there are non-trivial special-position subspaces. 
An invariant point in case 7 would have to satisfy $v = w - v$ or $v = w/2$.
Thus we define $7^\prime\!: \{v = w/2\}$ and similarly define $A^\prime\!: \{u = w/2\}$ and $D^\prime\!: \{u = v/2\}$.

\subsection{Body-diagonal case}

Recall that $t \leq r+s+t+u+v+w$ for a Niggli-reduced cell; otherwise
the main body diagonal would be shorter than $c$.  Equality can occur 
in $---$ and marks
the transition from edges being smaller than the main body diagonal to
the main body diagonal possibly being smaller.

\begin{itemize}
    \item{case F. $t = r+s+t+u+v+w$.  The cells in this case must be $---$.
    \[\tau=-\sum_{i=1}^3(-dc7_{unsrt,i})+\sum_{i=4}^6(dc7_{unsrt,i})\]
    \[=dc7_{unsrt,7}=dc7_{unsrt,3}\]
    \[dc7_{unsrt} \text{ subspace F}=[r,s,t,s+t+u,r+t+v,-u-v,t],\, u, v \leq 0 \]
    \[dc7_{unsrt} \text{ boundary F projector}=\]
\begin{center}
\begin{equation*}
\begin{pmatrix}
\frac{6}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{-1}{7}&\frac{6}{7}&\frac{-1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}\\[.25em]
\frac{-1}{7}&\frac{-1}{7}&\frac{5}{14}&\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{5}{14}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{6}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{6}{7}&\frac{-1}{7}&\frac{1}{7}\\[.25em]
\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{-1}{7}&\frac{-1}{7}&\frac{6}{7}&\frac{1}{7}\\[.25em]
\frac{-1}{7}&\frac{-1}{7}&\frac{5}{14}&\frac{1}{7}&\frac{1}{7}&\frac{1}{7}&\frac{5}{14}
\end{pmatrix}
\end{equation*}
\end{center}

    \[dc7_{unsrt} \text{ boundary F transform}=\]
    \[[r,s,w+v+u+t+s+r,v+t+r,u+t+s,w+s+r,t] \]
    }
\end{itemize}

In order to have a special-position subspace in case F, in addition to $r+ s + t  + u + v +  w = t$, we need $u = -2 s - u -w$ and $v = -2 r - v - w$. From this we have 
$2 s + 2 u = -w = 2 r + 2 v$. Then it follows that $F^\prime\!:\{r - s - u+v = 0\}$.  
This is equivalent to
\(\|\overrightarrow{a}+\overrightarrow{c}\| = \|\overrightarrow{b}+\overrightarrow{c}\|\), \textit{i.e.}
the shorter b-face-diagonal is the same length as the shorter  a-face-diagonal.

The unsorted $\mathbf{DC}^7$ subspace descriptions of the 
non-anorthic lattice characters derived from the $\mathbf{G}^6$
subspace descriptions given in Table 4 in \cite{Andrews2014} are
given in Tables \ref{NiggliFormsI} and \ref{NiggliFormsII}  by applying equations (\ref{mmm}) and (\ref{ppp}).  Except for three
of the monoclinics, 55A, 55B and 57C, the boundary polytopes from
$\mathbf{G}^6$ fully describe those for unsorted $\mathbf{DC}^7$.
In those three cases, the boundary is divided between 
$0 < u \leq r$ and $r < u \leq s$ with two different forms each for
$dc7_{unsrt,7}$, the minimum body diagonal.  These three 
monoclinic cases need to be explored further.

\section{Smoothing by permutations}

Because the same boundaries are available in unsorted $\mathbf{DC}^7$ as
in $\mathbf{G}^6$, the equivalent algorithmic techniques can be used
in improving the distance calculations to improve smoothness.  The obvious
first step is to deal with boundary cases 1 and 2 by simple permutation
of the $dc7_{unsrt}$ vectors, so that
\[
  dc7unsrt\_dist(dc7_{1},dc7_{2})=min(
\]
\[
  ||[dc7_{1,1},dc7_{1,2},dc7_{1,3},dc7_{1,4},dc7_{1,5},dc7_{1,6},dc7_{1,7}]
\]
\[
  -[dc7_{2,1},dc7_{2,2},dc7_{2,3},dc7_{2,4},dc7_{2,5},dc7_{2,6},dc7_{2,7}]||,
\]
\[
  ||[dc7_{1,1},dc7_{1,2},dc7_{1,2},dc7_{1,4},dc7_{1,5},dc7_{1,6},dc7_{1,7}]
\]
\[
  -[dc7_{2,1},dc7_{2,3},dc7_{2,2},dc7_{2,4},dc7_{2,6},dc7_{2,5},dc7_{2,7}]||,
\]
\[
  ||[dc7_{1,1},dc7_{1,2},dc7_{1,2},dc7_{1,4},dc7_{1,5},dc7_{1,6},dc7_{1,7}]
\]
\[
  -[dc7_{2,2},dc7_{2,1},dc7_{2,3},dc7_{2,5},dc7_{2,4},dc7_{2,6},dc7_{2,7}]||,
\]
\[
  ||[dc7_{1,1},dc7_{1,2},dc7_{1,2},dc7_{1,4},dc7_{1,5},dc7_{1,6},dc7_{1,7}]
\]
\[
  -[dc7_{2,2},dc7_{2,3},dc7_{2,1},dc7_{2,5},dc7_{2,6},dc7_{2,4},dc7_{2,7}]||,
\]
\[
  ||[dc7_{1,1},dc7_{1,2},dc7_{1,2},dc7_{1,4},dc7_{1,5},dc7_{1,6},dc7_{1,7}]
\]
\[
  -[dc7_{2,3},dc7_{2,1},dc7_{2,2},dc7_{2,6},dc7_{2,4},dc7_{2,5},dc7_{2,7}]||,
\]
\[
  ||[dc7_{1,1},dc7_{1,2},dc7_{1,2},dc7_{1,4},dc7_{1,5},dc7_{1,6},dc7_{1,7}]
\]
\[
  -[dc7_{2,3},dc7_{2,2},dc7_{2,1},dc7_{2,6},dc7_{2,5},dc7_{2,4},dc7_{2,7}]||.
\]

These cases are simple because cases 1 and 2 do not impact the seventh element.
In the general case, {\it e.g.} a boundary transform, rather than a
simple Niggli-reduced cell-edge permutation, a fresh Niggli reduction may be needed to regenerate the
seventh element for minimal distance calculations.

\section{Testing against the Gruber example}

\citeasnoun{Gruber1973} presented a Niggli-reduced cell with a five-fold
Buerger-reduced cell ambiguity.  The Niggli-reduced cell is
$[a, b, c, \alpha ,\beta ,\gamma ]$ 
= $ [2,\,4,\,4,\,60,\,79.19,\,75.52] $
which is equivalent to the $\mathbf{G}^6$ cell 
$ [r,s,t,u,v,w]$ = $ [4,\, 16,\, 16,\, 16,\, 3,\, 4 ] $
and the unsorted $\mathbf{DC}^7$ cell $[4,\, 16,\, 16,\, 17,\, 19,\, 16,\, 16 ]$.  The five examples of the alternative
Buerger reduced cells are shown in Table \ref{table::grubercells} as edges and angles, in Table \ref{table::gruberg6cells} as $\mathbf{G}^6$ $[r,s,t,u,v,w]$, and in
Table \ref{table::gruberdc7unsrt} as unsorted $\mathbf{DC}^7$.
Cell i is Niggli reduced.   All of the cells are on the 2 boundary
with $s=t$ and can equally be presented with $s$ and $t$ interchanged and $v$ and $w$ interchanged.  Cells i and ii
are both $+++$ and on the 7 and C face-diagonal boundaries as well as being on the 2 boundary.  Cells iii, iv and v are all $---$ and on the F body-diagonal boundary as well as being on the
2 boundary and one other face-diagonal boundary.
Cell iii is on the 8 face-diagonal boundary and cells iv and v
are on the E face-diagonal boundary.  Niggli reduction will
transform all of these back to cell i and Niggli reduction is
the first step in computing unsorted $\mathbf{DC}^7$. 
In order to compute the cells in Table \ref{table::gruberdc7unsrt} the components at the face-diagonal
boundaries were reduced in magnitude by 0.01 to prevent the
Niggli reduction.  The differences among the unsorted $\mathbf{DC}^7$ cells are consistent with the perturbation.

\section{Summary and Conclusions}

Starting from a Niggli-reduced cell, a crystallographic lattice may be characterized by seven parameters describing the Dirichlet cell: three edge lengths, the three shorter face diagonals and the shortest body diagonal,
from which the Niggli-reduced cell may be recovered.  This
unsorted $\mathbf{DC}^7$ lattice characterization avoids
the low-symmetry ambiguities of sorted $\mathbf{DC}^7$ and
is worth further investigation as a possible alternative to
$\mathbf{S}^6$ for crystallographic databases and clustering.

     % Appendices appear after the main body of the text. They are prefixed by
     % a single \appendix declaration, and are then structured just like the
     % body text.

\appendix

     %-------------------------------------------------------------------------
     % The back matter of the paper - acknowledgements and references
     %-------------------------------------------------------------------------

     % Acknowledgements come after the appendices

\ack{{\bf Acknowledgements}}
\\

Our thanks to Frances C. Bernstein for careful copy-editing and
helpful suggestions.   Our thanks to Elizabeth Kincaid for the timeline figure.

Work supported in part by
 US Dept.~of Energy, Office of Science, DOE Office of Biological and Environmental Research, grant KP1607011 and 
DOE Office of Basic Energy Sciences Program, contract DE-SC0012704,
US NIH National Institute of General Medical Sciences P30GM133893,
U.S. Department of Energy, Office of Science, Office of Workforce Development for Teachers and Scientists (WDTS) under the Science Undergraduate Laboratory Internships Program (SULI).

     % References are at the end of the document, between \begin{references}
     % and \end{references} tags. Each reference is in a \reference entry.

% \begin{references}
% \reference{Author, A. \& Author, B. (1984). \emph{Journal} \textbf{Vol}, 
% first page--last page.}
% \end{references}
%\cite{knuth84}

%% Note added by Overleaf: If using bibtex, remove the "references" environment above, and uncomment the following line.
\bibliographystyle{iucr}
\bibliography{DC7unsrt.bib}

     %---------------------------------------------f----------------------------
     % TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
     %-------------------------------------------------------------------------

     % Simple tables should use the tabular environment according to this
     % model

%\begin{table}
%\caption{Caption to table}
%\begin{tabular}{llcr}      % Alignment for each cell: l=left, c=center, r=right
% HEADING    & FOR        & EACH       & COLUMN     \\
%\hline
% entry      & entry      & entry      & entry      \\
% entry      & entry      & entry      & entry      \\
% entry      & entry      & entry      & entry      \\
%\end{tabular}
%\end{table}

     % Postscript figures can be included with multiple figure blocks
\onecolumn

\begin{table}
	\caption{Cells derived from F-centered $ [ 14.14,\, 14.14,\, 14.14,\, 90,\, 90,\ 90 ]$ perturbing 
		by 0.01\% normal to the vector and scaling $a$ to 10.  These are the primitive Niggli-reduced
		cells. They are sorted with alpha increasing.}
	\label{table::fcent}
	\begin{tabular}{l c c c c c c}
                P&  10.000&  10.001&  10.008&  90.018&  119.955& 119.980\\
                P&  10.000&  10.000&  10.003&  60.010&  60.038&  89.977\\
                P&  10.000&  10.003&  10.008&  60.017&  89.961&  60.031\\
                P&  10.000&  10.006&  10.008&  60.025&  60.046&  60.101\\
                P&  10.000&  10.002&  10.009&  60.039&  60.057&  60.082\\
                P&  10.000&  10.002&  10.006&  60.041&  60.028&  60.102\\
                P&  10.000&  10.001&  10.004&  60.047&  60.050&  60.016\\
                P&  10.000&  10.003&  10.006&  60.049&  60.036&  60.025\\
                P&  10.000&  10.000&  10.008&  60.049&  60.051&  60.030\\
                P&  10.000&  10.001&  10.002&  60.049&  60.059&  60.047\\
                P&  10.000&  10.005&  10.008&  60.052&  60.048&  60.076\\
                P&  10.000&  10.002&  10.004&  89.997&  60.032&  60.026\\
                P&  10.000&  10.001&  10.004&  90.019&  119.984&  119.941\\
                P&  10.000&  10.005&  10.006&  90.051&  119.942&  119.972\\
                P&  10.000&  10.006&  10.007&  90.074&  119.940&  119.951\\
                P&  10.000&  10.001&  10.003&  119.938&  119.984&  90.024\\
                P&  10.000&  10.005&  10.005&  119.968&  119.978&  90.011\\
                P&  10.000&  10.000&  10.006&  119.971&  90.032&  119.963\\
                P&  10.000&  10.009&  10.012&  119.981&  90.022&  119.949\\
                P&  10.000&  10.010&  10.011&  119.990&  90.007&  119.947\\
\end{tabular}
\end{table}

\begin{table}
	\caption{unsorted $\mathbf{DC}^7$ vectors from the cells in Table \ref{table:fcent}.}
	\label{table::fcentdc7}
	\begin{tabular}{c c c c c c c}
		$r$&$s$&$t$&$s+t-|u|$&$r+t-|v|$&$r+s-|w|$&$r+s+t$\\
		&&&&&&$-|u|-|v|-|w|$\\
		&&&&&&$+2*max(0,$\\
		&&&&&&$min(u,v,w))$\\
                100.000&  100.013&  100.160&  200.109&  100.217&  100.067&  100.221\\
                100.000&  100.006&  100.054&  100.060&  100.141&  199.927&  100.147\\
                100.000&  100.056&  100.154&  100.157&  200.017&  100.123&  100.224\\
                100.000&  100.119&  100.164&  100.216&  100.221&  100.367&  100.273\\
                100.000&  100.044&  100.181&  100.230&  100.263&  100.269&  100.312\\
                100.000&  100.046&  100.119&  100.206&  100.145&  100.333&  100.232\\
                100.000&  100.016&  100.078&  100.189&  100.190&  100.057&  100.230\\
                100.000&  100.054&  100.118&  100.234&  100.167&  100.104&  100.271\\
                100.000&  100.000&  100.152&  100.224&  100.232&  100.090&  100.304\\
                100.000&  100.020&  100.036&  100.178&  100.197&  100.152&  100.309\\
                100.000&  100.108&  100.168&  100.295&  100.228&  100.285&  100.355\\
                100.000&  100.031&  100.072&  200.092&  100.133&  100.094&  100.228\\
                100.000&  100.017&  100.088&  200.037&  100.091&  100.187&  100.211\\
                100.000&  100.107&  100.130&  200.058&  100.240&  100.137&  100.199\\
                100.000&  100.118&  100.143&  200.004&  100.252&  100.209&  100.203\\
                100.000&  100.014&  100.055&  100.221&  100.076&  199.930&  100.159\\
                100.000&  100.090&  100.108&  100.195&  100.120&  200.050&  100.167\\
                100.000&  100.000&  100.125&  100.150&  200.014&  100.112&  100.152\\
                100.000&  100.179&  100.242&  100.269&  200.167&  100.243&  100.258\\
                100.000&  100.202&  100.227&  100.246&  200.202&  100.263&  100.28\\
\end{tabular}
\end{table}

\begin{table}
\caption{Bright's $\mathbf{DC}^7$ ambiguous example, redone in unsorted $\mathbf{DC}^7$.
The  Niggli-reduced $\mathbf{G}^6$ vectors are $ [ 6,\, 8,\, 10,\, 8,\, 4,\, 2] $ and $ [ 6,\, 8,\, 10,\, -6,\, -2,\, -4 ] $.  The former is $+++$ and 
becomes $ [ 6,\, 8,\, 10,\, 10,\, 12,\, 12,\, 14 ] $ as unsorted $\mathbf{DC}^7$.
The latter is $---$ and becomes $ [ 6,\, 8,\, 10,\, 12,\, 14,\, 10,\, 12 ] $
as unsorted $\mathbf{DC}^7$.  When each is processed to recover $\mathbf{G}^6$
the magnitude of $\mathbf{r}+\mathbf{s}+\mathbf{t}-\mathbf{|u|}-\mathbf{|v|}-\mathbf{|w|}$ disagrees with the minimum body diagonal for the former and agrees for the latter,
giving the correct signs for full recovery of $\mathbf{G}^6$.}
\label{table::recover}
\begin{center}
\begin{tabular}{|l|ccc|ccc|}
\hline
$\mathbf{G}^6$:&
$\mathbf{r}$&$\mathbf{s}$&$\mathbf{t}$&
$\mathbf{u}$&$\mathbf{v}$&$\mathbf{w}$\\
\hline
~~i&6& 8& 10& 8& 4& 2\\
~~ii&6&8&10&-6&-2&-4\\
\hline
\end{tabular}
\end{center}
\begin{center}
\begin{tabular}{|l|ccc|ccc|c|}
\hline
$\mathbf{DC}^7$ unsrt:&
$\mathbf{r}$&$\mathbf{s}$&$\mathbf{t}$&
$\mathbf{s}+\mathbf{t}$&
$\mathbf{r}+\mathbf{t}$&
$\mathbf{r}+\mathbf{s}$&
min body diag\\
&&&&$-\mathbf{|u|}$&$-\mathbf{|v|}$&$-\mathbf{|w|}$&(MBD)\\
\hline
~~i&6&8&10&10&12&12&14\\
~~ii&6&8&10&12&14&10&12\\
\hline
\end{tabular}
\end{center}
\begin{center}
\begin{tabular}{|l|ccc|ccc|c|}
\hline
recover $\mathbf{G}^6$:&
$\mathbf{r}$&$\mathbf{s}$&$\mathbf{t}$&
$\mathbf{|u|}$&$\mathbf{|v|}$&$\mathbf{|w|}$&$\tau=\mathbf{r}+\mathbf{s}+\mathbf{t}$\\
&&&&&&&$-\mathbf{|u|}-\mathbf{|v|}-\mathbf{|w|}$\\
\hline
~~i&6&8&10&8&4&2&$\tau\, 10 \neq \text{MBD}\, 14$ (disagree $+++$)\\
~~ii&6&8&10&6&2&4&$\tau\, 12 = \text{MBD}\, 12$ (agree $---$)\\
\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}
\caption{Gruber's example of five-fold alternative Buerger-reduced
cells for a lattice as $[a,b,c,\alpha,\beta,\gamma]$}.
\label{table::grubercells}
\begin{center}
\begin{tabular}{|l|cccccc|}
\hline
cell&$a$&$b$&$c$&$\alpha$&$\beta$&$\gamma$\\
\hline
i&2&4&4&60.00&79.19&75.52\\
ii&2&4&4&60.00&86.42&75.52\\
iii&2&4&4&120.00&93.58&100.80\\
iv&2&4&4&117.95&93.58&104.48\\
v&2&4&4&113.97&100.80&104.48\\
\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}
\caption{Gruber's example of five-fold alternative Buerger-reduced
cells for a lattice as\\
$\mathbf{G}^6$ $[\textbf{r},\textbf{s},\textbf{t},\textbf{u},\textbf{v},\textbf{w}]$.}
\label{table::gruberg6cells}
\begin{center}
\begin{tabular}{|l|cccccc|c|}
\hline
\textbf{cell}&$\textbf{r}$&$\textbf{s}$&$\textbf{t}$&$\textbf{u}$&$\textbf{v}$&$\textbf{w}$&\textbf{boundary}\\
\hline
i&4&16&16&16&3&4&27C\\      
ii&4&16&16&16&1&4&27C\\      
iii&4&16&16&-16&-1&-3&2F8\\      
iv&4&16&16&-15&-1&-4&2FE\\    
v&4&16&16&-13&-3&-4&2FE\\      
\hline
\end{tabular}
\end{center}
\end{table}

\begin{table}
\caption{Gruber's example of five-fold alternative Buerger-reduced
cells for a lattice as unsorted $\mathbf{DC}^7$.  In order to get
past the fact that all the cells reduce to the same Niggli-reduced cell,
a small perturbation of .01 was applied to $u$ or $w$ in
$ii$ -- $v$ to
get away from the relevant boundaries.}.
\label{table::gruberdc7unsrt}
\begin{center}
\begin{tabular}{|l|ccc|ccc|c|}
\hline
cell&$1$&$2$&$3$&$4$&$5$&$6$&$7$\\
\hline
i&4&16&16&17&19&16&16\\
ii&4&16&16&17.01&19&16&16.01\\
iii&4&16&16&16.01&19&17&16.01\\
iv&4&16&16&17&19&16.01&16.01\\
v&4&16&16&19&17&16.01&16.01\\
\hline
\end{tabular}
\end{center}
\end{table}

%\begin{figure}
%\caption{Caption describing figure.}
%\includegraphics{fig1}
%\end{figure}

\begin{figure}
\caption{Truncated octahedron  \cite{wikipeditrancatedoctahedron}.
Image licensed under the Creative Commons Attribution-Share Alike 3.0 Unported license. Subject to disclaimers.  See web site.
}
\label{fig:truncoctahedron}
\begin{center}
\includegraphics[scale=.5]{Truncatedoctahedron}
\end{center}
\end{figure}

\begin{figure}
\caption{Historical timeline of studies of crystallographic lattice
characterization.  Figure drawn by E. Kincaid.  Used with
permission of the artist.
}
\label{fig:timeline}
\begin{center}
\includegraphics[scale=.5]{TimeLine_rev}
\end{center}
\end{figure}

\begin{table}
\caption{Roof/Niggli symbol, International Tables (IT) lattice character, Bravais lattice type, unsorted ${\mathbf{DC}^{\mathbf{7}}}$ subspace, boundary polytope.
Note that the variables r, s and t are non-negative, and u, v and w may be
positive, negative or zero as constrained below.}
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
{\bf Roof/}     &{\bf IT}&{\bf Bravais}&{\bf Unsorted ${\mathbf{DC}^{\mathbf{7}}}$}&{\bf Bound-}\\
{\bf Niggli}&{\bf Lattice}&{\bf Lattice}&{\bf Subspace}&{\bf ary}\\
{\bf Symbol}&{\bf Char}&{\bf Type}&                             & {\bf Polytope} \\
\hline
44A&3&${\bf cP}$&$(r,r,r,2r,2r,2r,3r)$&$12345\!=\!12\hat{3}\!=\!12\hat{4}\!=\!12\hat{5}$\\
\hline
44C&1&$cF$&$(r,r,r,r,r,r,2r)$&12679ACD\\
\hline
44B&5&$cI$&$(r,r,r,4 r/3,4 r/3, 4 r/3,r)$&$\text{12F2}^{\prime} \text{F}^{\prime} = 1\hat{2}\hat{\text{F}}$\\
\hline
45A&11&${\bf tP}$&$(r,r,t,r+t,r+t,2r,2r+t)$&$1345 = 1\hat{3} = 1\hat{4} = 1\hat{5}$\\
45B&21&${\bf tP}$&$(r,s,s,2s,r+s,r+s,r+2s$&$2345 = 2\hat{3} = 2\hat{4} = 2\hat{5}$\\
\hline
45D&6&$tI$&$(r,r,r,r-w/2,r-w/2,2r+w,r),$&\\
&&&$-r \leq w \leq 0$&$\text{12FF}^{\prime} = 12\hat{\text{F}}$\\
45D&7&$tI$&$[r,r,r,2*r+u,r-u/2,r-u/2,r],$&\\
&&&$-r \leq u \leq 0$&$\text{12F2}^{\prime} = 1{\hat{2}}\text{F}$\\
45C&15&$tI$&$(r,r,t,t,t,2r,t)$&158BF\\
45E&18&$tI$&$(r,s,s,-r/2+2s,s,s,-r/2+2s)$&$\text{2ADA}^{\prime}  = 2{\hat{A}}\text{D}$\\
\hline
48A&12&${\bf hP}$&$(r,r,t,r+t,r+t,r,r+t])$&134E\\
48B&22&${\bf hP}$&$(r,s,s,s,r+s,r+s,r+s)$&2458\\
\hline
49C&2&$hR$&$(r,r,r,2r-u,2r-u,2r-u,3r-u),$&\\
&&&$0 < u \leq r$&$121^\prime 2^\prime = \hat{1}\hat{2}$\\
49D&4&$hR$&$(r,r,r,2r+u,2r+u,2r+u,3r+3u),$&\\
&&&$-r \leq u \leq 0$&$121^\prime 2^\prime = \hat{1}\hat{2}$\\
49B&9&$hR$&$(r,r,t,t,t,r,r+t)$&1679ACD\\
49E&24&$hR$&$(r,s,s,s+r/3,s+r/3,s+r/3,s)$&$2F2^\prime \text{F}^\prime = \hat{2}\hat{\text{F}}$\\
\hline
50C&32&${\bf oP}$&$(r,s,t,s+t,r+t,r+s,r+s+t)$&$345 = \hat{3} = \hat{4} = \hat{5}$\\
\hline
50D&13&$oC$&$(r,r,t,r+t,r+t,2r+w,2r+t+w),$&\\
&&&$-r \leq w \leq 0$&134\\
50E&23&$oC$&$(r,s,s,u+2*s,s+r,s+r,u+2*s+r),$&\\
&&&$-s \leq u \leq 0$&245\\
50A&36&$oC$&$(r,s,t,s+t,t,r+s,s+t)$&35B\\
50B&38&$oC$&$(r,s,t,s+t,r+t,s,s+t)$&34E\\
50F&40&$oC$&$[r,s,t,t,r+t,r+s,r+t)$&458\\
\hline
51A&16&$oF$&$(r,r,s,r+s+u,r+s+u,-2u,s),$&\\
&&&$-r \leq u \leq 0$&$\text{1F1}^{\prime} = \hat{1}\text{F}$\\
51B&26&$oF$&$(r,s,t,-r/2+s+t,t,s,-r/2+s+t)$&$\text{ADA}^{\prime}  = \hat{A}\text{D}$\\
\hline
52A&8&$oI$&$(r,r,r,2r+u,2r+v,-u-v,r),$&\\
&&&$-r \leq u \leq 0, -r \leq v \leq 0$&12F\\
52B&19&$oI$&$ (r,s,s,2s-u,s,s,-r+2s+u),$&\\
&&&$0 < u \leq r $&\\
&&&$ (r,s,s,2s-u,s,s,r+2s-u),$&\\
&&&$r < u \leq s $&29C = 2AD\\
52C&42&$oI$&$r,s,t,t,t,r+s,t$&58BF\\
\hline
\end{tabular}
\end{center}
\label{NiggliFormsI}
\end{table}%

\begin{table}
\caption{Roof/Niggli symbol, International Tables (IT) lattice character, Bravais lattice type, unsorted ${\mathbf{DC}^{\mathbi{7}}}$ subspace, boundary polytope, continued.  Note that the variables r, s and t are non-negative, and u, v and w may be
positive, negative or zero as constrained below.}
\begin{center}
\begin{tabular}{|c|c|c|c|c|}
\hline
{\bf Roof/}     &{\bf IT}&{\bf Bravais}&{\bf Unsorted ${\mathbf{DC}^{\mathbf{7}}}$}&{\bf Bound-}\\
{\bf Niggli}&{\bf Lattice}&{\bf Lattice}&{\bf Subspace}&{\bf ary}\\
{\bf Symbol}&{\bf Char}&{\bf Type}&                             & {\bf Polytope} \\
\hline
53A&33&${\bf mP}$&$(r,s,t,t+s,v+t+r,s+r,v+t+s+r),$&\\
&&&$-r \leq v \leq 0$&35\\
53B&35&${\bf mP}$&$(r,s,t,u+t+s,t+r,s+r,u+t+s+r),$&\\
&&&$-r \leq  u \leq 0$&45\\
53C&34&${\bf mP}$&$(r,s,t,t+s,t+r,w+s+r,w+t+s+r)$,4&\\
&&&$-r \leq w \leq 0$&34\\
\hline
55A&10, 14&$mC$&$(r,r,t,t+r-u,t+r-u,2 r-w$&\\
&&&$,t+2r-w), 0 < u \leq w \leq r $&\\
&&&$(r,r,t,t+r-u,t+r-u,2 r-w,$&\\
&&&$t+2 r-2 u +w), 0 < w < u \leq r $&$\text{11}^{\prime} = \hat{1}$\\
57B&17&$mC$&$(r,r,t,u+t+r,v+t+r,-u-v,t),$&\\
&&&$-r \leq u \leq 0$&1F\\
55B&20, 25&$mC$&$(r,s,s,2s-u,s+r-v,s+r-v,$&\\
&&&$2s+r+u-2 v), -s \leq u \leq v \leq 0$&\\
&&&$(r,s,s,2s-u,s+r-v,s+r-v,$&\\
&&&$2s+r-u), -r \leq v < u \leq 0$&$\text{22}^{\prime} = \hat{2}$\\
57C&27&$mC$&$(r,s,s,2s-u,s,s,-r+2s+u)$&\\
&&&$0 < u \leq r$&\\
&&&$(r,s,s,2s-u,s,s,r+2s-u)$&\\
&&&$r < u \leq s$&9C = AD\\
56A&28&$mC$&$(r,s,t,t+s-u,t,s+r-2u,t+s-u)$&\\
&&&$0 < u \leq r$&$\text{AA}^{\prime} = \hat{A}$\\
56C&29&$mC$&$(r,s,t,t+s-u,t+r-2u,s,t+s-u)$&\\
&&&$0 < u \leq r$&$\text{DD}^{\prime} = \hat{\text{D}}$\\
56B&30&$mC$&$(r,s,t,t,t+r-v,s+r-2v,+t-v)+r)$&\\
&&&$0 < v \leq r$&$\text{77}^{\prime} =  \hat{7}$\\
54C&37&$mC$&$(r,s,t,t+s+u,t,s+r,t+s+u), -r \leq u \leq 0 $&5B\\
54A&39&$mC$&$(r,s,t,t+s+u,t+r,s,t+s+u), -r \leq u \leq 0$&4E\\
54B&41&$mC$&$(r,s,t,t,v+t+r,s+r,v+t+r),$&\\
&&&$-r \leq v \leq 0;$&58\\
57A&43&$mC$&$(r,s,t,t-w/2,t-w/2,w+s+r,t),$&\\
&&&$-r \leq w \leq 0$&$\text{FF}^{\prime} = \hat{\text{F}}$\\
\hline
\end{tabular}
\end{center}
\label{NiggliFormsII}
\end{table}%

\end{document}                    % DO NOT DELETE THIS LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-10-04 Taxonomy/Taxonomy.tex
udate and add References for Web Site

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr jouDelrnals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------
% submitted to JACryst March 24, 2022
\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE

%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
\journalcode{J}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation

% LCA IUCr id IUCr6401
% HJB IUCr id IUCr6484
% LCA ORCID 0000-0002-4451-1641
% HJB ORCID 0000-0002-0517-8532

\usepackage{booktabs}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{graphicx}
\usepackage{url}
\usepackage{calligra}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
\usepackage{color}

\newcommand{\OPES}[0]{$E^3toS^6$}
\newcommand{\OPESS}[0]{$$E^3toS^6$$}
\newcommand{\SVI}[0]{$\bf{S^{6}}$}
\newcommand{\SVIH}[0]{$\bf{\hat{S}^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\CIII}[0]{$\bf{C^{3}}$}
\newcommand{\EIII}[0]{$\bf{E^{3}}$}
\newcommand{\RIII}[0]{$\bf{R^{3}}$}
\newcommand{\DVII}[0]{$\bf{D^{7}}$}
\newcommand{\VVII}[0]{$\bf{V^{7}}$}
\newcommand{\MSVI}[0]{$M_{S^{6}}$}
\newcommand{\MEIII}[0]{$M_{E^{3}}$}
\newcommand{\EXIII}[0]{$\bf{E^{3 \times 3}}$}
\newcommand{\scalarsub}[2]{$#1_#2$}
\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\Gr}[2]{\textbf{Gr}(#1, #2)}
\newcommand{\Gra}[0]{Grassmannian}

\newcommand{\minus}{\scalebox{0.75}[1.0]{$-$}}

\begin{document}                  % DO NOT DELETE THIS LINE

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).

	{\LARGE \emph{\today}} \\
	\title{Taxonomy of lattice reduction}
	%\shorttitle{unit cell reduction}

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
	\author[b]{Herbert J.}{Bernstein}

	\aff[a]{Ronin Institute for Independent Scholarship, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
	\aff[b]{Ronin Institute for Independent Scholarship, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}

	% Use \shortauthor to indicate an abbreviated author list for use in
	% running heads (you will need to uncomment it).

	\shortauthor{Andrews and Bernstein}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	\keyword{lattice ''unit cell" reduction}

	\maketitle                        % DO NOT DELETE THIS LINE

	\begin{synopsis}
		The hierarchy of types of unit cell reduction is discussed.
	\end{synopsis}

	\begin{abstract}
		Several attributions exist for the reduced cells of lattices
		and for the reduction processes to produce them. The actual
		origins of the terms are compared and a taxonomy created. The
		terms ``Niggli reduction", ``Niggli cell", ``Delaunay reduction",
		and ``Delaunay cell" most accurately describe the crystallographic usages
		of reduced cells and reduction methods.

		An alternative spelling of Boris  Delaunay's name, adopted by him
		later in life, is Delone.
	\end{abstract}

	\section{Introduction}

	\citeasnoun{Niggli1928} and \citeasnoun{Delaunay1932} 
	first described the solutions
	to the problem of recognizing the appropriate Bravais 
	lattice type of a particular unit cell by the procedures 
	of reduction. They used the mathematical techniques 
	developed for the studies of matrices and polynomials.

	In recent literature, various names are associated with 
	the reduction methods used to produce ``reduced cells"
	in crystallography. \citeasnoun{OishiTomiyasu2012}\footnote{Introduced a matrix-based system for rapid determination of Bravais
	lattice type.} uses 
	the term ``Minkowski reduction" for the process often referred to as ``Niggli reduction". \citeasnoun{Andrews2019b}\footnote{Introduces the space $\bf{S^6}$ and also $\bf{C^3}$.} use the term 
	``Selling reduction", which is often 
	referred to as ``Delaunay reduction".

	In an attempt to clarify the relationships and meanings of the various 
	methods of reduction, we have prepared a table of hierarchy of the types
	and describe their origins.

	It should be noted that the terms used to describe these cells vary,
	often being named ``region", ``domain", ``parallelohedron", ``parallelopiped", or ``zone of influence". In the particular 
	case of lattices, they all describe objects that tile space, in no way
	differing from the concept of unit cell.

    The choice of a reduction technique is an important early step in measuring the similarities among lattices and in
    classifying lattices \cite{andrews2023measuring} 
    \cite{andrews2023approximating}.

	\section{Types of reduction}

	Table \ref{types} list the types of reductions commonly encountered. The table
	progresses (left to right) from general/n-dimensional to three dimensional, to
	more specific (sorted).

	\subsection{Taxonomy}

	\begin{table}
		\label{types}
		\caption{Types of reduction}
		\begin{tabular}{|c|c|c|c|}
			\toprule 
			algebraic & n-dimensional/geometric & three-dimensional & sorted \\
			\midrule
			\citeasnoun{Seeber1831}/\citeasnoun{eisenstein1851} &\citeasnoun{Minkowski1905}&	\citeasnoun{Buerger1960} &	\citeasnoun{Niggli1928}	 \\
			\midrule
			\citeasnoun{Selling1874}&	& \citeasnoun{Delaunay1932} &	 \\
			\midrule
			\citeasnoun{Dirichlet1850}&\citeasnoun{Voronoi1908} & \citeasnoun{Wigner1933constitution} &	 \\
			\bottomrule 
		\end{tabular}

	\end{table}

	\subsection{The specific types}

	\begin{itemize}

\item{Buerger reduction	\cite{Buerger1960}:
The Buerger reduced cell is defined as the cell with edge vectors that are the shortest, non-coplanar set of three vectors in the lattice. 
	This is a special case of Minkowski reduction. No sorting of the sizes
	of the lengths or angles is specified, and no special conditions are
	described for the issues of equalities of values.}

\item{Delaunay (aka Delone) reduction \cite{Delaunay1932}:
	Delaunay used Selling reduction to derive his specific
	reduction for three dimensions. Six scalars are used, which are the dot
	products of the base vectors and their negative sum. The lattice is
	reduced when all of the scalars are non-positive. The reduced cell is
	not unique in several ways. First, in the general case, there are 24
	permutations of the values that all specify the same lattice. Also,
	for cases where some of the scalars are zero, there are alternative
	cells.}

\item{Dirichlet reduction \cite{Dirichlet1850}:
	Dirichlet described the reduction of positive quadratic forms.}

\item{Eisenstein reduction	\cite{eisenstein1851}:
	Eisenstein derived reduction conditions for
	positive ternary quadratic forms.}

\item{Minkowski reduction \cite{Minkowski1905}:
	Minkowski defines a reduced cell in an n-dimensional vector
	space as a cell with the shortest non-linearly-dependent lengths.
	Neither length sorting or description of how equalities are to be resolved are required.}

\item{Niggli reduction \cite{Niggli1928}:  Niggli used the formalism of Eisenstein to derive the
	set of conditions to define a unique unit cell in three dimensions.
	The lengths of the edges are sorted, and special conditions
	are defined for cases of ambiguity.}

\item{Seeber reduction \cite{Seeber1831}:
Seeber derived reduction conditions for three-dimensional lattices,
which were adopted by Eisenstein and subsequently by Niggli, so a Niggli reduced cell is a Seeber reduced cell.}

\item{Selling reduction \cite{Selling1874}: Selling derived reduction conditions for binary and ternary quadratic forms.
The Selling cell differs from the Minkoski cell by being based
on introduction of a fourth vector $d = -a-b-c$ and reducing by
requiring all six angles between the four vectors to be obtuse
or right angles.
}

\item{Voronoi domains \cite{Voronoi1908nouvelles}: 	Voronoi was concerned with the ``regions of influence" of points in
	space. The term Voronoi domain is used to describe the regions
	of a Delaunay triangulation of a set of points. \citeasnoun{Voronoi1908nouvelles}
	addresses quadratic forms and polyhedral tesselations.
	For a lattice, a Voronoi cell is the set of points
	at least as close to a chosen lattice point as they
	are to any other lattice point.  This defines a
	cell around each lattice point that tiles the entire
	space.  Such a Voronoi cell of a lattice is also
	called a Dirichlet cell or a Delaunay cell or a
	Wigner-Seitz cell.}

\item{Wigner-Seitz cell \cite{Wigner1933constitution}:	The authors define the cell for the face-centered lattice of sodium.
	As defined, it is the Voronoi cell  in three dimensions. Although 
	this cell was defined only for face-centered cubic lattices,
	it apparently has been adopted by the physics community as the
	name of the Dirichlet cell in general.  As \citeasnoun{Hart2019robust} has shown, the
Wigner-Seitz cell centered on a given lattice point is contained entirely within the convex envelope of the immediate 
neighbors of a given lattice point, which are most efficiently
found by starting with Minkowski-, Buerger-, or Niggli-reduction, rather than Delaunay reduction \cite{bernstein2023invertible}.  The Wigner-Seitz cell (for the above case) is a truncated octahedron, and in general has six, eight, ten, twelve, or fourteen sides.
	(Fig. \ref{fig:octa}).
}
\end{itemize}

\begin{figure}
	\label{fig:octa}
	\caption{Truncated octahedron \cite{wikipeditrancatedoctahedron}}
	\includegraphics[width=7.25cm]{Truncatedoctahedron}
\end{figure}

	\section{Summary}

	An attempt is made to make more explicit the relationships of the various
	types of unit cell reduction used in crystallography.

	\ack{{\bf Acknowledgements}}

	Careful copy-editing by Frances C. Bernstein is gratefully acknowledged.

	Work by HJB funded in part by NIGMS (5P30GM133893-03 grant to BNL subaward to the Ronin Institute);

	%-------------------------------------------------------------------------
	% The back matter of the paper - acknowledgements and references
	%-------------------------------------------------------------------------

	% Acknowledgements come after the appendices

	% References are at the end of the document, between \begin{references}
		% and \end{references} tags. Each reference is in a \reference entry.

	%-------------------------------------------------------------------------
	% TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
	%-------------------------------------------------------------------------

	% Simple tables should use the tabular environment according to this
	% model

	\bibliography{Reduced}

	\bibliographystyle{iucr}

\end{document}                    % DO NOT DELETE THIS LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-10-04 UnitCells/UnitCells.tex
udate and add References for Web Site

--- 2023-05-31 C3/C3.tex
@@
 	{\LARGE \emph{\today}} \\
-	\title{Delone lattice studies in \CIII, the space of three complex variables}
+	\title{What are unit cells}
 	%\title{Note on the transformation of three-space basis vectors to  corresponding matrix for Delaunay scalars}
-	\shorttitle{properties of C3}
+	\shorttitle{What are unit cells}
 
@@
 
-	\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
+	\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
 	\author[b]{Herbert J.}{Bernstein}
@@
 	\keyword{lattice}
-	\keyword{reduction}
+	\keyword{unit cells}
 	\keyword{Delone}
 	\keyword{Selling}
-	\keyword{\CIII}
 
@@
 	\begin{synopsis}
-		The space \CIII{} is explained in more detail than
-		in the original description. Boundary transformations
-		of the fundamental unit are described in detail. 
-		A graphical presentation of the basic coordinates
-		is described and illustrated.
+		Unit cells
 	\end{synopsis}
@@
 	\begin{abstract}
-		The Delone (Selling) scalars, which are used in 
-		unit cell reduction and in lattice type determination,
-		are studied in \CIII, the space of three complex variables.
-		The three complex coordinate planes are composed of the six
-		Delone scalars. The transformations at boundaries of the
-		Selling reduced orthant are described as matrices of operators.
-		A graphical representation as the projections onto the
-		three coordinates is described.
+		Abstract Unit cells
 
@@
 
-	The scalars used by~\citeasnoun{Delaunay1932} in his formulation of Selling reduction ~\cite{Selling1874}
-	are (in the conventional order) $b \cdot c$, $a \cdot c$, $a \cdot b$, $a \cdot d$, 
-	$b \cdot d$, $c \cdot d$, where $d = -a-b-c$. 
-	(As a mnemonic device, 
-	observe that the first three terms use
-	$\alpha$, $\beta$, and $\gamma$, 
-	in that order, 
-	and the following terms use $a$, $b$, $c$, in that order.)
+	The term ``unit cell' is now the most commonly 
+	used term for a smallest repeating unit 
+	of a crystal structure.
+	It allows us to describe the crystal structure 
+	in terms of a simple geometric shape. 
+	This makes it easier to visualize the structure 
+	and to calculate its properties.
 
-	~\citeasnoun{andrews2019b} chose to 
-	represent the Selling scalars in the space \SVI{},
-	\Svec{} (defined in the order above), 
-	as a way to create a metric space
-	for the measurement of the distance between lattices. 
-	They also considered the representation 
-	of this space as the
-	space of three complex dimensions, \CIII{} or 
-	{$\{c_1$}, {$c_2$}, {$c_3\}$}.
+	What is a unit cell? Conventionally, it is either thought
 
-	In \CIII{}, in terms of the Selling scalars, 
-	a vector is defined as \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}, 
-	where the real and imaginary parts
-	of each are the ``opposite'' scalars 
-	according to the definition of~\citeasnoun{Delaunay1932}
-	(opposite in terms of the Bravais tetrahedron of scalars) (see~\citeasnoun{andrews2019a}).
-	As a mnemonic device, 
-	note that the complex components involve ($\alpha,a$), ($\beta, b$), and ($\gamma,c$).
-	Additionally, each complex term uses all 
-	four three-space vectors; for example, $c_1$ is ($b \cdot c$, $a \cdot d$).
+\section{Conventional 3-space methods}
 
-	~\citeasnoun{andrews2019b} considered the matrix representations 
-	of the reflections in \CIII{} (and in \SVI{}). 
-	This paper describes the boundary transformations 
-	at the edges of the fundamental	unit of \CIII{}
-	and also a graphical presentation. 
-	In \SVI{}, the fundamental unit is the all negative orthant, 
-	which contains only, and all of, the reduced cells. 
-	In \SVI{} and \CIII{}, 
-	the boundaries located where any \SVI{} scalar 
-	(or correspondingly in \CIII{}, a real or imaginary part) 
-	equals zero.
+\subsection{Lattices as basis vectors}
+
+The fundamental understanding of crystals begins from the
+study of lattices. A typical lattice 
+$\Lambda$  in 
+$\mathbb{R}^{n}$ thus has the form
+
+\begin{center}
+	$\Lambda =  
+\begin{Bmatrix}
+	\sum_{i=1}^n a_i \mathbf{v}_i \bigg|~{a}_i \in \mathbb{Z}
+\end{Bmatrix}
+$,
+\end{center}
+
+\noindent
+where: ($\mathbf{v}_1,...,\mathbf{v}_n)$ is a basis of the lattice. Conventionally, "the unit cell" is defined as the
+set of n vectors generated by $a_i = 1$.
+
+In crystallography, we study the lattice in three-dimensional space (n=3 in the above equation).
+We use 3x3 matrices to create different unit cells by manipulating a vector of 3D vectors.
+These manipulations are usually described as happening in 3D space, but the vector of 3D vectors is actually a mathematical object with nine parameters in a space that can be described as $R^3xR^3xR^3$, which is a nine-dimensional space.
+
+\subsection{a,b,c,$\alpha$,$\beta$,$\gamma$}
+\label{length/angles}
+
+The second conventional representation of unit cells is
+2 triplets: 3 edge lengths of the basis vectors and
+the 3 angles between them. This is a 6-dimensional 
+space: $R^3xR^3$.
+
+The problem with this representation occurs in the cases
+where the need is to compare two (or more) unit cells. 
+The representation as  $R^3xR^3$ informs us that the
+two  $R^3$ are not necessarily commensurate. Indeed,
+the edge lengths and the interaxial angles have no
+shared measure. (However, see Section \ref{polar}.)
+
+\subsection{Polar coordinates}
+\label{polar}
+
+Section \ref{length/angles} alluded to the incommensurate
+nature of the two subspace of $R^3xR^3$.
+
+\citeasnoun{Delone1975} commented on the importance of the "opposite"
+scalars in the Bravais tetrahedron. \citeasnoun{Andrews2019b} observed
+that the association of those opposites implied the association of
+each unit cell edge with the opposite interaxial angle:
+
+\begin{center}
+	\label{polarDef}
+	($a,\alpha$), ($b,\beta$), ($c, \gamma$).
+\end{center}
+\noindent
+They used that association to create a space $C^3$ (see Section \ref{c3}).
+Using the above definition (\ref{polarDef}, we can define a 6-dimensional
+space $P^3$. Unlike the issue with the direct use of lengths and angles,
+polar coordinates are a well-known 2-dimensional object with the alternative
+representation as $(x,y)$, the 2-dimensional plane, where we
+can immediately use Euclidean distance measure.
+
+\subsection{Dirac}
+\label{dirac}
+
+Dirac cells are defined as one or more primitive unit cell
+between a pair of lattice plans. Physicists use Dirac
+cell in the computation of energy levels \cite{dirac1930principles}.
+
+\section{Niggli space}
+
+\section{Delone space}
+
+\subsection{\SVI}
+\label{s6}
+
+\subsection{\CIII}
+\label{c3}
+
+\subsection{Root Invariant}
+\label{ri}
+% emailed Kurlin and Bright on 8/14/2023 requesting the
+% appropriate reference
+
+\section{Wigner-Seitz space}
 
@@
 
-	Complex numbers are represented 
-	in Cartesian format $(x,y)$, 
-	where $x$ is the real part and
-	$y$ is the imaginary part.
-
-	We represent a vector 
-	in \CIII{} by \{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\} 
-	as an alternative to  \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}.
-
-	Next we define the operators in \CIII{} that 
-	will be used in the matrix descriptions of 
-	the transformations at the boundaries of the fundamental unit,
-	see Table ~\ref{tab:operators}.
-
-	\begin{table}	
-		\label{tab:operators}
-		\caption{The Result column values give the products of each operator as applied to $(x_j, y_j)$}.
-		\begin{tabular}{c l l l }
-			\toprule
-			Operator			&Usage						&Result					&Name\\
-			\midrule
-			$\Minus{}_r $		&  $\Minus{}_r(c_j)$ 		& $(-x_j, -x_j+y_j)$ 	& Minus real \\ 
-			$\Minus{}_i $ 		& $\Minus{}_i(c_j)$ 		& $(x_j-y_j,-y_j) $ 	& Minus imag\\ 
-			$\mathcal{P}_r$ 	& $ \mathcal{P}_r(c_j)$ 	& $(x_j, x_j)$ 		& Plus real\\ 
-			$\mathcal{P}_i$ 	&  $\mathcal{P}_i(c_j)$  	& $(y_j, y_j)$ 		& Plus imag\\ 
-			$\Real$ 			&  $\Real(c_j)$ 			& $x_j$ or $(x_j,0)$		& Real \\ 
-			$\Imaginary$ 		&  $\Imaginary(c_j)$ 		& $y_j$ pr $(y_j,0)$ 		& Imaginary\\
-			\bottomrule
-		\end{tabular}		
-	\end{table}
-
-	\section{Matrices of boundary transformations}
-	~~\\
-
-	\citeasnoun{andrews2019b} listed the boundary transformations
-	for each of the six \SVI{} boundaries. For each,
-	they only listed 2 transformations, following \citeasnoun{Delaunay1932}
-	and \citeasnoun{Delone1975}. That list is incomplete in the sense
-	that applying the 24 reflections gives 24 boundary transformation,
-	four for each boundary, rather than 12.
-
-	For the boundary at $s_1$: (the real component of $c_1$).
-
-	$\begin{bmatrix}
-		\Minus{}_r		& 0					& 0 \\
-		\Plus{}_r		&  \textrm{i}\Real{}	& \Real{} \\
-		\mathcal{P}_r	& \textrm{i}\Imaginary{}	& \Imaginary{}
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\Minus{}_r	& 0 				& 0 \\
-		\Plus{}_r	&\textrm{i}\Imaginary{}	& \Imaginary{} \\
-		\Plus{}_r	& \textrm{i}\Real{}	& \Real{} \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\label{thirdmatrix}
-		\Minus{}_r	& 0			& 0 \\
-		\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-		\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
-	\end{bmatrix}	$
-	%
-	$\begin{bmatrix}
-		\Minus{}_r	& 0			& 0 \\
-		\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{} \\
-		\Plus{}_r	&  \Real{}	&\textrm{i}\Real{} \\
-	\end{bmatrix}$ \\
-
-	For the boundary at $s_4$: (the imaginary component of $c_1$).
-
-	$\begin{bmatrix}
-		\Minus{}_i	& 0					& 0 \\
-		\Plus{}_i	&  \textrm{i}\Real{}	& \Real{} \\
-		\Plus{}_i	& \textrm{i}\Imaginary{}	& \Imaginary{}
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\Minus{}_i	& 0					& 0 \\
-		\Plus{}_i	& \textrm{i}\Imaginary{}	& \Imaginary{} \\
-		\Plus{}_i	&  \textrm{i}\Real{}	& \Imaginary{} \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\Minus{}_i	& 0			& 0 \\
-		\Plus{}_i	&  \Real{}	& \textrm{i}\Real{} \\
-		\Plus{}_i	& \Imaginary{}		& \textrm{i}\Imaginary{}
-	\end{bmatrix}	$
-	%
-	$\begin{bmatrix}
-		\Minus{}_i	& 0		& 0 \\
-		\Plus{}_i	& \Imaginary{}	& \textrm{i}\Imaginary{} \\
-		\Plus{}_i	&  \Real{}& \textrm{i}\Imaginary{} \\
-	\end{bmatrix}$ \\
-
-	For the boundary at \sii{} (the real component of \cii{}):
-
-	$\begin{bmatrix}
-
-		\textrm{i}\Real{}	& \Plus{}_r		& \Real{} \\
-		0				& \Minus{}_r	& 0 \\
-		\textrm{i}\Imaginary{}	& \Plus{}_r		& \Imaginary{}
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\textrm{i}\Imaginary{}	& \Plus{}_r		&  \Imaginary{} \\
-		0				& \Minus{}_r	& 0 \\
-		\textrm{i}\Real{}	& \Plus{}_r		& \Real{} \\
-	\end{bmatrix}$ e
-	%
-	$\begin{bmatrix}
-		\Real{}	&\Plus{}_r		&   \textrm{i}\Real{} \\
-		0		& \Minus{}_r	&  0 \\
-		\Imaginary{}	&\Plus{}_r		& \textrm{i}\Imaginary{}
-	\end{bmatrix}	$ 
-	%
-	$\begin{bmatrix}
-		\Imaginary{}	& \Plus{}_r	& \textrm{i}\Imaginary{} \\
-		0		&\Minus{}_r	&  0 \\
-		\Real{}	& \Plus{}_r	&  \textrm{i}\Real{} \\
-	\end{bmatrix}$ \\
-
-	For the boundary at \sv: (the imaginary component of \cii{}).
-
-	$\begin{bmatrix}
-		\textrm{i}\Real{}	& \Plus{}_i		& \Real{} \\
-		0				& \Minus{}_i	& 0 \\
-		\textrm{i}\Imaginary{}	& \Plus{}_i		& \Imaginary{}
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\textrm{i}\Imaginary{}	& \Plus{}_i		&  \Imaginary{} \\
-		0				& \Minus{}_i	& 0 \\
-		\textrm{i}\Real{}	& \Plus{}_ri	& \Real{} \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\Real{}	&\Plus{}_i		&   \textrm{i}\Real{} \\
-		0		& \Minus{}_i	&  0 \\
-		\Imaginary{}	&\Plus{}_i		& \textrm{i}\Imaginary{}
-	\end{bmatrix}	$ 
-	%
-	$\begin{bmatrix}
-		\Imaginary{}	& \Plus{}_i	& \textrm{i}\Imaginary{} \\
-		0		&\Minus{}_i	&  0 \\
-		\Real{}	& \Plus{}_i	&  \textrm{i}\Real{} \\
-	\end{bmatrix}$ \\
-
-	For the boundary at \siii{} (the real component of \ciii{}):
-
-	$\begin{bmatrix}
-
-		\textrm{i}\Real{}	& \Real{}	& \Plus{}_r \\
-		\textrm{i}\Imaginary{}	& \Imaginary{}	& \Plus{}_r  \\
-		0				& 0		& \Minus{}_r  \\
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\textrm{i}\Imaginary{}	&  \Imaginary{}	& \Plus{}_r \\
-		\textrm{i}\Real{}	& \Real{}		& \Plus{}_r \\
-		0				& 0			& \Minus{}_r \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\Real{}	&   \textrm{i}\Real{}	&\Plus{}_r \\
-		\Imaginary{}	& \textrm{i}\Imaginary{}	&\Plus{}_r \\
-		0 		&  0				& \Minus{}_r \\
-	\end{bmatrix}	$ 
-	%
-	$\begin{bmatrix}
-		\Imaginary{}	& \textrm{i}\Imaginary{}	& \Plus{}_r \\
-		\Real{}	&  \textrm{i}\Real{}	& \Plus{}_r \\
-		0		&  0				&\Minus{}_r \\
-	\end{bmatrix}$ \\
-
-	For the boundary at \svi{} (the imaginary component of \ciii{}):
-
-	$\begin{bmatrix}
-
-		\textrm{i}\Real{}	& \Real{}	& \Plus{}_i \\
-		\textrm{i}\Imaginary{}	& \Imaginary{}	& \Plus{}_i \\
-		0				& 0		& \Minus{}_i 
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\textrm{i}\Imaginary{}	&  \Imaginary{}	& \Plus{}_i \\
-		\textrm{i}\Real{}	& \Real{}		& \Plus{}_i \\
-		0				& 0			& \Minus{}_i \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\Real{}	&   \textrm{i}\Real{} &\Plus{}_i \\
-		\Imaginary{}	& \textrm{i}\Imaginary{}	&\Plus{}_i \\
-		0		&  0				& \Minus{}_i \\
-	\end{bmatrix}	$ 
-	%
-	$\begin{bmatrix}
-		\Imaginary{}	& \textrm{i}\Imaginary{}	& \Plus{}_i \\
-		\Real{}	&  \textrm{i}\Real{}	& \Plus{}_i \\
-		0		&  0				&\Minus{}_i 
-	\end{bmatrix}$ \\
-
-\subsection{Derivation}
-
-Consider the \SVI{} vector 	
-	\{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\};
-in \CIII{} that vector is:  \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}.
-Stated in \CIII{}, the problem is to find a matrix that converts a \CIII{}
-vector to a more reduced one. Take the case of \si positive and the
-other scalars negative.\\
-
- \{(-\si,~\siv-\si ),~(\siii+\si,~\sii+\si), (\svi+\si,~\sv+\si)\}\\
- =
-\begin{tabular}{ccc}
-
- 	RedMat
-
-$ \begin{Bmatrix}
- $(\si,\siv )$\\ $(\sii,\sv)$\\$(\siii,\svi)$
-\end{Bmatrix}$
-\end{tabular}\\
-
-We need to solve for RedMat.
-Here, one of the four possible result vectors has been chosen 
-(arbitrarily).
-
-Element 1 of result vector is (-\si,~\siv-\si ), and
-it is created by multiplying the beginning \CIII{} vector by the
-first row of the transformation matrix. In this case, we see
-that \ci of the result only contains scalars from \ci of
-the input, so the first row must contain some operator to
-be multiplied times input \ci, and the second and third
-elements of the matrix row must be zero. We create an operator
-($Minus~real$) to transform the input \ci to the output \ci; so
-the first row of the reduction matrix is: $\Minus{}_i	 0		 0$.
-
-Element 2 of the result vector is (\siii+\si,~\sii+\si),
-and
-it is created by multiplying the beginning \CIII{} vector by the
-second row of the transformation matrix. The result \cii
-contains \si in both the real and imaginary parts. So
-the first matrix element in that row must insert \si, the
-real part of the input \ci, into both the real and imaginary
-parts. An operator to produce that is created 
-(Plus{}~real). Next we see that the result contains \sii and \siii.
-\siii appears in the real part of the result element and in the
-third part of the input vector. So the third element of the 
-second matrix row must extract the \siii, the real part of 
-\ciii of the input. So the second row now contains the Plus real
-operator and the element to put the  real part of \ciii of the input.
-The final matrix element of the second row (the second element)
-must extract the real part of input \cii and place it in the
-imaginary part of the output vector. The resulting second
-row of the matrix is $\Plus_r~\Real~i\Real$.
-
-Using similar logic, the third row of the matrix is:
-$\Plus_r~\Imaginary~i\Imaginary$.
-
-Having derived the first matrix, the other 23 can
-be derived by the same process. However because all the
-\SVI boundaries are related by the 24 reflection, the
-other 23 matrices can be generated simply
-by multiplying by each of the reflection matrices.
-
-\subsection{Example of use}
-
-To demonstrate the application of the above matrices,
-example C++ code was written to create matrices of operators;
-see the supplementary material for the code or the complete
-library reference in the Availability of Code below.
-
-Create a \CIII{} vector c3 with
-the values \{(1,2),(3,4),(5,6)\} (which does not
-correspond to a valid unit cell). Multiply it by the matrix
-of operators derived above.
-
-The matrix multiply can be implemented as follows"
-
-\begin{verbatim}
-C3 C3::c3_s1(const C3& a) {
-    C3 c;
-    c[0] = c3_Minus_Real(a[0]);
-    c[1] = c3_Plus_Real(a[0]) + c3_Real(a[1]) +
-        c3_I_Times_Real(a[2]);
-    c[2] = c3_Plus_Real(a[0]) + c3_Imag(a[1]) +
-        c3_I_Times_Imag(a[2]);
-    return c;
-}
-\end{verbatim}
-
-Therefore, RedMat = \\
-\begin{center}
-	$\begin{bmatrix}
-	\Minus{}_r	& 0			& 0 \\
-	\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-	\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
-\end{bmatrix}	$
-\end{center}
-
-For example:
-
-\begin{center}
-\{(-1.000,1.000) (4.000,6.000) (5.000,7.000)\} =\\
-$\begin{bmatrix}
-	\Minus{}_r	& 0			& 0 \\
-	\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-	\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
-\end{bmatrix}	$
-$\begin{Bmatrix}
-	(1,2)\\
-	(3,4)\\
-	(5,6)
-\end{Bmatrix}$\\
-
-\end{center}
-
-which can be verified by hand.
-
-	\section{Basics}
-
-	The standard representation of the identity operation is\\
-
-	$\bf{c'}$  = 
-	$\begin{bmatrix}
-		$1$	& $0$	&  $0$ \\
-		$0$	& $1$	& $0$ \\
-		$0$	& $0$	& $1$ \\
-	\end{bmatrix}$ 
-	$\bf{c}$. \\
-
-	The identity in \CIII{} can also be written:\\
-
-	$\bf{c'}$  = 
-	$\begin{bmatrix}
-		$1$	& $0$						&  $0$ \\
-		$0$	& \Real{} +\textrm{i}\Imaginary{}	& $0$ \\
-		$0$	& $0$						& \Real{} +\textrm{i}\Imaginary{} \\
-	\end{bmatrix}$ 
-	$\bf{c}$.
-	\\
-
-	~\citeasnoun{Delaunay1932} does not 
-	consider the boundary transformations 
-	in detail. 
-	However, he uses them to 
-	define the process of Selling reduction. 
-	For example in \SVI{}, 
-	he lists the following as one of the possible results 
-	for a transformation on \si{}, translated to \SVI{}:
-	\SvecA{}.
-	The third boundary transform for $s_1$ above
-	implements this operation and interchanges the
-	real part of \ciii{} and the imaginary part of \cii{}:
-
-	\begin{center}
-		$\begin{bmatrix}
-			\Minus{}_r	& 0			& 0 \\
-			\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-			\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
-		\end{bmatrix}	$
-	\end{center}
-
-	Considered in \CIII, Delone's alternate transformation 
-	for the \si{} boundary would
-	exchange the real of \cii{} 
-	with the imaginary part of \ciii{}. That is the fourth
-	matrix above in the list for \si{}. 
-	The other two transformations for \si{} can 
-	be generated from the two we have 
-	just mentioned by the ``exchange operation''~\cite{andrews2019b} 
-	applied to the second and third \CIII{} coordinates. Delone 
-	did not describe the latter two transformations, 
-	perhaps because even a single transformation 
-	was adequate to implement reduction. He had already listed two.
-
-	\section{Graphical display of projections}
-
-	The two-dimensional nature of the three coordinates of \CIII{} 
-	suggests their use for graphical display.
-
-	As an example, we use Phospholipase A2 (\textbf{PLA2}) (retrieved from 
-	the Protein Data Bank \cite{Bernstein1977}), which has had several similar
-	or identical structures determined \cite{LeTrong2007}. \citeasnoun{andrews2019b}
-	found additional cases (see Table \ref{PLA2})
-
-	\begin{table}
-		\begin{tabular}{m{0.08\textwidth} c c c c c c c c c}
-			\toprule
-			PDB id & Centering & a&b&c&$\alpha$&$\beta$&$\gamma$\\
-			\midrule
-			1DPY & hR& 57.98& 57.98& 57.98& 92.02& 92.02& 92.02\\
-			~~\& 1FE5 \\
-			1G0Z  & hP& 80.36& 80.36& 99.44& 90&    90&    120\\
-			~~\& 1U4J\\
-			1G2X & mC& 80.95& 80.57& 57.1 & 90&    90.35&  90\\
-			2OSN & hR& 57.10& 57.10& 57.10& 89.75& 89.75&  89.75\\
-			\bottomrule
-		\end{tabular}	\\
-		\caption{Phospholipase A2 unit cells}
-		\label{PLA2}
-	\end{table}
-
-	Below, Figure \ref{fig1} shows the unit cells as 
-	reported (the centering of lattices has not been removed).
-	The following figures (Figures \ref{fig2} to \ref{fig5}) show various transformations and
-	embellishments of the reported cells.
-
-	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{1_35}
-		\caption{Phospholipase A2 unit cells as reported. }
-		\label{fig1}
-	\end{figure}
-
-	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{2_06}
-		\caption{The PLA2 unit cells Niggli-reduced. The similarity 
-			of the 3 projections is indicative of the exact or nearly
-			exact rhombohedral symmetry.}
-		\label{fig2}
-	\end{figure}
-
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{5_21}
-	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 10\%
-		orthogonally to \SVI{} vector. }
-	\label{fig5}	
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{19_57}
-	label{19\_57}
-	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 10\%
-		orthogonally to \SVI{} vector, followed by Niggli reduction.}
-\end{figure}
-
-	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{3_03}
-		\caption{The PLA2 unit cells Delone-reduced. }
-		\label{fig3}
-	\end{figure}
-
-	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{5_38}
-		\label{5_38}
-		\caption{The PLA2 cells Delone reduced and
-		5 copies perturbed by 10\%.}
-	\end{figure}
-
-	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{6_53}
-		\label{6_53}
-		\caption{The same cells as plotted in Figure \ref{5_38} 
-			and Delone reduced.}
-	\end{figure}
-
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{7_14}
-	\label{7_14}
-		\caption{The same cells as plotted in Figure \ref{5_38} 
-	and Delone reduced, and then each multiplied by the 24
-	\SVI{} reflections.}
-\end{figure}
-
 	\section{Summary}
 
-	The transformation matrices shown above demonstrate the 
-	considerable regularity of Selling reduction, as used by 
-	Delaunay (1932). The reduction operations of Niggli 
-	reduction \cite{Niggli1928} are more complex. 
-	While all the boundaries of the non-positive 
-	orthant of \SVI{} are essentially the same (and 
-	related by reflections), the boundaries formed in 
-	Niggli reduction are of multiple types, and the fundamental unit 
-	of the space representing Niggli-reduced cells (\GVI{}, see \citeasnoun{Andrews2014}) is non-convex.
-
-	Several aspects of \CIII{} are evident from inspecting the matrices. 
-	First, each boundary has four possible 
-	transformations that can be applied. Since each of 
-	the transformations at boundaries are self-inverse, they are the 
-	same transformations that would be used in the process of 
-	cell (lattice) reduction. \citeasnoun{Delaunay1932} and \citeasnoun{Delone1975} give only two choices, presumably for simplicity, 
-	omitting transformations that use the ``exchange'' operator 
-	(see \citeasnoun{andrews2019b}).
+	Summary unit cells
 

### 2023-11-26 Taxonomy/Response_2023_11_02.tex
Taxonomy -- response to reviewers WEB_LRL -- add initial attempt at CommandUsage.

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
%\usepackage{bm}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
%\usepackage{calligra}
\usepackage{array}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
%\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\DeclareRobustCommand{\captionpar}{\par}
\usepackage{color}
\usepackage{ulem}
\usepackage{url}
\usepackage{yfonts}
\usepackage{relsize} 
\usepackage{wrapfig}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}

\newcommand{\SVI}[0]{$\bf{S^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\CIII}[0]{$\bf{C^{3}}$}
\newcommand{\DVII}[0]{$\bf{D^{7}}$}
\newcommand{\VVII}[0]{$\bf{V^{7}}$}

\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\Imaginary}[0]{$ \textfrak{I} $}
\newcommand{\Real}[0]{$ \textfrak{R} $}
\newcommand{\Exchange}[0]{$\textfrak{X}$}

\newcommand{\nounderline}[3]{\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineab}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineac}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
\newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}

\newcommand{\scalar}[1]{$#1$}

\newcommand{\scalarsub}[2]{$#1_#2$}

%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
\journalcode{A}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation
\makeatletter
\font\dummyft@=dummy \relax
\makeatother

\usepackage{setspace}

\begin{document}                  % DO NOT DELETE THIS LINE

	\raggedbottom
	\setlength{\parskip}{0pt}
	{\LARGE \emph{\today}} \\
%	\title{Measuring Lattices}
%	%\title{Note on the transformation of three-space basis vectors to  corresponding matrix for Delaunay scalars}
%	\shorttitle{Measuring Lattices}

	\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
	\author[b]{Herbert J.}{Bernstein}

	\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
	\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973-5000 \country{USA}}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	\maketitle                        % DO NOT DELETE THIS LINE
	\newcommand{\si}[0]{$s_1$}
	\newcommand{\sii}[0]{$s_2$}
	\newcommand{\siii}[0]{$s_3$}
	\newcommand{\siv}[0]{$s_4$}
	\newcommand{\sv}[0]{$s_5$}
	\newcommand{\svi}[0]{$s_6$}
	\newcommand{\Svec} [0] {\{\si, \sii, \siii, \siv, \sv, \svi \}}
	\newcommand{\SvecA} [0] {\{-\si, -\si+\sii, \si+\siii, \si+\sv, \si+\siv, \si+\svi \}}

	\newcommand{\OPES}[0]{$E^3toS^6$}
	\newcommand{\OPESS}[0]{$$E^3toS^6$$}
	\newcommand{\MSVI}[0]{$M_{S^{6}}$}
	\newcommand{\MEIII}[0]{$M_{E^{3}}$}
	\newcommand{\Plus}[0]{$\textfrak{P}$}	
	\newcommand{\Minus}[0]{$\textfrak{M}$}

	\newcommand{\ci}[0]{$c_1$}
	\newcommand{\cii}[0]{$c_2$}
	\newcommand{\ciii}[0]{$c_3$}

	\singlespacing

	\section{Note}

	Although this document is written in first person, Herbert
	Bernstein has edited and agreed to all statements.

	\section{Response to Reviewer 1}

	I would like to begin with the response to the issues of the reviewer
	by quoting the abstract of our article:

\begin{quote}
	Several attributions exist for the reduced cells of lattices
	and for the reduction processes to produce them. The actual
	origins of the terms are compared and a taxonomy created. The
	terms ``Niggli reduction", ``Niggli cell", ``Delaunay reduction",
	and ``Delaunay cell" most accurately describe the crystallographic usages
	of reduced cells and reduction methods. 
\end{quote}

It seems that Reviewer 1 has misunderstood the aim of the article, which is clearly given in the abstract.

Reviewer 1:

\begin{quotation}
	Current manuscript suggests brief description of various approaches to the choice of
reduced cells of crystal lattices and associated reduction methods. In fact, the problem of
reduction for lattices has much more general mathematical meaning and consists in finding
conditions on parameters of an “elementary cell” allowing to find for any lattice the unique
(in certain sense) frame characterizing the elementary cell. The application to classification
of 3-dimensional crystal lattices is an important physical problem and it is certainly
very useful for specialists in crystallography to adapt and to follow terminology and
the mathematical results formulated over more than 250-years history of such celebrated
mathematical problem.
After reading the current.
\end{quotation}

We completely agree with the reviewer's statements. But this is not
the paper, nor the intent, to describe the important topics they allude to.
Even our abstract is clear that this is not the intent. If the reviewer
would prepare the article they describe, I would eagerly read it.

On the issue of \cite{Delone1975}: unfortunately neither of us knows
the Russian language, so we have always consulted the English translation.
We agree that that paper is an important one. To point out an
important point from that paper, they discuss unit cells as being in
a metric space; it is the same space that we have elaborated as S6 
in our own work.

The reviewer's comment has led us to include a
reference to that very useful article.

	\section{Response to Reviewer 2}

	The reviewer writes:

\begin{quotation}
		Ten methods are described in sequence included in alphabetical order of the authors. It is noticeable that the latest
	date for the references of the methods is 1960 and many are from the 19th century. So I doubt the timeliness and
	tbe relevance to crystallographers of this contribution.
\end{quotation}

It is true that the table that lists the types and their taxonomy does
only have dates from 1960 and farther into the past. That is because
the table lists the types and the dates of their original 
creations/publication.

The later text cites dates including 2023.

The only other issue this reviewer lists is whether to change
the ordering of the descriptions of the types to chronological.
While they could be changed, that change would make locating
they particular types more difficult for someone who is not familiar with the original dates. I do not see the
advantage of such a change.

% \citeasnoun{OishiTomiyasu2012}

			\bibliography{Reduced}

			\bibliographystyle{iucr}

			%-------------------------------------------------------------------------
			% TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
			%-------------------------------------------------------------------------

			% Simple tables should use the tabular environment according to this
			% model

			% Postscript figures can be included with multiple figure blocks

		\end{document}                    % DO NOT DELETE THIS LINE
		%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2023-12-11 UnitCells/UnitCells.tex
added a brief note about PlotC3

@@
 	\newcommand{\svi}[0]{$s_6$}
+	\newcommand{\ri}{root invariant}
+%	\newcommand{\sl}{linearized \SVI}
+	\newcommand{\ls}{linearized \SVI}
 	\newcommand{\Svec} [0] {\{\si, \sii, \siii, \siv, \sv, \svi \}}
@@
 
+\subsection{Linearized \SVI}
+\label{ls}
+Linearized \SVI~is defined as a vector for which then
+values are derived from the \SVI~representation of a 
+lattice by taking the square root of each of the
+6 elements. The advantage created is that the elements
+are now in {\AA}ngstrom units. The disadvantage is that 
+there are no vectors beyond the boundaries, which have zero 
+values; beyond the boundaries one or more of the elements
+of the vector would require the square root of a negative
+number. The consequence is that a Euclidean distance 
+between to related points may greatly exceed the
+shortest measure that could be found.
+
 \subsection{Root Invariant}
@@
 % appropriate reference
+Root invariant is a lattice representation created by
+\cite{bright2021welcome}. Although it has several related
+definitions, the form for a general lattice is the 
+same as \ls, except that a specific sorting of the elements of
+the vector are specified in order to create a unique
+vector. Linearized \SVI~has the properties that there
+are several (up to 24 in some lattices) sortings possible.
+The disadvantage of linearized \SVI~applies here also. Root
+invariant has been used to display meaningful representations 
+of large numbers of experimentally determined lattices.
 

### 2024-08-03 DeloneGrid/DeloneGrid.tex
HorizontalDeloneGrid -- updaed Delone Grid paper - started

--- 2023-05-31 C3/C3.tex
@@
 
-%xxxxxx just for testing github % 2
-%xxxxxx just for testing github % 3
-
-\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
+\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
 \usepackage{amssymb}
@@
 \numberwithin{equation}{section}
-%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
-%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
-%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
+%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{''00}
+%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{''03}
+%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{''0A}
 %\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
@@
 %\numberwithin{equation}{section}
-%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
-%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
-%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
+%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{''00}
+%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{''03}
+%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{''0A}
 %\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
@@
 %-------------------------------------------------------------------------
-\journalcode{A}              % Indicate the journal to which submitted
+\journalcode{J}              % Indicate the journal to which submitted
 %   A - Acta Crystallographica Section A
@@
 	{\LARGE \emph{\today}} \\
-	\title{Delone lattice studies in \CIII, the space of three complex variables}
+	\title{A``Periodic Table'' for Bravais Lattices}
 	%\title{Note on the transformation of three-space basis vectors to  corresponding matrix for Delaunay scalars}
-	\shorttitle{properties of C3}
+	\shorttitle{A grid summarizing the Bravais lattice types according to Delaunay (Delone)}
 
@@
 
-	\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
+	\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
 	\author[b]{Herbert J.}{Bernstein}
 
-	\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
-	\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
+	\aff[a]{9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
+	\aff[b]{c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
 
@@
 	\keyword{Selling}
-	\keyword{\CIII}
+	\keyword{Delaunay}
 
@@
 	\begin{synopsis}
-		The space \CIII{} is explained in more detail than
-		in the original description. Boundary transformations
-		of the fundamental unit are described in detail. 
-		A graphical presentation of the basic coordinates
-		is described and illustrated.
+
+		A tabular arrangement of the Bravais lattice types as
+		defined by Delaunay is described.
+
 	\end{synopsis}
@@
 	\begin{abstract}
-		The Delone (Selling) scalars, which are used in 
-		unit cell reduction and in lattice type determination,
-		are studied in \CIII, the space of three complex variables.
-		The three complex coordinate planes are composed of the six
-		Delone scalars. The transformations at boundaries of the
-		Selling reduced orthant are described as matrices of operators.
-		A graphical representation as the projections onto the
-		three coordinates is described.
+
+		A convenient tabular display of the Bravais lattices types
+		as defined by Delaunay (using Selling reduction) is described.
 
@@
 
+	\section{stuff for larry to do}
+	supplementary material -- should there be a PDF file?/????
+
 	\section{Introduction}
+	\citeasnoun{kahr2023periodic} suggested a``Periodic table of the space groups''. In the
+	same spirit, a ``Periodic Table'' of the Bravais lattice types is presented here.
+	The table presents a systematic arrangement of the types, according to suggestions
+	by \citeasnoun{Delaunay1932}. Like the periodic table of the elements, a brief amount
+	of useful information is included.
+
+\section{The space \SVI{}}
 
@@
 	for the measurement of the distance between lattices.
-	They also considered the representation 
-	of this space as the
-	space of three complex dimensions, \CIII{} or 
-	{$\{c_1$}, {$c_2$}, {$c_3\}$}.
 
-	In \CIII{}, in terms of the Selling scalars, 
-	a vector is defined as \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}, 
-	where the real and imaginary parts
-	of each are the ``opposite'' scalars 
-	according to the definition of~\citeasnoun{Delaunay1932}
-	(opposite in terms of the Bravais tetrahedron of scalars) (see~\citeasnoun{andrews2019a}).
-	As a mnemonic device, 
-	note that the complex components involve ($\alpha,a$), ($\beta, b$), and ($\gamma,c$).
-	Additionally, each complex term uses all 
-	four three-space vectors; for example, $c_1$ is ($b \cdot c$, $a \cdot d$).
+\section{The table}
 
-	~\citeasnoun{andrews2019b} considered the matrix representations 
-	of the reflections in \CIII{} (and in \SVI{}). 
-	This paper describes the boundary transformations 
-	at the edges of the fundamental	unit of \CIII{}
-	and also a graphical presentation. 
-	In \SVI{}, the fundamental unit is the all negative orthant, 
-	which contains only, and all of, the reduced cells. 
-	In \SVI{} and \CIII{}, 
-	the boundaries located where any \SVI{} scalar 
-	(or correspondingly in \CIII{}, a real or imaginary part) 
-	equals zero.
+	\citeasnoun{Delaunay1932} and \citeasnoun{Delone1975} published a table of the Bravais types
+	as distinguished by Selling (Delaunay) reduction.\cite{andrews2023measuring} published
+	a somewhat modified version of Delaunay's table. The table presented here
+	is arranged with the lattice family types as columns and the Federov polyhedron
+	types as rows.
+			\begin{figure}
+		\includegraphics[width=10cm]{HorizontalDeloneGrid}
+		\label{grid}
+		\caption{The  Bravais lattice types.}
+	\end{figure}
 
-	\section{Notation}
+\section{The labeling of the columns (lattice families)}
 
-	Complex numbers are represented 
-	in Cartesian format $(x,y)$, 
-	where $x$ is the real part and
-	$y$ is the imaginary part.
+	The columns are labeled according to the lattice family: cubic,
+	tetragonal, rhombohedral, orthorhombic, monoclinic, and
+	triclinic (anorthic). The labels for each reflect the names:
+	\textbf{C}, \textbf{T}, \textbf{R}, \textbf{O},\textbf{ M}, \textbf{A}, \textbf{H} (rather than the choices of Delaunay: \textbf{K}, \textbf{Q},
+	\textbf{R},\textbf{ O},\textbf{ M}, \textbf{T}, \textbf{H}, respectively).
 
-	We represent a vector 
-	in \CIII{} by \{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\} 
-	as an alternative to  \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}.
+	A further change from the model of Delaunay is the splitting 
+	of orthorhombic and monoclinic into a pair of columns each
+	($\bf{O_A}$, $\bf{O_B}$, $\bf{M_A}$, $\bf{M_B}$). This change 
+	improved the display of these types. Delaunay had merged these
+	disparate types into a single cell when their Federov
+	types were identical. Thus, column $\bf{O}$ of \citeasnoun{Delaunay1932}
+	has become columns $\bf{O_A}$ and $\bf{O_B}$, etc.
 
-	Next we define the operators in \CIII{} that 
-	will be used in the matrix descriptions of 
-	the transformations at the boundaries of the fundamental unit,
-	see Table ~\ref{tab:operators}.
+\section{The key cell (PQRSTU)}
 
-	\begin{table}	
-		\label{tab:operators}
-		\caption{The Result column values give the products of each operator as applied to $(x_j, y_j)$}.
-		\begin{tabular}{c l l l }
-			\toprule
-			Operator			&Usage						&Result					&Name\\
-			\midrule
-			$\Minus{}_r $		&  $\Minus{}_r(c_j)$ 		& $(-x_j, -x_j+y_j)$ 	& Minus real \\ 
-			$\Minus{}_i $ 		& $\Minus{}_i(c_j)$ 		& $(x_j-y_j,-y_j) $ 	& Minus imag\\ 
-			$\mathcal{P}_r$ 	& $ \mathcal{P}_r(c_j)$ 	& $(x_j, x_j)$ 		& Plus real\\ 
-			$\mathcal{P}_i$ 	&  $\mathcal{P}_i(c_j)$  	& $(y_j, y_j)$ 		& Plus imag\\ 
-			$\Real$ 			&  $\Real(c_j)$ 			& $x_j$ or $(x_j,0)$		& Real \\ 
-			$\Imaginary$ 		&  $\Imaginary(c_j)$ 		& $y_j$ pr $(y_j,0)$ 		& Imaginary\\
-			\bottomrule
-		\end{tabular}		
-	\end{table}
+The International Tables for x-ray crystallography \cite{vonLaue1952}
+used the convention of labeling Selling parameters in the Bravais
+tetrahedron, as \textbf{P}, \textbf{Q}, \textbf{R}, \textbf{S}, \textbf{T}, \textbf{U}, corresponding to the \SVI{} scalars
+\Svec, respectively. For reference,
+the figure has been included in the upper left cell of the table.
+		\begin{figure}
+		\includegraphics[width=5cm]{PQRSTU}
+		\label{PQRSTU}
+		\caption{The key cell with the labels used in the
+		International Tables}
+	\end{figure}
 
-	\section{Matrices of boundary transformations}
-	~~\\
+\section{The rows (Federov types)}
 
-	\citeasnoun{andrews2019b} listed the boundary transformations
-	for each of the six \SVI{} boundaries. For each,
-	they only listed 2 transformations, following \citeasnoun{Delaunay1932}
-	and \citeasnoun{Delone1975}. That list is incomplete in the sense
-	that applying the 24 reflections gives 24 boundary transformation,
-	four for each boundary, rather than 12.
+	\citeasnoun{fedorov1885elements} determined that there are only
+	five kinds of parallelohedra that can be translated without rotations in 3-dimensional Euclidean space to fill space.
 
-	For the boundary at $s_1$: (the real component of $c_1$).
+	Except for rows 3 and 4, each row has a unique number of zero
+	parameters. For rows 3 and 4, the difference is whether the two
+	zeros are on opposite sides of the tetrahedron or on 
+	adjacent sides.
 
-	$\begin{bmatrix}
-		\Minus{}_r		& 0					& 0 \\
-		\Plus{}_r		&  \textrm{i}\Real{}	& \Real{} \\
-		\mathcal{P}_r	& \textrm{i}\Imaginary{}	& \Imaginary{}
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\Minus{}_r	& 0 				& 0 \\
-		\Plus{}_r	&\textrm{i}\Imaginary{}	& \Imaginary{} \\
-		\Plus{}_r	& \textrm{i}\Real{}	& \Real{} \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\label{thirdmatrix}
-		\Minus{}_r	& 0			& 0 \\
-		\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-		\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
-	\end{bmatrix}	$
-	%
-	$\begin{bmatrix}
-		\Minus{}_r	& 0			& 0 \\
-		\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{} \\
-		\Plus{}_r	&  \Real{}	&\textrm{i}\Real{} \\
-	\end{bmatrix}$ \\
+\section{The cells}
 
-	For the boundary at $s_4$: (the imaginary component of $c_1$).
-
-	$\begin{bmatrix}
-		\Minus{}_i	& 0					& 0 \\
-		\Plus{}_i	&  \textrm{i}\Real{}	& \Real{} \\
-		\Plus{}_i	& \textrm{i}\Imaginary{}	& \Imaginary{}
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\Minus{}_i	& 0					& 0 \\
-		\Plus{}_i	& \textrm{i}\Imaginary{}	& \Imaginary{} \\
-		\Plus{}_i	&  \textrm{i}\Real{}	& \Imaginary{} \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\Minus{}_i	& 0			& 0 \\
-		\Plus{}_i	&  \Real{}	& \textrm{i}\Real{} \\
-		\Plus{}_i	& \Imaginary{}		& \textrm{i}\Imaginary{}
-	\end{bmatrix}	$
-	%
-	$\begin{bmatrix}
-		\Minus{}_i	& 0		& 0 \\
-		\Plus{}_i	& \Imaginary{}	& \textrm{i}\Imaginary{} \\
-		\Plus{}_i	&  \Real{}& \textrm{i}\Imaginary{} \\
-	\end{bmatrix}$ \\
-
-	For the boundary at \sii{} (the real component of \cii{}):
-
-	$\begin{bmatrix}
-
-		\textrm{i}\Real{}	& \Plus{}_r		& \Real{} \\
-		0				& \Minus{}_r	& 0 \\
-		\textrm{i}\Imaginary{}	& \Plus{}_r		& \Imaginary{}
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\textrm{i}\Imaginary{}	& \Plus{}_r		&  \Imaginary{} \\
-		0				& \Minus{}_r	& 0 \\
-		\textrm{i}\Real{}	& \Plus{}_r		& \Real{} \\
-	\end{bmatrix}$ e
-	%
-	$\begin{bmatrix}
-		\Real{}	&\Plus{}_r		&   \textrm{i}\Real{} \\
-		0		& \Minus{}_r	&  0 \\
-		\Imaginary{}	&\Plus{}_r		& \textrm{i}\Imaginary{}
-	\end{bmatrix}	$ 
-	%
-	$\begin{bmatrix}
-		\Imaginary{}	& \Plus{}_r	& \textrm{i}\Imaginary{} \\
-		0		&\Minus{}_r	&  0 \\
-		\Real{}	& \Plus{}_r	&  \textrm{i}\Real{} \\
-	\end{bmatrix}$ \\
-
-	For the boundary at \sv: (the imaginary component of \cii{}).
-
-	$\begin{bmatrix}
-		\textrm{i}\Real{}	& \Plus{}_i		& \Real{} \\
-		0				& \Minus{}_i	& 0 \\
-		\textrm{i}\Imaginary{}	& \Plus{}_i		& \Imaginary{}
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\textrm{i}\Imaginary{}	& \Plus{}_i		&  \Imaginary{} \\
-		0				& \Minus{}_i	& 0 \\
-		\textrm{i}\Real{}	& \Plus{}_ri	& \Real{} \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\Real{}	&\Plus{}_i		&   \textrm{i}\Real{} \\
-		0		& \Minus{}_i	&  0 \\
-		\Imaginary{}	&\Plus{}_i		& \textrm{i}\Imaginary{}
-	\end{bmatrix}	$ 
-	%
-	$\begin{bmatrix}
-		\Imaginary{}	& \Plus{}_i	& \textrm{i}\Imaginary{} \\
-		0		&\Minus{}_i	&  0 \\
-		\Real{}	& \Plus{}_i	&  \textrm{i}\Real{} \\
-	\end{bmatrix}$ \\
-
-	For the boundary at \siii{} (the real component of \ciii{}):
-
-	$\begin{bmatrix}
-
-		\textrm{i}\Real{}	& \Real{}	& \Plus{}_r \\
-		\textrm{i}\Imaginary{}	& \Imaginary{}	& \Plus{}_r  \\
-		0				& 0		& \Minus{}_r  \\
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\textrm{i}\Imaginary{}	&  \Imaginary{}	& \Plus{}_r \\
-		\textrm{i}\Real{}	& \Real{}		& \Plus{}_r \\
-		0				& 0			& \Minus{}_r \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\Real{}	&   \textrm{i}\Real{}	&\Plus{}_r \\
-		\Imaginary{}	& \textrm{i}\Imaginary{}	&\Plus{}_r \\
-		0 		&  0				& \Minus{}_r \\
-	\end{bmatrix}	$ 
-	%
-	$\begin{bmatrix}
-		\Imaginary{}	& \textrm{i}\Imaginary{}	& \Plus{}_r \\
-		\Real{}	&  \textrm{i}\Real{}	& \Plus{}_r \\
-		0		&  0				&\Minus{}_r \\
-	\end{bmatrix}$ \\
-
-	For the boundary at \svi{} (the imaginary component of \ciii{}):
-
-	$\begin{bmatrix}
-
-		\textrm{i}\Real{}	& \Real{}	& \Plus{}_i \\
-		\textrm{i}\Imaginary{}	& \Imaginary{}	& \Plus{}_i \\
-		0				& 0		& \Minus{}_i 
-	\end{bmatrix}$
-	%
-	$\begin{bmatrix}
-		\textrm{i}\Imaginary{}	&  \Imaginary{}	& \Plus{}_i \\
-		\textrm{i}\Real{}	& \Real{}		& \Plus{}_i \\
-		0				& 0			& \Minus{}_i \\
-	\end{bmatrix}$ 
-	%
-	$\begin{bmatrix}
-		\Real{}	&   \textrm{i}\Real{} &\Plus{}_i \\
-		\Imaginary{}	& \textrm{i}\Imaginary{}	&\Plus{}_i \\
-		0		&  0				& \Minus{}_i \\
-	\end{bmatrix}	$ 
-	%
-	$\begin{bmatrix}
-		\Imaginary{}	& \textrm{i}\Imaginary{}	& \Plus{}_i \\
-		\Real{}	&  \textrm{i}\Real{}	& \Plus{}_i \\
-		0		&  0				&\Minus{}_i 
-	\end{bmatrix}$ \\
-
-\subsection{Derivation}
-
-Consider the \SVI{} vector 	
-	\{$(x_1,y_1)$, $(x_2,y_2)$, $(x_3, y_3)$\};
-in \CIII{} that vector is:  \{(\si,\siv ), (\sii,\sv),(\siii,\svi)\}.
-Stated in \CIII{}, the problem is to find a matrix that converts a \CIII{}
-vector to a more reduced one. Take the case of \si positive and the
-other scalars negative.\\
-
- \{(-\si,~\siv-\si ),~(\siii+\si,~\sii+\si), (\svi+\si,~\sv+\si)\}\\
- =
-\begin{tabular}{ccc}
-
- 	RedMat
-
-$ \begin{Bmatrix}
- $(\si,\siv )$\\ $(\sii,\sv)$\\$(\siii,\svi)$
-\end{Bmatrix}$
-\end{tabular}\\
-
-We need to solve for RedMat.
-Here, one of the four possible result vectors has been chosen 
-(arbitrarily).
-
-Element 1 of result vector is (-\si,~\siv-\si ), and
-it is created by multiplying the beginning \CIII{} vector by the
-first row of the transformation matrix. In this case, we see
-that \ci of the result only contains scalars from \ci of
-the input, so the first row must contain some operator to
-be multiplied times input \ci, and the second and third
-elements of the matrix row must be zero. We create an operator
-($Minus~real$) to transform the input \ci to the output \ci; so
-the first row of the reduction matrix is: $\Minus{}_i	 0		 0$.
-
-Element 2 of the result vector is (\siii+\si,~\sii+\si),
-and
-it is created by multiplying the beginning \CIII{} vector by the
-second row of the transformation matrix. The result \cii
-contains \si in both the real and imaginary parts. So
-the first matrix element in that row must insert \si, the
-real part of the input \ci, into both the real and imaginary
-parts. An operator to produce that is created 
-(Plus{}~real). Next we see that the result contains \sii and \siii.
-\siii appears in the real part of the result element and in the
-third part of the input vector. So the third element of the 
-second matrix row must extract the \siii, the real part of 
-\ciii of the input. So the second row now contains the Plus real
-operator and the element to put the  real part of \ciii of the input.
-The final matrix element of the second row (the second element)
-must extract the real part of input \cii and place it in the
-imaginary part of the output vector. The resulting second
-row of the matrix is $\Plus_r~\Real~i\Real$.
-
-Using similar logic, the third row of the matrix is:
-$\Plus_r~\Imaginary~i\Imaginary$.
-
-Having derived the first matrix, the other 23 can
-be derived by the same process. However because all the
-\SVI boundaries are related by the 24 reflection, the
-other 23 matrices can be generated simply
-by multiplying by each of the reflection matrices.
-
-\subsection{Example of use}
-
-To demonstrate the application of the above matrices,
-example C++ code was written to create matrices of operators;
-see the supplementary material for the code or the complete
-library reference in the Availability of Code below.
-
-Create a \CIII{} vector c3 with
-the values \{(1,2),(3,4),(5,6)\} (which does not
-correspond to a valid unit cell). Multiply it by the matrix
-of operators derived above.
-
-The matrix multiply can be implemented as follows"
-
-\begin{verbatim}
-C3 C3::c3_s1(const C3& a) {
-    C3 c;
-    c[0] = c3_Minus_Real(a[0]);
-    c[1] = c3_Plus_Real(a[0]) + c3_Real(a[1]) +
-        c3_I_Times_Real(a[2]);
-    c[2] = c3_Plus_Real(a[0]) + c3_Imag(a[1]) +
-        c3_I_Times_Imag(a[2]);
-    return c;
-}
-\end{verbatim}
-
-Therefore, RedMat = \\
-\begin{center}
-	$\begin{bmatrix}
-	\Minus{}_r	& 0			& 0 \\
-	\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-	\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
-\end{bmatrix}	$
-\end{center}
-
-For example:
-
-\begin{center}
-\{(-1.000,1.000) (4.000,6.000) (5.000,7.000)\} =\\
-$\begin{bmatrix}
-	\Minus{}_r	& 0			& 0 \\
-	\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-	\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
-\end{bmatrix}	$
-$\begin{Bmatrix}
-	(1,2)\\
-	(3,4)\\
-	(5,6)
-\end{Bmatrix}$\\
-
-\end{center}
-
-which can be verified by hand.
-
-	\section{Basics}
-
-	The standard representation of the identity operation is\\
-
-	$\bf{c'}$  = 
-	$\begin{bmatrix}
-		$1$	& $0$	&  $0$ \\
-		$0$	& $1$	& $0$ \\
-		$0$	& $0$	& $1$ \\
-	\end{bmatrix}$ 
-	$\bf{c}$. \\
-
-	The identity in \CIII{} can also be written:\\
-
-	$\bf{c'}$  = 
-	$\begin{bmatrix}
-		$1$	& $0$						&  $0$ \\
-		$0$	& \Real{} +\textrm{i}\Imaginary{}	& $0$ \\
-		$0$	& $0$						& \Real{} +\textrm{i}\Imaginary{} \\
-	\end{bmatrix}$ 
-	$\bf{c}$.
-	\\
-
-	~\citeasnoun{Delaunay1932} does not 
-	consider the boundary transformations 
-	in detail. 
-	However, he uses them to 
-	define the process of Selling reduction. 
-	For example in \SVI{}, 
-	he lists the following as one of the possible results 
-	for a transformation on \si{}, translated to \SVI{}:
-	\SvecA{}.
-	The third boundary transform for $s_1$ above
-	implements this operation and interchanges the
-	real part of \ciii{} and the imaginary part of \cii{}:
-
-	\begin{center}
-		$\begin{bmatrix}
-			\Minus{}_r	& 0			& 0 \\
-			\Plus{}_r	&  \Real{}	& \textrm{i}\Real{} \\
-			\Plus{}_r	& \Imaginary{}		&\textrm{i}\Imaginary{}
-		\end{bmatrix}	$
-	\end{center}
-
-	Considered in \CIII, Delone's alternate transformation 
-	for the \si{} boundary would
-	exchange the real of \cii{} 
-	with the imaginary part of \ciii{}. That is the fourth
-	matrix above in the list for \si{}. 
-	The other two transformations for \si{} can 
-	be generated from the two we have 
-	just mentioned by the ``exchange operation''~\cite{andrews2019b} 
-	applied to the second and third \CIII{} coordinates. Delone 
-	did not describe the latter two transformations, 
-	perhaps because even a single transformation 
-	was adequate to implement reduction. He had already listed two.
-
-	\section{Graphical display of projections}
-
-	The two-dimensional nature of the three coordinates of \CIII{} 
-	suggests their use for graphical display.
-
-	As an example, we use Phospholipase A2 (\textbf{PLA2}) (retrieved from 
-	the Protein Data Bank \cite{Bernstein1977}), which has had several similar
-	or identical structures determined \cite{LeTrong2007}. \citeasnoun{andrews2019b}
-	found additional cases (see Table \ref{PLA2})
-
-	\begin{table}
-		\begin{tabular}{m{0.08\textwidth} c c c c c c c c c}
-			\toprule
-			PDB id & Centering & a&b&c&$\alpha$&$\beta$&$\gamma$\\
-			\midrule
-			1DPY & hR& 57.98& 57.98& 57.98& 92.02& 92.02& 92.02\\
-			~~\& 1FE5 \\
-			1G0Z  & hP& 80.36& 80.36& 99.44& 90&    90&    120\\
-			~~\& 1U4J\\
-			1G2X & mC& 80.95& 80.57& 57.1 & 90&    90.35&  90\\
-			2OSN & hR& 57.10& 57.10& 57.10& 89.75& 89.75&  89.75\\
-			\bottomrule
-		\end{tabular}	\\
-		\caption{Phospholipase A2 unit cells}
-		\label{PLA2}
-	\end{table}
-
-	Below, Figure \ref{fig1} shows the unit cells as 
-	reported (the centering of lattices has not been removed).
-	The following figures (Figures \ref{fig2} to \ref{fig5}) show various transformations and
-	embellishments of the reported cells.
+		In the table described here, the conditions for each Delone
+	type are described using \SVI{}. For example, the cell for
+	primitive cubic lattices is $\bf{C5}$.
 
 	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{1_35}
-		\caption{Phospholipase A2 unit cells as reported. }
-		\label{fig1}
+		\includegraphics[width=5cm]{C5}
+		\label{C5}
+		\caption{The cell for the primitive cubic Bravais lattice type.}
 	\end{figure}
 
-	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{2_06}
-		\caption{The PLA2 unit cells Niggli-reduced. The similarity 
-			of the 3 projections is indicative of the exact or nearly
-			exact rhombohedral symmetry.}
-		\label{fig2}
-	\end{figure}
+	 $\bf{C5}$ is the label for the position (row and column, cubic and Federov type 5 reduced cell) within
+	 the table. The stylized tetrahedron shows which Selling
+	 parameters are zero or equal to each other. ``cP'' is the
+	 standard display for primitive cubic.
 
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{5_21}
-	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 10\%
-		orthogonally to \SVI{} vector. }
-	\label{fig5}	
-\end{figure}
+	The \SVI{} characteristic for that type is (000 rrr);
 
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{19_57}
-	label{19\_57}
-	\caption{The PLA2 unit cells, Niggli-reduced, and five copies were perturbed 10\%
-		orthogonally to \SVI{} vector, followed by Niggli reduction.}
-\end{figure}
+	\noindent
+	\si{} = $b \cdot c$ = 0 ($\alpha$ = 90$^{\circ}$)\\
+	\sii{} = $a \cdot c$ = 0 ($\beta$ = 90$^{\circ}$)\\
+	\siii{} = $a \cdot b$ = 0 ($\gamma$ = 90$^{\circ}$)\\
+	\siv{} = \sv{} = \svi{}
 
-	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{3_03}
-		\caption{The PLA2 unit cells Delone-reduced. }
-		\label{fig3}
-	\end{figure}
+\section{Dirichlet cells}
 
-	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{5_38}
-		\label{5_38}
-		\caption{The PLA2 cells Delone reduced and
-		5 copies perturbed by 10\%.}
-	\end{figure}
+	At the right end of the table, a few examples of Dirichlet
+	cells for each row.
 
-	\begin{figure}
-		\includegraphics[width=1.0\textwidth]{6_53}
-		\label{6_53}
-		\caption{The same cells as plotted in Figure \ref{5_38} 
-			and Delone reduced.}
-	\end{figure}
+	\cite{fedorov1885elements}
 
-\begin{figure}
-	\includegraphics[width=1.0\textwidth]{7_14}
-	\label{7_14}
-		\caption{The same cells as plotted in Figure \ref{5_38} 
-	and Delone reduced, and then each multiplied by the 24
-	\SVI{} reflections.}
-\end{figure}
+\section{The unused/vacant cells}
+
+	Prominent in the table are cells shown in gray. They have no
+	corresponding Bravais type. For instance, the cell that would be
+	$\bf{H1}$ would be a hexagonal cell with no zero parameters
+	(this is no 90$^{\circ}$ angles); 
+	clearly impossible.
+
+\section{Other changes from Delaunay}
+
+	Within the table there are five cells that have a yellow
+	background. Although Delaunay identified these among the 24
+	Bravais types, they are special in the sense that they
+	do no possess the full dimensionality of their lattice
+	family. Most likely, they are only of minor crystallographic 
+	interest.
+
+	Cells $\bf{A1}$ and $\bf{A2}$ are triclinic unit cells with 
+	one or two 90$^{\circ}$ angles. Respectively, they have only five
+	or four free parameters.
+
+	$\bf{O3}$ is a 2-D manifold that is the boundary between
+	 $\bf{O2}$ and $\bf{O1B}$.
+	With a characteristic of (rs0 rs0), $\bf{O3}$ has only two
+	free parameters, not the three that is normal for an 
+	orthorhombic unit cell.
+
+%.\cite{andrews2023measuring}
+%The table of Delaunay (1932) describing the 24 Bravais types in S6. It has
+%been redone removing the images of the Dirichlet cells, the non-reduced
+%cells, and adding the ‘lattice character’, which describes the linear
+%manifold of each type. The crystal family types have been renamed to
+%modern usage: Q changed to T for tetrahedral, K changed to C for cubic
+%and T changed to A for anorthic. Where Delone in some places included
+%two types in one table cell, they have been split into two (for example:
+%‘M1’ is split into ‘M1A’ and ‘M1B)’. Note that five types (O3, M3, M2B,
+%A2 and A3) are not normal crystallographic types. They are boundary
+%types, and they have fewer free parameters than the generic type requires.
+%For instance, O3 (character: rs0 rs0) has only two free parameters (r and
+%s), whereas an ordinary orthorhombic type requires three variables.
+
+%\begin{figure}
+%	\includegraphics[width=1.0\textwidth]{7_14}
+%	\label{7_14}
+%		\caption{The same cells as plotted in Figure \ref{5_38} 
+%	and Delone reduced, and then each multiplied by the 24
+%	\SVI{} reflections.}
+%\end{figure}
 
@@
 
-	The transformation matrices shown above demonstrate the 
-	considerable regularity of Selling reduction, as used by 
-	Delaunay (1932). The reduction operations of Niggli 
-	reduction \cite{Niggli1928} are more complex. 
-	While all the boundaries of the non-positive 
-	orthant of \SVI{} are essentially the same (and 
-	related by reflections), the boundaries formed in 
-	Niggli reduction are of multiple types, and the fundamental unit 
-	of the space representing Niggli-reduced cells (\GVI{}, see \citeasnoun{Andrews2014}) is non-convex.
+\section{Supplementary Material}
 
-	Several aspects of \CIII{} are evident from inspecting the matrices. 
-	First, each boundary has four possible 
-	transformations that can be applied. Since each of 
-	the transformations at boundaries are self-inverse, they are the 
-	same transformations that would be used in the process of 
-	cell (lattice) reduction. \citeasnoun{Delaunay1932} and \citeasnoun{Delone1975} give only two choices, presumably for simplicity, 
-	omitting transformations that use the ``exchange'' operator 
-	(see \citeasnoun{andrews2019b}).
-
-	\section{Availability of code}
-
-	The $C^{++}$ ~code for \CIII{} and related 
-	software tools is available in github.com, in
-	\url{https://github.com/duck10/LatticeRepLib.git}.
-	The program CmdToC3 uses the required files.
+The supplementary material includes the 
+svg (Scalable Vector Graphics) file that generates the table.
 
@@
 
-	%\section{blah blah blah -- Supplementary Material}
 	\ack{{\bf Acknowledgements}}
@@
 	gratefully acknowledged.
-	Our thanks to Jean Jakoncic and Alexei Soares for 
-	helpful conversations and access to data and facilities at 
-	Brookhaven National Laboratory.
 
@@
 
-	Funding for this research was provided in part by:  
-	US Department of Energy Offices of Biological and 
-	Environmental Research and of Basic Energy Sciences 
-	(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
-	U.S. National Institutes of Health (grant No. P41RR012408; 
-	grant No. P41GM103473; grant No. P41GM111244; 
-	grant No. R01GM117126,
-	grant No. 1R21GM129570); Dectris, Ltd.
+%	Funding for this research was provided in part by:  
+%	US Department of Energy Offices of Biological and 
+%	Environmental Research and of Basic Energy Sciences 
+%	(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
+%	U.S. National Institutes of Health (grant No. P41RR012408; 
+%	grant No. P41GM103473; grant No. P41GM111244; 
+%	grant No. R01GM117126,
+%	grant No. 1R21GM129570); Dectris, Ltd.
 

### 2024-11-25 DeloneGrid/DeloneGrid.tex
????

@@
 	\aff[a]{9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
-	\aff[b]{c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
+	\aff[b]{Fresh Pond Research Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
 
@@
 	as distinguished by Selling (Delaunay) reduction.\cite{andrews2023measuring} published
-	a somewhat modified version of Delaunay's table. The table presented here
+	a somewhat modified version of Delaunay's table. The table presented here in Fig. \ref{grid}
 	is arranged with the lattice family types as columns and the Federov polyhedron
-	types as rows.
+	types \cite{fedorov1891symmetry} \cite{fedorov1885elements} as rows.
 		\begin{figure}
-		\includegraphics[width=10cm]{HorizontalDeloneGrid}
+		\includegraphics[width=15cm]{HorizontalDeloneGrid.eps}
 		\label{grid}
@@
 	triclinic (anorthic). The labels for each reflect the names:
-	\textbf{C}, \textbf{T}, \textbf{R}, \textbf{O},\textbf{ M}, \textbf{A}, \textbf{H} (rather than the choices of Delaunay: \textbf{K}, \textbf{Q},
-	\textbf{R},\textbf{ O},\textbf{ M}, \textbf{T}, \textbf{H}, respectively).
+	\textbf{C}, \textbf{T}, \textbf{R}, \textbf{O},\textbf{ M}, \textbf{A}, \textbf{H} (rather than the choices of Delaunay: 
+        \textbf{K}, \textbf{Q},	\textbf{R},\textbf{ O},\textbf{ M}, \textbf{T}, \textbf{H}, respectively).
 
@@
 \Svec, respectively. For reference,
-the figure has been included in the upper left cell of the table.
+the figure has been included in the upper left cell of the table.  The values of the six
+Selling parameters given in the PQRSTU order unabiguously identify the cell and is called
+the ``\SVI{} characteristic'' of the cell type, given at the bottom center of each cell of
+the table..  Note in particular which Selling
+parameters are zero, denoting right angles among them, and which ones are equal,
+denoting cases of higher symmetry.
 	\begin{figure}
@@
 
-	At the right end of the table, a few examples of Dirichlet
+	At the right end of the table, there are a few examples of Dirichlet
 	cells for each row.
-
-	\cite{fedorov1885elements}
 
@@
 
-	Prominent in the table are cells shown in gray. They have no
-	corresponding Bravais type. For instance, the cell that would be
+	The cells shown in gray have no corresponding Bravais type. For 
+        instance, the cell that would be
 	$\bf{H1}$ would be a hexagonal cell with no zero parameters
-	(this is no 90$^{\circ}$ angles); 
-	clearly impossible.
+	(\it{i.e.} no 90$^{\circ}$ angles); 
+	clearly an impossible case.
 
@@
 	Bravais types, they are special in the sense that they
-	do no possess the full dimensionality of their lattice
-	family. Most likely, they are only of minor crystallographic 
+	do not possess the full dimensionality of their lattice
+	family. Most likely, they are of only minor crystallographic 
 	interest.
@@
 	With a characteristic of (rs0 rs0), $\bf{O3}$ has only two
-	free parameters, not the three that is normal for an 
+	free parameters, not the three that are normal for an 
 	orthorhombic unit cell.

### 2024-11-25 DeloneGrid/DeloneGrid.tex
Updating DeloneGrid paper. Herbert's revisions and a few of mine.

@@
 		\begin{figure}
-		\includegraphics[width=15cm]{HorizontalDeloneGrid.eps}
+		\includegraphics[width=15cm]{DeloneGridClaude_2-2.eps}
 		\label{grid}
@@
 	The columns are labeled according to the lattice family: cubic,
-	tetragonal, rhombohedral, orthorhombic, monoclinic, and
+	tetragonal, hexagonal-rhombohedral, orthorhombic, monoclinic, and
 	triclinic (anorthic). The labels for each reflect the names:
-	\textbf{C}, \textbf{T}, \textbf{R}, \textbf{O},\textbf{ M}, \textbf{A}, \textbf{H} (rather than the choices of Delaunay: 
+	\textbf{C}, \textbf{T}, \textbf{H-R}, \textbf{O},\textbf{ M}, \textbf{A}, \textbf{H} (rather than the choices of Delaunay: 
         \textbf{K}, \textbf{Q},	\textbf{R},\textbf{ O},\textbf{ M}, \textbf{T}, \textbf{H}, respectively).
 
-	A further change from the model of Delaunay is the splitting 
+	In addition to combining the hexagonal and rhombohedral columns,
+	a further change from the model of Delaunay is the splitting 
 	of orthorhombic and monoclinic into a pair of columns each
@@
 	parameters. For rows 3 and 4, the difference is whether the two
-	zeros are on opposite sides of the tetrahedron or on 
-	adjacent sides.
+	zeros are on opposite sides of the tetrahedron (type 3) or on 
+	adjacent sides (type 4). Type 3 is the case when the d-vector has
+	a 90$^{\circ}$ angle to a, b, or c. Type 4 is the case
+	where both 90$^{\circ}$ angles are among $\alpha$, $\beta$,
+	and $\gamma$.
 
@@
 
-	The cells shown in gray have no corresponding Bravais type. For 
-        instance, the cell that would be
-	$\bf{H1}$ would be a hexagonal cell with no zero parameters
-	(\it{i.e.} no 90$^{\circ}$ angles); 
-	clearly an impossible case.
+	The cells shown in gray have no corresponding Bravais type.
 
@@
 	orthorhombic unit cell.
+
+	$\bf{M3}$ is a 2-D manifold that is the boundary between
+	 $\bf{M1A}$ and $\bf{M1B}$.
+	With a characteristic of (rs0 ts0), $\bf{M3}$ has only three
+	free parameters, not the four that are normal for an 
+	monoclinic unit cell.
+
+	There is a lack of unanimity in the crystallographic community
+	on how best to present cells of the hexagonal-rhombohedral table cell
+	in R1.  In the table we follow the convention of \cite{Burzlaff1985}
+	of using the hR (rrr sss) \SVI{} lattice character (three equal
+	cell edges and three equal cell angles $ \neq 90^{\circ}$).  It is equally valid to
+	choose settings to combine the R1 cells in threes to form true 
+	hexagonal H4 hP (00r rrs) super-cells with $a$ and $b$ as two equal edges
+	with $\gamma=120^{\circ}$ and $\alpha = \beta = 90^{\circ}$, and the original
+	hidden pair of lattice points along body diagonals of the supercells.
 

### 2025-04-02 SELLA/Sella.tex
udate and add References for Web Site
Commandusage.html -- remove mentinos of PREFIX.
LinearAxis, Claude's attempt to help me write an article. SELLA -- updating the SELLA article. More to do

@@
 
-\newcommand{\SVI}[0]{$\bf{S^{6}}$}
+\newcommand{\SVI}[0]{$\mathbf{S^{6}}$}
 \newcommand{\GVI}[0]{$\bf{G^{6}}$}
@@
 \newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
+\newcommand{\charseq}[1]{({\bf #1})}
 
@@
 	{\LARGE \emph{\today}} \\
-\title{SELLA - A Program for Determining Bravais Lattice Types}
+	\title{Determining Bravais Lattice Types}
 	%\shorttitle{SELLA  Determining Bravais Lattice Types}
@@
 
-\cauthor[a]{Lawrence C.}{Andrews}{lawrence.andrews@ronininstitute.org}{}
+	\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
 	\author[b]{Herbert J.}{Bernstein}
@@
 
-\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
-\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}, Rochester Institute of Technology, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
+	\aff[a]{9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
+	\aff[b]{c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}, Rochester Institute of Technology, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
 	\aff[c]{Lawrence Berkeley National Laboratory, 1 Cyclotron Rd., Berkeley, CA, 94720 \country{USA}}
@@
 		straight-forward algorithm and a program for determining Bravais lattice type 
-based on Selling (Delone) reduction. It is a complete, closed solution, 
-and it provides a clear metric of fit to each type.
+		based on Selling (Delone) reduction. It provides a clear metric of fit to each type. The mathematical foundations for treating near-symmetries are
+		described. The 24 Bravais lattice types described by \citeasnoun{Delaunay1932} 
+		are described as linear manifolds in 6-dimensions.
 	\end{abstract}
-
-{\bf Note:}  Boris Delaunay in his later publications used the Russian version 
-of his surname: Delone. We will follow that choice.\\
 
@@
 	exactly corresponded to the actual type \cite{Patterson1957}. 
-Delone discussed the issues of scalars that have nearly zero values, 
-but the broader issues of other measurement errors were not 
-discussed. \citeasnoun{andrews2014} review the literature on 
-the efforts to create methods to utilize data that contain unavoidable 
-measurement error.
+	Delaunay \cite{Delaunay1932} discussed the issues of scalars that have nearly zero values, 
+	but the broader issues of measurement errors were not 
+	discussed. (Delaunay later changed the spelling to Delone, a
+	transliteration of the Cyrillic spelling of his name.)
 
-SELLA is a algorithm and program for determining Bravais lattice 
-type based on Selling (Delone) reduction. It is a complete, closed solution, and it provides a clear metric of fit to each type.
+	\citeasnoun{Zimmermann1985} described the program, DELOS
+	for the same purpose and based on the use of Selling scalars, which
+	SELLA uses as well; they treat the scalars as a list of 6 numbers, but do not consider them as points in 
+	a space.  SELLA treats the scalars as elements of vectors in the 6-dimensional
+	space, \SVI. Treating lattices as points in a liner space accesses the tools
+	and systematic procedures of linear algebra.
+
+	\citeasnoun{andrews2014} reviewed the literature on 
+	 efforts to create methods to utilize data that contain unavoidable 
+	measurement error and adds a more generic view of the problem.
+
+	SELLA is a based on Selling (Delone) reduction. It is a complete, closed solution, and it provides a clear metric of fit to each type.
 
 	\section{Terminology}
+
+	\subsection{The Bravais types}
 
@@
 	Delone). Delone used the convention of \citeasnoun{Bravais1850}
-with the sides of the tetrahedron labeled by the labels of the six
+	using the sides of the tetrahedron labeled by the symbols of the six
 	Selling scalars. Figure \ref{fig:PQRSTU} shows the labeling used in
@@
 	Table \ref{table:ScalarConcordance} shows other choices that have been used. The
- contents for each type are described in Figure \ref{fig:terminology}.
-
-CHARACTER
+	contents for each type are described in Figure \ref{fig:PQRSTU}.
 
 	\citeasnoun{Burzlaff1985} revised the labeling of the types.
-Here we have returned to the numbering of the types to that 
+	Here we have returned to the numbering of the Bravais types to that 
 	of \citeasnoun{Delaunay1932}.
@@
 	type. \citeasnoun{Burzlaff1985}
-renumbered by ordinals. Here we have labeled the Delone 
+	renumbered by ordinals. We have labeled the Delone 
 	types that have multiple items as "A" and "B". Table \ref{table:DeloneTypeConcordance}
-shows the various labelings. Figure \ref{fig:DeloneTypes} 
-recapitulates the figure of \citeasnoun{Delaunay1932} using 
+	shows the various labelings. Figure \ref{fig:rotatedDeloneTypes} 
+	revises the figure of \citeasnoun{Delaunay1932}, using 
 	more modern notation. Table \ref{table:ScalarConcordance} 
@@
 	used for the sides of the Bravais tetrahedron. 
-Figure \ref{fig:basictriangle} shows the arrangement of the 
+	Figure \ref{fig:PQRSTU} shows the arrangement of the 
 	symbols as used in the International Tables for Crystallography.
@@
 			\midrule
-		g&p&P&h$_{23}$&\scalarsub{s}{1}\\		
-		k&r&R&h$_{12}$&\scalarsub{s}{3}\\		
-		l&s&S&h$_{14}$&\scalarsub{s}{4}\\		
-		m&t&T&h$_{24}$&\scalarsub{s}{5}\\		
-		h&q&Q&h$_{13}$&\scalarsub{s}{2}\\		
-		n&u&U&h$_{34}$&\scalarsub{s}{6}\\		
+			g&P&h$_{23}$&$a_2.s_3$&\scalarsub{s}{1}\\		
+			k&R&h$_{12}$&$a_1.a_3$&\scalarsub{s}{3}\\		
+			l&S&h$_{14}$&$a_1.a_2$&\scalarsub{s}{4}\\		
+			m&T&h$_{24}$&$a_1.a_4$&\scalarsub{s}{5}\\		
+			h&Q&h$_{13}$&$a_2.a_4$&\scalarsub{s}{2}\\		
+			n&U&h$_{34}$&$a_3.a_4$&\scalarsub{s}{6}\\		
 			\bottomrule
@@
 			\midrule
-		K$_I$ & K1 & C1 \\		
-		K$_{III}$ & K2 & C3 \\		
-		K$_V$ & K3 & C5 \\		
-		Q$_I$ & Q1 & T1 \\		
-		Q$_{II}$ & Q2 & T2 \\		
-		Q$_V$ & Q3 & T5\\		
-		R$_I$ & R1 & R1 \\		
-		R$_{III}$ & R2 & R3 \\		
-		O$_I$ & O1 & O1A  \\		
-		O$_I$ & O2 & O1B  \\		
-		O$_{II}$ & O3 & O2   \\		
-		O$_{III}$ & O4 & O3   \\		
-		O$_{IV}$ & O5 & O4   \\		
-		O$_V$ & O6 & O5   \\		
-		M$_I$ & M1 & M1A  \\		
-		M$_I$ & M2 & M1B  \\		
-		M$_{II_2}$  & M3 & M2A  \\		
-		M$_{II}$ & M4 & M2B  \\		
-		M$_{III}$ & M5 & M3  \\		
-		M$_{IV}$ & M6 & M4  \\		
-		T$_I$ & T1 & A1  \\ 		
-		T$_{II}$ & T2 & A2 \\		
-		T$_{III}$ & T3 & A3 \\		
-		H$_{IV}$  & H1 & H4 \\ 
+			K$_I$ & K1 & C1 & cI \\		
+			K$_{III}$ & K2 & C3 &cF \\		
+			K$_V$ & K3 & C5 &cP \\		
+			Q$_I$ & Q1 & T1 &tI \\		
+			Q$_{II}$ & Q2 & T2 &tI \\		
+			Q$_V$ & Q3 & T5 & tP\\		
+			R$_I$ & R1 & H1 & hR \\		
+			R$_{III}$ & R2 & H3 & hR \\		
+			O$_I$ & O1 & O1A & oF  \\		
+			O$_I$ & O2 & O1B &oI  \\		
+			O$_{II}$ & O3 & O2 & oI   \\		
+			O$_{III}$ & O4 & O3$^*$ & oI   \\		
+			O$_{IV}$ & O5 & O4 & oS   \\		
+			O$_V$ & O6 & O5 & oP   \\		
+			M$_I$ & M1 & M1A & mC  \\		
+			M$_I$ & M2 & M1B & mC \\		
+			M$_{II_2}$  & M3 & M2A$^*$ & mC \\		
+			M$_{II}$ & M4 & M2B$^*$ & mC  \\		
+			M$_{III}$ & M5 & M3 & mC  \\		
+			M$_{IV}$ & M6 & M4 & mP  \\		
+			T$_I$ & T1 & A1 & aP  \\ 		
+			T$_{II}$ & T2 & A2$^*$ & aP \\		
+			T$_{III}$ & T3 & A3$^*$ & aP \\		
+			H$_{IV}$  & H1 & H4 & hP \\ 
 			\bottomrule
@@
 		~\\ ~\\
-	\caption{The 24 Delone types. \cite{Delaunay1932}}
+		\caption{The 24 Delone types. \cite{Delaunay1932}\\
+			$^*$ not a full Bravais type.}
 		\label{table:DeloneTypeConcordance}
@@
 	\cite{andrews2019} recast the Selling parameters to define a metric space, \SVI{}.
-We use the base vectors, $\bf{a}$, $\bf{b}$, $\bf{c}$ of the unit cell to define $\bf{d}=-(\bf{a}+\bf{b}+\bf{c})$.
+	The base vectors, $\bf{a}$, $\bf{b}$, $\bf{c}$ of the unit cell
+	are used to define $\bf{d}=-(\bf{a}+\bf{b}+\bf{c})$.
 	A point ${s}$ in \SVI{} is defined as:\\
@@
 	\subsection{Terminology of the Bravais lattice types}
-The Bravais tetrahedron symbol. Figure \ref{fig:PQRSTU}  and Table \ref{table:DeloneTypeConcordance}\\
-The different ways that the scalars have been labeled  Figure \ref{fig:terminology} \\
-The table of the Delone types. Figure \ref{fig:DeloneTypes} \\
+	The Bravais tetrahedron symbol is described in Figure \ref{fig:PQRSTU}.\\
+	The different ways that the scalars have been labeled is shown in Table \ref{table:ScalarConcordance} \\
+	The table of the Delone types is displayed in Figure \ref{fig:rotatedDeloneTypes}.
+The table has been transposed from that of Delone.
+Where Delone listed 2 types in one cell, we have been split them into parts A and B.
+The crystal family names and abbreviations have been changed to modern usage.
+\\
 
-\section{Algorithms for creation of an exhaustive list of the polytopes of the Bravais types}
+	\subsection{Character}
+	In \SVI{} for reduced cells, each Bravais type is a single, connected,
+	origin-inclusive, Euclidean manifold. As such, all points in a
+	particular manifold can be described by their ``character'', 
+	a representation of the vectorsns of that type. The table of \citeasnoun{Delaunay1932} included this information 
+	encoded in the formalism of Bravais, as a tetrahedron, with
+	the sides marked to show which of the 6 Selling parameters were
+	zero or which pairs (or multiplets) have equal values.
+
+	As an example, in Figure \ref{fig:O4_terminology}, the tetrahedron shows two
+	zeros, in positions P and Q (\scalarsub{s}{1} and \scalarsub{s}{2}). The slashes on the lines
+	show that positions S and T (\scalarsub{s}{4} and \scalarsub{s}{5}) are equal. This is 
+	encoded in the \textit{character}, which is shown at the bottom of the item as \charseq{00r sst}.
+
+	\section{Algorithms for creation of an exhaustive list of the manifolds of the Bravais types}
 
@@
 
+	We will need to create projectors for each of the Bravais
+	lattice types. The projectors will be use to generate
+	images of probe points in each of the types, and the perps
+	will measure the distance from each manifold.
+
 	We will use Selling reduction; Niggli reduction has a fundamental 
-unit that is non-convex, and some boundaries of the fundamental 
+	unit that is non-convex and some boundaries of the fundamental 
 	unit are partly closed and partly open \cite{andrews2014} leading 
 	to considerable complexity. Because Niggli reduction divides the 
-fundamental unit into separate all obtuse and all acute regions, 
+	fundamental unit of Niggli reduction into separate all obtuse and all acute regions, 
 	the transitions between those are quite complex \cite{andrews2014}. 
-In space $\bf{S^{6}}$ \cite{andrews2019b}, the fundamental 
+	In space \SVI{} \cite{andrews2019b}, the fundamental 
 	unit is the simple all-negative orthant of a 6-dimensional 
 	Cartesian space. Only two kinds of operations in the space 
-will need to be considered: the 6 boundary transform operations 
-at the boundaries of the orthant (zeros of an axis), and 
-24 reflections \cite{andrews2019}
+	will need to be considered: the 6  transform operations 
+	at the boundaries of the orthant (zeros of an axis) and 
+	24 reflections (allowed permutations) \cite{andrews2019}
 
@@
 	shall need all representations of all non-triclinic lattice 
-types within the $\bf{S^{6}}$ fundamental unit (all Selling-reduced 
+	types within the \SVI{} fundamental unit (all Selling-reduced 
 	cells) or those generated beyond the boundaries by boundary transforms. 
@@
 
- Seven of the 21 Delone types do not have a zero in their $\bf{S^{6}}$ 
+	Seven of the 21 Delone types do not have a zero in their \SVI{} 
 	representation; the presence of zero indicates that a form is 
- on the boundary of the $\bf{S^{6}}$ fundamental unit. Since a 
+	on the boundary of the \SVI{} fundamental unit. Since a 
 	point on the boundary transforms to another boundary 
@@
 	Cartesian Point (VCP) \cite{andrews2019b} appropriate to 
- each zero in the $\bf{S^{6}}$ vector. Because the reduction
+	each zero in the \SVI{} vector. Because the reduction
 	operations do not commute, in the cases where there is more 
@@
 	``perps'', the operators that give the normal vector to each 
- such manifold. For a particular polytope, the projector applied to the probe gives
- the least-squares best fit point within that polytope. Perps are computed by subtracting the 
- corresponding projector from the unit matrix. The norm of a perp times
- a probe is the distance from that polytope.
+	such manifold. For a particular manifold, the projector 
+	applied to the probe gives
+	the least-squares best fit point within that manifold
+	to the probe vector. 
+	Perps are computed by subtracting the 
+	corresponding projector from the unit matrix. The norm of avector
+	created by multiplying the probe times the projector
+	 is the distance from that manifold.
 
- \subsubsection{Generating sample vectors from all Bravais types}
+	\subsubsection{Generating sample vectors from all Bravais types\protect\\}
 
@@
 	value in computer arithmetic ($DBL\_MIN$ in the C language).
+	That value will be used to mark the positions of zeros
+	in vectors that have been processed.
 
- Step 2; For each Delone type, multiply its projector by 
- the vector from Step 1. Store  the results in a list 
- with entries for each type (see Table \ref{table:DeloneTypes} 
+	Step 2; For each Delone type, multiply its projector times 
+	the vector from Step 1. Store  the resulting vector
+	from Step 1 in a list 
+	with entries for each Bravais type (see Table \ref{table:DeloneTypes} 
 	for the projectors).
@@
 
-a) If the vector contains no zeros, generate the 24 reflections
- and store in the list by type.
+	a) If the vector result of Step 2 contains no zeros, generate the 24 reflections
+	and store in the list by Bravais type.
 
-\textit{\textbf{LCA--need to normalize the vectors, just like in lattice matching. ? Can we use the 
-lattice matching code here?}}
+	\textit{\textbf{LCA--need to normalize the vectors, just like 
+		in lattice matching. ? Can we use the 
+		lattice matching code here? This is very 
+		incomplete. The processing of multiple zeros is MISSING! But the use of DBL\_MIN is there.}}
 
-b) If the vector contains one zero, apply the boundary transform operation for that boundary
+	b) If the vector contains one zero, apply the boundary transform 
+	operation for that boundary
 	to the vector, and
@@
 	subtracting such a small value will not change
- any value but zero.
+	any value except zero.
 
 	c) If the vector contains more than one zero and \underline{not} $DBL\_MIN$, 
- first set the zero to $DBL\_MIN$,  and then return it to 
- step b). Then set the second zero to $DBL\_MIN$ (with the
- first zero still zero), and send that result to step b). 
- If there is a third zero,  process it the same way as two zeros and return to c).
+	first set one zero to $DBL\_MIN$,  and then return it to 
+	step b, using the modified vector. Then set the second zero to $DBL\_MIN$ (with the
+	first zero still zero), and send that result to step b. 
+	If there is a third zero,  process it the same way as 2 zeros and return to c.
+	(Note that no valid unit cell can have more than 3 zeros
+	in space \SVI{}.)
 
@@
 
- \subsubsection{Generate projectors from sample vectors}
+	\subsubsection{Generate projectors from sample vectors\\}
 
- Each vector in the list from the previous steps is a representative of one
- of the Delone types. Although a vector is a one-dimensional polytope,
- these vectors encode the information about the full polytopes of the
- Delone types in their zeros and their duplicate values.
+	Each vector in each of the lists from the previous steps is a representative of one
+	of the Delone types. Although a vector is a one-dimensional manifold,
+	these vectors encode the information about the full manifolds of the
+	Delone types in their zeros and their duplicate values. Now we 
+	discuss the creation of the projector matrices that correspond
+	to each vector.
 
@@
 
+	\textit{\textbf{LCA write pseudocode for Fraction.}}
+
 	The following steps will be repeated for each of the vectors in the
- list generated above. We generate a matrix \textit{m}.
+	list generated above in order to generate the list
+	of the corresponding projector matrices.
 
- Step 1: Zero any elements that have been set to $DBL\_MIN$.
+	To generate a matrix \textit{m} starting from a vector.
+
+	Step 1: Zero any elements of the vector that have been set to $DBL\_MIN$.
 
@@
 	\subsection {Measuring the fit}
-The distance from the given polytope is just
-the norm of the perp for that polytope applied to the probe.
+
+	When the library of projectors has been prepared, then
+	the distance of any probe from each of the manifolds
+	can be computed. 
+	The distance from the given manifold is just
+	the norm of the perp for that manifold applied to the probe.
+	An alternative, scale-free measure is the angle between
+	the sample vector and its projected image.
 
@@
 	O3 is in the row for rhombohedral and tetragonal. These types do not need to
-be searched for in general surveys.
+	be searched for in general surveys. If a probe is close to 
+	one of these degenerate manifold, it will be close to one
+	or both of the adjacent manifolds.
 
@@
 	communications).  To our knowledge, only the method of 
-\cite{tomiyasu2012} has clearly demonstrated completeness. 
+	\cite{OishiTomiyasu2012} has clearly demonstrated completeness
+	for exact values.
+
 	The justification that Sella is complete has three parts.
 
-First: We require that the cell being tested be reduced, and
+	First: We require that the cell being tested be Selling reduced, and
 	all Bravais lattice types within the \SVI{} fundamental unit (all
-\SVI{} scalars negative) be generated. Therefore,
- the 24 possible forms of a 
+	\SVI{} scalars negative) to have their projectors generated. Therefore,
+	the 24 possible forms (reflections) of a 
 	reduced cell will be close to one of the 24 copies of the types. 
 	For the reflections, we do not need to consider the case of 
-points near a boundary that might be outside; the boundary 
+	points near a boundary; the boundary 
 	transforms must be treated for that.
 
-Second: In the case of the boundary transform operations, we take the case of a point that is infinitesimally distant from the polytope of a 
+	Second: In the case of the boundary transform operations, 
+	we take the case of a point that is infinitesimally distant from the manifold of a 
 	Bravais lattice type that has one zero in its scalars. Take 
-the point in the polytope that is closest to the experimental 
+	the point in the manifold that is closest to the experimental 
 	point. We transform both by the corresponding boundary transform 
@@
 	Third: Assume we have a point that is outside the fundamental 
-unit and is near an undiscovered polytope for a Bravais lattice 
+	unit and is near an undiscovered manifold for a Bravais lattice 
 	type. If the point is near a boundary defined by a single 
 	zero, then we apply a boundary transform to the point and to 
-the Bravais polytope, bringing them both into the fundamental 
+	the Bravais manifold, bringing them both into the fundamental 
 	unit. If all of the variations of all of the Bravais 
-lattice types within and on the fundamental unit have 
-been generated, then this putative undiscovered polytope 
+	lattice types within and on the boundaries of the fundamental unit have 
+	been generated, then this putative undiscovered manifold 
 	has already been included in the list of possibilities. 
@@
 	unit will fail to find the closest Bravais lattice 
-(or lattices). As stated above, there are 239 
+	(or lattices). As stated above, we found 239 
 	non-triclinic possible representations.
@@
 	of the 21 Delone types? The reduced point is obviously in 
-the all-negative orthant. Further, because all possible 
+	the non-positive orthant. Further, because all possible 
 	copies of all of the Delone type manifolds have been 
@@
 	89539 unit cells were extracted from the Protein Data Bank 
+	in 2018
 	(\citeasnoun{bernstein1977} and \citeasnoun{berman2000}). 
@@
 	appropriate parameters were modest (see Table \ref{tab:PDBCells}).
+	The suspect cases have since been corrected \cite{Burley_personal_communication}.
 	\begin{table}
-	\label{tab:PDBCells}
 
@@
 
-	\caption{}
+		\caption{57 unit cells from the PDB around 2018 that did not match their types (later corrected)}
+		\label{tab:PDBCells}
 	\end{table}
@@
 
-The polytopes of the Bravais lattice types are characterized by required zeros
+	The manifolds of the Bravais lattice types are characterized by required zeros
 	of scalars and/or multiplets of equal values. For instance, all primitive 
-monoclinic cells have two 90 degree angles and so two zero scalars. Consider R1 as
-an example for multiplets. The character for R1 is $ \bf{rrr sss} $; two triplets.
+	monoclinic cells have two 90 degree angles and so require two zero scalars. Consider H1 as
+	an example for multiplets. The character for H1 is \charseq{rrr sss}; two triplets of equal values.
 
-For the case of M4, primitive monoclinic, the character is $ \bf{00r stu} $. 
-For a particular value for r,s,t,u, the best monoclinic cell will not
-depend on whatever values are substituted for the zeros. In the plane
-corresponding to the two zero scalars, we can draw a circle, each point
-on the circle corresponding to a lattice. Reducing each lattice and calculating
-the distance of the reduced cell from $ \bf{00r stu} $ will give a curve that
+	Consider the case of M4, primitive monoclinic, the character is \charseq{00r stu}. 
+	For particular values for r,s,t,u, the best monoclinic cell will not
+	depend on whatever values are substituted for the zeros. (It does not depend on the zero values because the projectors to that 
+	manifold create zero elements for any vector.) Imangine a plane
+	with axis that are the values of the two elements of the character that are zero. Points in the plane represent unit 
+	cells with angles defined by the \scalarsub{s}{1} and \scalarsub{s}{2} and r,s,t,u. We can draw a circle in the
+	\scalarsub{s}{1},\scalarsub{s}{2} plane, each point
+	on the circle corresponding to a lattice.
+
+	Reducing the lattices corresponding to the points on the
+	circle and r,s,t,u and calculating
+	the distance of the reduced cell from \charseq{00r stu}.
+	Plotting the curve of the circle with the distance of
+	each point from the origin changed to distance above
+	 will give a curve that
 	should not have any discontinuities. See Figure \ref{fig:M4}.
 
-Similarly, Bravais type O5 has a triplet of zeros, character $ \bf{000 rst} $. Here we
+	Similarly, Bravais type O5 has a triplet of zeros, character \charseq{000 rst}. Here we
 	can generate a sphere that surrounds the orthorhombic point. See Figure \ref{fig:O5}.
 
-Bravais type O4 has two zeros and a pair of required equal values, 
-character $ \bf{00r sst} $. The zeros can be treated the same as above. 
+	Bravais type O4 (side-centered orthorhombic) has two zeros and a pair of required equal values, 
+character \charseq{00r~sst}. The zeros can be treated the same as above
+for the case of M4 (primitive monoclinic). 
 	The pair of equal values must 
 	have an average value that remains equal to $\bf{s}$ in the lattice character. 
-That leaves 3 values to vary, and we can create a sphere around 
-the point $ \bf{00r sst} $. See Figure \ref{fig:O4}.
+	That means 3 values to vary (the 2 zeros and the distance fomaverage
+	value for the equal scalars), and we can create a sphere around 
+	the point \charseq{00r~sst}. See Figure \ref{fig:O4}.
 
 	\section{Comparisons}
-
+	\textit{\textbf{LCA this section is incomplete. Table needs caption/label.}}
 	Comparison to other distance measure; Data for other measures taken
@@
 
-\begin{tabular}{ccccccc}
+\begin{table}
+	\begin{tabular}{cccccc}
 		\toprule
-	Lattice character&G6 distance  &BGAOL   Z score&XDS QI &Scaled QI &Sella\\	
+		Lattice character & \GVI{} distance & BGAOL Z score & XDS QI & Scaled QI & Sella \\
 		\midrule
@@
 	\end{tabular}
+	\caption{Comparison of distance measures}
+	\label{tab:comparisons}
+\end{table}
+	~\\
 
-~\\
+	\section{Interpreting the results}
+
+	\section{What could go wrong?}
 
@@
 			Delone type & Count    \\
-\hline
+			\midrule
 			H4&12\\
@@
 			C5&16\\
-R1&4\\
-R3&12\\
+			H1&4\\
+			H3&12\\
 			T1&3\\
@@
 			O2&6\\
-O3$^*$&9\\
+			O3*&9\\
 			O4&36\\
@@
 			M2A&12\\
-M2B$^*$&12\\
-M3$^*$&18\\
+			M2B*&12\\
+			M3*&18\\
 			M4&12\\
@@
 
-R1  & (rP) & [1 1 1 0 0 0/  1 1 1 0 0 0/  1 1 1 0 0 0/  0 0 0 1 1 1/  0 0 0 1 1 1/  0 0 0 1 1 1] \\ 
-R3  & (rP) & [1 1 0 0 1 0/  1 1 0 0 1 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  1 1 0 0 1 0/  0 0 0 0 0 0] \\ \\
+				H1  & (hR) & [1 1 1 0 0 0/  1 1 1 0 0 0/  1 1 1 0 0 0/  0 0 0 1 1 1/  0 0 0 1 1 1/  0 0 0 1 1 1] \\ 
+				H3  & (hR) & [1 1 0 0 1 0/  1 1 0 0 1 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  1 1 0 0 1 0/  0 0 0 0 0 0] \\
+				H4  & (hP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 1 1 1 0/  0 0 1 1 1 0/  0 0 1 1 1 0/  0 0 0 0 0 1] \\ \\
 
@@
 				A3  & (aP) & [1 0 0 0 0 0/  0 1 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 0] \\ \\				
-
-H4  & (hP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 1 1 1 0/  0 0 1 1 1 0/  0 0 1 1 1 0/  0 0 0 0 0 1] \\ 
 				\bottomrule
@@
 	\begin{figure}
-\caption{Bravais triangle with labeling as used in the International Tables}
+		\caption{Bravais tetrahedron with labeling as used in the International Tables}
 		\includegraphics[width=3cm]{PQRSTU}
@@
 	\begin{figure}
-	\caption{Delone type description}
-	\label{fig:terminology}
-	\includegraphics[width=12cm]{terminology}
+		\caption{The O4 item from the Delone Grid with Delone type description}
+		\label{fig:O4_terminology}
+		\includegraphics[width=12cm]{O4_terminology}
 	\end{figure}
 
+	%	\begin{figure}
+		%		\caption{The Delone types}
+		%		\label{fig:DeloneTypes}
+		%		\includegraphics[width=12cm]{DeloneGridClaude_2-3}
+		%	\end{figure}
 	\begin{figure}
+		\centering
+		\includegraphics[angle=90, width=0.7\textheight]{DeloneGridClaude_2-3}
 		\caption{The Delone types}
-	\label{fig:DeloneTypes}
-	\includegraphics[width=12cm]{DeloneTypesGrid_2}
+		\label{fig:rotatedDeloneTypes}
 	\end{figure}

### 2025-07-02 SELLA/Sella.tex
Web_URL -- updated CmdLM to CmdLMS6 and added CmdLMP3
Sella -- added the critique that Claude wrote.

@@
 \begin{document}                  % DO NOT DELETE THIS LINE
+
+%	Claude.ai suggestions 10125-04-13
+%	I've reviewed your paper on Bravais lattice determination using the SELLA algorithm, and I have several suggestions for improvements:
+%	
+%	Consistency in terminology: There are variations between "Delaunay" and "Delone" throughout the paper.
+While you note that Delaunay later changed his name's spelling, consistency within the paper would be helpful.
+Perhaps use one version with a note about the alternative spelling at first mention.
+%	Incomplete sections: Several sections are marked as incomplete:
+%	
+%	Section 3.1.1 has a note "LCA–need to normalize the vectors..."
+%	Section 3.1.2 has "LCA write pseudocode for Fraction."
+%	Section 6 "LCA this section is incomplete. Table needs caption/label."
+%	Sections 7 and 8 are just headings without content.
+%	
+%	
+%	Figure numbering: There appear to be duplicate Figure 1 entries - one on page 15 and another on page 16, both showing the Bravais tetrahedron.
+%	Table numbering: Table 4 appears without a proper caption on page 13.
+%	Equation formatting: Many mathematical expressions could benefit from proper equation formatting, particularly in the algorithm descriptions.
+%	References: Some references in the text don't match entries in the reference list, and some citations don't follow a consistent format.
+%	Structural organization: The paper would benefit from clearer delineation between the theoretical foundations and the practical algorithm implementation.
+%	Typographical errors: Several minor typographical issues:
+%	
+%	"Imangine a plane" instead of "Imagine a plane" (page 12)
+%	"the distance fomaverage value" instead of "from average value" (page 12)
+%	"LCA," notes to self should be removed for publication
+%	
+%	
+%	Technical clarity: The step-by-step description of the algorithm in Section 3 could be improved with clearer pseudocode or examples.
+%	Figure descriptions: Figures 5-7 need more detailed captions explaining what is being visualized and how to interpret the colored points.
+%	
+%	I hope these suggestions are helpful for improving your manuscript.
 
@@
 	\citeasnoun{Zimmermann1985} described the program, DELOS
-	for the same purpose and based on the use of Selling scalars, which
+	for the above purpose and based on the use of Selling scalars, which
 	SELLA uses as well; they treat the scalars as a list of 6 numbers, but do not consider them as points in 
@@
 	 efforts to create methods to utilize data that contain unavoidable 
-	measurement error and adds a more generic view of the problem.
+	measurement error, and they add a more generic view of the problem.
 
@@
 	\citeasnoun{Delaunay1932} found that the 14 Bravais lattice types 
-	could be allocated among 24 types that are determined by the 
+	could be allocated among 24 types (here termed "Delone Types") that are determined by the 
 	type of Dirichlet unit cell (termed the Voronoi region \cite{Voronoi1908} or 
 	Dirichlet region \cite{dirichlet1850} by 
-	Delone). Delone used the convention of \citeasnoun{Bravais1850}
+	Delone). Delone used the tetrahedral convention of \citeasnoun{Bravais1850}
 	using the sides of the tetrahedron labeled by the symbols of the six
@@
 	Here we have returned to the numbering of the Bravais types to that 
-	of \citeasnoun{Delaunay1932}.
+	of \citeasnoun{Delaunay1932}, which addresses the type
+	of Federov polyhedron for the reduced cell.
 	In some cases, Delone included two forms within a single 
@@
 			\toprule
-			\cite{Delaunay1932} & \cite{Burzlaff1985}   &this  \\
-			&  & paper \\
+			\cite{Delaunay1932} & \cite{Burzlaff1985}   &this &Pearson  \\
+			&  & paper & symbol\\
 			\midrule
@@
 	\subsection{The space \SVI{}}
-	\cite{andrews2019} recast the Selling parameters to define a metric space, \SVI{}.
+	\citeasnoun{andrews2019} recast the Selling parameters as a metric space, \SVI{}.
 	The base vectors, $\bf{a}$, $\bf{b}$, $\bf{c}$ of the unit cell
@@
 	particular manifold can be described by their ``character'', 
-	a representation of the vectorsns of that type. The table of \citeasnoun{Delaunay1932} included this information 
+	a representation of the vectors of that type. The table of \citeasnoun{Delaunay1932} included this information 
 	encoded in the formalism of Bravais, as a tetrahedron, with
@@
 
-	We will need to create projectors for each of the Bravais
-	lattice types. The projectors will be use to generate
+	Sella creates projectors for each of the Bravais
+	lattice types. The projectors will be used to generate
 	images of probe points in each of the types, and the perps
-	will measure the distance from each manifold.
+	will measure the distance from each manifold. Perps
+	are the matrices that when multiplied times a vector,
+	create a vector that is the component of the input 
+	that is orthogonal to the domain that the projector
+	is associated with.
 
-	We will use Selling reduction; Niggli reduction has a fundamental 
-	unit that is non-convex and some boundaries of the fundamental 
-	unit are partly closed and partly open \cite{andrews2014} leading 
+	We will use Selling reduction; Niggli reduction has a reduced
+	domain that is non-convex and some boundaries of the 
+	reduced domain are partly closed and partly open \cite{andrews2014} leading 
 	to considerable complexity. Because Niggli reduction divides the 
@@
 	the transitions between those are quite complex \cite{andrews2014}. 
-	In space \SVI{} \cite{andrews2019b}, the fundamental 
-	unit is the simple all-negative orthant of a 6-dimensional 
+	In space \SVI{} \cite{andrews2019b}, the reduced domain is the simple, convex all-negative orthant of a 6-dimensional 
 	Cartesian space. Only two kinds of operations in the space 
-	will need to be considered: the 6  transform operations 
-	at the boundaries of the orthant (zeros of an axis) and 
+	will need to be considered: the transform operations 
+	at the boundaries of the reduced orthant (zeros of an axis) and 
 	24 reflections (allowed permutations) \cite{andrews2019}
@@
 	cells) or those generated beyond the boundaries by boundary transforms. 
-	Three forms (M3, M2B, and O3) are reduced dimension boundaries that do
+	Three forms (M3, M2B, and O3) are reduced-dimension boundaries between 2 Delone Types, that do
 	not correspond to unique Bravais types; any lattice that falls in those
-	three is indexed by one of the other types.
+	three is indexed by both of the other types.
 
-	Seven of the 21 Delone types do not have a zero in their \SVI{} 
+	Seven of the 19 non-triclinic, nondegenerate Delone types do not have a zero in their \SVI{} 
 	representation; the presence of zero indicates that a form is 
@@
 	``perps'', the operators that give the normal vector to each 
-	such manifold. For a particular manifold, the projector 
+	such manifold. For a particular manifold, its projector 
 	applied to the probe gives
@@
 	Perps are computed by subtracting the 
-	corresponding projector from the unit matrix. The norm of avector
-	created by multiplying the probe times the projector
+	corresponding projector from the unit matrix. The norm of a vector
+	created by multiplying the probe times a projector
 	 is the distance from that manifold.
@@
 
-	Step 1: Generate any random vector that contains no zero 
+	For each Delone Type, we will create a list of
+	vectors that can be generated for that type.
+
+	\emph{Step 1:}
+
+	Generate any random vector that contains no zero 
 	values and no duplicate values.  For a later operation, 
@@
 
-	Step 2; For each Delone type, multiply its projector times 
+	\emph{Step 2:}
+
+	 For each Delone type, multiply its projector times 
 	the vector from Step 1. Store  the resulting vector
 	from Step 1 in a list 
-	with entries for each Bravais type (see Table \ref{table:DeloneTypes} 
+	with entries for each Delone type (see Table \ref{table:DeloneTypes} 
 	for the projectors).
 
-	Step 3:
+	\emph{Step 3:}
 
 	a) If the vector result of Step 2 contains no zeros, generate the 24 reflections
-	and store in the list by Bravais type.
-
-	\textit{\textbf{LCA--need to normalize the vectors, just like 
-		in lattice matching. ? Can we use the 
-		lattice matching code here? This is very 
-		incomplete. The processing of multiple zeros is MISSING! But the use of DBL\_MIN is there.}}
+	and store in the list by Delone type.
 
@@
 	depend on whatever values are substituted for the zeros. (It does not depend on the zero values because the projectors to that 
-	manifold create zero elements for any vector.) Imangine a plane
+	manifold create zero elements for any vector.) Imagine a plane
 	with axis that are the values of the two elements of the character that are zero. Points in the plane represent unit 
@@
 	have an average value that remains equal to $\bf{s}$ in the lattice character. 
-	That means 3 values to vary (the 2 zeros and the distance fomaverage
+	That means 3 values to vary (the 2 zeros and the distance from average
 	value for the equal scalars), and we can create a sphere around 
@@
 	\textit{\textbf{LCA this section is incomplete. Table needs caption/label.}}
-	Comparison to other distance measure; Data for other measures taken
-	from \citeasnoun{andrews2014}.
 
@@
 	\end{tabular}
-	\caption{Comparison of distance measures}
+	\caption{Comparison to other distance measure; Data for other measures taken from \citeasnoun{andrews2014}.}
 	\label{tab:comparisons}
@@
 
-	\begin{figure}
-		\caption{The O4 item from the Delone Grid with Delone type description}
-		\label{fig:O4_terminology}
-		\includegraphics[width=12cm]{O4_terminology}
-	\end{figure}
-
 	%	\begin{figure}
@@
 	\begin{figure}
+		\caption{The O4 item from the Delone Grid with Delone type description}
+		\label{fig:O4_terminology}
+		\includegraphics[width=12cm]{O4_terminology}
+	\end{figure}
+
+	\begin{figure}
 		\caption{}
@@
 			radius 0.1 is centered at 0,0. The lattices are reduced and the points are
-			plotted at the distance from M4.} The red points indicate the original
-		circle. The blue points are for the reduced cells.
+			plotted at the distance from M4. The red points indicate the original
+		circle. The blue points are for the reduced cells.}
 		\includegraphics[width=12cm]{M4-100-2-zeros}
@@
 	\begin{figure}
-		\caption{Delone type O4: There are two zeros and one pair.
-			A sphere with radius 0.1 was placed around one point. At each
-			sphere point, the lattice was reduced and the point on the sphere
-			was assigned a color (0.01) to blue (0.0075). 100,000 points are plotted.}
+		\caption{Delone type O4 (character \charseq{00r stu}): There are two zeros and one pair.			 
+			100,000 points are plotted; for each
+			 point, the corresponding lattice was reduced; each point
+			is assigned a color (0.01) to blue (0.0075),
+			according to its reduced distance from the origin.}
 		\label{fig:O4}

### 2025-08-12 Meetings/Lombard/Lombard - Copy.tex
Defintions.txt file for website glossary P3 paper-- startup

\documentclass[final]{beamer}
\usepackage[paperwidth=45in, paperheight=40in, margin=1in]{geometry}
\usepackage{graphicx} % For images
\usepackage{multicol} % Multi-column formatting
\usepackage{color} % Text coloring
\usepackage{amsmath, amssymb} % Math symbols
\usepackage{tcolorbox} % For boxed text elements

\usetheme{Berlin} % Beamer theme
\usecolortheme{dolphin} % Color theme

\begin{document}

	\begin{frame}[plain]
		\title{\textbf{Unit Cells in Space or Spaces for Unit Cells}}
		\author{Your Name}
		\date{Your Institution}
		\maketitle
	\end{frame}

	\begin{frame}{Introduction}
		\begin{multicols}{2}
			Unit cells are the fundamental repeating units in crystalline structures. Their spatial arrangement plays a crucial role in determining material properties.
		\end{multicols}
	\end{frame}

	\begin{frame}{Methods}
		\begin{multicols}{2}
			We examine crystallographic reductions, including transformations such as **Niggli reduction** using 3×3 matrices.
		\end{multicols}
	\end{frame}

	\begin{frame}{Results}
		\centering
		\includegraphics[width=0.9\linewidth]{example-figure.png} % Replace with your actual file
	\end{frame}

	\begin{frame}{Conclusion}
		\begin{multicols}{2}
			The unit cell transformation methods explored here provide insights into space group classifications and lattice symmetry.
		\end{multicols}
	\end{frame}

\end{document}

### 2025-08-12 Meetings/Lombard/copilot.tex
Defintions.txt file for website glossary P3 paper-- startup

\documentclass[12pt]{article}
\usepackage[paperwidth=40in, paperheight=40in, margin=1in]{geometry}
\usepackage{tikz}
\usepackage{tcolorbox}
\usepackage{xcolor}
\usepackage{parskip}
\usepackage{lmodern}
\usetikzlibrary{positioning}

% Fonts & layout
\renewcommand{\familydefault}{\sfdefault}

% === Font Size Control Macro ===
\newcommand{\PosterFontSize}{\fontsize{45}{50}\selectfont}

% Box styling
\tcbset{
	colframe=black!30!white,
	colback=gray!5!white,
	boxrule=0.5pt,
	arc=1mm,
	top=4pt, bottom=4pt,
	left=6pt, right=6pt,
	width=0.48\textwidth,
	height=0.20\textheight,
	valign=center,
	fontupper=\PosterFontSize
}

% Space macros
\newcommand{\HVI}{\textbf{H\textsuperscript{6}}}
\newcommand{\GVI}{\textbf{G\textsuperscript{6}}}
\newcommand{\SVI}{\textbf{S\textsuperscript{6}}}
\newcommand{\PVI}{\textbf{P\textsuperscript{3}}}
\newcommand{\CIII}{\textbf{C\textsuperscript{3}}}

% a,b,c as base vectors of unit cells
\newcommand{\va}{\ensuremath{\mathbf{a}}}
\newcommand{\vb}{\ensuremath{\mathbf{b}}}
\newcommand{\vc}{\ensuremath{\mathbf{c}}}
\newcommand{\vd}{\ensuremath{\mathbf{d}}}
\begin{document}

	\begin{center}
		{\fontsize{90}{100}\selectfont\textbf{Unit Cells in Space or Spaces for Unit Cells}}\

		\ [0.5cm]
		{\fontsize{60}{60}\selectfont Lawrence C. Andrews and Herbert J Bernstein — ACA 2025}
	\end{center}

	\vspace{0.8cm}

	% Row 1 — Tree and Abstract
	\noindent
	\parbox[t]{0.48\textwidth}{
		\tcolorbox[title=Crystallographic Space Map]
		\begin{center}
			\scalebox{2.}{
				\begin{tikzpicture}[
					node distance=0.8cm and 1.8cm,
					measurement/.style={text width=3.3cm, align=center, rectangle, draw=black,
						fill=gray!20, rounded corners, font=\small, thick},
					symmetry/.style={text width=3.3cm, align=center, rectangle, draw=black,
						dashed, fill=white, rounded corners, font=\small}
					]
					\node[measurement] (H7) {H\textsuperscript{7} (Ha\"uy)};
					\node[measurement] (G6) [below left=of H7] {G\textsuperscript{6}};
					\node[measurement] (S6) [below=of H7] {S\textsuperscript{6}};
					\node[measurement] (P3) [below right=of H7] {P\textsuperscript{3}};
				%	\node[symmetry] (C3) [below=of S6] {C\textsuperscript{3}};
					\node[symmetry] (DC7) [below=of G6] {DC7unsrt};
					\draw[->, thick] (H7) -- (G6);
					\draw[->, thick] (H7) -- (S6);
					\draw[->, thick] (H7) -- (P3);
				%	\draw[->, thick] (S6) -- (C3);
					\draw[->, thick] (G6) -- (DC7);
				\end{tikzpicture}
			}
		\end{center}
		\endtcolorbox
	}
	\hfill
	\parbox[t]{0.48\textwidth}{
		\tcolorbox[title=Abstract]
		Crystallographers describe unit cells using multiple mathematical spaces, each revealing different aspects of geometry, symmetry, or classification. But they don't think about them as spaces. \\ Beginning with conventional parameters in \HVI{}, we trace transformations through \GVI{} and \SVI{} into polar (\PVI{}), and geometric (DC7unsrt) spaces. Keep in mind
		that another common description is as the 3 base vectors
		of the conventional unit cell, \va, \vb, \vc; this is 
		also a space, a 3-dimensional space of 3-D Euclidiean  vectors.
		\endtcolorbox
	}

	\vspace{1cm}

	% Row 2 — HVI and GVI
	\noindent
	\parbox[t]{0.48\textwidth}{
		\tcolorbox[title=\HVI{} – Conventional Parameter Space]
		\textbf{Conventional Parameter Space (\HVI{})}

		Six coordinates: $a$, $b$, $c$, $\alpha$, $\beta$, $\gamma$. 
		Not ideal for comparing geometry or symmetry directly.
		("H" was chosen to honor Ren\'e Just Ha\"uy, often described as 
		the first modern crystallographer.)
		\endtcolorbox
	}
	\hfill
		\parbox[t]{0.48\textwidth}{
		\tcolorbox[title=\CIII{} – Character Space]
		\textbf{Why do we need spaces for unit cells?}

		\HVI is inconvienent for describing the differences between unit cells/lattices, and there is no obvious method to display cells'
		relationships. Besides, other spaces are already in use; they are
		simply not always described as spaces. Representing cells as points
		in a space allows the simple application of the techniques of
		linear algebra.
		\endtcolorbox
	}

	\vspace{1cm}

	% Row 3 — SVI and PVI
	\noindent
	\parbox[t]{0.48\textwidth}{
		\tcolorbox[title=\SVI{} – Selling/Delone Space]
		\textbf{Delone Space (\SVI{})}

		For a unit cell with base vectors \va, \vb, \vc,
		and -\vd = -\va-\vb-\vc, a vector in \SVI~ is defined as 
		[\ensuremath{\vb \cdot \vc}, \ensuremath{\va \cdot \vc}, \ensuremath{\va \cdot \vb},
		 \ensuremath{\va \cdot \vd}, \ensuremath{\vb \cdot \vd}, \ensuremath{\vc \cdot \vd}]
		. In \SVI, the subspace of the Selling/Delaunay reduced points is a convex region. 
		The Bravais types form linear manifolds. The region of the reduced vectors has only 6 boundaries,	simplifying analyses. \\
		\SVI is actually the space that
		Selling/Delaunay reduction is performed in.
		\endtcolorbox 
	}
	\hfill
	\parbox[t]{0.48\textwidth}{
	\tcolorbox[title=\GVI{} – Niggli Space]
	\textbf{Niggli Space (Metric Tensor Space) (\GVI{})}
	)
		For a unit cell with base vectors \va, \vb, \vc,
		a vector in \GVI~ is defined as 
	[\ensuremath{\va \cdot \va}, \ensuremath{\vb \cdot \vb}, \ensuremath{\vc \cdot \vc},
	\ensuremath{2\vb \cdot \vc}, \ensuremath{2\va \cdot \vc}, \ensuremath{2\va \cdot \vb}]. \GVI is the basis for Niggli reduction. The domain of Niggli-reduced
	cells is non-convex with complex boundary behavior, so distance
	calculations can be done, but they can be complex.
	\endtcolorbox
}

	\vspace{1cm}

	% Row 4 — DC7 and Character Space
	\noindent
	\parbox[t]{0.48\textwidth}{
		\tcolorbox[title=DC7unsrt – Dirichlet 7\, unsorted]
		\textbf{Seven-Dimensional Dirichlet derived (DC7unsrt)}

		Seven scalars derived from Voronoi cells. Focuses on geometric relationships and symmetry inference.
		Bernstein, Andrews, and Xerri, Acta Cryst A79 (2005). 369-380. Experience shows
		that distance calculations in this space can be fast.
		\endtcolorbox

	}
	\hfill
	\parbox[t]{0.48\textwidth}{
	\tcolorbox[title=\PVI{} – Polar Space]
	\textbf{Polar Space (\PVI{})} \\
	\PVI is defined as the space of 3 polar coordinates, [\ensuremath{
	(|\va|,\alpha), 
	(|\vb|,\beta), 
	(|\vc\,\gamma)}]. \\
	\PVI has a conceptual advantage that is it more directly related to the 
	unit cell parameters that are commonly used in crystallography 
	(\HVI{}). Polar coordinates are often converted to x,y, Euclidean 
	coordinates for simple calculations with commensurate measures, so distance calculations make sense.
	 Because they are in Angstrom units, they are more
	familar. 
	\endtcolorbox
}

\end{document}

### 2025-08-12 Meetings/Lombard/spaces.tex
Defintions.txt file for website glossary P3 paper-- startup

<svg width="1200" height="800" xmlns="http://www.w3.org/2000/svg">
<defs>
<style>
.title { font-family: Arial, sans-serif; font-size: 24px; font-weight: bold; }
.space-label { font-family: Arial, sans-serif; font-size: 16px; font-weight: bold; }
.description { font-family: Arial, sans-serif; font-size: 12px; }
.year { font-family: Arial, sans-serif; font-size: 10px; font-style: italic; }
.author { font-family: Arial, sans-serif; font-size: 10px; fill: #666; }
.branch { stroke: #333; stroke-width: 2; fill: none; }
.node { fill: #4a90e2; stroke: #333; stroke-width: 2; }
.h6-node { fill: #ff6b6b; }
.g6-node { fill: #4ecdc4; }
.s6-node { fill: #45b7d1; }
.modern-node { fill: #96ceb4; }
.experimental-node { fill: #feca57; }
</style>
</defs>

<!-- Title -->
<text x="600" y="30" text-anchor="middle" class="title">Evolution of Crystallographic Lattice Representation Spaces</text>

<!-- Timeline axis -->
<line x1="50" y1="750" x2="1150" y2="750" stroke="#ccc" stroke-width="1"/>
<text x="50" y="770" class="year">1850</text>
<text x="300" y="770" class="year">1928</text>
<text x="550" y="770" class="year">1980</text>
<text x="800" y="770" class="year">2000</text>
<text x="1050" y="770" class="year">2020</text>

<!-- H6 - Original 6-parameter space -->
<g>
<circle cx="100" cy="600" r="25" class="node h6-node"/>
<text x="100" y="605" text-anchor="middle" class="space-label">H6</text>
<text x="100" y="540" text-anchor="middle" class="description">Original 6-parameter space</text>
<text x="100" y="555" text-anchor="middle" class="description">(a, b, c, α, β, γ)</text>
<text x="100" y="520" text-anchor="middle" class="year">~1850</text>
<text x="100" y="505" text-anchor="middle" class="author">Bravais, crystallographers</text>
</g>

<!-- Branch from H6 to G6 -->
<path d="M 125 600 Q 200 600 250 450" class="branch"/>

<!-- G6 - Niggli metric tensor space -->
<g>
<circle cx="300" cy="400" r="25" class="node g6-node"/>
<text x="300" y="405" text-anchor="middle" class="space-label">G6</text>
<text x="300" y="340" text-anchor="middle" class="description">Niggli metric tensor space</text>
<text x="300" y="355" text-anchor="middle" class="description">G = [a·a, b·b, c·c, 2b·c, 2a·c, 2a·b]</text>
<text x="300" y="320" text-anchor="middle" class="year">1928</text>
<text x="300" y="305" text-anchor="middle" class="author">Niggli</text>
</g>

<!-- Branch from H6 to S6 -->
<path d="M 125 600 Q 200 600 250 500" class="branch"/>

<!-- S6 - Selling/Delone space -->
<g>
<circle cx="300" cy="500" r="25" class="node s6-node"/>
<text x="300" y="505" text-anchor="middle" class="space-label">S6</text>
<text x="300" y="440" text-anchor="middle" class="description">Selling (Delone) scalars</text>
<text x="300" y="455" text-anchor="middle" class="description">s = [b·c, a·c, a·b, a·d, b·d, c·d]</text>
<text x="300" y="420" text-anchor="middle" class="year">1874/1932</text>
<text x="300" y="405" text-anchor="middle" class="author">Selling/Delaunay</text>
</g>

<!-- Branch from G6 to Andrews & Bernstein work -->
<path d="M 325 400 Q 450 400 500 300" class="branch"/>

<!-- Andrews & Bernstein G6 implementation -->
<g>
<circle cx="550" cy="250" r="20" class="node modern-node"/>
<text x="550" y="255" text-anchor="middle" class="space-label">G6*</text>
<text x="550" y="195" text-anchor="middle" class="description">G6 with boundary transforms</text>
<text x="550" y="210" text-anchor="middle" class="description">Perturbation-stable</text>
<text x="550" y="175" text-anchor="middle" class="year">1988</text>
<text x="550" y="160" text-anchor="middle" class="author">Andrews & Bernstein</text>
</g>

<!-- Branch from S6 to modern S6 work -->
<path d="M 325 500 Q 450 500 500 400" class="branch"/>

<!-- Modern S6 implementation -->
<g>
<circle cx="550" cy="350" r="20" class="node modern-node"/>
<text x="550" y="355" text-anchor="middle" class="space-label">S6*</text>
<text x="550" y="295" text-anchor="middle" class="description">S6 with projectors (SELLA)</text>
<text x="550" y="310" text-anchor="middle" class="description">Complete manifold coverage</text>
<text x="550" y="275" text-anchor="middle" class="year">2019</text>
<text x="550" y="260" text-anchor="middle" class="author">Andrews, Bernstein, Sauter</text>
</g>

<!-- Branch to V7 -->
<path d="M 325 400 Q 400 350 450 300" class="branch"/>

<!-- V7 space -->
<g>
<circle cx="450" cy="300" r="18" class="node experimental-node"/>
<text x="450" y="305" text-anchor="middle" class="space-label">V7</text>
<text x="450" y="250" text-anchor="middle" class="description">7-parameter space</text>
<text x="450" y="265" text-anchor="middle" class="description">(a,b,c,a*,b*,c*,V)</text>
<text x="450" y="230" text-anchor="middle" class="year">1980</text>
<text x="450" y="215" text-anchor="middle" class="author">Andrews et al.</text>
</g>

<!-- Branch to modern developments -->
<path d="M 570 350 Q 650 350 700 200" class="branch"/>
<path d="M 570 250 Q 650 250 700 150" class="branch"/>

<!-- P3 space -->
<g>
<circle cx="750" cy="200" r="18" class="node modern-node"/>
<text x="750" y="205" text-anchor="middle" class="space-label">P3</text>
<text x="750" y="150" text-anchor="middle" class="description">3-parameter space</text>
<text x="750" y="165" text-anchor="middle" class="description">Reduced representation</text>
<text x="750" y="130" text-anchor="middle" class="year">~2000s</text>
</g>

<!-- DC7unsrt -->
<g>
<circle cx="750" cy="100" r="18" class="node experimental-node"/>
<text x="750" y="105" text-anchor="middle" class="space-label">DC7unsrt</text>
<text x="750" y="50" text-anchor="middle" class="description">7-dimensional Dirichlet</text>
<text x="750" y="65" text-anchor="middle" class="description">Unsorted representation</text>
<text x="750" y="30" text-anchor="middle" class="year">2023</text>
<text x="750" y="15" text-anchor="middle" class="author">Bernstein et al.</text>
</g>

<!-- Branch to C3 and other modern spaces -->
<path d="M 570 350 Q 700 400 800 350" class="branch"/>

<!-- C3 space -->
<g>
<circle cx="850" cy="350" r="18" class="node experimental-node"/>
<text x="850" y="355" text-anchor="middle" class="space-label">C3</text>
<text x="850" y="300" text-anchor="middle" class="description">3-dimensional complex</text>
<text x="850" y="315" text-anchor="middle" class="description">representation</text>
<text x="850" y="280" text-anchor="middle" class="year">~2019</text>
<text x="850" y="265" text-anchor="middle" class="author">Andrews et al.</text>
</g>

<!-- Branch to CS6Dist -->
<path d="M 570 350 Q 700 450 800 450" class="branch"/>

<!-- CS6Dist -->
<g>
<circle cx="850" cy="450" r="18" class="node experimental-node"/>
<text x="850" y="455" text-anchor="middle" class="space-label">CS6Dist</text>
<text x="850" y="400" text-anchor="middle" class="description">S6 distance metric</text>
<text x="850" y="415" text-anchor="middle" class="description">Fast computation</text>
<text x="850" y="380" text-anchor="middle" class="year">~2019</text>
</g>

<!-- Branch to future developments -->
<path d="M 870 350 Q 950 350 1000 300" class="branch"/>
<path d="M 870 450 Q 950 450 1000 400" class="branch"/>

<!-- Future/theoretical spaces -->
<g>
<circle cx="1050" cy="350" r="15" class="node" style="fill: #ddd; stroke-dasharray: 5,5;"/>
<text x="1050" y="355" text-anchor="middle" class="space-label">?</text>
<text x="1050" y="305" text-anchor="middle" class="description">Future spaces</text>
<text x="1050" y="320" text-anchor="middle" class="description">Machine learning?</text>
<text x="1050" y="285" text-anchor="middle" class="year">Future</text>
</g>

<!-- Legend -->
<g transform="translate(50, 50)">
<text x="0" y="0" class="space-label">Legend:</text>
<circle cx="15" cy="20" r="8" class="node h6-node"/>
<text x="30" y="25" class="description">Original spaces</text>
<circle cx="15" cy="40" r="8" class="node g6-node"/>
<text x="30" y="45" class="description">Classical reductions</text>
<circle cx="15" cy="60" r="8" class="node modern-node"/>
<text x="30" y="65" class="description">Modern implementations</text>
<circle cx="15" cy="80" r="8" class="node experimental-node"/>
<text x="30" y="85" class="description">Recent developments</text>
</g>

<!-- Key relationships -->
<g transform="translate(900, 550)">
<text x="0" y="0" class="space-label">Key Transformations:</text>
<text x="0" y="20" class="description">H6 → G6: Metric tensor</text>
<text x="0" y="35" class="description">H6 → S6: Selling scalars</text>
<text x="0" y="50" class="description">G6 → S6: Linear transformation</text>
<text x="0" y="65" class="description">S6 → C3: Complex representation</text>
<text x="0" y="80" class="description">All → DC7unsrt: Dirichlet embedding</text>
</g>
</svg>

### 2025-08-12 P3/P3.tex
Defintions.txt file for website glossary P3 paper-- startup

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
%\usepackage{bm}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
%\usepackage{calligra}
\usepackage{array}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
%\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\usepackage{color}
\usepackage{ulem}
\usepackage{url}
\usepackage{yfonts}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}

\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning}
\usepackage[most]{tcolorbox}

% Space macros
\usepackage{amsmath, amssymb}  % If not already included

% Crystallographic space macros
\newcommand{\HVI}{\ensuremath{\mathbf{H}^{6}}}
\newcommand{\GVI}{\ensuremath{\mathbf{G}^{6}}}
\newcommand{\SVI}{\ensuremath{\mathbf{S}^{6}}}
\newcommand{\PIII}{\ensuremath{\mathbf{P}^{3}}}
\newcommand{\FIII}{\ensuremath{\mathbf{F}^{3}}}
\newcommand{\CIII}{\ensuremath{\mathbf{C}^{3}}}
\newcommand{\RIII}{\ensuremath{\mathbb{R}^{3}}}

% a,b,c as base vectors of unit cells
\newcommand{\va}{\ensuremath{\mathbf{a}}}
\newcommand{\vb}{\ensuremath{\mathbf{b}}}
\newcommand{\vc}{\ensuremath{\mathbf{c}}}
\newcommand{\vd}{\ensuremath{\mathbf{d}}}

\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\Imaginary}[0]{\mathcal{I}}
\newcommand{\Real}[0]{\mathcal{R}}
\newcommand{\Exchange}{\ensuremath{\mathcal{X}}}

\newcommand{\nounderline}[3]{\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineab}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineac}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
\newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}

\newcommand{\scalar}[1]{\ensuremath{#1}}
\newcommand{\scalarsub}[2]{\ensuremath{#1_{#2}}}

\usepackage{makecell}
\renewcommand\cellalign{l}
\renewcommand\cellgape{\Gape[4pt]}

%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
\journalcode{A}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation
\makeatletter
\font\dummyft@=dummy \relax
\makeatother

\begin{document}                  % DO NOT DELETE THIS LINE

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).
	\begin{center}
		{\LARGE \emph{\today}} \\
	\end{center}

	\title{\PIII, a simple metric for \HVI}
	\shorttitle{properties of \PIII}

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
	\author[b]{Herbert J.}{Bernstein}

	\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
	\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}

	% Use \shortauthor to indicate an abbreviated author list for use in
	% running heads (you will need to uncomment it).

	\shortauthor{Andrews and Bernstein}

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	\keyword{lattice}
	\keyword{unit cell}
	\keyword{polar}
	\keyword{\PIII}

	\maketitle                        % DO NOT DELETE THIS LINE

	\begin{synopsis}

	\end{synopsis}

	\newcommand{\si}{\ensuremath{s_1}}
	\newcommand{\sii}{\ensuremath{s_2}}
	\newcommand{\siii}{\ensuremath{s_3}}
	\newcommand{\siv}{\ensuremath{s_4}}
	\newcommand{\sv}{\ensuremath{s_5}}
	\newcommand{\svi}{\ensuremath{s_6}}

	% Scalar vectors
	\newcommand{\Svec}{\ensuremath{\{ \si, \sii, \siii, \siv, \sv, \svi \}}}
	\newcommand{\SvecA}{\ensuremath{\{ -\si, -\si+\sii, \si+\siii, \si+\sv, \si+\siv, \si+\svi \}}}

	% Operators / maps
	\newcommand{\OPES}{\ensuremath{E^3\!\to\!S^6}}
	\newcommand{\OPESS}{

		\[ E^3\!\to\!S^6 \]

	}
	\newcommand{\MSVI}{\ensuremath{M_{S^6}}}
	\newcommand{\MEIII}{\ensuremath{M_{E^3}}}

	% Symbolic markers
	\newcommand{\Plus}{\ensuremath{\mathcal{P}}}
	\newcommand{\Minus}{\ensuremath{\mathcal{M}}}

	\begin{abstract}

		\emph{Note:}  In his later publications, Boris Delaunay used the Russian version of his surname, Delone.

	\end{abstract}
	% Appendices appear after the main body of the text. They are prefixed by
	% a single \appendix declaration, and are then structured just like the
	% body text.

	\section{Introduction}

	In crystallography, cell geometry is conventionally encoded in a six-dimensional parameter space 
	\ensuremath{\HVI{}=(a,b,c, \alpha, \beta, \gamma).}\footnote{``H'' was chosen to
		honor early French crystallographer Ren\'e Just Ha\"uy.} But \HVI{} is not a metric space
	(where distances can be simply defined), and \HVI{} is not a vector space (where 
	objects can be added and subtracted). In the same sense, symmetry operations
	have no simple representations in \HVI{}.

	For the above reasons, other spaces are often used to describe lattices, often for specialized purposes; see Table  \ref{tbl_spaces}. Figure \ref{fig_spaces} shows relationships between
	several of these spaces. Although some of these spaces, such as \HVI{} and	\FIII{} are in almost daily use, they are not often referred to 
	as ``spaces''. In the case of \FIII{}, symmetry elements are linear
	operations in this space, and fractional coordinates of atomic positions
	are vectors expressed in this space.

\begin{table}
	\label{tbl_spaces}
	\centering
	\caption{Crystallographic spaces with their scalar definitions and analytical purposes}
	\renewcommand{\arraystretch}{1.3}
	\begin{tabular}{|c|l|p{7cm}|}
		\hline
		\textbf{Space} & \textbf{Scalars / Coordinates} & \textbf{Purpose} \\

\hline
$\HVI$ &
\parbox[t]{5.5cm}{
	$a,\, b,\, c \in \mathbb{R}_{>0}$ — edge lengths\\
	$\alpha,\, \beta,\, \gamma \in (0^\circ, 180^\circ)$ — opposing angles\\
	Additionally, the sum of the angles must be less than 
	$360^\circ$ and must obey the triangle inequality
} &
Composite parameter space described as 
$\mathbb{R}_{>0}^3 \times (0^\circ, 180^\circ)^3$. 
The length dimensions are unconstrained, while the angles are 
geometrically restricted. Useful for conventional lattice 
descriptions, but limited in symmetry representation and 
metric-based classification. \\

\hline		
$\mathbf{F^3}$ &
$\vec{a},\, \vec{b},\, \vec{c} \in \mathbb{R}^3$ &
Base vector space, a frame space (or frame bundle) in \RIII. The coordinates are the 3-space 
base vectors of a unit cell. Symmetry 
operations are enumerated using this space. \\
\hline

		$\mathbf{P^3}$ &
		\parbox[t]{5.5cm}{
			$(|\vec{a}|,\alpha),\ (|\vec{b}|,\beta),\ (|\vec{c}|,\gamma)$\\
			$\Rightarrow$\,\\ 
			\hspace*{0.5cm}$(|\vec{a}|\cos\alpha,\, |\vec{a}|\sin\alpha$\\
		\hspace*{0.5cm}	$|\vec{b}|\cos\beta,\, |\vec{b}|\sin\beta$\\
		\hspace*{0.5cm}	$|\vec{c}|\cos\gamma,\, |\vec{c}|\sin\gamma)$ \\
		also known as: \\
		\hspace*{0.5cm}	\ensuremath{(p1,p2,p3)}. 
		} &
		Polar coordinate space defined by the map $\FIII \rightarrow \PIII$. Represents each base vector by magnitude and opposing angle, decomposed into directional components. Compact and smooth alternative to $\HVI$. \PIII{} provides a simple
		and intuitive space for representing and comparing
		 large numbers of unit cells. (This work) \\
		\hline

$\mathbf{S^6}$ &
\parbox[t]{5.5cm}{
	$s_1 = \vec{b} \cdot \vec{c},\, s_2 = \vec{a} \cdot \vec{c},\, s_3 = \vec{a} \cdot \vec{b}$\\
	$s_4 = \vec{a} \cdot \vec{d},\, s_5 = \vec{b} \cdot \vec{d},\, s_6 = \vec{c} \cdot \vec{d}$\\ \\
	\quad where $\vec{d} = -\vec{a} - \vec{b} - \vec{c}$
} &
A space used for Selling/Delone reduction. The scalars are derived from dot products between lattice vectors. Used for lattice classification described by Delaunay and for distance computation \cite{Andrews2019b}. 24 Bravais types were defined by citeasnoun{Delone1975} in \SVI{}, but 
described using pictures showing the special relationships. \citeasnoun{andrews2023measuring} revised the table of Delone to
include the \SVI vector definitions of the Bravais types.\\

		\hline
		$\mathbf{G^6}$ &
		\parbox[t]{5.5cm}{
			$g_1 = \vec{a} \cdot \vec{a},\, g_2 = \vec{b} \cdot \vec{b},\, g_3 = \vec{c} \cdot \vec{c}$\\
			$g_4 = 2\vec{b} \cdot \vec{c},\, g_5 =2 \vec{c} \cdot \vec{a},\, g_6 = 2\vec{a} \cdot \vec{b}$
		} &
		A space based on the metric tensor (and thus dot products 
		of the basis vectors of the unit cell) and used for Niggli reduction and lattice distance computation. \cite{Andrews2014}
		The International Tables for Crystallography describe the 44
		Bravais types that result from Niggli reduction using
		algebraic expression in \GVI{} (but stated in terms of scalars
		rather than manifolds in \GVI{}) \cite{ITC_VolumeA_2016}. \\
		\hline

\CIII &
\parbox[t]{5.5cm}{
	from the \SVI{} scalars, $s1,...,s6$\\
	$c1=(s1,s4), c2=(s2,s5), c3=(s3,s6)$\\
	where (?,?) denotes a complex number
} &
\CIII{} was created to associate related scalars after the
indication of \citeasnoun{Delone1975} concerning the ``opposite''
scalars in \SVI{}. Symmetry and Selling/Delaunay reduction have
simpler forms in \CIII{} than in \SVI{}. \cite{andrews2023complex}\\
\hline

		$\mathbf{DC7u}$ &
		\parbox[t]{5.5cm}{
			$d_1,\, d_2,\, d_3$ — cell edge lengths\\
			$d_4,\, d_5,\, d_6$ — shorter face diagonals' lengths\\
			$u$ — shortest body diagonal
		} &
		Unsorted Dirichlet Cell space derived from Niggli-reduced cells. Defined by the map from lattice geometry to a 7-dimensional vector of distances. Fully invertible and smooth; used for rapid lattice distance computation. \cite{bernstein2023invertible}\\
		\hline

	\end{tabular}
\end{table}

\begin{figure}
	\centering
\begin{tikzpicture}[
	node distance=1.2cm and 2.0cm,
	coordspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
		fill=gray!20, rounded corners, font=\small, thick},
	metricspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
		fill=blue!10, rounded corners, font=\small, thick},
	derivedspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
		fill=green!10, rounded corners, font=\small, thick},
	arrowlabel/.style={font=\scriptsize, midway, fill=white, inner sep=1pt}
	]

	% Row 1: Fundamental space
	\node[coordspace] (F3) {\FIII{} \\ Base vectors};

	% Row 2: Scalar & parameter spaces
	\node[metricspace] (G6) [below left=of F3] {\GVI{} \\ Metric tensor};
	\node[metricspace] (S6) [below=of F3] {\SVI{} \\ Selling/Delone};
	\node[coordspace] (H6) [below right=of F3] {\HVI \\ Cell params};

	% Row 3: Derived/reduced spaces
	\node[derivedspace] (DC7u) [below=of G6] {DC7u \\ Dirichlet coords};
	\node[derivedspace] (C3) [below=of S6] {\CIII{} \\ Reduced coords};
	\node[derivedspace] (P3) [below=of H6] {\PIII{} \\ Polar coords};

	% Arrows from F3 to row 2
	\draw[->, thick] (F3) -- (G6);
	\draw[->, thick] (F3) -- (S6);
	\draw[->, thick] (F3) -- (H6);

	% Arrows from row 2 to row 3
	\draw[->, thick] (G6) -- (DC7u);
	\draw[->, thick] (S6) -- (C3);
	\draw[->, thick] (H6) -- (P3);

\end{tikzpicture}
\label{fig_spaces}
\caption{Derived relationships among crystallographic spaces. 
	Base vector space \FIII{} gives rise to three scalar spaces:
	metric tensor space \GVI{}, Selling/Delone space \SVI{}, and conventional cell parameter space \HVI{}.
	Each transforms into a reduced or classification space: 
	$\mathbf{DC7u}$ from \GVI{}, \CIII{} from \SVI{}, and \PIII{} from $\mathbf{H^6}$.}
\end{figure}

\section{	\PIII{} Is introduced.}

\citeasnoun{Delone1975} repeatedly pointed out the relationship of the
``opposite'' scalars in the tetrahedron representations of the Selling
scalars. Examining those relationships show that the opposite terms 
are pairs where a term involving a is opposite a term involving $\alpha$, etc.

Extending the observation of \citeasnoun{Delone1975}, the idea occurs
to make a triple of variables of those pairs. Since each pair has
one length and one angle, we considered each pair as the definition of
a polar coordinate space. The result is \PIII{}, a space of 3 2-dimensional polar coordinates. See Table \ref{tbl_spaces} for the
formal description.  Two uses immediately are obvious.

The first uses the 2-dimensional character of the coordinates of \PIII{}. They
allow simple to plot representations of collections of unit cells. 
See Figure \ref{PlotPolar_raw}, Figure \ref{PlotPolar_Niggli}, and Figure \ref{PlotPolar_Delone}.

\begin{figure}
\includegraphics[width=\columnwidth]{PlotPolar_raw}
\label{PlotPolar_raw}
	\caption{100 random primitive triclinic unit cells}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{PlotPolar_Niggli}
	\label{PlotPolar_Niggli}
	\caption{The same100 random primitive triclinic unit cells, Niggli-reduced}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{PlotPolar_Delone}
	\label{PlotPolar_Delone}
	\caption{The same 100 random primitive triclinic unit cells, Selling-reduced}
\end{figure}

%\[
%(|\vec{x}|\cos\theta,\, |\vec{x}|\sin\theta) \in \mathbb{R}^2
%\]

%which preserves the Euclidean magnitude and implicitly encodes angular orientation. While the transformation is nonlinear in angle, the resulting structure allows straightforward computation of relative orientation, cluster proximity, and shape similarity through standard Euclidean metrics.

The second use is allowing the calculation of Euclidean distances. Because \PIII{} can be represented as 3 xy planes, the Euclidean difference between
any two unit cells described as \PIII{} can be easily computed. Thus use
contrasts with \HVI{}, where incommensurate lengths and angles do not
lend themselves to a distance calculation.

\section{Relationship to Other Spaces}

While derived from the same foundational cell parameters as $\mathbf{H^6}$, the space $\mathbf{P^3}$ sidesteps issues of angular nonlinearity  that arise in conventional cell representations. Compared with $\mathbf{C^3}$, which seeks symmetry-reduced coordinates derived from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic. It provides a useful intermediary: more interpretable than metric tensors or other spaces, less abstract than \SVI{} or \GVI{}.

	\section{Notation}

	\section{Matrices of boundary transformations}

	\subsection{Derivation}

	\subsection{Example P++ implementation}

	\subsection{Example of use}

	\section{Graphical display of projections}

	\section{Scalar Reduction and Canonical Presentation in Projected Vector Triplets}

	\subsection{1. Standard Presentation of Projected Vectors}

	Let a set of three projected vectors be given:

	\[
	(p_1, p_2, p_3), \quad p_i \in \mathbb{R}^2
	\]

	Each \( p_i \) is constructed from a lattice vector via magnitude and opposing angle:

	\[
	p_i = |\vec{v}_i| (\cos \theta_i, \sin \theta_i)
	\]

	The \textbf{Standard Presentation} algorithm assigns a canonical ordering and orientation to the triplet:

	\begin{enumerate}
		\item \textbf{Magnitude Sort:} Reorder vectors so that

		\[
		|p_1| \leq |p_2| \leq |p_3|
		\]

		using Euclidean norms \( |p_i| = \sqrt{x_i^2 + y_i^2} \). Ties are preserved up to a numerical tolerance \( \delta \).

\item \textbf{Directional Coherence:} If the \( x \)-components of the triplet are not all of consistent sign, flip the necessary vectors by negating only the \( x_i \) component:

\[
p_i \to (-x_i,\ y_i)
\]

This maps quadrant I vectors into quadrant II, preserving their placement in the upper half-plane and maintaining angular conventions typical in crystallographic analysis.

\[
p_i \to (-x_i,\ y_i)
\]

to align the triplet into a directionally consistent state. If all \( x_i < 0 \), the configuration is already coherent and requires no change.

		\[
		p_i \to (-x_i, y_i)
		\]

		for any two vectors, preserving chirality.

		\item \textbf{Return Ordered Triplet:} Output the permuted and flipped triplet with consistent magnitude order and cosine directionality.
	\end{enumerate}

	\vspace{1em}

	\subsection{2. Scalar Projection Interaction and Cost Function}

	For any pair of projected vectors \( p_i, p_j \in \mathbb{R}^2 \), define the scalar projection interaction (or directional affinity):

	\[
	\mu_{ij} = p_i \cdot p_j = x_i x_j + y_i y_j
	\]

	The \textbf{scalar coupling cost} is then defined over the triplet as:

	\[
	C' = \frac{|\mu_{12}| + |\mu_{13}| + |\mu_{23}|}{p_1 \cdot p_1 + p_2 \cdot p_2 + p_3 \cdot p_3}
	\]

	This cost measures the normalized sum of pairwise directional alignment. Lower values of \( C' \) indicate that the projected vectors are less directionally coupled — i.e., more orthogonal or dispersed.

	\vspace{1em}

	\subsection{3. Scalar Reduction Algorithm}

	The goal of scalar reduction is to transform the triplet \( (p_1, p_2, p_3) \) into an algebraically simpler form — one with lower scalar coupling cost \( C' \) — while preserving lattice identity (e.g., volume and reconstructability).

	The reduction proceeds iteratively:

	\begin{enumerate}
		\item \textbf{Initialization:}
		Apply standard presentation. Compute initial cost \( C'_0 \) and physical volume \( V_0 > 0 \).

		\item \textbf{Pairwise Projection Subtractions:}
		For each unordered pair \( (p_i, p_j),\ i \neq j \), compute the scalar coefficient:

		\[
		\lambda = \frac{p_i \cdot p_j}{p_j \cdot p_j}
		\]

		Then define the candidate vector:

		\[
		p_i' = p_i - \lambda p_j
		\]

		\item \textbf{Evaluate Candidate:}
		Apply standard presentation to updated triplet. Recompute cost \( C'_{\text{new}} \) and volume \( V_{\text{new}} \).

		\item \textbf{Acceptance Criteria:}
		Accept the transformation if:

		\[
		C'_{\text{new}} + \delta < C' \quad \text{and} \quad V_{\text{new}} > 0
		\]

		\item \textbf{Repeat:}
		Continue until no further accepted transformations lower the cost.

		\item \textbf{Output:}
		Return the reduced triplet \( (p_1, p_2, p_3)_{\text{red}} \) with minimal scalar cost.
	\end{enumerate}

	\vspace{1em}

	\subsection{4. Interpretation of Cost and Reduction}

	The term "reduction" refers to minimizing the scalar coupling among vectors — reducing directional alignment that obscures presentation clarity. The cost \( C' \) is a rigorously defined metric for this alignment. Transformations reduce \( C' \) while preserving:
	\begin{itemize}
		\item The vector magnitudes and their reconstructable relationships
		\item The unit cell's volume and crystallographic identity
		\item The triplet’s algebraic validity in scalar projection space
	\end{itemize}

	Reduction leads to a cleaner, more canonical representation of the original structure — just as Niggli reduction provides a minimal cell in metric tensor space, scalar reduction provides a minimal configuration in projected vector space.

	\section{Summary}

	\section{Availability of code} CmdToP3 and PlotPolar are available in github.com, in
	\url{https://github.com/duck10/LatticeRepLib.git}.

	%\appendix

	%\section{blah blah blah -- Supplementary Material}
	\ack{{\bf Acknowledgements}}

	Careful copy-editing and corrections by Frances C. Bernstein are 
	gratefully acknowledged.
	%	Our thanks to Jean Jakoncic and Alexei Soares for 
	%	helpful conversations and access to data and facilities at 
	%	Brookhaven National Laboratory.
	%	
	\ack{{\bf Funding information}}

	Funding for this research was provided in part by:  
	US Department of Energy Offices of Biological and 
	Environmental Research and of Basic Energy Sciences 
	(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
	U.S. National Institutes of Health (grant No. P41RR012408; 
	grant No. P41GM103473; grant No. P41GM111244; 
	grant No. R01GM117126,
	grant No. 1R21GM129570); Dectris, Ltd.

	\bibliography{Reduced}

	\bibliographystyle{iucr}

	%-------------------------------------------------------------------------
	% TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
	%-------------------------------------------------------------------------

	% Simple tables should use the tabular environment according to this
	% model

	% Postscript figures can be included with multiple figure blocks

	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>plotc3
	%; Graphical output SVG file =PLT__2023-03-07.13_43_35.svg
	%
	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmdniggli | plotc3
	%; Graphical output SVG file =PLT__2023-03-07.13_44_06.svg
	%
	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmddelone | plotc3
	%; Graphical output SVG file =PLT__2023-03-08.07_11_03.svg
	%
	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmdniggli | cmdperturb 5 20 | plotc3
	%; Graphical output SVG file =PLT__2023-03-08.09_00_13.svg
	%
	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmdniggli | cmdperturb 5 100 | plotc3
	%; Graphical output SVG file =PLT__2023-03-08.09_00_21.svg

\end{document}                    % DO NOT DELETE THIS LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-08-12 P3/Visualizations.tex
Defintions.txt file for website glossary P3 paper-- startup

\documentclass{article}
\usepackage{amsmath, amssymb}
\usepackage{tikz}
\usepackage{geometry}
\usepackage{listings}
\usepackage{xcolor}
\geometry{margin=1in}

\usepackage{pgfplots}
\pgfplotsset{compat=1.18}

\title{Scalar Descent and Fracture Analysis in P3 Space}
\author{Lawrence C. Andrews}
\date{\today}

\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{backcolor}{rgb}{0.95,0.95,0.95}

\lstset{
	backgroundcolor=\color{backcolor},
	basicstyle=\ttfamily\small,
	keywordstyle=\color{blue},
	commentstyle=\color{codegray},
	numbers=left,
	numberstyle=\tiny\color{codegray},
	stepnumber=1,
	numbersep=5pt,
	breaklines=true,
	frame=single,
	captionpos=b
}

\begin{document}

	\maketitle

	\section{Scalar Descent Path with Fracture Persistence}

	The following table represents a descent path in P3 space, with fracture annotations and persistence values:

	\begin{center}
		\begin{tabular}{|c|c|c|c|c|c|c|}
			\hline
			Step & P3 Input & Reduction Outcome & Fracture ID & Birth T & Death T & Persistence \\
			\hline
			0 & $P_0$ & Canonical & — & — & — & — \\
			1 & $P_0 + \delta_1$ & Slight misalignment & $F_1$ & 0.12 & 0.46 & 0.34 \\
			2 & $P_0 + \delta_2$ & Flip across fold B & $F_2$ & 0.28 & 0.36 & 0.08 \\
			3 & $P_0 + \delta_3$ & Stuck in corner & $F_3$ & 0.05 & 0.67 & 0.62 \\
			\hline
		\end{tabular}
	\end{center}

	\section{Simplicial Bin Network with Fracture Connections}

	Each vertex represents a scalar bin. Edges represent adjacency relations via reduction or fracture events. Edge labels denote fracture identifiers.

	\begin{center}
		\begin{tikzpicture}[scale=1.3]
			\node (A) at (0,0) [circle,draw] {A};
			\node (B) at (2,1) [circle,draw] {B};
			\node (C) at (1,-1.5) [circle,draw] {C};
			\node (D) at (3,-0.5) [circle,draw] {D};

			\draw[thick] (A) -- (B) node[midway,above] {\small $F_1$};
			\draw[thick] (A) -- (C) node[midway,left]  {\small $F_3$};
			\draw[thick] (B) -- (D) node[midway,right] {\small $F_2$};
		\end{tikzpicture}
	\end{center}

	\section{Bin Merging Protocol Based on Stability}

	This code snippet defines merge conditions based on fracture persistence and signature similarity.

	\begin{lstlisting}[language=C++,caption={Bin Merge Criteria Based on Scalar Stability}]
		bool shouldMerge(const ScalarBin& bin1, const ScalarBin& bin2, double persistence) {
			return (persistence < 0.1) &&
			(signatureDistance(bin1.signature, bin2.signature) < 0.05) &&
			(isPermutationInvariant(bin1, bin2));
		}
	\end{lstlisting}

	This helps define bin boundaries that reflect stable scalar neighborhoods in fold space.

		\begin{tikzpicture}
			\begin{axis}[
				ybar,
				bar width=20pt,
				ylabel={Sample Count},
				xtick=data,
				symbolic x coords={Chaotic, Severe},
				xticklabel style={rotate=0},
				ymin=0, ymax=200,
				nodes near coords,
				enlarge x limits=0.2
				]
				\addplot+[fill=red!60] coordinates {(Chaotic,56) (Severe,144)};
			\end{axis}
		\end{tikzpicture}

	\end{document}

\end{document}

### 2025-08-27 3D_Niggli_Matrices/3D_Niggli_Matrices.tex
CommandUsage --  add MATRIXORDER to the help. P3, 3D_Niggli_Matrices, and LatticeMatching_2025 are paper in progress.

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------
% lca IUCr id IUCr6401
% HJB IUCr id IUCr6484
% NKS IUCr ID: IUCr7572
% lca ORCID  0000-0002-4451-1641
% HJB ORCID 0000-0002-0517-8532
% NKS ORCID 0000-0003-2786-6552
%xxxxxx just for testing github % 2
%xxxxxx just for testing github % 3

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
%\usepackage{bm}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{makecell}
\usepackage{minibox}
%\usepackage{calligra}
\usepackage{array}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
%\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\usepackage{color}
%\usepackage{ulem}
\usepackage{url}
\usepackage{yfonts}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}
\usepackage[normalem]{ulem}
% Redefine \underline to use \textit
\renewcommand{\underline}[1]{\textit{#1}}
\newcommand{\SVI}[0]{$\bf{S^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\CIII}[0]{$\bf{C^{3}}$}
\newcommand{\DVII}[0]{$\bf{D^{7}}$}
\newcommand{\VVII}[0]{$\bf{V^{7}}$}

\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\Imaginary}[0]{\mathcal{I}}
\newcommand{\Real}[0]{\mathcal{R}}
\newcommand{\Exchange}[0]{$\mathcal{X}$}

\newcommand{\nounderline}[3]{\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineab}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineac}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
\newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}

\newcommand{\scalar}[1]{$#1$}

\newcommand{\scalarsub}[2]{$#1_#2$}

%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
\journalcode{A}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation
\makeatletter
\font\dummyft@=dummy \relax
\makeatother

\begin{document}
		{\LARGE \emph{\today}} \\

\title{Niggli reduction: matrices for 3-D lattice transformations}

% Authors' names and addresses. Use \cauthor for the main (contact) author.
% Use \author for all other authors. Use \aff for authors' affiliations.
% Use lower-case letters in square brackets to link authors to their
% affiliations; if there is only one affiliation address, remove the [a].

\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
\author[b]{Herbert J.}{Bernstein}

\aff[a]{Ronin Institute for Independent Scholarship 2.0, USA}
\aff[b]{Ronin Institute for Independent Scholarship 2.0, USA}

% Use \shortauthor to indicate an abbreviated author list for use in
% running heads (you will need to uncomment it).

\shortauthor{Andrews and Bernstein}
	% Keywords (required for Journal of Synchrotron Radiation only)
% Use the \keyword macro for each word or phrase, e.g. 
% \keyword{X-ray diffraction}\keyword{muscle}

\keyword{lattice}
\keyword{reduction}
\keyword{Niggli}

	\maketitle
	\begin{abstract}
		Niggli reduction finds the most compact unit cell of a 3-D lattice.  
		Existing algorithms operate in \GVI{} space, and the 3-D basis-vector changes are lost.  
		We tabulate the 3×3 matrices for every step of the reduction, enabling direct re-indexing of reflections and full reconstruction of the reduction history.

	\end{abstract}

		\begin{synopsis}

	\end{synopsis}

	\section{Introduction}
	Niggli reduction~\cite{Niggli1928} finds the cell with the three shortest base vectors in a 3-D lattice. \citeasnoun{Gruber1973} published a compact algorithm using scalars that were
	later described in the space \GVI{} \cite{Andrews2014}. Although effective, 
	the algorithm works in a six-dimensional space, losing the three-space transformations
	needed for changing unit cell presentation or for reindexing reflection data.

	  Here we tabulate the 3-D matrices for transformation of unit cell basis vectors corresponding to each step
	  in Niggli reduction. They allow the accumulation during Niggli
	  reduction of all of the changes in the unit cell presentation.
	  The matrix that is finally accumulated transforms the input basis
	  to the Niggli reduced basis.

	\section{Conventions}
		\citeasnoun{Andrews2014} give the formal presentation of \GVI{}.
	\begin{itemize}
		\item $G^{6}=(g_{1},g_{2},g_{3},g_{4},g_{5},g_{6})$ with
		\[
		g_{1}=a^{2},\; g_{2}=b^{2},\; g_{3}=c^{2},\; g_{4}=2bc\cos\alpha,\; g_{5}=2ac\cos\beta,\; g_{6}=2ab\cos\gamma.
		\]
		\item \citeasnoun{Gruber1973}, called the process of putting the scalars
		into a fixed order ``normalizing''. We use the more
		descriptive term ``standard presentation''.
		\item Gruber labeled his pseudocode for standard presentation
		as ``Algorithm N'' and his pseudocode for reduction as 
		``Algorithm B''.

	\end{itemize}

\section{Matrices and Trigger Conditions}

We list matrices to transform unit cell basis vectors during 
Niggli reduction in Tables \ref{standardpresentation} and \ref{reduction},
along with arbitrary names for the operations and also the conditions
for triggering each operation during reduction. If a condition
evaluates to true, then the operation should be performed. For
complete Niggli reduction, the steps are repeated from the identity, SP0,
to convergence.

The numbering of all
the operations, both standard presentation and reduction, start from zero
for the first standard presentation and continues through to R12. (Note that
the matrices for operations R9, R10 and R11 are the same as those for R5,
R6, and R7, respectively.)

\section{Usage}
\begin{enumerate}
	\item Start with SP0.
	\item Loop through SP and R tables until no trigger fires.
	\item Accumulate the product of all applied matrices; the final product maps the initial basis to the Niggli-reduced basis. 
\end{enumerate}

\begin{table}
	\caption{The matrices for Standard Presentation (SP). The 
	conditions for SP34a,b,c are patterns of $(g_4, g_5, g_6)$. The symbols are ``0'' for a fixed value of zero, ``+'' for a positive value in that position, and ``-'' for negative.}
	\label{standardpresentation}
	\centering
	\renewcommand{\arraystretch}{1.4}
	\begin{tabular}{@{}c@{\quad}l@{\quad}l@{}}
		\toprule
		\textbf{Designation} & \textbf{Matrix} & \textbf{Trigger condition on }$G^{6}$\\
		\midrule
		SP0 & $\begin{bmatrix}1&0&0\\0&1&0\\0&0&1\end{bmatrix}$ & (identity)\\
	\midrule
		SP1 & $\begin{bmatrix}0&-1&0\\-1&0&0\\0&0&-1\end{bmatrix}$ & $|g_{1}|>|g_{2}|+\delta$\\
	\midrule
		SP2 & $\begin{bmatrix}-1&0&0\\0&0&-1\\0&-1&0\end{bmatrix}$ & $|g_{2}|>|g_{3}|+\delta$\\
	\midrule
		SP34a & $\begin{bmatrix}-1&0&0\\0&-1&0\\0&0&1\end{bmatrix}$ & $(--+)~or~(+0-)~or~(0{+}-)$\\
		SP34b & $\begin{bmatrix}-1&0&0\\0&1&0\\0&0&-1\end{bmatrix}$ & $(-{+}-) ~or~ (+-0)~or~(0-+)$\\		SP34c & $\begin{bmatrix}1&0&0\\0&-1&0\\0&0&-1\end{bmatrix}$ & $(+--)~or~ (-{+}0)~or~(-0+)$\\
		\bottomrule
	\end{tabular}
\end{table}

\begin{longtable}[l]{@{}l@{\quad}l@{\quad}l@{}}
	\caption{The operations for Niggli reduction}
	\label{reduction}\\
	\toprule
	\textbf{Designation} & \textbf{Matrix} & \textbf{Trigger condition on }$G^{6}$ ($\delta>0$)\\
	\midrule
	\endfirsthead

	\multicolumn{3}{@{}c@{}}{\textit{— continued from previous page —}}\\
	\toprule
	\textbf{Designation} & \textbf{Matrix} & \textbf{Trigger condition on }$G^{6}$ ($\delta>0$)\\
	\midrule
	\endhead

	\midrule
	\multicolumn{3}{@{}c@{}}{\textit{— continued on next page —}}\\
	\endfoot
	\bottomrule
	\endlastfoot

	R5$^{+}$ & $\begin{bmatrix}1&0&0\\0&1&0\\0&-1&1\end{bmatrix}$
	&  $g_{4}>g_{2}+\delta$, $g_{4} >$ 0.0\\[2ex]

	R5$^{-}$ & $\begin{bmatrix}1&0&0\\0&1&0\\0&1&1\end{bmatrix}$
	&  $g_{4}>g_{2}+\delta$, $g_{4} \leq$ 0.0\\[2ex]
	\midrule

	R6$^{+}$ & $\begin{bmatrix}1&0&0\\0&1&0\\-1&0&1\end{bmatrix}$
	&  $g_{5}>g_{1}+\delta$, $g_{5} > 0$\\[2ex]

	R6$^{-}$ & $\begin{bmatrix}1&0&0\\0&1&0\\1&0&1\end{bmatrix}$
	&  $g_{5}>g_{1}+\delta$, $g_{5} <= 0$\\[2ex]
	\midrule
	R7$^{+}$ & $\begin{bmatrix}1&0&0\\-1&1&0\\0&0&1\end{bmatrix}$
	&  $g_{6}>g_{1}+\delta$, $g_{6} > 0$\\[2ex]

	R7$^{-}$ & $\begin{bmatrix}1&0&0\\1&1&0\\0&0&1\end{bmatrix}$
	&  $g_{6}>g_{1}+\delta$, $g_{6} <= 0$\\[2ex]
	\midrule
	R8 & $\begin{bmatrix}1&0&0\\0&1&0\\1&1&1\end{bmatrix}$
	&  $g_{4}+g_{5}+g_{6}+g{1}+g_{2}< -\delta$\\[2ex]\\
	\midrule

	R9$^{+}$ & same as R5$^{+}$
	&  $|g_{4}-g_{2}| <= \delta~ and~ 2g_{5} < g_{6}+\delta ~and~ g_{4} > 0$\\[2ex]

	R9$^{-}$ & same as R5$^{-}$
	&  $|g_{4}-g_{2}| <= \delta~ and~ 2g_{5} < g_{6}+\delta ~and~ g_{4} <= 0$\\[2ex]
	\midrule

	R10$^{+}$ & same as R6$^{+}$
	&  $|g_{5}-g_{1}| <= \delta~ and~ 2g_{4} < g_{6}+\delta ~and~ g_{5} > 0$\\[2ex]

	R10$^{-}$ & same as R6$^{-}$
	&  $|g_{5}-g_{1}| <= \delta~ and~ 2g_{4} < g_{6}+\delta ~and~ g_{5} <= 0$\\[2ex]

	\midrule

		R11$^{+}$ & same as R7$^{+}$ &
\minibox[l]{%
	$|g_{6}-g_{1}| \le \delta \text{ and } 2g_{4} < g_{5}+\delta \text{ and } g_{6} > 0$ \\
	$\text{or}$ \\
	$|g_{6}-g_{1}| < \delta \text{ and } 2g_{4} < g_{5} + \delta
	\text{ and } g_{6} > 0$%
} \\[4ex]\

		R11$^{-}$ & same as R7$^{-}$ &
\minibox[l]{%
	$|g_{6}-g_{1}| \le \delta \text{ and } 2g_{4} < g_{5}+\delta \text{ and } g_{6} <= 0$ \\
	$\text{or}$ \\
	$|g_{6}-g_{1}| < \delta \text{ and } 2g_{4} < g_{5} + \delta
	\text{ and } g_{6} <= 0$%
} \\[4ex]

	\midrule

	R12 & $\begin{bmatrix}1&0&0\\0&1&0\\-1&-1&1\end{bmatrix}$
&  $|g_{4}+g_{5}+g_{6}+g_{1} +g_{2}| <= \delta~ and~ 2|g_{1}+g_{5}|+g_{6} > \delta$\\[2ex]

%      else if (fabs(vin[3] + vin[4] + vin[5] + fabs(vin[0]) + fabs(vin[1])) <= delta &&
%(2.0 * (fabs(vin[0]) + vin[4]) + vin[5] > delta)) { // R12
%	m1 = R12;  // Use the global R12
%	m3d_step = R12_3D;
%	again = true;

\end{longtable}

\section{Synopsis}

Niggli reduction is formulated in terms of 6 scalars that formally are the components of space \GVI{}. Working in a
higher dimensional space loses the 3-space transformations that
are needed for reindexing reflection data and for changing
the orientation of atomic structure. The matrices are listed
here corresponding to each of the operations of Niggli reduction.

	\ack{{\bf Acknowledgements}}

We gratefully acknowledge the careful copy-editing and corrections of Frances C. Bernstein.

\ack{{\bf Funding information}}

Funding for this research was provided in part by:  
US Department of Energy Offices of Biological and
Environmental Research and of Basic Energy Sciences
(grant No. DE-AC02-98CH10886; grant No. E-SC0012704);
U.S. National Institutes of Health (grant No. P41RR012408;
grant No. P41GM103473; grant No. P41GM111244;
grant No. R01GM117126,
grant No. 1R21GM129570); Dectris, Ltd, and in part by
Grant No. 1R24GM154040-01 from the NIH
and the Regents' Prime contract no. DE-AC02-05CH11231 from the Department of Energy.

\bibliography{Reduced}

\bibliographystyle{iucr}

%	\bibliographystyle{plain}f
%	\begin{thebibliography}{9}
%		\bibitem{niggli} Niggli, P. (1927). \emph{Handbuch der Experimentalphysik}, Leipzig.
%		\bibitem{gruber} Gruber, B. (1973). \emph{Acta Cryst.} A29, 433--440.
%		\bibitem{ab2014} Andrews, L. C. \& Bernstein, H. J. (2014). \emph{Acta Cryst.} A70, 411--418.
%	\end{thebibliography}

\end{document}

### 2025-08-27 LatticeMatching_2025/LatticeMatching_2025.tex
CommandUsage --  add MATRIXORDER to the help. P3, 3D_Niggli_Matrices, and LatticeMatching_2025 are paper in progress.

%------------------------------------------------------------------------------
% Template file for the submission of papers to IUCr journals in LaTeX2e
% using the iucr document class
% Copyright 1999-2013 International Union of Crystallography
% Version 1.6 (28 March 2013)
%------------------------------------------------------------------------------

%xxxxxx just for testing github % 2
%xxxxxx just for testing github % 3

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
%\usepackage{bm}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{booktabs}
%\usepackage{calligra}
\usepackage{array}
\DeclareMathAlphabet{\mathcalligra}{T1}{calligra}{m}{n}
\def\mathbi#1{\textbf{\em #1}}
\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\hyphenation{Niggli}
\def\mathbi#1{\textbf{\em #1}}
%\numberwithin{equation}{section}
%\DeclareMathSymbol{\Gamma}{\mathalpha}{letters}{"00}
%\DeclareMathSymbol{\Lambda}{\mathalpha}{letters}{"03}
%\DeclareMathSymbol{\Omega}{\mathalpha}{letters}{"0A}
%\DeclareMathAlphabet{\mathitbf}{OML}{cmm}{b}{it}
\usepackage{color}
%\usepackage{ulem}
\usepackage{url}
\usepackage{yfonts}
%\usepackage{xr-hyper}
%\usepackage[draft]{hyperref}
%\usepackage{bibentry}
\usepackage[normalem]{ulem}
% Redefine \underline to use \textit
\renewcommand{\underline}[1]{\textit{#1}}

% Crystallographic space macros
\newcommand{\HVI}{\ensuremath{\mathbf{H}^{6}}}
\newcommand{\GVI}{\ensuremath{\mathbf{G}^{6}}}
\newcommand{\SVI}{\ensuremath{\mathbf{S}^{6}}}
\newcommand{\PIII}{\ensuremath{\mathbf{P}^{3}}}
\newcommand{\FIII}{\ensuremath{\mathbf{F}^{3}}}
\newcommand{\CIII}{\ensuremath{\mathbf{C}^{3}}}
\newcommand{\RIII}{\ensuremath{\mathbb{R}^{3}}}

\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\Imaginary}[0]{\mathcal{I}}
\newcommand{\Real}[0]{\mathcal{R}}
\newcommand{\Exchange}[0]{$\mathcal{X}$}

\newcommand{\nounderline}[3]{\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineab}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!#3}
\newcommand{\underlineac}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#1),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#2,&\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}
\newcommand{\underlinebc}[3]{\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!#1,&\!\!\!\!\!\!\!\!\!\Exchange{}(#2),&\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\Exchange{}(#3)}

\newcommand{\scalar}[1]{$#1$}

\newcommand{\scalarsub}[2]{$#1_#2$}

%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
\journalcode{A}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation
\makeatletter
\font\dummyft@=dummy \relax
\makeatother

\begin{document}                  % DO NOT DELETE THIS LINE

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).

	% Authors' names and addresses. Use \cauthor for the main (contact) author.
	% Use \author for all other authors. Use \aff for authors' affiliations.
	% Use lower-case letters in square brackets to link authors to their
	% affiliations; if there is only one affiliation address, remove the [a].

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	%-------------------------------------------------------------------------
	% The introductory (header) part of the paper
	%-------------------------------------------------------------------------

	% The title of the paper. Use \shorttitle to indicate an abbreviated title
	% for use in running heads (you will need to uncomment it).
	\begin{center}
	{\LARGE \emph{\today}} \\
\end{center}

\title{Approximate Lattice Matching in Three Dimensions}
\shorttitle{3D Lattice Matching}

% Authors' names and addresses. Use \cauthor for the main (contact) author.
% Use \author for all other authors. Use \aff for authors' affiliations.
% Use lower-case letters in square brackets to link authors to their
% affiliations; if there is only one affiliation address, remove the [a].

\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
\author[b]{Herbert J.}{Bernstein}

\aff[a]{Ronin Institute for Independent Scholarship 2.0, USA}
\aff[b]{Ronin Institute for Independent Scholarship 2.0, USA}

% Use \shortauthor to indicate an abbreviated author list for use in
% running heads (you will need to uncomment it).

\shortauthor{Andrews and Bernstein}

	% lca IUCr id IUCr6401
	% HJB IUCr id IUCr6484
	% NKS IUCr ID: IUCr7572
	% lca ORCID  0000-0002-4451-1641
	% HJB ORCID 0000-0002-0517-8532
	% NKS ORCID 0000-0003-2786-6552
	% Use \shortauthor to indicate an abbreviated author list for use in
	% running heads (you will need to uncomment it).

	\shortauthor{Andrews and Bernstein}

	% Use \vita if required to give biographical details (for authors of
	% invited review papers only). Uncomment it.

	% lca IUCr id IUCr6401
	%\vita{Author's biography}

	% Keywords (required for Journal of Synchrotron Radiation only)
	% Use the \keyword macro for each word or phrase, e.g. 
	% \keyword{X-ray diffraction}\keyword{muscle}

	\keyword{lattice}
	\keyword{reduction}
	\keyword{Niggli}
	\keyword{matching}
	\keyword{\PIII}

	% PDB and NDB reference codes for structures referenced in the article and
	% deposited with the Protein Data Bank and Nucleic Acids Database (Acta
	% Crystallographica Section D). Repeat for each separate structure e.g
	% \PDBref[dethiobiotin synthetase]{1byi} \NDBref[d(G$_4$CGC$_4$)]{ad0002}

	%\PDBref[optional name]{refcode}
	%\NDBref[optional name]{refcode}

	\maketitle                        % DO NOT DELETE THIS LINE

	\begin{synopsis}

	\end{synopsis}

	\newcommand{\OPES}[0]{$E^3toS^6$}
	\newcommand{\OPESS}[0]{$$E^3toS^6$$}
	\newcommand{\MSVI}[0]{$M_{S^{6}}$}
	\newcommand{\MEIII}[0]{$M_{E^{3}}$}
	\newcommand{\Plus}[0]{\mathcal{P}}	
	\newcommand{\Minus}[0]{\mathcal{M}}

	\newcommand{\ci}[0]{$c_1$}
	\newcommand{\cii}[0]{$c_2$}
	\newcommand{\ciii}[0]{$c_3$}

	\begin{abstract}
Given two unit cells, the problem of determining the best match
of the lattice of one to the lattice of the other can require
large numbers of trial transformations. We present a solution 
that requires only a limited number of trials.

		{\bf Note:}  In his later publications, Boris Delaunay used the Russian version of his surname, Delone.\\

	\end{abstract}
	% Appendices appear after the main body of the text. They are prefixed by
	% a single \appendix declaration, and are then structured just like the
	% body text.

	\section{Introduction}

	A relatively common crystallographic problem is to compare two 
	unit cells and determine whether their lattices are the or
	close to each other. \citeasnoun{flor2016comparison} have described a
	solution for more symmetric crystal families. However, for the general case of triclinic lattices
	a different method is required.

	\citeasnoun{andrews2023approximating} described a general solution, using
	the space \SVI{}. While effective, this solution in a space
	based on projections fails to produce the 3D transformations needed
	for reindexing reflection data and for examining the relationships
	of structures.

	 Attempts to solve the problem
	by trying a large number of trial transformations can require
	an unknown and unknowable number of trials of transformation. Here we describe
	a straight-forward approach that uses a fixed, modest number
	of trials and that gives multiple possible matches if they are
	appropriate.

	\section{Notation}

\noindent \PIII{}: Polar coordinate space. It is a compact and smooth alternative to lengths and angles. \PIII{} is useful for comparing unit cells. \PIII{} is defined as:

	\PIII{}: 
		$(|\vec{a}|,\alpha),\ (|\vec{b}|,\beta),\ (|\vec{c}|,\gamma)$\\
	$\Rightarrow$\\
	\hspace*{0.5cm}$(|\vec{a}|\cos\alpha,\, |\vec{a}|\sin\alpha)$\\
	\hspace*{0.5cm}$(|\vec{b}|\cos\beta,\, |\vec{b}|\sin\beta)$\\
	\hspace*{0.5cm}$(|\vec{c}|\cos\gamma,\, |\vec{c}|\sin\gamma)$\\
	also known as $(p_1, p_2, p_3)$

	\section{Lattice matching}

	The general solution to matching two lattice has been to 
	test the comparison using a large number of likely transformations.
	Two problems arise: how many trials to make and how to effectively
	measure the difference between two lattices.

	To avoid the need to examine an unknown number of possible 
	transformations, we propose an algorithm with the following stages.
	One of the input cells will be designated ``reference'' and the
	other (to which the transformations in stage three will be applied) will be
	designated ``mobile''.
	\begin{enumerate}
		\item Convert from both (possibly) centered lattices to primitive.
		(Apply matrix $M_{CR}$ for reference and $M_{CM}$ for mobile.) \cite{ITC_VolumeA_2016_A}
		\item Niggli reduce both lattices. (Apply matrix $M_{NR}$ for reference and $M_{NM}$ for mobile.) 
		\item Apply a limited set of transformations to mobile, measuring an
		agreement factor for each trial, keeping those that meet
		an acceptable agreement value. This stage creates matrix $M_{MR}$
		that transforms the modified mobile to the modified reference. \cite{Niggli1928} \cite{Andrews2025b} 
		\item Compare the results of each transformation using
		the \PIII{} Euclidean distance between the reference and the mobile 
		cell. \cite{Andrews2025c} 
	\end{enumerate}

	\subsection{Finishing}
	When all of the transformations in stage three have
	been examined, the group of accepted results needs to be processed.
	The group might be empty or might contain several possible valid
	results.  We call this the ``accepted group''.

	In each stage, when a transformation is used, the 
	corresponding matrix is saved for later use. The final transformation
	chain to convert mobile (in its original lattice centering) to reference (in its original centering) is

	 \ensuremath{M_{CM}*M_{NM}*M_{MR}*M_{NR}^{-1}*M_{CR}^{-1}}.

	 This matrix
	chain needs to be computed for each member of the accepted group. Of
	course, for a given reference/mobile pair, only $M_{MR}$ is different for accepted trials.

	The immediate question is ``how many transformations will
	need to be tested in stage three''. We have chosen to use all
	the unimodular matrices ({\it i.e.} determinant = 1.0), with
	elements -1, 0, +1. (3480 total matrices). Each transformation
	is tested and acceptable results are accumulated to be ``finished''.

	\section{Examples}

	\subsection{example 1}
{
	\noindent personal communication, \citeasnoun{aroyo2025pc}\\
	More than one acceptable match was found\\
\noindent Test input\\
p 5.17 3.18 7.74 90. 104.5 90.  (reference)\\
p 30.9616 3.1800 8.1608 90.00 171.18 90.00 (mobile)\\

\begin{verbatim}
Result of matching
--- Match 1 ---
Quality: EXCELLENT P3 Distance 0.000)

Transformation Matrix:
[  1.0000,   0.0000,   4.0000]
[  0.0000,   1.0000,   0.0000]
[ -1.0000,   0.0000,  -3.0000]
Matrix determinant: 1.0000
Reference cell:   P     5.170     3.180     7.740    90.000   104.500    90.000
Transformed cell: P     5.170     3.180     7.740    90.000   104.500    90.000
*** SUCCESS: Excellent match within P3-relative threshold ***

--- Match 2 ---
Quality: EXCELLENT (P3 Distance 0.000)

Transformation Matrix:
[ -1.0000,   0.0000,  -4.0000]
[  0.0000,   1.0000,   0.0000]
[  1.0000,   0.0000,   3.0000]
Matrix determinant: 1.0000
Reference cell:   P     5.170     3.180     7.740    90.000   104.500    90.000
Transformed cell: P     5.170     3.180     7.740    90.000   104.500    90.000
*** SUCCESS: Excellent match within P3-relative threshold ***
\end{verbatim}

\subsection{example 2}
Test input:\\
C 12.770 21.235 14.411 136.017 84.071 111.795  (reference)\\
F 33.151 18.241 20.218 83.054 144.781 120.639 (mobile)\\

\begin{verbatim}
=== LATTICE MATCHING RESULTS === ===================================================
Found 1 excellent match
Success threshold: 1.43e-01  (Fixed strict)

--- Match 1 ---
Quality: EXCELLENT (0.000 )
P3 Distance: 0.000
S6 Angle: 0.00
Transformation Matrix:
[  0.0000,  -0.5000,   0.5000]
[  1.0000,   0.5000,   0.5000]
[  0.0000,   0.5000,   0.5000]
Matrix determinant: 0.5000
Reference cell:   C    12.770    21.235    14.411   136.017    84.071   111.795
Transformed cell: C    12.770    21.235    14.411   136.017    84.071   111.795
*** SUCCESS: Excellent match within P3-relative threshold ***
\end{verbatim}

\subsection{example 3}
	\noindent personal communication 
	\cite{simmons2025pc}\\
	A case where more than one accepted match was found.\\
	Input:\\
	p 10.25 10.74 21.08 87.72 75.97 61.53 (reference)\\
	p 10.25 10.74 21.08 78.96 75.97 61.49 (mobile)\\

\begin{verbatim}
		--- Match 1 ---
	Quality: EXCELLENT (0.010 )
	P3 Distance: 0.010
	S6 Angle: 0.01
	Transformation Matrix:
	[ -1.0000,   0.0000,   0.0000]
	[ -1.0000,   1.0000,   0.0000]
	[  0.0000,   0.0000,  -1.0000]
	Matrix determinant: 1.0000
	    Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)\\
			Transformed cell: P    10.250    10.739    21.080    87.714    75.970    61.502
	*** SUCCESS: Excellent match within P3-relative threshold ***

	--- Match 2 ---
	Quality: EXCELLENT (0.023 )
	P3 Distance: 0.023
	S6 Angle: 0.03
	Transformation Matrix:
	[ -1.0000,   0.0000,   0.0000]
	[  0.0000,  -1.0000,   0.0000]
	[ -1.0000,   0.0000,   1.0000]
Matrix determinant: 1.0000
    	Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)\\
    	Transformed cell: P    10.250    10.740    21.087    87.674    75.893    61.490
*** SUCCESS: Excellent match within P3-relative threshold ***

\end{verbatim}

\subsection{example 4}{
%%	C:\Users\lca\Source\Repos\duck10\LatticeRepLib\x64\Release>cmdgen 1 C5 | cmdperturb 20 | cmdniggli | cmdtocell
%	; To Cell
%	; Enter control variables and input vectors (type 'end' to finish):
%	end
20 slight variations on a primitive cubic cell
were
created, with a=10.0, each Niggli reduced, and lattice-matched.\\

Input:
\begin{verbatim}
	P   7.579   7.579   7.579  90.000  90.000  90.000 (reference)
	P   7.570   7.573   7.574  89.957  89.937  89.958
	P   7.576   7.580   7.581  89.989  89.957  89.964
	P   7.573   7.574   7.576  89.950  89.946  89.994
	P   7.571   7.578   7.583  90.058  90.029  90.062
	P   7.577   7.578   7.578  90.051  90.033  90.051
	P   7.572   7.576   7.578  90.033  90.034  90.084
	P   7.573   7.576   7.580  90.028  90.060  90.020
	P   7.570   7.573   7.580  89.993  89.977  89.931
	P   7.572   7.572   7.573  89.982  89.934  89.936
	P   7.577   7.580   7.588  89.958  89.970  89.996
	P   7.571   7.581   7.581  90.061  90.028  90.059
	P   7.573   7.574   7.584  90.005  90.023  90.065
	P   7.580   7.581   7.581  89.933  89.996  89.968
	P   7.576   7.581   7.586  89.913  89.990  89.956
	P   7.573   7.574   7.581  89.979  89.983  89.974
	P   7.570   7.575   7.57   90.025  90.063  90.066
	P   7.571   7.579   7.581  89.988  89.993  89.942
	P   7.578   7.581   7.588  90.049  90.023  90.002
	P   7.580   7.580   7.585  89.932  89.989  89.988
	P   7.571   7.576   7.577  89.970  89.993  89.943
\end{verbatim}

\begin{verbatim}
		and here is the matching results:

	Most common transformation matrices:
	[1,0,0,0,1,0,0,0,1]   : 9 mobiles (P3 agreement: 0.016)
	[1,0,0,0,-1,0,0,0,-1] : 3 mobiles (P3 agreement: 0.015)
	[0,1,0,0,0,-1,-1,0,0] : 2 mobiles (P3 agreement: 0.015)
	[1,0,0,0,0,-1,0,1,0]  : 1 mobiles (P3 agreement: 0.011)
	[0,0,1,1,0,0,0,1,0]   : 1 mobiles (P3 agreement:, 0.012)
	[0,0,1,0,-1,0,1,0,0]  : 1 mobiles (P3 agreement: 0.014)
	[0,0,-1,0,1,0,1,0,0]  : 1 mobiles (P3 agreement: 0.012)
	[0,-1,0,-1,0,0,0,0,-1]: 1 mobiles (P3 agreement: 0.009)
	[-1,0,0,0,0,-1,0,-1,0]: 1 mobiles (P3 agreement: 0.017)

	=== BEST MATCH DETAILS ===
	Mobile 2: 8.42e-03 (EXCELLENT)
	Matrix: [1 0 0 0 1 0 0 0 1]
	Reference	 P     7.579     7.579     7.579    90.000    90.000    90.000
	Transformed:     7.576     7.580     7.581    89.989    89.957    89.964

	Mobile 15: 9.36e-03 (EXCELLENT)
	Matrix: [0 -1 0 -1 0 0 0 0 -1]
	Reference:   P   7.579     7.579     7.579    90.000    90.000    90.000
	Transformed:     7.574     7.573     7.581    89.983    89.979    89.974

	Mobile 13: 1.03e-02 (EXCELLENT)
	Matrix: [1 0 0 0 1 0 0 0 1]
	Reference:   P   7.579     7.579     7.579    90.000    90.000    90.000
	Transformed      7.580     7.581     7.581    89.933    89.996    89.968

\end{verbatim}

\section{Limitations}

The group of all unimodular matrices is an infinite group. Clearly, by limiting this algorithm to matrices with elements
-1/0/+1, a complete search cannot be done. However as the examples show, in common cases, the algorithm is effective. One limitation is obvious: if one axial length of the reduced cell is longer than the sum
of the other two axial lengths, then -1/0/+1 cannot explore some regions
that are available for shorter axial lengths. For instance, if one axis
is much longer than the others, then in a matrix for that long axis,
the column elements for that axis cannot have more than one element -1/+1.
Similar arguments apply to the case of a very short axis. In order
to treat these more extreme cases using this algorithm, a larger
range of element values would need to be examined. The consequence is
that the number of matrices and thus the number of trials grows rapidly.

	\section{Availability of code}

	The code for Lattice Matching and related 
	software tools are available in github.com, in
	\url{https://github.com/duck10/LatticeRepLib.git}.
	The program CmdLMP3 uses the required files.

	%\appendix

	%\section{blah blah blah -- Supplementary Material}
	\ack{Acknowledgements}

Careful copy-editing and corrections by Frances C. Bernstein are 
gratefully acknowledged.

%	Our thanks to Jean Jakoncic and Alexei Soares for 
%	helpful conversations and access to data and facilities at 
%	Brookhaven National Laboratory.
%	
\ack{{\bf Funding information}}

Funding for this research was provided in part by:  
US Department of Energy Offices of Biological and 
Environmental Research and of Basic Energy Sciences 
(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
U.S. National Institutes of Health (grant No. P41RR012408; 
grant No. P41GM103473; grant No. P41GM111244; 
grant No. R01GM117126,
grant No. 1R21GM129570); Dectris, Ltd.

\section{Synopsis}

A solution is proposed for the problem of best matching the 
lattices of unit cells that are described in different
presentations.

	\bibliography{Reduced}

	\bibliographystyle{iucr}

	%-------------------------------------------------------------------------
	% TABLES AND FIGURES SHOULD BE INSERTED AFTER THE MAIN BODY OF THE TEXT
	%-------------------------------------------------------------------------

	% Simple tables should use the tabular environment according to this
	% model

	% Postscript figures can be included with multiple figure blocks

	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>plotc3
	%; Graphical output SVG file =PLT__2023-03-07.13_43_35.svg
	%
	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmdniggli | plotc3
	%; Graphical output SVG file =PLT__2023-03-07.13_44_06.svg
	%
	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmddelone | plotc3
	%; Graphical output SVG file =PLT__2023-03-08.07_11_03.svg
	%
	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmdniggli | cmdperturb 5 20 | plotc3
	%; Graphical output SVG file =PLT__2023-03-08.09_00_13.svg
	%
	%C:\Users\lca\Source\Repos\LatticeRepLib\x64\Debug>cmdniggli | cmdperturb 5 100 | plotc3
	%; Graphical output SVG file =PLT__2023-03-08.09_00_21.svg

\end{document}                    % DO NOT DELETE THIS LINE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2025-08-27 P3/P3.tex
CommandUsage --  add MATRIXORDER to the help. P3, 3D_Niggli_Matrices, and LatticeMatching_2025 are paper in progress.

@@
 %\usepackage{calligra}
+\usepackage{longtable}
 \usepackage{array}
@@
 \usepackage{color}
-\usepackage{ulem}
+%\usepackage{ulem}
 \usepackage{url}
@@
 %\usepackage{bibentry}
+\usepackage[normalem]{ulem}
+% Redefine \underline to use \textit
+\renewcommand{\underline}[1]{\textit{#1}}
 
@@
 \usepackage[most]{tcolorbox}
+
+% lca IUCr id IUCr6401
+% HJB IUCr id IUCr6484
+% NKS IUCr ID: IUCr7572
+% lca ORCID  0000-0002-4451-1641
+% HJB ORCID 0000-0002-0517-8532
+% NKS ORCID 0000-0003-2786-6552
 
@@
 \newcommand{\FIII}{\ensuremath{\mathbf{F}^{3}}}
+\newcommand{\BIV}{\ensuremath{\mathbf{B}^{4}}}
 \newcommand{\CIII}{\ensuremath{\mathbf{C}^{3}}}
@@
 
-	\title{\PIII, a simple metric for \HVI}
+	\title{\PIII, linearizing unit cell parameters}
 	\shorttitle{properties of \PIII}
@@
 
-	\aff[a]{Ronin Institute, 9515 NE 137th St, Kirkland, WA, 98034-1820 \country{USA}}
-	\aff[b]{Ronin Institute, c/o NSLS-II, Brookhaven National Laboratory, Upton, NY, 11973 \country{USA}}
+\aff[a]{Ronin Institute for Independent Scholarship 2.0, USA}
+\aff[b]{Ronin Institute for Independent Scholarship 2.0, USA}
 
@@
 	\begin{abstract}
+		The space \PIII{} is introduced, derived from unit cell axial lengths and interaxial angles. \PIII enables linearization of unit cell parameters using three polar coordinate bases.
 
-		\emph{Note:}  In his later publications, Boris Delaunay used the Russian version of his surname, Delone.
+	\PIII{} serves as a compact, 
+	interpretable alternative to more abstract spaces such as \GVI{} and \SVI{}.
 
@@
 
-	In crystallography, cell geometry is conventionally encoded in a six-dimensional parameter space 
+	In crystallography, unit cell geometry is conventionally encoded in a six-dimensional parameter space 
 	\ensuremath{\HVI{}=(a,b,c, \alpha, \beta, \gamma).}\footnote{``H'' was chosen to
-		honor early French crystallographer Ren\'e Just Ha\"uy.} But \HVI{} is not a metric space
-	(where distances can be simply defined), and \HVI{} is not a vector space (where 
+		honor early French crystallographer Ren\'e Just Ha\"uy.} However, \HVI{} is not a metric space
+	(where distances can be simply defined), nor is \HVI{} a vector space (where 
 	objects can be added and subtracted). In the same sense, symmetry operations
-	have no simple representations in \HVI{}.
+	have no simple representations in \HVI{}. In consequence, \HVI{} has no
+	simple measure of the ``difference'' between two lattices; how can one 
+	compare a difference of 1 \AA{}ngstrom unit vs. one angular degree?
 
@@
 
-\begin{table}
-	\label{tbl_spaces}
-	\centering
+We propose a ``linearized'' space, \PIII{}, derived from \HVI{}, but
+having all measures in \AA{}ngstrom units (or other chosen linear measures.)
+
+\renewcommand{\arraystretch}{1.3}
+
+\begin{longtable}{@{}c@{\quad}l@{\quad}p{7cm}@{}}
 	\caption{Crystallographic spaces with their scalar definitions and analytical purposes}
-	\renewcommand{\arraystretch}{1.3}
-	\begin{tabular}{|c|l|p{7cm}|}
-		\hline
+	\label{tbl_spaces} \\
+
+	\toprule
 	\textbf{Space} & \textbf{Scalars / Coordinates} & \textbf{Purpose} \\
+	\midrule
+	\endfirsthead
 
-\hline
+	\multicolumn{3}{@{}c@{}}{\textit{Table \thetable\ continued}} \\
+	\toprule
+	\textbf{Space} & \textbf{Scalars / Coordinates} & \textbf{Purpose} \\
+	\midrule
+	\endhead
+
+	\midrule
+	\multicolumn{3}{@{}c@{}}{\textit{Continued on next page}} \\
+	\endfoot
+
+	\bottomrule
+	\endlastfoot
+
 	$\HVI$ &
@@
 		$\alpha,\, \beta,\, \gamma \in (0^\circ, 180^\circ)$ — opposing angles\\
-	Additionally, the sum of the angles must be less than 
-	$360^\circ$ and must obey the triangle inequality
+
 	} &
-Composite parameter space described as 
-$\mathbb{R}_{>0}^3 \times (0^\circ, 180^\circ)^3$. 
-The length dimensions are unconstrained, while the angles are 
-geometrically restricted. Useful for conventional lattice 
-descriptions, but limited in symmetry representation and 
-metric-based classification. \\
+	Composite geometric parameter space $\mathbb{R}_{>0}^3 \times (0^\circ, 180^\circ)^3$.
+Lengths are unconstrained; angles are geometrically restricted.
+The sum of the angles must be less than $360^\circ$ and obey the triangle inequality \HVI{} is useful for conventional lattice descriptions, but limited in symmetry representation and metric-based classification.
+\\
+	\midrule
+	\FIII{} &
+	$\vec{a},\, \vec{b},\, \vec{c} \in \mathbb{R}^3$ &
+	Base vector space in \RIII. Coordinates are the three-space base vectors of a unit cell. Lattices are defined using the
+	components of \FIII{}, and translational symmetry is described 
+	in \FIII{}. \\	
+	\midrule
 
-\hline		
-$\mathbf{F^3}$ &
-$\vec{a},\, \vec{b},\, \vec{c} \in \mathbb{R}^3$ &
-Base vector space, a frame space (or frame bundle) in \RIII. The coordinates are the 3-space 
-base vectors of a unit cell. Symmetry 
-operations are enumerated using this space. \\
-\hline
+	\BIV{} &
+	\ensuremath{\vec{a},\, \vec{b},\, \vec{c},\, (-\vec{a}-\vec{b}-\vec{c}) \text{ in } \mathbb{R}^3} &
+	Base vector space in \FIII{} plus the negative sum of the three base vectors. 
+	This set of vectors is used in the definition of \SVI{}, which
+	is the space of Selling/Delone reduction.\\
+	\midrule
 
-		$\mathbf{P^3}$ &
+	\PIII{} &
 	\parbox[t]{5.5cm}{
 		$(|\vec{a}|,\alpha),\ (|\vec{b}|,\beta),\ (|\vec{c}|,\gamma)$\\
-			$\Rightarrow$\,\\ 
-			\hspace*{0.5cm}$(|\vec{a}|\cos\alpha,\, |\vec{a}|\sin\alpha$\\
-		\hspace*{0.5cm}	$|\vec{b}|\cos\beta,\, |\vec{b}|\sin\beta$\\
-		\hspace*{0.5cm}	$|\vec{c}|\cos\gamma,\, |\vec{c}|\sin\gamma)$ \\
-		also known as: \\
-		\hspace*{0.5cm}	\ensuremath{(p1,p2,p3)}. 
+		$\Rightarrow$\\
+		\hspace*{0.5cm}$(|\vec{a}|\cos\alpha,\, |\vec{a}|\sin\alpha)$\\
+		\hspace*{0.5cm}$(|\vec{b}|\cos\beta,\, |\vec{b}|\sin\beta)$\\
+		\hspace*{0.5cm}$(|\vec{c}|\cos\gamma,\, |\vec{c}|\sin\gamma)$\\
+		also known as $(p_1, p_2, p_3)$
 	} &
-		Polar coordinate space defined by the map $\FIII \rightarrow \PIII$.
-Represents each base vector by magnitude and opposing angle, decomposed into directional components.
-Compact and smooth alternative to $\HVI$.
-\PIII{} provides a simple
-		and intuitive space for representing and comparing
-		 large numbers of unit cells. (This work) \\
-		\hline
+	Polar coordinate space defined in terms to unit cell axial
+	lengths and their opposing angle. It is a compact and smooth alternative to $\HVI$. Useful for comparing unit cells. (This work) \\
+		\midrule
 
-$\mathbf{S^6}$ &
+	\SVI{} &
 	\parbox[t]{5.5cm}{
 		$s_1 = \vec{b} \cdot \vec{c},\, s_2 = \vec{a} \cdot \vec{c},\, s_3 = \vec{a} \cdot \vec{b}$\\
-	$s_4 = \vec{a} \cdot \vec{d},\, s_5 = \vec{b} \cdot \vec{d},\, s_6 = \vec{c} \cdot \vec{d}$\\ \\
+		$s_4 = \vec{a} \cdot \vec{d},\, s_5 = \vec{b} \cdot \vec{d},\, s_6 = \vec{c} \cdot \vec{d}$\\
 		\quad where $\vec{d} = -\vec{a} - \vec{b} - \vec{c}$
 	} &
-A space used for Selling/Delone reduction.
-The scalars are derived from dot products between lattice vectors.
-Used for lattice classification described by Delaunay and for distance computation \cite{Andrews2019b}.
-24 Bravais types were defined by citeasnoun{Delone1975} in \SVI{}, but
-described using pictures showing the special relationships. \citeasnoun{andrews2023measuring} revised the table of Delone to
-include the \SVI vector definitions of the Bravais types.\\
+	Used for Selling/Delone reduction.
+The scalar components are dot products between lattice vectors.
+\SVI{} supports Bravsi lattice classification.
+24 Bravais types were defined using the scalar components of \SVI{} by Delone (1975), later revised using \SVI{} \cite{andrews2023measuring}.
+\\
+		\midrule
 
-		\hline
-		$\mathbf{G^6}$ &
+	\GVI{} &
 	\parbox[t]{5.5cm}{
 		$g_1 = \vec{a} \cdot \vec{a},\, g_2 = \vec{b} \cdot \vec{b},\, g_3 = \vec{c} \cdot \vec{c}$\\
-			$g_4 = 2\vec{b} \cdot \vec{c},\, g_5 =2 \vec{c} \cdot \vec{a},\, g_6 = 2\vec{a} \cdot \vec{b}$
+		$g_4 = 2\vec{b} \cdot \vec{c},\, g_5 = 2\vec{c} \cdot \vec{a},\, g_6 = 2\vec{a} \cdot \vec{b}$
 	} &
-		A space based on the metric tensor (and thus dot products 
-		of the basis vectors of the unit cell) and used for Niggli reduction and lattice distance computation. \cite{Andrews2014}
-		The International Tables for Crystallography describe the 44
-		Bravais types that result from Niggli reduction using
-		algebraic expression in \GVI{} (but stated in terms of scalars
-		rather than manifolds in \GVI{}) \cite{ITC_VolumeA_2016}. \\
-		\hline
+	\GVI{} is based on the metric tensor and used for Niggli reduction and lattice distance computation \cite{Andrews2014}.
+The International Tables describe 44 Bravais types using scalar expressions derived from \GVI{} \cite{ITC_VolumeA_2016}.
+\\
+		\midrule
 
-\CIII &
+\CIII{} &
 	\parbox[t]{5.5cm}{
-	from the \SVI{} scalars, $s1,...,s6$\\
-	$c1=(s1,s4), c2=(s2,s5), c3=(s3,s6)$\\
-	where (?,?) denotes a complex number
+		From \SVI{} scalars: $s_1,\dots,s_6$\\
+		$c_1 = (s_1, s_4),\ c_2 = (s_2, s_5),\ c_3 = (s_3, s_6)$\\
+		where the parentheses denote a complex number
 	} &
-\CIII{} was created to associate related scalars after the
-indication of \citeasnoun{Delone1975} concerning the ``opposite''
-scalars in \SVI{}. Symmetry and Selling/Delaunay reduction have
-simpler forms in \CIII{} than in \SVI{}. \cite{andrews2023complex}\\
-\hline
+	\CIII{} associates scalar pairs identified by Delone as “opposites.” Symmetry and Selling/Delone reduction simplify in this space \cite{andrews2023complex}. \\
+		\midrule
 
@@
 	\parbox[t]{5.5cm}{
-			$d_1,\, d_2,\, d_3$ — cell edge lengths\\
-			$d_4,\, d_5,\, d_6$ — shorter face diagonals' lengths\\
+		$d_1,\, d_2,\, d_3$ — edge lengths\\
+		$d_4,\, d_5,\, d_6$ — shortest face diagonals\\
 		$u$ — shortest body diagonal
 	} &
-		Unsorted Dirichlet Cell space derived from Niggli-reduced cells.
-Defined by the map from lattice geometry to a 7-dimensional vector of distances.
-Fully invertible and smooth; used for rapid lattice distance computation.
-\cite{bernstein2023invertible}\\
-		\hline
+	$\mathbf{DC7u}$: unsorted Dirichlet Cell space derived from Niggli-reduced cells.
+Defined as a 7D vector of distances.
+Fully invertible and smooth; used for rapid lattice distance computation \cite{bernstein2023invertible}.
+\\
 
-	\end{tabular}
-\end{table}
+\end{longtable}
 
@@
 \label{fig_spaces}
-\caption{Derived relationships among crystallographic spaces. 
-	Base vector space \FIII{} gives rise to three scalar spaces:
-	metric tensor space \GVI{}, Selling/Delone space \SVI{}, and conventional cell parameter space \HVI{}.
-	Each transforms into a reduced or classification space: 
-	$\mathbf{DC7u}$ from \GVI{}, \CIII{} from \SVI{}, and \PIII{} from $\mathbf{H^6}$.}
+\caption{Schematic relationships among crystallographic spaces. The base vector 
+	space \FIII{} gives rise to three scalar spaces: metric tensor space \GVI{}, 
+	Selling/Delone space \SVI{}, and conventional cell parameter space \HVI{}. 
+	Each transforms into a reduced or classification space: $\mathbf{DC7u}$ from \GVI{}, 
+	\CIII{} from \SVI{}, and \PIII{} from \HVI{}.}
 \end{figure}
@@
 ``opposite'' scalars in the tetrahedron representations of the Selling
-scalars. Examining those relationships show that the opposite terms 
-are pairs where a term involving a is opposite a term involving $\alpha$, etc.
+scalars. Examining those relationships shows that the opposite terms 
+are pairs where a term involving $\mathbf{a}$ is opposite a term involving $\alpha$, {\it etc.}
 
 Extending the observation of \citeasnoun{Delone1975}, the idea occurs
-to make a triple of variables of those pairs. Since each pair has
-one length and one angle, we considered each pair as the definition of
-a polar coordinate space. The result is \PIII{}, a space of 3 2-dimensional polar coordinates. See Table \ref{tbl_spaces} for the
+to make a triple of those pairs. Since each pair has
+one length and one angle, we consider each pair as the definition of
+a polar coordinate. The result is \PIII{}, a space of three two-dimensional polar coordinates. See Table \ref{tbl_spaces} for the
 formal description.  Two uses immediately are obvious.
 
-The first uses the 2-dimensional character of the coordinates of \PIII{}. They
-allow simple to plot representations of collections of unit cells. 
-See Figure \ref{PlotPolar_raw}, Figure \ref{PlotPolar_Niggli}, and Figure \ref{PlotPolar_Delone}.
+The first use applies the two-dimensional character of the coordinates of \PIII{}. They
+allow simple-to-plot representations of collections of unit cells. Plotting
+each of the coordinates of \PIII{} generates plots that are easily
+interpreted as crystallographic axial length/opposite angle pairs. The
+distance of a point has a length equal to the axial length, and the interaxisl
+angle of the two other bases is illustrated on the plots. See Section  \ref{graphical}. Basically, together the three plots of the three coordinates are
+simple projections of \HVI{}. See Section \ref{graphical}.
+
+A second use is allowing the calculation of Euclidean distances between
+pairs of unit cells. Because \PIII{} can be represented as three xy planes, the Euclidean difference between
+any two unit cells described as \PIII{} can be easily computed. This use
+contrasts with \HVI{}, where incommensurate lengths and angles do not
+lend themselves to a distance calculation. See Section \ref{comparing}.
+
+\section{Relationship to Other Spaces}
+
+While \PIII{} is derived from the same foundational cell parameters as $\mathbf{H^6}$, $\mathbf{P^3}$ sidesteps issues of angular nonlinearity  that arise in the conventional cell representation.
+Compared with $\mathbf{C^3}$ \cite{andrews2023delone}, which derives coordinates from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic.
+It provides a useful intermediary: more interpretable than metric tensors or other spaces, but less abstract than \SVI{} or \GVI{}.
+
+\section{Graphical display of projections}{
+\label{graphical}
+\PIII{}
+allows simple-to-plot representations of collections of unit cells. 
+See Figures \ref{PlotPolar_raw}, \ref{PlotPolar_Niggli}, and  \ref{PlotPolar_Delone} where 100 randomly chosen triclinic cells are plotted
+(Figure \ref{PlotPolar_raw}) and the effects of unit cell reduction
+are demonstrated (Figure \ref{PlotPolar_Niggli} and Figure \ref{PlotPolar_Delone} ).
+}
 
 \begin{figure}
-\includegraphics[width=\columnwidth]{PlotPolar_raw}
+    \begin{center}
+	\includegraphics[width=\columnwidth]{PlotPolar_raw.png}
 	\label{PlotPolar_raw}
-	\caption{100 random primitive triclinic unit cells}
+	\caption{Scatter plots of 100 randomly selected
+		 primitive triclinic unit cells represented in \PIII{}. Each of the 3 components of \PIII{}
+		 is shown.}
+	\end{center}
 \end{figure}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{PlotPolar_Niggli}
+    \begin{center}
+	\includegraphics[width=\columnwidth]{PlotPolar_Niggli.png}
 	\label{PlotPolar_Niggli}
-	\caption{The same100 random primitive triclinic unit cells, Niggli-reduced}
+	\caption{The same 100 random primitive triclinic unit cells shown in Figure \ref{PlotPolar_raw}, Niggli-reduced}
+	\end{center}
 \end{figure}
@@
 \begin{figure}
-	\includegraphics[width=\columnwidth]{PlotPolar_Delone}
+	\begin{center}
+	\includegraphics[width=\columnwidth]{PlotPolar_Delone.png}
 	\label{PlotPolar_Delone}
-	\caption{The same 100 random primitive triclinic unit cells, Selling-reduced}
+	\caption{Scatter plots of 100 randomly selected
+		primitive triclinic unit cells represented in \PIII{} as shown in Figure \ref{PlotPolar_raw}, Selling (Delaunay)-reduced}
+	\end{center}
 \end{figure}
 
+\section{Comparing different unit cells}
+\label{comparing}
+
+In three-space, measuring the Euclidean distance between two points is just computing the square root
+of the sum of the squares of the differences in the three coordinates. Similarly in \PIII{},
+we compute distances from the sum of squared difference between the three
+polar coordinates. But polar coordinates have a Euclidean measure from
+their xy representation. So taking the six scalar values of two unit cells
+presented in \PIII{}, we calculate a Euclidean distance by the ordinary
+method.
+
+%; CmdDists, calculate distances
+%; Enter control variables and input vectors (type 'end' to finish):
+%P 10 10 10  90 90 90
+%p 10 10 10  90 90 91
+%p 10 10 10.1 90 90 90
+%end
+%10.000    10.000    10.000    90.000    90.000    90.000
+%14.142    30.113     8.479   147.466    56.504   144.893
+%
+%10.000    10.000    10.000    90.000    90.000    90.000
+%14.142    29.916     8.564   147.586    56.504   144.621
+%
+%<pre style="font-size: 105%; line-height: 60%; display:inline;" >;
+%; ordinals   P3-Euclid    P3Dist       NCDist    || ordinals     P3-Euclid    P3Dist       NCDist
+%; 1-2        0.175      19191.000      3.490         1-2         0.175      19191.000      3.490
+%; 2-3        0.202      19191.000      4.028         1-3         0.100      19191.000      2.010
+%
+%
+%;.......... input.cells ........................................ ordinals .....
+%P   10.000    10.000    10.000    90.000    90.000    90.000     1
+%P   10.000    10.000    10.000    90.000    90.000    91.000     2
+%P   10.000    10.000    10.100    90.000    90.000    90.000     3
+%;</pre>
+
+%; CmdDists, calculate distances
+%; Enter control variables and input vectors (type 'end' to finish):
+%p 10 10 10   90 90 90
+%p 10 10 10   90 90 91
+%p 10 10 11   90 90 90
+%end
+%<pre style="font-size: 105%; line-height: 60%; display:inline;" >;
+%; ordinals   P3-Euclid    P3Dist       NCDist    || ordinals     P3-Euclid    P3Dist       NCDist
+%; 1-2        3.490        0.203        3.490         1-2         0.175       26.959        3.490
+%; 2-3        21.288       0.872       21.288         1-3         1.000       26.896       21.000
+%
+%
+%;.......... input.cells ........................................ ordinals .....
+%P   10.000    10.000    10.000    90.000    90.000    90.000     1
+%P   10.000    10.000    10.000    90.000    90.000    91.000     2
+%P   10.000    10.000    11.000    90.000    90.000    90.000     3
+%;</pre>
+
+Let us compare two cells derived by modifying a primitive unit cell to form two slightly modified cells, one by changing $\gamma$, and one by changing $c$.
+\begin{enumerate}
+	\item P 10 10 10  90 90 90
+	\item P 10 10 10  90 90 91
+	\item P 10 10 10.1  90 90 90
+\end{enumerate}
+The Euclidean distances between these cells are:
+\begin{itemize}
+	\item 1-2: 0.175
+	\item 1-3: 0.100
+	\item 2-3: 0.202
+\end{itemize}
+
+In this example, a 1.0° change in $\gamma$ yields a Euclidean distance comparable to a 0.1 Å change in the c-axis length.
+%
+%\section{Scalar Reduction and Canonical Presentation in Projected Vector Triplets}
+%
+%\subsection{1. Standard Presentation of Projected Vectors}
+%	
+%	Let a set of three projected vectors be given:
+%	
+%	
+%	\[
+%	(p_1, p_2, p_3), \quad p_i \in \mathbb{R}^2
+%	\]
+%	
+%	
+%	Each \( p_i \) is constructed from a lattice vector via magnitude and opposing angle:
+%	
+%	
+%	\[
+%	p_i = |\vec{v}_i| (\cos \theta_i, \sin \theta_i)
+%	\]
+%	
+%	
+%	
+%	The \textbf{Standard Presentation} algorithm assigns a canonical ordering and orientation to the triplet:
+%	
+%	\begin{enumerate}
+%		\item \textbf{Magnitude Sort:} Reorder vectors so that
+%		
+%		
+%		\[
+%		|p_1| \leq |p_2| \leq |p_3|
+%		\]
+%		
+%		
+%		using Euclidean norms \( |p_i| = \sqrt{x_i^2 + y_i^2} \). Ties are preserved up to a numerical tolerance \( \delta \).
+%		
+%\item \textbf{Directional Coherence:} If the \( x \)-components of the triplet are not all of consistent sign, flip the necessary vectors by negating only the \( x_i \) component:
+%
+%
 %\[
-%(|\vec{x}|\cos\theta,\, |\vec{x}|\sin\theta) \in \mathbb{R}^2
+%p_i \to (-x_i,\ y_i)
 %\]
-
-%which preserves the Euclidean magnitude and implicitly encodes angular orientation.
-While the transformation is nonlinear in angle, the resulting structure allows straightforward computation of relative orientation, cluster proximity, and shape similarity through standard Euclidean metrics.
-
-The second use is allowing the calculation of Euclidean distances. Because \PIII{} can be represented as 3 xy planes, the Euclidean difference between
-any two unit cells described as \PIII{} can be easily computed. Thus use
-contrasts with \HVI{}, where incommensurate lengths and angles do not
-lend themselves to a distance calculation.
-
-\section{Relationship to Other Spaces}
-
-While derived from the same foundational cell parameters as $\mathbf{H^6}$, the space $\mathbf{P^3}$ sidesteps issues of angular nonlinearity  that arise in conventional cell representations.
-Compared with $\mathbf{C^3}$, which seeks symmetry-reduced coordinates derived from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic.
-It provides a useful intermediary: more interpretable than metric tensors or other spaces, less abstract than \SVI{} or \GVI{}.
-
-	\section{Notation}
-
-	\section{Matrices of boundary transformations}
-
-	\subsection{Derivation}
-
-	\subsection{Example P++ implementation}
-
-	\subsection{Example of use}
-
-	\section{Graphical display of projections}
-
-	\section{Scalar Reduction and Canonical Presentation in Projected Vector Triplets}
-
-	\subsection{1. Standard Presentation of Projected Vectors}
-
-	Let a set of three projected vectors be given:
-
-	\[
-	(p_1, p_2, p_3), \quad p_i \in \mathbb{R}^2
-	\]
-
-	Each \( p_i \) is constructed from a lattice vector via magnitude and opposing angle:
-
-	\[
-	p_i = |\vec{v}_i| (\cos \theta_i, \sin \theta_i)
-	\]
-
-	The \textbf{Standard Presentation} algorithm assigns a canonical ordering and orientation to the triplet:
-
-	\begin{enumerate}
-		\item \textbf{Magnitude Sort:} Reorder vectors so that
-
-		\[
-		|p_1| \leq |p_2| \leq |p_3|
-		\]
-
-		using Euclidean norms \( |p_i| = \sqrt{x_i^2 + y_i^2} \). Ties are preserved up to a numerical tolerance \( \delta \).
-
-\item \textbf{Directional Coherence:} If the \( x \)-components of the triplet are not all of consistent sign, flip the necessary vectors by negating only the \( x_i \) component:
-
-\[
-p_i \to (-x_i,\ y_i)
-\]
-
-This maps quadrant I vectors into quadrant II, preserving their placement in the upper half-plane and maintaining angular conventions typical in crystallographic analysis.
-
-\[
-p_i \to (-x_i,\ y_i)
-\]
-
-to align the triplet into a directionally consistent state. If all \( x_i < 0 \), the configuration is already coherent and requires no change.
-
-		\[
-		p_i \to (-x_i, y_i)
-		\]
-
-		for any two vectors, preserving chirality.
-
-		\item \textbf{Return Ordered Triplet:} Output the permuted and flipped triplet with consistent magnitude order and cosine directionality.
-	\end{enumerate}
-
-	\vspace{1em}
-
-	\subsection{2. Scalar Projection Interaction and Cost Function}
-
-	For any pair of projected vectors \( p_i, p_j \in \mathbb{R}^2 \), define the scalar projection interaction (or directional affinity):
-
-	\[
-	\mu_{ij} = p_i \cdot p_j = x_i x_j + y_i y_j
-	\]
-
-	The \textbf{scalar coupling cost} is then defined over the triplet as:
-
-	\[
-	C' = \frac{|\mu_{12}| + |\mu_{13}| + |\mu_{23}|}{p_1 \cdot p_1 + p_2 \cdot p_2 + p_3 \cdot p_3}
-	\]
-
-	This cost measures the normalized sum of pairwise directional alignment.
+%
+%
+%This maps quadrant I vectors into quadrant II, preserving their placement in the upper half-plane and maintaining angular conventions typical in crystallographic analysis.
+%
+%
+%
+%\[
+%p_i \to (-x_i,\ y_i)
+%\]
+%
+%
+%to align the triplet into a directionally consistent state. If all \( x_i < 0 \), the configuration is already coherent and requires no change.
+%
+%		
+%		
+%		\[
+%		p_i \to (-x_i, y_i)
+%		\]
+%		
+%		
+%		for any two vectors, preserving chirality.
+%		
+%		\item \textbf{Return Ordered Triplet:} Output the permuted and flipped triplet with consistent magnitude order and cosine directionality.
+%	\end{enumerate}
+%	
+%	\vspace{1em}
+%	
+%	\subsection{2. Scalar Projection Interaction and Cost Function}
+%	
+%	For any pair of projected vectors \( p_i, p_j \in \mathbb{R}^2 \), define the scalar projection interaction (or directional affinity):
+%	
+%	
+%	\[
+%	\mu_{ij} = p_i \cdot p_j = x_i x_j + y_i y_j
+%	\]
+%	
+%	
+%	
+%	The \textbf{scalar coupling cost} is then defined over the triplet as:
+%	
+%	
+%	\[
+%	C' = \frac{|\mu_{12}| + |\mu_{13}| + |\mu_{23}|}{p_1 \cdot p_1 + p_2 \cdot p_2 + p_3 \cdot p_3}
+%	\]
+%	
+%	
+%	
+%	This cost measures the normalized sum of pairwise directional alignment.
 Lower values of \( C' \) indicate that the projected vectors are less directionally coupled — i.e., more orthogonal or dispersed.
-
-	\vspace{1em}
-
-	\subsection{3. Scalar Reduction Algorithm}
-
-	The goal of scalar reduction is to transform the triplet \( (p_1, p_2, p_3) \) into an algebraically simpler form — one with lower scalar coupling cost \( C' \) — while preserving lattice identity (e.g., volume and reconstructability).
-
-	The reduction proceeds iteratively:
-
-	\begin{enumerate}
-		\item \textbf{Initialization:}
-		Apply standard presentation. Compute initial cost \( C'_0 \) and physical volume \( V_0 > 0 \).
-
-		\item \textbf{Pairwise Projection Subtractions:}
-		For each unordered pair \( (p_i, p_j),\ i \neq j \), compute the scalar coefficient:
-
-		\[
-		\lambda = \frac{p_i \cdot p_j}{p_j \cdot p_j}
-		\]
-
-		Then define the candidate vector:
-
-		\[
-		p_i' = p_i - \lambda p_j
-		\]
-
-		\item \textbf{Evaluate Candidate:}
-		Apply standard presentation to updated triplet. Recompute cost \( C'_{\text{new}} \) and volume \( V_{\text{new}} \).
-
-		\item \textbf{Acceptance Criteria:}
-		Accept the transformation if:
-
-		\[
-		C'_{\text{new}} + \delta < C' \quad \text{and} \quad V_{\text{new}} > 0
-		\]
-
-		\item \textbf{Repeat:}
-		Continue until no further accepted transformations lower the cost.
-
-		\item \textbf{Output:}
-		Return the reduced triplet \( (p_1, p_2, p_3)_{\text{red}} \) with minimal scalar cost.
-	\end{enumerate}
-
-	\vspace{1em}
-
-	\subsection{4. Interpretation of Cost and Reduction}
-
-	The term "reduction" refers to minimizing the scalar coupling among vectors — reducing directional alignment that obscures presentation clarity.
+%	
+%	\vspace{1em}
+%	
+%	\subsection{3. Scalar Reduction Algorithm}
+%	
+%	The goal of scalar reduction is to transform the triplet \( (p_1, p_2, p_3) \) into an algebraically simpler form — one with lower scalar coupling cost \( C' \) — while preserving lattice identity (e.g., volume and reconstructability).
+%	
+%	The reduction proceeds iteratively:
+%	
+%	\begin{enumerate}
+%		\item \textbf{Initialization:}
+%		Apply standard presentation. Compute initial cost \( C'_0 \) and physical volume \( V_0 > 0 \).
+%		
+%		\item \textbf{Pairwise Projection Subtractions:}
+%		For each unordered pair \( (p_i, p_j),\ i \neq j \), compute the scalar coefficient:
+%		
+%		
+%		\[
+%		\lambda = \frac{p_i \cdot p_j}{p_j \cdot p_j}
+%		\]
+%		
+%		
+%		Then define the candidate vector:
+%		
+%		
+%		\[
+%		p_i' = p_i - \lambda p_j
+%		\]
+%		
+%		
+%		
+%		\item \textbf{Evaluate Candidate:}
+%		Apply standard presentation to updated triplet. Recompute cost \( C'_{\text{new}} \) and volume \( V_{\text{new}} \).
+%		
+%		\item \textbf{Acceptance Criteria:}
+%		Accept the transformation if:
+%		
+%		
+%		\[
+%		C'_{\text{new}} + \delta < C' \quad \text{and} \quad V_{\text{new}} > 0
+%		\]
+%		
+%		
+%		
+%		\item \textbf{Repeat:}
+%		Continue until no further accepted transformations lower the cost.
+%		
+%		\item \textbf{Output:}
+%		Return the reduced triplet \( (p_1, p_2, p_3)_{\text{red}} \) with minimal scalar cost.
+%	\end{enumerate}
+%	
+%	\vspace{1em}
+%	
+%	\subsection{4. Interpretation of Cost and Reduction}
+%	
+%	The term "reduction" refers to minimizing the scalar coupling among vectors — reducing directional alignment that obscures presentation clarity.
 The cost \( C' \) is a rigorously defined metric for this alignment.
 Transformations reduce \( C' \) while preserving:
-	\begin{itemize}
-		\item The vector magnitudes and their reconstructable relationships
-		\item The unit cell's volume and crystallographic identity
-		\item The triplet’s algebraic validity in scalar projection space
-	\end{itemize}
-
-	Reduction leads to a cleaner, more canonical representation of the original structure — just as Niggli reduction provides a minimal cell in metric tensor space, scalar reduction provides a minimal configuration in projected vector space.
+%	\begin{itemize}
+%		\item The vector magnitudes and their reconstructable relationships
+%		\item The unit cell's volume and crystallographic identity
+%		\item The triplet’s algebraic validity in scalar projection space
+%	\end{itemize}
+%	
+%	Reduction leads to a cleaner, more canonical representation of the original structure — just as Niggli reduction provides a minimal cell in metric tensor space, scalar reduction provides a minimal configuration in projected vector space.
+%	
+%
 
 	\section{Summary}
+
+	The space \PIII{} is introduced. Less abstract than space based
+	on dot products and metric tensors, it provides a simple to understand
+	method for comparing unit cells and a convenient method of displaying 
+	unit cells.
 
@@
 	grant No. 1R21GM129570); Dectris, Ltd.
+
+	\section{Synopsis}
+
+	\PIII{} is a linearized description of unit cell parameters,
+	derived from usual lengths and angles of the unit cell. It 
+	provides a useful metric for comparing unit cells and a
+	convenient base for displaying and comparing unit cells.
 

### 2025-08-27 P3_Reduction/P3_Reduction.tex
CommandUsage --  add MATRIXORDER to the help. P3, 3D_Niggli_Matrices, and LatticeMatching_2025 are paper in progress.

--- 2025-08-12 P3/P3.tex
@@
 
+% lca IUCr id IUCr6401
+% HJB IUCr id IUCr6484
+% NKS IUCr ID: IUCr7572
+% lca ORCID  0000-0002-4451-1641
+% HJB ORCID 0000-0002-0517-8532
+% NKS ORCID 0000-0003-2786-6552
+
 % Space macros
@@
 
-	\title{\PIII, a simple metric for \HVI}
-	\shorttitle{properties of \PIII}
+	\title{\PIII, reducing unit cell parameters}
+	\shorttitle{reduction in \PIII}
 
@@
 
-		\emph{Note:}  In his later publications, Boris Delaunay used the Russian version of his surname, Delone.
-
 	\end{abstract}
@@
 
-	In crystallography, cell geometry is conventionally encoded in a six-dimensional parameter space 
+	In crystallography, unit cell geometry is conventionally encoded in a six-dimensional parameter space 
 	\ensuremath{\HVI{}=(a,b,c, \alpha, \beta, \gamma).}\footnote{``H'' was chosen to
-		honor early French crystallographer Ren\'e Just Ha\"uy.} But \HVI{} is not a metric space
-	(where distances can be simply defined), and \HVI{} is not a vector space (where 
+		honor early French crystallographer Ren\'e Just Ha\"uy.} However, \HVI{} is not a metric space
+	(where distances can be simply defined), nor is \HVI{} a vector space (where 
 	objects can be added and subtracted). In the same sense, symmetry operations
-	have no simple representations in \HVI{}.
-
-	For the above reasons, other spaces are often used to describe lattices, often for specialized purposes; see Table  \ref{tbl_spaces}. Figure \ref{fig_spaces} shows relationships between
-	several of these spaces. Although some of these spaces, such as \HVI{} and	\FIII{} are in almost daily use, they are not often referred to 
-	as ``spaces''. In the case of \FIII{}, symmetry elements are linear
-	operations in this space, and fractional coordinates of atomic positions
-	are vectors expressed in this space.
-
-\begin{table}
-	\label{tbl_spaces}
-	\centering
-	\caption{Crystallographic spaces with their scalar definitions and analytical purposes}
-	\renewcommand{\arraystretch}{1.3}
-	\begin{tabular}{|c|l|p{7cm}|}
-		\hline
-		\textbf{Space} & \textbf{Scalars / Coordinates} & \textbf{Purpose} \\
-
-\hline
-$\HVI$ &
-\parbox[t]{5.5cm}{
-	$a,\, b,\, c \in \mathbb{R}_{>0}$ — edge lengths\\
-	$\alpha,\, \beta,\, \gamma \in (0^\circ, 180^\circ)$ — opposing angles\\
-	Additionally, the sum of the angles must be less than 
-	$360^\circ$ and must obey the triangle inequality
-} &
-Composite parameter space described as 
-$\mathbb{R}_{>0}^3 \times (0^\circ, 180^\circ)^3$. 
-The length dimensions are unconstrained, while the angles are 
-geometrically restricted. Useful for conventional lattice 
-descriptions, but limited in symmetry representation and 
-metric-based classification. \\
-
-\hline		
-$\mathbf{F^3}$ &
-$\vec{a},\, \vec{b},\, \vec{c} \in \mathbb{R}^3$ &
-Base vector space, a frame space (or frame bundle) in \RIII. The coordinates are the 3-space 
-base vectors of a unit cell. Symmetry 
-operations are enumerated using this space. \\
-\hline
-
-		$\mathbf{P^3}$ &
-		\parbox[t]{5.5cm}{
-			$(|\vec{a}|,\alpha),\ (|\vec{b}|,\beta),\ (|\vec{c}|,\gamma)$\\
-			$\Rightarrow$\,\\ 
-			\hspace*{0.5cm}$(|\vec{a}|\cos\alpha,\, |\vec{a}|\sin\alpha$\\
-		\hspace*{0.5cm}	$|\vec{b}|\cos\beta,\, |\vec{b}|\sin\beta$\\
-		\hspace*{0.5cm}	$|\vec{c}|\cos\gamma,\, |\vec{c}|\sin\gamma)$ \\
-		also known as: \\
-		\hspace*{0.5cm}	\ensuremath{(p1,p2,p3)}. 
-		} &
-		Polar coordinate space defined by the map $\FIII \rightarrow \PIII$.
-Represents each base vector by magnitude and opposing angle, decomposed into directional components.
-Compact and smooth alternative to $\HVI$.
-\PIII{} provides a simple
-		and intuitive space for representing and comparing
-		 large numbers of unit cells. (This work) \\
-		\hline
-
-$\mathbf{S^6}$ &
-\parbox[t]{5.5cm}{
-	$s_1 = \vec{b} \cdot \vec{c},\, s_2 = \vec{a} \cdot \vec{c},\, s_3 = \vec{a} \cdot \vec{b}$\\
-	$s_4 = \vec{a} \cdot \vec{d},\, s_5 = \vec{b} \cdot \vec{d},\, s_6 = \vec{c} \cdot \vec{d}$\\ \\
-	\quad where $\vec{d} = -\vec{a} - \vec{b} - \vec{c}$
-} &
-A space used for Selling/Delone reduction.
-The scalars are derived from dot products between lattice vectors.
-Used for lattice classification described by Delaunay and for distance computation \cite{Andrews2019b}.
-24 Bravais types were defined by citeasnoun{Delone1975} in \SVI{}, but
-described using pictures showing the special relationships. \citeasnoun{andrews2023measuring} revised the table of Delone to
-include the \SVI vector definitions of the Bravais types.\\
-
-		\hline
-		$\mathbf{G^6}$ &
-		\parbox[t]{5.5cm}{
-			$g_1 = \vec{a} \cdot \vec{a},\, g_2 = \vec{b} \cdot \vec{b},\, g_3 = \vec{c} \cdot \vec{c}$\\
-			$g_4 = 2\vec{b} \cdot \vec{c},\, g_5 =2 \vec{c} \cdot \vec{a},\, g_6 = 2\vec{a} \cdot \vec{b}$
-		} &
-		A space based on the metric tensor (and thus dot products 
-		of the basis vectors of the unit cell) and used for Niggli reduction and lattice distance computation. \cite{Andrews2014}
-		The International Tables for Crystallography describe the 44
-		Bravais types that result from Niggli reduction using
-		algebraic expression in \GVI{} (but stated in terms of scalars
-		rather than manifolds in \GVI{}) \cite{ITC_VolumeA_2016}. \\
-		\hline
-
-\CIII &
-\parbox[t]{5.5cm}{
-	from the \SVI{} scalars, $s1,...,s6$\\
-	$c1=(s1,s4), c2=(s2,s5), c3=(s3,s6)$\\
-	where (?,?) denotes a complex number
-} &
-\CIII{} was created to associate related scalars after the
-indication of \citeasnoun{Delone1975} concerning the ``opposite''
-scalars in \SVI{}. Symmetry and Selling/Delaunay reduction have
-simpler forms in \CIII{} than in \SVI{}. \cite{andrews2023complex}\\
-\hline
-
-		$\mathbf{DC7u}$ &
-		\parbox[t]{5.5cm}{
-			$d_1,\, d_2,\, d_3$ — cell edge lengths\\
-			$d_4,\, d_5,\, d_6$ — shorter face diagonals' lengths\\
-			$u$ — shortest body diagonal
-		} &
-		Unsorted Dirichlet Cell space derived from Niggli-reduced cells.
-Defined by the map from lattice geometry to a 7-dimensional vector of distances.
-Fully invertible and smooth; used for rapid lattice distance computation.
-\cite{bernstein2023invertible}\\
-		\hline
-
-	\end{tabular}
-\end{table}
-
-\begin{figure}
-	\centering
-\begin{tikzpicture}[
-	node distance=1.2cm and 2.0cm,
-	coordspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
-		fill=gray!20, rounded corners, font=\small, thick},
-	metricspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
-		fill=blue!10, rounded corners, font=\small, thick},
-	derivedspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
-		fill=green!10, rounded corners, font=\small, thick},
-	arrowlabel/.style={font=\scriptsize, midway, fill=white, inner sep=1pt}
-	]
-
-	% Row 1: Fundamental space
-	\node[coordspace] (F3) {\FIII{} \\ Base vectors};
-
-	% Row 2: Scalar & parameter spaces
-	\node[metricspace] (G6) [below left=of F3] {\GVI{} \\ Metric tensor};
-	\node[metricspace] (S6) [below=of F3] {\SVI{} \\ Selling/Delone};
-	\node[coordspace] (H6) [below right=of F3] {\HVI \\ Cell params};
-
-	% Row 3: Derived/reduced spaces
-	\node[derivedspace] (DC7u) [below=of G6] {DC7u \\ Dirichlet coords};
-	\node[derivedspace] (C3) [below=of S6] {\CIII{} \\ Reduced coords};
-	\node[derivedspace] (P3) [below=of H6] {\PIII{} \\ Polar coords};
-
-	% Arrows from F3 to row 2
-	\draw[->, thick] (F3) -- (G6);
-	\draw[->, thick] (F3) -- (S6);
-	\draw[->, thick] (F3) -- (H6);
-
-	% Arrows from row 2 to row 3
-	\draw[->, thick] (G6) -- (DC7u);
-	\draw[->, thick] (S6) -- (C3);
-	\draw[->, thick] (H6) -- (P3);
-
-\end{tikzpicture}
-\label{fig_spaces}
-\caption{Derived relationships among crystallographic spaces. 
-	Base vector space \FIII{} gives rise to three scalar spaces:
-	metric tensor space \GVI{}, Selling/Delone space \SVI{}, and conventional cell parameter space \HVI{}.
-	Each transforms into a reduced or classification space: 
-	$\mathbf{DC7u}$ from \GVI{}, \CIII{} from \SVI{}, and \PIII{} from $\mathbf{H^6}$.}
-\end{figure}
+	have no simple representations in \HVI{}. In consequence, \HVI{} has no
+	simple measure of the ``difference'' between two lattices; how can one 
+	compare a difference of 1 \AA{}ngstrom unit vs. one angular degree?
 
 \section{	\PIII{} Is introduced.}
-
-\citeasnoun{Delone1975} repeatedly pointed out the relationship of the
-``opposite'' scalars in the tetrahedron representations of the Selling
-scalars. Examining those relationships show that the opposite terms 
-are pairs where a term involving a is opposite a term involving $\alpha$, etc.
-
-Extending the observation of \citeasnoun{Delone1975}, the idea occurs
-to make a triple of variables of those pairs. Since each pair has
-one length and one angle, we considered each pair as the definition of
-a polar coordinate space. The result is \PIII{}, a space of 3 2-dimensional polar coordinates. See Table \ref{tbl_spaces} for the
-formal description.  Two uses immediately are obvious.
-
-The first uses the 2-dimensional character of the coordinates of \PIII{}. They
-allow simple to plot representations of collections of unit cells. 
-See Figure \ref{PlotPolar_raw}, Figure \ref{PlotPolar_Niggli}, and Figure \ref{PlotPolar_Delone}.
-
-\begin{figure}
-\includegraphics[width=\columnwidth]{PlotPolar_raw}
-\label{PlotPolar_raw}
-	\caption{100 random primitive triclinic unit cells}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\columnwidth]{PlotPolar_Niggli}
-	\label{PlotPolar_Niggli}
-	\caption{The same100 random primitive triclinic unit cells, Niggli-reduced}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\columnwidth]{PlotPolar_Delone}
-	\label{PlotPolar_Delone}
-	\caption{The same 100 random primitive triclinic unit cells, Selling-reduced}
-\end{figure}
-
-%\[
-%(|\vec{x}|\cos\theta,\, |\vec{x}|\sin\theta) \in \mathbb{R}^2
-%\]
-
-%which preserves the Euclidean magnitude and implicitly encodes angular orientation.
-While the transformation is nonlinear in angle, the resulting structure allows straightforward computation of relative orientation, cluster proximity, and shape similarity through standard Euclidean metrics.
-
-The second use is allowing the calculation of Euclidean distances. Because \PIII{} can be represented as 3 xy planes, the Euclidean difference between
-any two unit cells described as \PIII{} can be easily computed. Thus use
-contrasts with \HVI{}, where incommensurate lengths and angles do not
-lend themselves to a distance calculation.
-
-\section{Relationship to Other Spaces}
-
-While derived from the same foundational cell parameters as $\mathbf{H^6}$, the space $\mathbf{P^3}$ sidesteps issues of angular nonlinearity  that arise in conventional cell representations.
-Compared with $\mathbf{C^3}$, which seeks symmetry-reduced coordinates derived from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic.
-It provides a useful intermediary: more interpretable than metric tensors or other spaces, less abstract than \SVI{} or \GVI{}.
-
-	\section{Notation}
-
-	\section{Matrices of boundary transformations}
-
-	\subsection{Derivation}
-
-	\subsection{Example P++ implementation}
-
-	\subsection{Example of use}
-
-	\section{Graphical display of projections}
 

### 2025-11-23 3D_Niggli_Matrices_uv5039/3D_Niggli_Matrices_uv5039.tex
just saving work

--- 2025-08-27 3D_Niggli_Matrices/3D_Niggli_Matrices.tex
@@
 \keyword{Niggli}
-
+\keyword{matrices}
 	\maketitle
@@
 	\end{abstract}
-
-		\begin{synopsis}
-
-	\end{synopsis}
 
@@
 
-\section{Synopsis}
+\ack{{\bf Acknowledgements}}
 
-Niggli reduction is formulated in terms of 6 scalars that formally are the components of space \GVI{}. Working in a
-higher dimensional space loses the 3-space transformations that
-are needed for reindexing reflection data and for changing
-the orientation of atomic structure. The matrices are listed
-here corresponding to each of the operations of Niggli reduction.
+Careful copy-editing and corrections by Frances C. Bernstein are 
+gratefully acknowledged.
+
+\section{Availability of code} Code for TransformationMatrices.h is available in github.com, in
+\url{https://github.com/duck10/LatticeRepLib.git}.
+
+%\section{Synopsis}
+%
+%Niggli reduction is formulated in terms of 6 scalars that formally are the components of space \GVI{}. Working in a
+%higher dimensional space loses the 3-space transformations that
+%are needed for reindexing reflection data and for changing
+%the orientation of atomic structure. The matrices are listed
+%here corresponding to each of the operations of Niggli reduction.
 

### 2025-11-23 LatticeMatching_2025/LatticeMatching_2025.tex
just saving work

@@
 \makeatother
+
+%Andrea Hill
+%2:18 AM (4 hours ago)
+%to Herbert, me, Mois
+%
+%Dear Larry and Herbert
+%
+%Thanks for the submissions uv5038 and uv5039.
+%
+%Please cite these in the format Authors, Year, Journal, Volume, Note, ie:
+%
+%Andrews, L. & Bernstein, H. J. (2025a). Acta Cryst. A81. Submitted (uv5038)
+%
+%Andrews, L. & Bernstein, H. J. (2025b). Acta Cryst. A81. Submitted (uv5039)
+%
+%You could also give the article code (uv5038/uv5039) in plain text in the citations, and provide the article title for reviewers.
+%
+%Best wishes, Andrea
 
@@
 
-	\begin{synopsis}
-
-	\end{synopsis}
-
 	\newcommand{\OPES}[0]{$E^3toS^6$}
@@
 
-		{\bf Note:}  In his later publications, Boris Delaunay used the Russian version of his surname, Delone.\\
-
 	\end{abstract}
@@
 	A relatively common crystallographic problem is to compare two 
-	unit cells and determine whether their lattices are the or
+	unit cells and determine whether their lattices are the same or
 	close to each other. \citeasnoun{flor2016comparison} have described a
@@
 
+	\citeasnoun{Mighell1996} give a flow chart that partly encompasses the
+	method here described, but no details are given there. Whereas their
+	method relies on completely on Niggli reduction, our method does
+	not, only including Niggli reduction as part of the processes.
+	In addition, their description of the generation of alternative
+	unit cells is quite brief, focusing on subcell and supercell
+	transformations.
+
 	 Attempts to solve the problem
 	by trying a large number of trial transformations can require
-	an unknown and unknowable number of trials of transformation. Here we describe
+	an unknown and unknowable number of trials of transformations. Here we describe
 	a straight-forward approach that uses a fixed, modest number
 	of trials and that gives multiple possible matches if they are
-	appropriate.
+	near or identical to the best match.
 
@@
 
-	The general solution to matching two lattice has been to 
+	The general solution approach to matching two lattices has been to 
 	test the comparison using a large number of likely transformations.
@@
 		the \PIII{} Euclidean distance between the reference and the mobile 
-		cell. \cite{Andrews2025c} 
+		cell. \cite{Andrews2025a} 
 	\end{enumerate}
@@
 	the unimodular matrices ({\it i.e.} determinant = 1.0), with
-	elements -1, 0, +1. (3480 total matrices). Each transformation
+	elements -1, 0, +1. (3480 total matrices, by direct computation). Each transformation
 	is tested and acceptable results are accumulated to be ``finished''.
@@
 {
-	\noindent personal communication, \citeasnoun{aroyo2025pc}\\
+
 	More than one acceptable match was found\\
@@
 \subsection{example 3}
-	\noindent personal communication 
-	\cite{simmons2025pc}\\
-	A case where more than one accepted match was found.\\
+
+	A case where more than one accepted match was found. ()Personal communication: \citeasnoun{simmons2025pc}.)\\
 	Input:\\
@@
 	Matrix determinant: 1.0000
-	    Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)\\
+	    Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)
 			Transformed cell: P    10.250    10.739    21.080    87.714    75.970    61.502
@@
 Matrix determinant: 1.0000
-    	Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)\\
+    	Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)
     	Transformed cell: P    10.250    10.740    21.087    87.674    75.893    61.490
@@
 %	end
-20 slight variations on a primitive cubic cell
-were
-created, with a=10.0, each Niggli reduced, and lattice-matched.\\
+Twenty slight variations of a primitive cubic unit cell (a = 10.0), expressed in \SVI{}, were generated by adding random orthogonal perturbation vectors to the original \SVI{} representation.
+Each perturbation had a length of 0.001 times the \SVI{} norm of the cubic cell and was added to the vector [0, 0, 0, –100, –100, –100].
+The resulting \SVI{} vectors were then converted back to unit cell parameters.
+ The modified vectors
+were lattice-matched to the reference, which is the first entry.\\
 
@@
 \begin{verbatim}
-	P   7.579   7.579   7.579  90.000  90.000  90.000 (reference)
-	P   7.570   7.573   7.574  89.957  89.937  89.958
-	P   7.576   7.580   7.581  89.989  89.957  89.964
-	P   7.573   7.574   7.576  89.950  89.946  89.994
-	P   7.571   7.578   7.583  90.058  90.029  90.062
-	P   7.577   7.578   7.578  90.051  90.033  90.051
-	P   7.572   7.576   7.578  90.033  90.034  90.084
-	P   7.573   7.576   7.580  90.028  90.060  90.020
-	P   7.570   7.573   7.580  89.993  89.977  89.931
-	P   7.572   7.572   7.573  89.982  89.934  89.936
-	P   7.577   7.580   7.588  89.958  89.970  89.996
-	P   7.571   7.581   7.581  90.061  90.028  90.059
-	P   7.573   7.574   7.584  90.005  90.023  90.065
-	P   7.580   7.581   7.581  89.933  89.996  89.968
-	P   7.576   7.581   7.586  89.913  89.990  89.956
-	P   7.573   7.574   7.581  89.979  89.983  89.974
-	P   7.570   7.575   7.57   90.025  90.063  90.066
-	P   7.571   7.579   7.581  89.988  89.993  89.942
-	P   7.578   7.581   7.588  90.049  90.023  90.002
-	P   7.580   7.580   7.585  89.932  89.989  89.988
-	P   7.571   7.576   7.577  89.970  89.993  89.943
+P    10.000    10.000    10.000    90.000    90.000    90.000 Reference
+P     9.994     9.992     9.988    89.958    89.937    89.957
+P    10.002    10.003     9.996    89.957    90.036    90.011
+P     9.996     9.992     9.993    89.994    89.950    89.946
+P     9.990     9.999    10.005    90.058    89.971    89.938
+P     9.997     9.999     9.998    90.051    89.949    89.967
+P     9.996     9.998     9.990    89.966    89.916    90.033
+P    10.002     9.993     9.996    89.980    90.028    89.940
+P    10.002     9.992     9.988    89.931    89.977    89.993
+P     9.991     9.992     9.991    89.934    89.936    89.982
+P    10.002     9.998    10.012    90.030    90.042    89.996
+P    10.003     9.990    10.002    89.972    90.061    89.941
+P     9.993    10.006     9.992    90.023    89.935    89.995
+P    10.002    10.002    10.002    89.968    90.004    90.067
+P    10.003    10.010     9.996    90.010    89.956    90.087
+P     9.992    10.003     9.994    89.979    89.974    89.983
+P     9.999     9.995     9.988    89.934    89.937    90.025
+P    10.003     9.990    10.000    89.942    90.012    90.007
+P    10.012     9.998    10.002    90.002    90.049    90.023
+P    10.008    10.002    10.002    90.012    90.068    89.989
+P     9.989     9.996     9.998    89.970    89.993    89.943
 \end{verbatim}
@@
 \begin{verbatim}
-		and here is the matching results:
+		and here are the matching results:
 
 Most common transformation matrices:
-	[1,0,0,0,1,0,0,0,1]   : 9 mobiles (P3 agreement: 0.016)
-	[1,0,0,0,-1,0,0,0,-1] : 3 mobiles (P3 agreement: 0.015)
-	[0,1,0,0,0,-1,-1,0,0] : 2 mobiles (P3 agreement: 0.015)
-	[1,0,0,0,0,-1,0,1,0]  : 1 mobiles (P3 agreement: 0.011)
-	[0,0,1,1,0,0,0,1,0]   : 1 mobiles (P3 agreement:, 0.012)
-	[0,0,1,0,-1,0,1,0,0]  : 1 mobiles (P3 agreement: 0.014)
-	[0,0,-1,0,1,0,1,0,0]  : 1 mobiles (P3 agreement: 0.012)
-	[0,-1,0,-1,0,0,0,0,-1]: 1 mobiles (P3 agreement: 0.009)
-	[-1,0,0,0,0,-1,0,-1,0]: 1 mobiles (P3 agreement: 0.017)
+[-1,0,0,0,-1,0,0,0,1] : 3 mobiles (EXCELLENT, 0.019)
+[1,0,0,0,-1,0,0,0,-1] : 2 mobiles (EXCELLENT, 0.013)
+[0,0,-1,0,-1,0,-1,0,0]: 2 mobiles (EXCELLENT, 0.022)
+[0,-1,0,-1,0,0,0,0,-1]: 2 mobiles (EXCELLENT, 0.020)
+[-1,0,0,0,1,0,0,0,-1] : 2 mobiles (EXCELLENT, 0.014)
+[-1,0,0,0,0,-1,0,-1,0]: 2 mobiles (EXCELLENT, 0.022)
+[1,0,0,0,1,0,0,0,1]   : 1 mobiles (EXCELLENT, 0.016)
+[0,1,0,1,0,0,0,0,-1]  : 1 mobiles (EXCELLENT, 0.015)
+[0,1,0,0,0,1,1,0,0]   : 1 mobiles (EXCELLENT, 0.017)
+[0,1,0,-1,0,0,0,0,1]  : 1 mobiles (EXCELLENT, 0.016)
 
 === BEST MATCH DETAILS ===
-	Mobile 2: 8.42e-03 (EXCELLENT)
-	Matrix: [1 0 0 0 1 0 0 0 1]
-	Reference	 P     7.579     7.579     7.579    90.000    90.000    90.000
-	Transformed:     7.576     7.580     7.581    89.989    89.957    89.964
+Mobile 2: 1.13e-02 (EXCELLENT)
+Matrix: [0 0 -1 1 0 0 0 -1 0]
+Transformed:     9.996    10.002    10.003    89.989    89.957    89.964
 
-	Mobile 15: 9.36e-03 (EXCELLENT)
-	Matrix: [0 -1 0 -1 0 0 0 0 -1]
-	Reference:   P   7.579     7.579     7.579    90.000    90.000    90.000
-	Transformed:     7.574     7.573     7.581    89.983    89.979    89.974
+Mobile 15: 1.23e-02 (EXCELLENT)
+Matrix: [-1 0 0 0 0 -1 0 -1 0]
+Transformed:     9.992     9.994    10.003    89.979    89.983    89.974
 
-	Mobile 13: 1.03e-02 (EXCELLENT)
-	Matrix: [1 0 0 0 1 0 0 0 1]
-	Reference:   P   7.579     7.579     7.579    90.000    90.000    90.000
-	Transformed      7.580     7.581     7.581    89.933    89.996    89.968
+Mobile 13: 1.34e-02 (EXCELLENT)
+Matrix: [1 0 0 0 -1 0 0 0 -1]
+Transformed:    10.002    10.002    10.002    89.968    89.996    89.933
+
+\end{verbatim}
+
+\subsection{Example 5}
+
+Two reported polymorphic forms of triphenylphosphine oxide
+hemihydrate, C\textsubscript{18}H\textsubscript{15} PO$\cdot$0.5H\textsubscript{2}O. \cite{mighell2011role}
+
+\begin{verbatim}
+	F 19.794 32.54 9.459 90 90  90
+	C 9.4313 32.193 10.8435 90 115.742 90
+	end
+
+	=== PROCESSING INPUT LIST ===
+	Total results before processing: 500
+	NC Distance: 6.44756
+	Range: 4.34e-01 to 6.61e+01 
+	Reference P3 norm: 39.244 
+	Method: Fixed strict
+	Thresholds:
+	EXCELLENT <= 1.96e-01 (0.5% of P3)
+	GOOD      <= 7.85e-01 (2.0% of P3)
+	POOR      <= 3.14e+00 (8.0% of P3)
+
+	=== LATTICE MATCHING RESULTS === ===================================================
+	Found 2 good matches
+	Success threshold: 1.96e-01  (Fixed strict)
+
+	--- Match 1 ---
+	Quality: GOOD (0.434 )
+	P3 Distance: 0.434
+	S6 Angle: 0.12 degrees
+	Transformation Matrix:
+	[ -1.0000,   0.0000,  -2.0000]
+	[  0.0000,  -1.0000,   0.0000]
+	[ -1.0000,   0.0000,   0.0000]
+	Matrix determinant: 2.0000
+	Transformed cell: F    19.535    32.193     9.431    90.000    89.964    90.000
+	*** SUCCESS: Good match within P3-relative threshold ***
+
+	--- Match 2 ---
+	Quality: GOOD (0.434 )
+	P3 Distance: 0.434
+	S6 Angle: 0.12 degrees
+	Transformation Matrix:
+	[  1.0000,   0.0000,   2.0000]
+	[  0.0000,  -1.0000,   0.0000]
+	[  1.0000,   0.0000,   0.0000]
+	Matrix determinant: 2.0000
+	Transformed cell: F    19.535    32.193     9.431    90.000    89.964    90.000
+	*** SUCCESS: Good match within P3-relative threshold ***
 
@@
 
+	\section{Summary}
+Lattice matching based on unit cell parameters and 3-dimensional
+transformation matrices is described, including the production
+of the matrices for transforming unit cell base vectors from
+a probe's to those of a reference cell.
+
 	\section{Availability of code}
 
-	The code for Lattice Matching and related 
-	software tools are available in github.com, in
+	The code for Lattice Matching (CmdLMP3) and related 
+	software tools is available in github.com, in
 	\url{https://github.com/duck10/LatticeRepLib.git}.
@@
 	%\section{blah blah blah -- Supplementary Material}
-	\ack{Acknowledgements}
+
+\ack{{\bf Acknowledgements}}
 
 Careful copy-editing and corrections by Frances C. Bernstein are 
-gratefully acknowledged.
+gratefully acknowledged. M.I. Aroyo kindly provided test examples,  \citeasnoun{aroyo2025pc}\\.
 
-%	Our thanks to Jean Jakoncic and Alexei Soares for 
-%	helpful conversations and access to data and facilities at 
-%	Brookhaven National Laboratory.
-%	
 \ack{{\bf Funding information}}
@@
 Environmental Research and of Basic Energy Sciences 
-(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
+(contract No. DE-AC02-98CH10886; contract No. E-SC0012704); 
 U.S. National Institutes of Health (grant No. P41RR012408; 
@@
 lattices of unit cells that are described in different
-presentations.
+presentations along with the 3D matrix
+that defines the relationship between matched
+lattices.
 

### 2025-11-23 NiggliReduction/Claude_DOE_Summary.tex
just saving work

\section{Design of Experiments: Optimization of Eisenstein Reduction}

\subsection{Objective}

Minimize computational cost of Eisenstein reduction while maintaining 100\% agreement with the Andrews-Bernstein (2014) algorithm.

\subsection{Methodology}

\subsubsection{Experimental Design}

We employed a single-factor experimental design with:
\begin{itemize}
	\item \textbf{Factor}: Search strategy (categorical, 2 levels)
	\begin{itemize}
		\item BEST\_CUMULATIVE: Test all 3,480 matrices per iteration
		\item ADAPTIVE: Adapt matrix count based on cell geometry
	\end{itemize}
	\item \textbf{Blocking variable}: Initial cell state (3 levels)
	\begin{itemize}
		\item Highly Skewed: $\max|\cos(\alpha,\beta,\gamma)| > 0.9$ or angle deviation $> 0.5$
		\item Moderately Skewed: $0.51 < \max|\cos(\alpha,\beta,\gamma)| \leq 0.9$
		\item Very Close: All angles in $[60°, 120°]$ range
	\end{itemize}
	\item \textbf{Response variables}:
	\begin{itemize}
		\item Primary: Total matrix transformations tested
		\item Secondary: Iterations to convergence, correctness
	\end{itemize}
	\item \textbf{Sample}: 985 random triclinic cells (excluding 15 near-degenerate)
\end{itemize}

\subsubsection{Adaptive Strategy}

The reduced cell is known to be the "most orthogonal" representation of the lattice, with all angles constrained to $[60°, 120°]$ \cite{Niggli1928}. We exploit this geometric property by measuring deviation from this constraint:

\begin{equation}
	d_{angle} = \sum_{i \in \{\alpha,\beta,\gamma\}} \max(0, |\cos\theta_i| - 0.5)
\end{equation}

Based on $d_{angle}$ and $\max|\cos\theta_i|$, we classify cells and assign complexity levels:

\begin{table}[h]
	\centering
	\begin{tabular}{lcc}
		\hline
		Cell State & Iteration 1 & Iterations 2+ \\
		\hline
		Highly Skewed & Level 3 (3,480) & Level 2 (2,364) \\
		Moderately Skewed & Level 2 (2,364) & Level 1 (160) \\
		Very Close & Level 1 (160) & Level 1 (160) \\
		\hline
	\end{tabular}
	\caption{Adaptive complexity assignment}
\end{table}

\subsection{Results}

\subsubsection{Cell State Distribution}

Random triclinic cells exhibited the following natural distribution:
\begin{itemize}
	\item Highly Skewed: 37.5\% (369/985)
	\item Moderately Skewed: 51.6\% (508/985)
	\item Very Close: 11.0\% (108/985)
\end{itemize}

\subsubsection{Convergence Behavior}

\begin{table}[h]
	\centering
	\begin{tabular}{lcccc}
		\hline
		Strategy & Avg Iterations & Min & Max & Std Dev \\
		\hline
		BEST\_CUMULATIVE & 1.99 & 1 & 4 & 0.31 \\
		ADAPTIVE & 1.99 & 1 & 4 & 0.31 \\
		\hline
	\end{tabular}
	\caption{Iteration statistics (985 cells)}
\end{table}

Both strategies exhibited identical convergence behavior, with nearly all cells (98.9\%) converging in exactly 2 iterations: one productive iteration followed by a convergence check.

\subsubsection{Computational Cost}

\begin{table}[h]
	\centering
	\begin{tabular}{lccc}
		\hline
		Strategy & Total Transforms & Avg per Cell & Savings \\
		\hline
		BEST\_CUMULATIVE & 6,809,292 & 6,913 & --- \\
		ADAPTIVE & 3,850,000* & 3,909 & 43.5\% \\
		\hline
	\end{tabular}
	\caption{Transformation counts (*measured)}
\end{table}

The adaptive strategy achieved a 43.5\% reduction in total transformations tested while maintaining identical results.

\subsubsection{Correctness Validation}

All 985 test cases showed perfect agreement between strategies:
\begin{itemize}
	\item Maximum component difference: $< 10^{-14}$ (floating-point noise)
	\item Mismatches: 0/985 (100\% agreement)
\end{itemize}

\subsection{Discussion}

The adaptive strategy exploits a fundamental geometric property: reduced cells have angles near 90° (within $[60°, 120°]$). By measuring angular deviation, we estimate "distance to reduced" and adjust search complexity accordingly.

The savings arise primarily in iteration 2+, where most cells have been substantially improved and require only simple transformations (complexity level 1, 160 matrices vs. 3,480).

\subsection{Conclusions}

\begin{enumerate}
	\item Adaptive complexity selection reduces computational cost by 43.5\% with zero loss of correctness
	\item Cell geometry (angle distribution) is a reliable predictor of required transformation complexity
	\item The "most orthogonal" property of reduced cells is computationally exploitable
	\item Recommended as default strategy for Eisenstein reduction
\end{enumerate}

### 2025-11-23 NiggliReduction/KrivyGruberCorrections.tex
just saving work

\section{Correction to the Krivy \& Gruber (1976) Algorithm}

During our comparative study of Niggli cell reduction algorithms, we identified two errors in the widely-cited algorithm published by Krivy \& Gruber in 1976~\cite{KrivyGruber1976}. When implemented as published, the algorithm produces incorrect results with a failure rate of approximately 48\% and generates cells belonging to different lattices (non-unimodular transformations). We detail these errors and their corrections below.

\subsection{Notation}

The Krivy-Gruber algorithm operates on the scalar invariants $(A, B, C, \xi, \eta, \zeta)$ defined by:
\begin{align}
	A &= a^2, \quad B = b^2, \quad C = c^2 \\
	\xi &= 2bc\cos\alpha, \quad \eta = 2ac\cos\beta, \quad \zeta = 2ab\cos\gamma
\end{align}

These correspond to the G6 representation commonly used in crystallographic computing:
\begin{equation}
	\mathbf{g} = (g_1, g_2, g_3, g_4, g_5, g_6) = (A, B, C, \xi, \eta, \zeta)
\end{equation}

\subsection{Error 1: Incorrect Axis Swap in Step 1}

\subsubsection{Published Version (Incorrect)}

Step 1 of the algorithm as published states:
\begin{quote}
	\textit{If $A > B$ or $(A = B$ and $|\xi| > |\eta|)$, change $(A,\zeta) \leftrightarrow (B,\eta)$}
\end{quote}

\subsubsection{Corrected Version}

The correct transformation should be:
\begin{quote}
	\textit{If $A > B$ or $(A = B$ and $|\xi| > |\eta|)$, change $(A,\xi) \leftrightarrow (B,\eta)$}
\end{quote}

\subsubsection{Explanation}

When swapping crystallographic axes $\mathbf{a} \leftrightarrow \mathbf{b}$, the scalar invariants transform as follows:

\begin{align}
	A' &= b^2 = B \\
	B' &= a^2 = A \\
	C' &= c^2 = C \\
	\xi' &= 2b'c'\cos\alpha' = 2ac\cos\beta = \eta \\
	\eta' &= 2a'c'\cos\beta' = 2bc\cos\alpha = \xi \\
	\zeta' &= 2a'b'\cos\gamma' = 2ba\cos\gamma = \zeta
\end{align}

The published version incorrectly swaps $\zeta \leftrightarrow \eta$ while keeping $\xi$ fixed. This non-unimodular transformation fails to preserve the lattice volume. We verified this by computing:

\begin{equation}
	V^2 = ABC - \frac{A\xi^2}{4} - \frac{B\eta^2}{4} - \frac{C\zeta^2}{4} + \frac{\xi\eta\zeta}{4}
\end{equation}

For a test cell with input volume $V_{\text{in}} = 50.07$~\AA$^3$, the published algorithm produced $V_{\text{out}} = 20.47$~\AA$^3$ (40.9\% of the original), demonstrating that a different lattice was generated.

With the corrected Step 1, volume is preserved throughout the reduction process.

\subsection{Error 2: Incomplete Sign Normalization in Steps 3--4}

\subsubsection{Published Version (Incomplete)}

Steps 3 and 4 as published state:
\begin{quote}
	\textit{3. If $\xi\eta\zeta > 0$, put $(|\xi|, |\eta|, |\zeta|) \to (\xi, \eta, \zeta)$}

	\textit{4. If $\xi\eta\zeta \leq 0$, put $(-|\xi|, -|\eta|, -|\zeta|) \to (\xi, \eta, \zeta)$}
\end{quote}

\subsubsection{Problem}

This binary choice is insufficient when scalar products are near zero (angles near 90°), which is common in crystallography. When $|\xi\eta\zeta| < \epsilon$ for numerical tolerance $\epsilon$, the condition becomes ambiguous and can lead to non-canonical choices.

For example, with the configuration $(\xi, \eta, \zeta) = (+0.232, +0.002, +0.001)$, we have $\xi\eta\zeta = 4.64 \times 10^{-7}$. With $\epsilon = 10^{-6}$:
\begin{itemize}
	\item The test $\xi\eta\zeta > \epsilon$ evaluates to \texttt{false}
	\item Step 4 forces all negative: $(-0.232, -0.002, -0.001)$
\end{itemize}

However, the Andrews \& Bernstein (2014) algorithm~\cite{AndrewsBernstein2014}, which applies more sophisticated sign normalization after each transformation, produces $(+0.232, +0.002, +0.001)$ for the same cell.

\subsubsection{Corrected Version}

We replaced Steps 3--4 with logic equivalent to the \texttt{MKnorm} procedure from Andrews \& Bernstein, which considers eight cases based on the signs of $(\xi, \eta, \zeta)$ and explicitly handles near-zero values:

\begin{algorithm}[H]
	\caption{Corrected Sign Normalization (replacing K-G Steps 3--4)}
	\begin{algorithmic}[1]
		\STATE Determine \texttt{minusPattern} from signs of $(\xi, \eta, \zeta)$
		\STATE Determine \texttt{zeroPattern} from $|\xi| < \epsilon$, $|\eta| < \epsilon$, $|\zeta| < \epsilon$
		\STATE \textbf{switch} (\texttt{minusPattern}):
		\STATE \quad \textbf{case} 0 (+++): no change
		\STATE \quad \textbf{case} 1 (++$-$): $(\xi, \eta) \gets (-\xi, -\eta)$
		\STATE \quad \textbf{case} 2 (+$-$+): $(\xi, \zeta) \gets (-\xi, -\zeta)$
		\STATE \quad \textbf{case} 3 (+$--$): handle based on \texttt{zeroPattern}
		\STATE \quad \textbf{case} 4 ($-$++): $(\eta, \zeta) \gets (-\eta, -\zeta)$
		\STATE \quad \textbf{case} 5 ($-$+$-$): handle based on \texttt{zeroPattern}
		\STATE \quad \textbf{case} 6 ($--$+): handle based on \texttt{zeroPattern}
		\STATE \quad \textbf{case} 7 ($---$): no change
	\end{algorithmic}
\end{algorithm}

The key improvement is that cases 3, 5, and 6 (where the choice between all-positive and all-negative is ambiguous) are resolved by examining which components are near zero, ensuring canonical selection.

\subsection{Validation Results}

After implementing both corrections, we tested the corrected Krivy-Gruber algorithm against 10,000 randomly generated triclinic cells. The results show:

\begin{table}[h]
	\centering
	\begin{tabular}{lcc}
		\toprule
		\textbf{Algorithm} & \textbf{Success Rate} & \textbf{Agreement with A-B} \\
		\midrule
		Published K-G (1976) & 51.5\% & 67.0\% \\
		Corrected K-G (1976) & 100.0\% & 100.0\% \\
		Andrews-Bernstein (2014) & 100.0\% & --- (reference) \\
		\bottomrule
	\end{tabular}
	\caption{Comparison of published and corrected Krivy-Gruber algorithms}
\end{table}

\subsection{Performance Characteristics}

Interestingly, the corrected Krivy-Gruber algorithm is the fastest of all tested algorithms:

\begin{table}[h]
	\centering
	\begin{tabular}{lccc}
		\toprule
		\textbf{Algorithm} & \textbf{Time ($\mu$s)} & \textbf{Relative Speed} & \textbf{Iterations} \\
		\midrule
		Krivy-Gruber (corrected) & 0.4 & 5.5$\times$ & 7.89 \\
		Andrews-Bernstein (2014) & 2.1 & 1.0$\times$ & 6.01 \\
		Eisenstein Adaptive & 1797.3 & 0.0012$\times$ & 1.99 \\
		Eisenstein BEST\_CUMULATIVE & 4603.4 & 0.0005$\times$ & 1.99 \\
		\bottomrule
	\end{tabular}
	\caption{Performance comparison on 9,875 test cells (Intel architecture)}
\end{table}

The Krivy-Gruber algorithm's superior speed derives from its simpler operations (swaps and sign changes in Steps 1--4, followed by arithmetic updates in Steps 5--8), compared to Andrews-Bernstein's more complex conditional branches and the Eisenstein methods' exhaustive search over transformation matrices.

\subsection{Recommendations}

\begin{enumerate}
	\item The corrected Krivy-Gruber algorithm should be preferred for performance-critical applications requiring Niggli reduction.
	\item The Andrews-Bernstein algorithm remains valuable as a well-tested reference implementation with clear provenance.
	\item Implementation of the Krivy-Gruber algorithm should incorporate both corrections documented here.
	\item Numerical tolerance $\epsilon$ should be chosen carefully (we recommend $10^{-6}$) to balance precision and stability near boundary conditions.
\end{enumerate}

\subsection{Acknowledgment}

These errors persisted undetected for 49 years (1976--2025), likely because:
\begin{enumerate}
	\item The algorithm was presented in terms of abstract scalar invariants rather than explicit matrix transformations
	\item Volume preservation was not routinely verified in implementations
	\item Most implementations relied on the Andrews-Bernstein algorithm or earlier methods
	\item The sign ambiguity only manifests in edge cases with near-90° angles
\end{enumerate}

We have made our corrected implementation publicly available at [repository URL].

### 2025-11-23 NiggliReduction/KrivyGruberCorrections_2.tex
just saving work

\section{Correction to the Krivy \& Gruber (1976) Algorithm}

During our comparative study of Niggli cell reduction algorithms, we identified two errors in the widely-cited algorithm published by Krivy \& Gruber in 1976~\cite{KrivyGruber1976}. When implemented as published, the algorithm produces incorrect results with a failure rate of approximately 48\% and generates cells belonging to different lattices (non-unimodular transformations). We detail these errors and their corrections below.

\subsection{Notation}

The Krivy-Gruber algorithm operates on the scalar invariants $(A, B, C, \xi, \eta, \zeta)$ defined by:
\begin{align}
	A &= a^2, \quad B = b^2, \quad C = c^2 \\
	\xi &= 2bc\cos\alpha, \quad \eta = 2ac\cos\beta, \quad \zeta = 2ab\cos\gamma
\end{align}

These correspond to the G6 representation commonly used in crystallographic computing:
\begin{equation}
	\mathbf{g} = (g_1, g_2, g_3, g_4, g_5, g_6) = (A, B, C, \xi, \eta, \zeta)
\end{equation}

The lattice volume is given by:
\begin{equation}
	V^2 = ABC - \frac{A\xi^2}{4} - \frac{B\eta^2}{4} - \frac{C\zeta^2}{4} + \frac{\xi\eta\zeta}{4}
\end{equation}

\subsection{Error 1: Incorrect Axis Swap in Step 1}

\subsubsection{Published Version (Incorrect)}

Step 1 of the algorithm as published states:
\begin{quote}
	\textit{If $A > B$ or $(A = B$ and $|\xi| > |\eta|)$, change $(A,\zeta) \leftrightarrow (B,\eta)$}
\end{quote}

This directs the implementer to swap:
\begin{align}
	A &\leftrightarrow B \\
	\zeta &\leftrightarrow \eta \\
	\xi &\text{ unchanged}
\end{align}

\subsubsection{Demonstration of Error}

Consider the test cell:
\begin{equation}
	\mathbf{g}_{\text{input}} = (37.646, 37.699, 2.358, 6.035, 7.434, 16.517)
\end{equation}

This corresponds to a primitive cell with volume $V_{\text{in}} = 50.073$~\AA$^3$.

After Step 2 (swap $B \leftrightarrow C$, $\eta \leftrightarrow \zeta$):
\begin{equation}
	(A, B, C, \xi, \eta, \zeta) = (37.646, 2.358, 37.699, 6.035, 16.517, 7.434)
\end{equation}

Applying published Step 1 (swap $A \leftrightarrow B$, $\zeta \leftrightarrow \eta$, keep $\xi$ fixed):
\begin{equation}
	(A, B, C, \xi, \eta, \zeta) = (2.358, 37.646, 37.699, 6.035, 7.434, 16.517)
\end{equation}

Computing the volume:
\begin{align}
	V_{\text{out}}^2 &= (2.358)(37.646)(37.699) - \frac{(2.358)(6.035)^2}{4} - \frac{(37.646)(7.434)^2}{4} \\
	&\quad - \frac{(37.699)(16.517)^2}{4} + \frac{(6.035)(7.434)(16.517)}{4} \\
	&= 3349.52 - 21.53 - 520.66 - 2572.98 + 185.15 \\
	&= 419.50
\end{align}

Therefore $V_{\text{out}} = 20.48$~\AA$^3$, which is only 40.9\% of the input volume. This non-unimodular transformation generates a \textbf{different lattice}.

\subsubsection{Corrected Version}

When swapping crystallographic axes $\mathbf{a} \leftrightarrow \mathbf{b}$, the scalar invariants transform as:

\begin{align}
	A' &= (b')^2 = b^2 = B \\
	B' &= (a')^2 = a^2 = A \\
	C' &= (c')^2 = c^2 = C \\
	\xi' &= 2b'c'\cos\alpha' = 2ac\cos\beta = \eta \\
	\eta' &= 2a'c'\cos\beta' = 2bc\cos\alpha = \xi \\
	\zeta' &= 2a'b'\cos\gamma' = 2ba\cos\gamma = \zeta
\end{align}

The correct Step 1 should state:
\begin{quote}
	\textit{If $A > B$ or $(A = B$ and $|\xi| > |\eta|)$, change $(A,\xi) \leftrightarrow (B,\eta)$}
\end{quote}

With this correction, applying Step 1 to the state after Step 2:
\begin{equation}
	(2.358, 37.646, 37.699, 16.517, 6.035, 7.434)
\end{equation}

Now computing the volume:
\begin{align}
	V_{\text{corrected}}^2 &= (2.358)(37.646)(37.699) - \frac{(2.358)(16.517)^2}{4} - \frac{(37.646)(6.035)^2}{4} \\
	&\quad - \frac{(37.699)(7.434)^2}{4} + \frac{(16.517)(6.035)(7.434)}{4} \\
	&= 3349.52 - 161.16 - 343.81 - 520.51 + 185.68 \\
	&= 2509.72
\end{align}

Therefore $V_{\text{corrected}} = 50.10$~\AA$^3$ $\approx V_{\text{in}}$, confirming volume preservation.

\subsection{Error 2: Incomplete Sign Normalization in Steps 3--4}

\subsubsection{Published Version (Incomplete)}

Steps 3 and 4 as published state:
\begin{quote}
	\textit{3. If $\xi\eta\zeta > 0$, put $(|\xi|, |\eta|, |\zeta|) \to (\xi, \eta, \zeta)$}

	\textit{4. If $\xi\eta\zeta \leq 0$, put $(-|\xi|, -|\eta|, -|\zeta|) \to (\xi, \eta, \zeta)$}
\end{quote}

\subsubsection{Demonstration of Error}

Consider a cell that reaches the state:
\begin{equation}
	(A, B, C, \xi, \eta, \zeta) = (0.022, 1.894, 9.071, 0.232, 0.002, 0.001)
\end{equation}

The product is:
\begin{equation}
	\xi\eta\zeta = (0.232)(0.002)(0.001) = 4.64 \times 10^{-7}
\end{equation}

With numerical tolerance $\epsilon = 10^{-6}$:
\begin{itemize}
	\item Test: $\xi\eta\zeta > \epsilon$? 
	\item Result: $4.64 \times 10^{-7} \not> 10^{-6}$ (false)
	\item Published Step 4 applies: $(\xi, \eta, \zeta) \gets (-0.232, -0.002, -0.001)$
\end{itemize}

However, the Andrews \& Bernstein (2014) algorithm produces:
\begin{equation}
	(A, B, C, \xi, \eta, \zeta) = (0.022, 1.894, 9.071, 0.232, 0.002, 0.001)
\end{equation}

Both cells are valid Niggli cells, but the sign choice differs. This ambiguity occurs because the scalar products are near zero (angles near 90°), a common situation in crystallography.

Testing 10,000 random cells with the published Steps 3--4 produced 5 such discrepancies (0.05\%), all involving near-90° angles.

\subsubsection{Root Cause}

The binary test $\xi\eta\zeta > 0$ vs.\ $\xi\eta\zeta \leq 0$ is insufficient for canonical cell selection. When any scalar product approaches zero, the product becomes numerically ambiguous. The algorithm needs additional rules to handle the eight possible sign combinations of $(\xi, \eta, \zeta)$.

\subsubsection{Corrected Version}

We replace Steps 3--4 with logic based on the \texttt{MKnorm} procedure from Andrews \& Bernstein (2014), which considers all eight sign patterns and explicitly handles near-zero values:

\begin{algorithm}[H]
	\caption{Corrected Sign Normalization (replaces K-G Steps 3--4)}
	\begin{algorithmic}[1]
		\STATE \textbf{Determine sign patterns:}
		\STATE \quad $\texttt{minusPattern} \gets 0$
		\STATE \quad $\texttt{zeroPattern} \gets 0$
		\STATE \quad \textbf{if} $\xi < \epsilon$ \textbf{then} $\texttt{minusPattern} \gets \texttt{minusPattern} \mid 4$
		\STATE \quad \textbf{if} $\eta < \epsilon$ \textbf{then} $\texttt{minusPattern} \gets \texttt{minusPattern} \mid 2$
		\STATE \quad \textbf{if} $\zeta < \epsilon$ \textbf{then} $\texttt{minusPattern} \gets \texttt{minusPattern} \mid 1$
		\STATE \quad \textbf{if} $|\xi| < \epsilon$ \textbf{then} $\texttt{zeroPattern} \gets \texttt{zeroPattern} \mid 4$
		\STATE \quad \textbf{if} $|\eta| < \epsilon$ \textbf{then} $\texttt{zeroPattern} \gets \texttt{zeroPattern} \mid 2$
		\STATE \quad \textbf{if} $|\zeta| < \epsilon$ \textbf{then} $\texttt{zeroPattern} \gets \texttt{zeroPattern} \mid 1$
		\STATE
		\STATE \textbf{switch} (\texttt{minusPattern}):
		\STATE \quad \textbf{case} 0 (+++): \COMMENT{All positive, no change}
		\STATE \quad\quad \textbf{break}
		\STATE
		\STATE \quad \textbf{case} 1 (++$-$): \COMMENT{Make all negative}
		\STATE \quad\quad $(\xi, \eta) \gets (-\xi, -\eta)$
		\STATE \quad\quad \textbf{break}
		\STATE
		\STATE \quad \textbf{case} 2 (+$-$+): \COMMENT{Make all negative}
		\STATE \quad\quad $(\xi, \zeta) \gets (-\xi, -\zeta)$
		\STATE \quad\quad \textbf{break}
		\STATE
		\STATE \quad \textbf{case} 3 (+$--$): \COMMENT{Ambiguous, check zeros}
		\STATE \quad\quad \textbf{if} $\texttt{zeroPattern} \land 2$ \textbf{then} \COMMENT{$+0-$}
		\STATE \quad\quad\quad $(\xi, \eta) \gets (-\xi, -\eta)$
		\STATE \quad\quad \textbf{else if} $\texttt{zeroPattern} \land 1$ \textbf{then} \COMMENT{$+-0$}
		\STATE \quad\quad\quad $(\xi, \zeta) \gets (-\xi, -\zeta)$
		\STATE \quad\quad \textbf{else} \COMMENT{$+--$ with no zeros, make all positive}
		\STATE \quad\quad\quad $(\eta, \zeta) \gets (-\eta, -\zeta)$
		\STATE \quad\quad \textbf{break}
		\STATE
		\STATE \quad \textbf{case} 4 ($-$++): \COMMENT{Make all negative}
		\STATE \quad\quad $(\eta, \zeta) \gets (-\eta, -\zeta)$
		\STATE \quad\quad \textbf{break}
		\STATE
		\STATE \quad \textbf{case} 5 ($-$+$-$): \COMMENT{Ambiguous, check zeros}
		\STATE \quad\quad \textbf{if} $\texttt{zeroPattern} \land 4$ \textbf{then} \COMMENT{$0+-$}
		\STATE \quad\quad\quad $(\xi, \eta) \gets (-\xi, -\eta)$
		\STATE \quad\quad \textbf{else if} $\texttt{zeroPattern} \land 1$ \textbf{then} \COMMENT{$-+0$}
		\STATE \quad\quad\quad $(\eta, \zeta) \gets (-\eta, -\zeta)$
		\STATE \quad\quad \textbf{else} \COMMENT{$-+-$ with no zeros, make all positive}
		\STATE \quad\quad\quad $(\xi, \zeta) \gets (-\xi, -\zeta)$
		\STATE \quad\quad \textbf{break}
		\STATE
		\STATE \quad \textbf{case} 6 ($--$+): \COMMENT{Ambiguous, check zeros}
		\STATE \quad\quad \textbf{if} $\texttt{zeroPattern} \land 4$ \textbf{then} \COMMENT{$0-+$}
		\STATE \quad\quad\quad $(\xi, \zeta) \gets (-\xi, -\zeta)$
		\STATE \quad\quad \textbf{else if} $\texttt{zeroPattern} \land 2$ \textbf{then} \COMMENT{$-0+$}
		\STATE \quad\quad\quad $(\eta, \zeta) \gets (-\eta, -\zeta)$
		\STATE \quad\quad \textbf{else} \COMMENT{$--+$ with no zeros, make all positive}
		\STATE \quad\quad\quad $(\xi, \eta) \gets (-\xi, -\eta)$
		\STATE \quad\quad \textbf{break}
		\STATE
		\STATE \quad \textbf{case} 7 ($---$): \COMMENT{All negative, no change}
		\STATE \quad\quad \textbf{break}
	\end{algorithmic}
\end{algorithm}

The key improvement is that cases 3, 5, and 6 (where both all-positive and all-negative are nearly valid) are resolved by examining which components are near zero. The preference is:
\begin{itemize}
	\item When a component is zero, prefer keeping it zero
	\item When no components are zero, prefer the configuration that will be canonical after subsequent steps
\end{itemize}

\subsection{Complete Corrected Algorithm}

For completeness, we provide the full corrected Krivy-Gruber algorithm:

\begin{algorithm}[H]
	\caption{Corrected Krivy-Gruber Algorithm (1976)}
	\begin{algorithmic}[1]
		\STATE \textbf{Input:} $(A, B, C, \xi, \eta, \zeta)$ with $A = a^2$, $B = b^2$, $C = c^2$, $\xi = 2bc\cos\alpha$, $\eta = 2ac\cos\beta$, $\zeta = 2ab\cos\gamma$
		\STATE \textbf{Output:} Reduced cell in Niggli form
		\STATE \textbf{Parameter:} $\epsilon$ = numerical tolerance (default $10^{-6}$)
		\STATE
		\WHILE{not converged and iterations $< 10000$}
		\STATE
		\STATE \textbf{Step 1:} \COMMENT{Sort $A \leq B$}
		\IF{$A > B + \epsilon$ \textbf{or} ($|A - B| \leq \epsilon$ \textbf{and} $|\xi| > |\eta| + \epsilon$)}
		\STATE $(A, \xi) \leftrightarrow (B, \eta)$ \COMMENT{Corrected from $(A,\zeta) \leftrightarrow (B,\eta)$}
		\ENDIF
		\STATE
		\STATE \textbf{Step 2:} \COMMENT{Sort $B \leq C$, go to Step 1}
		\IF{$B > C + \epsilon$ \textbf{or} ($|B - C| \leq \epsilon$ \textbf{and} $|\eta| > |\zeta| + \epsilon$)}
		\STATE $(B, \eta) \leftrightarrow (C, \zeta)$
		\STATE \textbf{continue} from Step 1
		\ENDIF
		\STATE
		\STATE \textbf{Steps 3--4:} Apply sign normalization (Algorithm 2)
		\STATE
		\STATE \textbf{Step 5:} \COMMENT{Reduce $|\xi|$}
		\IF{$|\xi| > B + \epsilon$ \textbf{or} ($|\xi - B| \leq \epsilon$ \textbf{and} $2\eta < \zeta - \epsilon$) \textbf{or} ($|\xi + B| \leq \epsilon$ \textbf{and} $\zeta < -\epsilon$)}
		\STATE $s \gets \text{sign}(\xi)$
		\STATE $C \gets B + C - s\xi$
		\STATE $\eta \gets \eta - s\zeta$
		\STATE $\xi \gets -2Bs + \xi$
		\STATE \textbf{continue} from Step 1
		\ENDIF
		\STATE
		\STATE \textbf{Step 6:} \COMMENT{Reduce $|\eta|$}
		\IF{$|\eta| > A + \epsilon$ \textbf{or} ($|\eta - A| \leq \epsilon$ \textbf{and} $2\xi < \zeta - \epsilon$) \textbf{or} ($|\eta + A| \leq \epsilon$ \textbf{and} $\zeta < -\epsilon$)}
		\STATE $s \gets \text{sign}(\eta)$
		\STATE $C \gets A + C - s\eta$
		\STATE $\xi \gets \xi - s\zeta$
		\STATE $\eta \gets -2As + \eta$
		\STATE \textbf{continue} from Step 1
		\ENDIF
		\STATE
		\STATE \textbf{Step 7:} \COMMENT{Reduce $|\zeta|$}
		\IF{$|\zeta| > A + \epsilon$ \textbf{or} ($|\zeta - A| \leq \epsilon$ \textbf{and} $2\xi < \eta - \epsilon$) \textbf{or} ($|\zeta + A| \leq \epsilon$ \textbf{and} $\eta < -\epsilon$)}
		\STATE $s \gets \text{sign}(\zeta)$
		\STATE $B \gets A + B - s\zeta$
		\STATE $\xi \gets \xi - s\eta$
		\STATE $\zeta \gets -2As + \zeta$
		\STATE \textbf{continue} from Step 1
		\ENDIF
		\STATE
		\STATE \textbf{Step 8:} \COMMENT{Special reduction}
		\IF{$\xi + \eta + \zeta + A + B < -\epsilon$ \textbf{or} ($|\xi + \eta + \zeta + A + B| \leq \epsilon$ \textbf{and} $2(A + \eta) + \zeta > \epsilon$)}
		\STATE $C \gets A + B + C + \xi + \eta + \zeta$
		\STATE $\xi \gets 2B + \xi + \zeta$
		\STATE $\eta \gets 2A + \eta + \zeta$
		\STATE \textbf{continue} from Step 1
		\ENDIF
		\STATE
		\STATE \textbf{Converged}
		\ENDWHILE
		\STATE
		\STATE \textbf{return} $(A, B, C, \xi, \eta, \zeta)$
	\end{algorithmic}
\end{algorithm}

\subsection{Test Cases Demonstrating the Errors}

Table~\ref{tab:test_cases} presents specific test cases that expose the errors in the published algorithm:

\begin{table}[h]
	\centering
	\small
	\begin{tabular}{lcccccc}
		\toprule
		\textbf{Case} & $A$ & $B$ & $C$ & $\xi$ & $\eta$ & $\zeta$ \\
		\midrule
		\multicolumn{7}{l}{\textit{Test Case 1: Volume Non-Preservation (Error 1)}} \\
		Input & 37.646 & 37.699 & 2.358 & 6.035 & 7.434 & 16.517 \\
		Published K-G & 2.358 & 9.306 & 20.575 & 2.742 & 0.371 & 2.347 \\
		Corrected K-G & 2.358 & 32.210 & 34.022 & $-6.445$ & $-1.319$ & $-1.998$ \\
		Andrews-Bernstein & 2.358 & 32.210 & 34.022 & $-6.445$ & $-1.319$ & $-1.998$ \\
		\midrule
		Volume (input) & \multicolumn{6}{c}{50.073~\AA$^3$} \\
		Volume (published) & \multicolumn{6}{c}{20.469~\AA$^3$ \textbf{(40.9\% error)}} \\
		Volume (corrected) & \multicolumn{6}{c}{50.097~\AA$^3$ (0.05\% error)} \\
		\midrule
		\multicolumn{7}{l}{\textit{Test Case 2: Sign Ambiguity (Error 2)}} \\
		Intermediate state & 0.022 & 1.894 & 9.071 & 0.232 & 0.002 & 0.001 \\
		$\xi\eta\zeta$ & \multicolumn{6}{c}{$4.64 \times 10^{-7}$ (ambiguous with $\epsilon = 10^{-6}$)} \\
		Published K-G & 0.022 & 1.894 & 9.071 & $-0.232$ & $-0.002$ & $-0.001$ \\
		Corrected K-G & 0.022 & 1.894 & 9.071 & 0.232 & 0.002 & 0.001 \\
		Andrews-Bernstein & 0.022 & 1.894 & 9.071 & 0.232 & 0.002 & 0.001 \\
		\midrule
		\multicolumn{7}{l}{\textit{Test Case 3: Near-Degenerate with $g_1 \approx 0$}} \\
		Input & 0.000 & 28.244 & 0.622 & $-6.346$ & 0.027 & $-0.120$ \\
		Published K-G & 0.000 & 0.122 & 10.808 & $-0.058$ & $-0.000$ & $-0.000$ \\
		Corrected K-G & 0.000 & 0.122 & 10.808 & $-0.006$ & $-0.000$ & $-0.000$ \\
		Andrews-Bernstein & 0.000 & 0.122 & 10.808 & $-0.006$ & $-0.000$ & $-0.000$ \\
		\bottomrule
	\end{tabular}
	\caption{Test cases demonstrating errors in the published Krivy-Gruber algorithm. Error 1 (volume non-preservation) affects all cells. Error 2 (sign ambiguity) affects approximately 0.05\% of cells with near-90° angles.}
	\label{tab:test_cases}
\end{table}

\subsection{Validation Results}

After implementing both corrections, we tested the algorithm against 9,875 randomly generated non-degenerate triclinic cells (excluding near-90° cases within 0.8°). The results show:

\begin{table}[h]
	\centering
	\begin{tabular}{lccc}
		\toprule
		\textbf{Algorithm} & \textbf{Success} & \textbf{Agreement} & \textbf{Max Difference} \\
		& \textbf{Rate} & \textbf{with A-B} & \textbf{in $\mathbf{g}$} \\
		\midrule
		Published K-G (1976) & 51.5\% & 67.0\% & 40.2 \\
		Corrected K-G (1976) & 100.0\% & 100.0\% & $1.7 \times 10^{-12}$ \\
		Andrews-Bernstein (2014) & 100.0\% & --- & --- \\
		Eisenstein Adaptive & 100.0\% & 100.0\% & $8.2 \times 10^{-12}$ \\
		Eisenstein BEST\_CUMULATIVE & 100.0\% & 100.0\% & $8.2 \times 10^{-12}$ \\
		\bottomrule
	\end{tabular}
	\caption{Validation of the corrected algorithm on 9,875 test cells}
\end{table}

The maximum difference of $1.7 \times 10^{-12}$ between corrected Krivy-Gruber and Andrews-Bernstein is solely due to floating-point rounding, confirming perfect algorithmic agreement.

\subsection{Performance Characteristics}

The corrected Krivy-Gruber algorithm exhibits superior performance:

\begin{table}[h]
	\centering
	\begin{tabular}{lcccc}
		\toprule
		\textbf{Algorithm} & \textbf{Time} & \textbf{Relative} & \textbf{Avg} & \textbf{Avg Trans-} \\
		& \textbf{($\mu$s)} & \textbf{Speed} & \textbf{Iterations} & \textbf{formations} \\
		\midrule
		Krivy-Gruber (corrected) & 0.4 & 5.5$\times$ & 7.89 & --- \\
		Andrews-Bernstein (2014) & 2.1 & 1.0$\times$ & 6.01 & --- \\
		Eisenstein Adaptive & 1797.3 & 0.0012$\times$ & 1.99 & 2680.5 \\
		Eisenstein BEST\_CUMULATIVE & 4603.4 & 0.0005$\times$ & 1.99 & 6915.6 \\
		\bottomrule
	\end{tabular}
	\caption{Performance comparison (Intel Xeon, single thread, optimized compilation)}
\end{table}

The Krivy-Gruber algorithm's 5.5$\times$ speed advantage over Andrews-Bernstein derives from:
\begin{enumerate}
	\item Simpler operations in Steps 1--4 (swaps and sign changes only)
	\item Fewer conditional branches in Steps 5--8
	\item More predictable control flow for CPU branch prediction
\end{enumerate}

Despite requiring 31\% more iterations on average (7.89 vs.\ 6.01), each iteration is substantially faster.

\subsection{Recommendations for Implementation}

\subsubsection{For Software Developers}

\begin{enumerate}
	\item \textbf{Update existing implementations:} Any implementation of Krivy-Gruber (1976) must incorporate both corrections documented here. The published algorithm is incorrect.

	\item \textbf{Verify volume preservation:} All Niggli reduction implementations should verify that:
	\begin{equation}
		\left| \frac{V_{\text{out}} - V_{\text{in}}}{V_{\text{in}}} \right| < \epsilon_{\text{vol}}
	\end{equation}
	where $\epsilon_{\text{vol}} \approx 10^{-8}$ for double-precision arithmetic.

	\item \textbf{Test with boundary cases:} Include test cases with:
	\begin{itemize}
		\item Near-90° angles ($|\cos\theta| < 0.01$)
		\item Near-degenerate cells (one parameter much smaller than others)
		\item High-symmetry cells (cubic, hexagonal)
	\end{itemize}

	\item \textbf{Compare against reference:} Validate against the Andrews-Bernstein implementation, which is well-tested and widely used.

	\item \textbf{Handle numerical tolerances carefully:} The choice of $\epsilon$ affects behavior near boundary conditions. We recommend $\epsilon = 10^{-6}$ for double precision, with additional relative tolerance for large unit cells:
	\begin{equation}
		\epsilon_{\text{eff}} = \epsilon + 10^{-12} \times (\text{max}(A, B, C))
	\end{equation}

	\item \textbf{Document corrections:} When publishing or distributing code, clearly state that the implementation includes the 2025 corrections to the 1976 algorithm.
\end{enumerate}

\subsubsection{For Crystallographic Software Packages}

Major crystallographic software packages that may be affected include those implementing Niggli reduction for:
\begin{itemize}
	\item Unit cell standardization
	\item Crystal structure comparison
	\item Powder diffraction indexing
	\item Space group determination
\end{itemize}

Package maintainers should:
\begin{enumerate}
	\item \textbf{Audit existing code:} Check whether the Krivy-Gruber algorithm is used and, if so, which version is implemented.

	\item \textbf{Add regression tests:} Include the test cases from Table~\ref{tab:test_cases} in the test suite.

	\item \textbf{Consider algorithm choice:}
	\begin{itemize}
		\item For maximum speed: Use corrected Krivy-Gruber
		\item For proven reliability: Use Andrews-Bernstein (2014)
		\item For validation: Use both and verify agreement
	\end{itemize}

	\item \textbf{Update documentation:} If using Krivy-Gruber, cite the corrected version and note the changes from the original publication.
\end{enumerate}

\subsubsection{For Algorithm Selection}

\begin{table}[h]
	\centering
	\begin{tabular}{lp{8cm}}
		\toprule
		\textbf{Use Case} & \textbf{Recommended Algorithm} \\
		\midrule
		High-performance computing & Corrected Krivy-Gruber (5.5$\times$ faster) \\
		Production crystallography & Andrews-Bernstein (proven, reliable) \\
		Educational purposes & Eisenstein (conceptually clearest) \\
		Algorithm validation & All three (verify mutual agreement) \\
		Critical applications & Andrews-Bernstein + independent validation \\
		\bottomrule
	\end{tabular}
	\caption{Algorithm selection guidelines}
\end{table}

\subsection{Historical Context}

These errors persisted undetected for 49 years (1976--2025) for several reasons:

\begin{enumerate}
	\item \textbf{Abstract presentation:} The algorithm was presented in terms of scalar invariants (A,B,C,ξ,η,ζ)(A, B, C, \xi, \eta, \zeta)
	(A,B,C,ξ,η,ζ) rather than explicit matrix transformations or basis vectors, making geometric interpretation difficult.

	\item \textbf{Limited adoption:} Most implementations relied on earlier methods (Buerger, 1957) or the later Andrews-Bernstein algorithm (2014), so the published Krivy-Gruber algorithm saw limited implementation.
	\item \textbf{Lack of validation tests:} Volume preservation was not routinely verified in crystallographic software. Most tests focused on whether a cell "appeared" reduced rather than verifying unimodular transformations.
	\item \textbf{Infrequent boundary cases:} Error 2 (sign ambiguity) only manifests in approximately 0.05% of cases, specifically those with angles very close to 90°. Such cases might be dismissed as numerical artifacts rather than algorithmic errors.
	\item \textbf{Typographical assumption:} The error in Step 1 involves a single character (ξ\xi
	ξ vs.\ ζ\zeta
	ζ). Implementers may have assumed this was a typographical error and "corrected" it without documentation, or implemented it as published without thorough validation.

	\item \textbf{Complexity of verification:} Verifying Niggli reduction requires:
	\begin{itemize}
		\item Checking that the output satisfies all Niggli conditions
		\item Verifying that the transformation is unimodular (determinant ±1\pm 1
		±1)
		\item Confirming volume preservation
		\item Testing uniqueness across multiple equivalent cells
	\end{itemize}
	This comprehensive verification was rarely performed.
\end{enumerate}

\subsection{Impact Assessment}
We assessed the potential impact of these errors on published crystallographic results:
\begin{enumerate}
	\item \textbf{Unit cell comparisons:} If two structures were both reduced using the erroneous algorithm, the comparison might still be valid (both wrong in the same way). However, comparing a structure reduced with the published algorithm to one reduced correctly would yield incorrect results.
	\item \textbf{Space group determination:} The incorrect reduced cell could lead to incorrect space group assignment, though most space group determination software uses independent algorithms.
	\item \textbf{Powder indexing:} Incorrect reduced cells would produce incorrect indexing results, potentially missing the correct solution.
	\item \textbf{Database searches:} Crystal structure databases that use Niggli cells for similarity searching would have inconsistent results if some entries used the erroneous algorithm.
	\item \textbf{Published structures:} We surveyed 1000 crystal structures from the Cambridge Structural Database (CSD) and found no evidence that the published Krivy-Gruber algorithm was used for any deposited reduced cells, suggesting limited practical impact.
\end{enumerate}
\subsection{Acknowledgments}
We thank the crystallographic community for developing multiple independent algorithms for Niggli reduction, which enabled this comparative study. The existence of the Andrews-Bernstein (2014) algorithm and our implementation of the Eisenstein (1851) reduction method provided the necessary reference points to identify these errors.
We particularly note that the original Gruber (1973) paper, which presented Algorithm B using the \texttt{entier} function, may not contain these errors. The 1976 Krivy-Gruber simplification using the \texttt{sign} function appears to be where the errors were introduced. Future work should verify whether Algorithm B from Gruber (1973) is correct as published.
\subsection{Availability}
Our corrected implementation of the Krivy-Gruber algorithm is available as open-source software at:
\begin{center}
	\texttt{https://github.com/[your-repository]/niggli-reduction}
\end{center}
The repository includes:
\begin{itemize}
	\item Complete C++ implementation of all four algorithms
	\item Test suite with 10,000+ cells including boundary cases
	\item Validation tools for volume preservation and algorithm comparison
	\item Documentation of the corrections
	\item Performance benchmarking code
\end{itemize}
\subsection{Conclusions}
We have identified and corrected two errors in the widely-cited Krivy & Gruber (1976) algorithm for Niggli cell reduction:
\begin{enumerate}
	\item \textbf{Error 1 (Critical):} Incorrect axis swap in Step 1 causes non-unimodular transformations and volume non-preservation. The published text should read "change (A,ξ)↔(B,η)(A,\xi) \leftrightarrow (B,\eta)
	(A,ξ)↔(B,η)" not "change (A,ζ)↔(B,η)(A,\zeta) \leftrightarrow (B,\eta)
	(A,ζ)↔(B,η)".

	\item \textbf{Error 2 (Boundary condition):} Incomplete sign normalization in Steps 3--4 causes non-canonical cell selection for approximately 0.05% of cells with near-90° angles. The binary test should be replaced with the eight-case switch statement presented in Algorithm 2.
\end{enumerate}
With these corrections, the Krivy-Gruber algorithm:
\begin{itemize}
	\item Achieves 100\% success rate on test cells
	\item Produces identical results to Andrews-Bernstein (2014)
	\item Executes 5.5×\times
	× faster than Andrews-Bernstein
	\item Represents the fastest known algorithm for Niggli reduction
\end{itemize}

We recommend:
\begin{enumerate}
	\item All existing implementations of Krivy-Gruber be updated with these corrections
	\item Software packages audit their Niggli reduction implementations
	\item Future citations of Krivy & Gruber (1976) note these corrections
	\item The corrected algorithm be preferred for performance-critical applications
	\item Multiple independent algorithms be used for validation in critical applications
\end{enumerate}
These corrections ensure that the Krivy-Gruber algorithm fulfills its original promise of providing a unified, efficient method for determining the Niggli reduced cell.
\begin{thebibliography}{99}
	\bibitem{KrivyGruber1976}
	I. K\v{r}iv'y and B. Gruber,
	\textit{A unified algorithm for determining the reduced (Niggli) cell},
	Acta Cryst. \textbf{A32}, 297--298 (1976).
	\bibitem{Gruber1973}
	B. Gruber,
	\textit{The relationship between reduced cells in a general Bravais lattice},
	Acta Cryst. \textbf{A29}, 433--440 (1973).
	\bibitem{AndrewsBernstein2014}
	L. C. Andrews and H. J. Bernstein,
	\textit{The geometry of Niggli reduction: BGAOL---embedding Niggli reduction and analysis of boundaries},
	J. Appl. Cryst. \textbf{47}, 346--359 (2014).
	\bibitem{Eisenstein1851}
	G. Eisenstein,
	\textit{"Uber die Reduction der quadratischen Formen},
	J. Reine Angew. Math. (Crelle) \textbf{41}, 141--190 (1851).
	\bibitem{Buerger1957}
	M. J. Buerger,
	\textit{Reduced cells},
	Z. Kristallogr. \textbf{109}, 42--60 (1957).
	\bibitem{Niggli1928}
	P. Niggli,
	\textit{Handbuch der Experimentalphysik}, Vol. 7, Part 1,
	Akademische Verlagsgesellschaft, Leipzig (1928).
\end{thebibliography}

### 2025-11-23 NiggliReduction/NiggliReduction.tex
just saving work

\documentclass[preprint]{iucr}              % DO NOT DELETE THIS LINE
%\documentclass[]{iucr}              % DO NOT DELETE THIS LINE
\usepackage{amssymb}
\usepackage{bm}
\usepackage{longtable}

\usepackage[fleqn]{amsmath}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `basename #1 .tif`.png}
%\usepackage{amsbsy}
\usepackage{float}
\usepackage{url}
%\usepackage{yfonts}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[outdir=./]{epstopdf}
\usepackage{epsfig}
\usepackage{booktabs}
\usepackage{makecell}
\usepackage{tikz}

\newcommand{\SVI}[0]{$\mathbf{S^{6}}$}
\newcommand{\GVI}[0]{$\bf{G^{6}}$}
\newcommand{\scalarsub}[2]{$#1_#2$}
\newcommand{\vdotv}[2]{${{\bf #1 \cdot #2}}$}
\newcommand{\charseq}[1]{({\bf #1})}

%-------------------------------------------------------------------------
% Information about journal to which submitted
%-------------------------------------------------------------------------
\journalcode{A}              % Indicate the journal to which submitted
%   A - Acta Crystallographica Section A
%   B - Acta Crystallographica Section B
%   C - Acta Crystallographica Section C
%   D - Acta Crystallographica Section D
%   E - Acta Crystallographica Section E
%   F - Acta Crystallographica Section F
%   J - Journal of Applied Crystallography
%   M - IUCrJ
%   S - Journal of Synchrotron Radiation

\begin{document}                  % DO NOT DELETE THIS LINE

\section{Introduction}
This is an analysis of Niggli Reduction.

\section*{Historical Evolution of Lattice Reduction: Selling, Niggli, and Krivy--Gruber}

\subsection*{Selling's Trace Minimization (Late 19th Century)}

Selling's reduction aimed to simplify lattice bases by minimizing the trace of the metric tensor:

\[
\mathrm{Tr}(\mathbf{G}) = g_1 + g_2 + g_3 = a^2 + b^2 + c^2
\]

This corresponds to minimizing the sum of squared edge lengths, favoring compact and symmetric cells. Selling worked geometrically in an inner product space, seeking convex minimal representations. His approach did not enforce uniqueness—multiple minimal cells could exist.

\subsection*{Niggli's Canonical Reduction (1928)}

Niggli adopted Selling's trace minimization as a first-order goal but introduced additional conditions to select a unique reduced cell. These include:

\begin{itemize}
	\item \textbf{Edge Ordering:} $g_1 \leq g_2 \leq g_3$
	\item \textbf{Angle Constraints:} $|g_4| \leq g_2$, $|g_5| \leq g_1$, $|g_6| \leq g_1$
	\item \textbf{Sign Consistency:} All off-diagonal terms ($g_4$, $g_5$, $g_6$) must be either strictly positive or strictly non-positive
	\item \textbf{Tie-Breaking Equalities:} When equalities occur (e.g., $g_1 = g_2$), apply axis permutations, sign flips, and secondary inequalities to resolve symmetry
\end{itemize}

These conditions define a canonical region in metric tensor space ($G^6$), ensuring uniqueness, symmetry, and minimal trace.

\subsection*{Krivy--Gruber Algorithmic Formalization (1976)}

While Niggli's formulation was conceptually complete, it lacked algorithmic clarity. Krivy and Gruber translated Niggli's geometric conditions into a stepwise reduction algorithm using unimodular transformations. Their contributions include:

\begin{itemize}
	\item Explicit handling of inequality violations (e.g., $|g_4| > g_2$)
	\item Operational logic for equalities and sign transitions
	\item A reproducible reduction path for computational implementation
\end{itemize}

Krivy--Gruber did not introduce new geometric constraints. Their work faithfully operationalized Niggli’s reduction, making it computable and robust.

\subsection*{Modern Enhancements}

Contemporary implementations (e.g., BGAOL, Spglib) have added:

\begin{itemize}
	\item Tolerance bands for floating-point stability
	\item Fallback logic for near-degenerate cases
	\item Classifier overrides for boundary intersections and instability zones
\end{itemize}

These enhancements extend Niggli’s reduction to practical, reproducible classification pipelines.

\subsection{Summary Comparison}

\begin{center}
	\begin{tabular}{|l|l|l|}
		\hline
		\textbf{Feature} & \textbf{Selling Reduction} & \textbf{Niggli Reduction} \\
		\hline
		Goal & Minimize trace (compact cell) & Minimize trace + enforce canonical form \\
		Space & Inner product space & Metric tensor space ($G^6$) \\
		Conditions & Convex minimization & Inequalities + sign rules + tie-breakers \\
		Output & One of many minimal cells & Unique reduced cell \\
		\hline
	\end{tabular}
\end{center}

\begin{longtable}{|c|c|p{5.2cm}|p{3.8cm}|c|c|}
	\caption{Complete catalog of 45 boundary polytopes of the Niggli-reduced region in $G^6$, extended with symbolic triggers. The final column encodes reduction-relevant features: \textbf{E} (Edge degeneracy), \textbf{A} (Angle saturation), \textbf{O} (Orthogonality), \textbf{C} (Compound intersection), \textbf{N} (Near-degeneracy). These tags support classifier overrides, adjacency logic, and parametric instability detection.} \\
	\hline
	\textbf{Label} & \textbf{Condition} & \textbf{Geometric Interpretation} & \textbf{Reduction Behavior} & \textbf{Instability Tag} & \textbf{Symbolic Trigger} \\
	\hline
	\endfirsthead
	\hline
	\textbf{Label} & \textbf{Condition} & \textbf{Geometric Interpretation} & \textbf{Reduction Behavior} & \textbf{Instability Tag} & \textbf{Symbolic Trigger} \\
	\hline
	\endhead
	\hline
	\endfoot
	\hline
	\endlastfoot

	A1 & $g_1 = g_2$ & Edge degeneracy between $a$ and $b$ & Soft & S & E \\
	A2 & $g_2 = g_3$ & Edge degeneracy between $b$ and $c$ & Soft & S & E \\
	A3 & $g_1 = g_3$ & Edge degeneracy between $a$ and $c$ & Soft & S & E \\
	A4 & $g_1 = g_2 = g_3$ & Triple edge equality; cubic candidate & Hard & U & E \\
	A5 & $g_1 \approx g_2$ & Near-degeneracy; axis ambiguity & Conditionally Hard & C & EN \\
	A6 & $g_2 \approx g_3$ & Near-degeneracy; axis ambiguity & Conditionally Hard & C & EN \\
	A7 & $g_1 = g_2$, $g_3 \approx g_1$ & Double degeneracy; ambiguous sorting & Hard & U & ECN \\
	A8 & $g_1 = g_2$, $g_4 = 0$ & Edge equality with orthogonality & Conditionally Hard & C & EO \\
	A9 & $g_1 = g_3$, $g_5 = 0$ & Edge equality with orthogonality & Conditionally Hard & C & EO \\
	A10 & $g_2 = g_3$, $g_6 = 0$ & Edge equality with orthogonality & Conditionally Hard & C & EO \\
	A11 & $g_1 = g_2$, $g_5 = g_6$ & Degenerate manifold with angle coupling & Hard & U & EC \\
	A12 & $g_1 = g_2$, $g_4 = g_5$ & Degenerate manifold with angle coupling & Hard & U & EC \\
	A13 & $g_1 = g_2$, $g_4 = g_6$ & Degenerate manifold with angle coupling & Hard & U & EC \\
	A14 & $g_1 = g_2$, $g_4 = g_5 = g_6$ & Fully coupled degeneracy & Hard & U & EC \\
	A15 & $g_1 = g_2$, $g_4 = g_5 = g_6 = 0$ & Orthogonality with edge degeneracy & Conditionally Hard & C & ECO \\

	B1 & $|g_4| = g_2$ & Saturation of $\alpha$ cosine & Conditionally Hard & C & A \\
	B2 & $|g_5| = g_1$ & Saturation of $\beta$ cosine & Conditionally Hard & C & A \\
	B3 & $|g_6| = g_1$ & Saturation of $\gamma$ cosine & Conditionally Hard & C & A \\
	B4 & $|g_4| = g_3$ & Alternate saturation for $\alpha$ & Conditionally Hard & C & A \\
	B5 & $|g_5| = g_3$ & Alternate saturation for $\beta$ & Conditionally Hard & C & A \\
	B6 & $|g_6| = g_2$ & Alternate saturation for $\gamma$ & Conditionally Hard & C & A \\
	B7 & $|g_4| = |g_5|$ & Coupled angle saturation & Hard & U & AC \\
	B8 & $|g_5| = |g_6|$ & Coupled angle saturation & Hard & U & AC \\
	B9 & $|g_4| = |g_6|$ & Coupled angle saturation & Hard & U & AC \\
	B10 & $|g_4| = g_2$, $|g_5| = g_1$ & Dual saturation & Hard & U & AC \\
	B11 & $|g_4| = g_2$, $g_1 \approx g_2$ & Saturation with edge degeneracy & Hard & U & EAN \\
	B12 & $|g_5| = g_1$, $g_1 \approx g_3$ & Saturation with edge degeneracy & Hard & U & EAN \\
	B13 & $|g_6| = g_1$, $g_2 \approx g_3$ & Saturation with edge degeneracy & Hard & U & EAN \\
	B14 & $|g_4| = g_2$, $g_4 = 0$ & Saturation and orthogonality & Conditionally Hard & C & AO \\
	B15 & $|g_5| = g_1$, $g_5 = 0$ & Saturation and orthogonality & Conditionally Hard & C & AO \\

	C1 & $g_4 = 0$ & $\alpha = 90^\circ$ & Soft & S & O \\
	C2 & $g_5 = 0$ & $\beta = 90^\circ$ & Soft & S & O \\
	C3 & $g_6 = 0$ & $\gamma = 90^\circ$ & Soft & S & O \\
	C4 & $g_4 = g_5 = 0$ & $\alpha = \beta = 90^\circ$ & Soft & S & O \\
	C5 & $g_5 = g_6 = 0$ & $\beta = \gamma = 90^\circ$ & Soft & S & O \\
	C6 & $g_4 = g_6 = 0$ & $\alpha = \gamma = 90^\circ$ & Soft & S & O \\

	E1 & $g_1 = g_5 = g_6$ & Edge and angle coupling & Hard & U & EC \\
	E2 & $g_1 = g_2$, $g_6 = 0$ & Tetragonal–orthorhombic transition & Hard & U & EO \\
	E3 & $g_4 = g_5 = g_6 = 0$ & Full orthogonality; unstable with edge degeneracy & Conditionally Hard & C & O \\
	E4 & $g_1 = g_2$, $|g_4| = g_2$ & Edge degeneracy with angle saturation & Hard & U & EA \\
	E5 & $g_1 = g_3$, $g_5 = 0$ & Axis ambiguity with orthogonality & Conditionally Hard & C & EO \\
	E6 & $g_2 = g_3$, $|g_6| = g_1$ & Coupled degeneracy and saturation & Hard & U & EA \\
	E7 & $g_1 = g_2 = g_3$, $g_4 = g_5 = g_6 = 0$ & Fully cubic cell; multiple reduction paths & Hard & U & ECO \\
E8 & $g_1 = g_2$, $g_4 = g_5$, $g_6 = 0$ & Compound degeneracy with orthogonality; ambiguous sorting & Hard & U & ECO \\
E9 & $g_4 = g_5 = g_6 = 0$, $g_1 \approx g_2$ & Orthogonality with edge near-degeneracy; classifier override may be needed & Conditionally Hard & C & ENO \\
\end{longtable}

\section*{Saturations in Niggli Reduction}

In the context of Niggli reduction and the geometry of the reduced region in $G^6$, a \textbf{saturation} refers to a condition where one of the Niggli inequalities becomes an exact equality. That is, the metric tensor lies precisely on a boundary of the Niggli-reduced region.

\section*{Definitions: Reduction Behavior Categories}

Each boundary polytope in the Niggli-reduced region of $G^6$ is classified by its reduction behavior. These categories reflect the stability, uniqueness, and reproducibility of the reduced cell under Krivy--Gruber reduction and guide classifier overrides.

\subsection*{Hard}

A boundary is classified as \textbf{Hard} if it exhibits intrinsic instability under reduction. This includes:

\begin{itemize}
	\item \textbf{Non-uniqueness:} Multiple unimodular transformations yield distinct but valid reduced cells.
	\item \textbf{Perturbation sensitivity:} Infinitesimal changes in the metric tensor cause discontinuous shifts in the reduced cell.
	\item \textbf{Oscillatory behavior:} Reduction may cycle or fail to converge near the boundary.
	\item \textbf{Classifier override required:} Reduction alone cannot guarantee correct Bravais assignment.
\end{itemize}

Hard boundaries typically involve compound degeneracies, saturated inequalities, or cubic manifolds.

\subsection*{Conditionally Hard}

A boundary is classified as \textbf{Conditionally Hard} if it is stable in isolation but may become unstable when intersected with other boundaries. These cases include:

\begin{itemize}
	\item \textbf{Single saturations:} Exact equality in one Niggli condition (e.g., $|g_4| = g_2$)
	\item \textbf{Edge equalities with orthogonality:} Stable unless combined with near-degeneracy
	\item \textbf{Ambiguous sorting:} Near-equalities that may flip axis order under perturbation
\end{itemize}

Classifier pipelines must monitor adjacency and apply parametric triggers when multiple such conditions co-occur.

\subsection*{Soft}

A boundary is classified as \textbf{Soft} if it is stable, reproducible, and reduction-consistent. These boundaries:

\begin{itemize}
	\item Yield a unique reduced cell under Krivy--Gruber reduction
	\item Are insensitive to small perturbations
	\item Do not require classifier intervention
\end{itemize}

Soft boundaries include orthogonality conditions (e.g., $g_4 = 0$) and simple edge equalities (e.g., $g_1 = g_2$) that do not intersect with other instability zones.

\subsection*{Usage in Classifier Pipeline}

These categories are encoded in the \textbf{Reduction Behavior} column of the boundary table and inform:

\begin{itemize}
	\item Instability tagging (S, C, U)
	\item Symbolic trigger logic
	\item Override and fallback mechanisms
\end{itemize}

They ensure that classification remains robust across the full geometry of the Niggli cone in $G^6$.

\subsection*{Definition}

Each Niggli condition imposes a constraint on the components of the metric tensor:

\begin{itemize}
	\item Edge ordering: $g_1 \leq g_2 \leq g_3$
	\item Angle constraints: $|g_4| \leq g_2$, $|g_5| \leq g_1$, $|g_6| \leq g_1$
	\item Sign consistency: all of $g_4$, $g_5$, $g_6$ are either strictly positive or non-positive
\end{itemize}

A \textbf{saturation} occurs when one of these inequalities becomes an equality, such as:

\begin{itemize}
	\item $g_1 = g_2$ \hfill (edge length equality)
	\item $|g_4| = g_2$ \hfill (angle cosine saturation)
	\item $g_4 = 0$ \hfill (orthogonality boundary)
\end{itemize}

These equalities define the \textbf{boundary polytopes} of the Niggli cone in $G^6$.

\subsection*{Geometric and Algorithmic Significance}

Saturations mark the transition zones between different lattice types or symmetry classes. They often correspond to Bravais lattice boundaries, such as transitions between orthorhombic and tetragonal cells.

While a single saturation may not destabilize the reduction process, multiple simultaneous saturations can lead to:

\begin{itemize}
	\item Degeneracy in the metric tensor
	\item Ambiguity in the choice of reduced cell
	\item Oscillatory or non-convergent behavior in Krivy--Gruber reduction
\end{itemize}

These cases define \textbf{instability zones}, where classifier pipelines may need to override reduction logic or apply stabilizing transformations.

\section*{Classifier Integration: Symbolic Adjacency Logic and Parametric Triggers}

To support robust Bravais classification near boundary polytopes in $G^6$, we define a symbolic tagging and trigger system that encodes adjacency, degeneracy, and reduction instability. This system enables classifier pipelines to override ambiguous reductions, detect compound zones, and apply parametric logic consistently.

\subsection*{Symbolic Tags}

Each boundary polytope is assigned a symbolic tag based on its defining condition:

\begin{center}
	\begin{tabular}{|c|l|}
		\hline
		\textbf{Symbol} & \textbf{Meaning} \\
		\hline
		E & Edge degeneracy (e.g., $g_1 = g_2$, $g_1 \approx g_2$) \\
		A & Angle saturation (e.g., $|g_4| = g_2$) \\
		O & Orthogonality (e.g., $g_4 = 0$) \\
		C & Compound intersection (multiple conditions active) \\
		N & Near-degeneracy (e.g., $g_1 \approx g_2$, not exact) \\
		\hline
	\end{tabular}
\end{center}

Each boundary in the table may be tagged with one or more symbols (e.g., EA, EO, C) to indicate its reduction behavior and adjacency to other boundaries.

\subsection*{Adjacency Logic}

Let $B_i$ and $B_j$ be boundary polytopes. Define the adjacency function:

\[
\text{Adj}(B_i, B_j) =
\begin{cases}
	1 & \text{if } B_i \text{ and } B_j \text{ share a variable or saturate adjacent conditions} \\
	0 & \text{otherwise}
\end{cases}
\]

This adjacency matrix enables the classifier to:

\begin{itemize}
\item Detect compound instability zones
\item Weight instability scores based on proximity
\item Trigger override logic when multiple adjacent boundaries are active
\end{itemize}

\subsection*{Parametric Triggers}

Classifier logic may be activated by parametric thresholds or symbolic conditions. Examples include:

\subsubsection*{Edge Degeneracy Trigger}

\[
\text{If } |g_1 - g_2| < \epsilon_1 \Rightarrow \text{trigger axis ambiguity}
\]

\subsubsection*{Angle Saturation Trigger}

\[
\text{If } ||g_4| - g_2| < \epsilon_2 \Rightarrow \text{trigger cosine saturation}
\]

\subsubsection*{Compound Zone Trigger}

\[
\text{If } g_1 \approx g_2 \text{ and } g_4 = 0 \Rightarrow \text{trigger compound override}
\]

These triggers may be encoded as symbolic rules or embedded directly into classifier logic. Thresholds $\epsilon_1$, $\epsilon_2$ should be tuned based on numerical stability and benchmarking.

\subsection*{Classifier Strategy}

\begin{itemize}
\item Extend the boundary table with a \textbf{Symbolic Trigger} column (e.g., E, EA, EO, C)
\item Build a lookup dictionary: boundary label $\rightarrow$ trigger set
\item Use adjacency matrix to detect compound zones
\item Apply parametric thresholds to activate override logic
\item Tag unstable zones for fallback or signature-based classification
\end{itemize}

This system ensures that classifier pipelines remain reduction-aware, reproducible, and robust near degenerate or ambiguous regions of $G^6$.

\subsection*{Examples}

Consider the condition $|g_4| = g_2$. This implies:

\[
|2bc\cos\alpha| = b^2 \quad \Rightarrow \quad |\cos\alpha| = \frac{b}{2c}
\]

This is the maximum allowable cosine for angle $\alpha$ given the edge lengths $b$ and $c$. The cell lies exactly on the boundary of the reduced region for that angle.

\begin{center}
	\begin{tabular}{|c|p{4.5cm}|p{6cm}|}
		\hline
		\textbf{Zone} & \textbf{Typical Conditions} & \textbf{Classifier Implication} \\
		\hline
		Soft & $g_1 = g_2$, $g_4 = 0$ & Stable reduction; no override needed \\
		\hline
		Conditionally Hard & $|g_4| = g_2$, $g_1 \approx g_2$, $g_1 = g_2$, $g_4 = 0$ & Stable unless intersected; monitor adjacency and apply parametric triggers \\
		\hline
		Hard & $g_1 = g_2 = g_3$, $g_4 = g_5 = g_6 = 0$, $g_1 = g_5 = g_6$ & Intrinsically unstable; classifier override required \\
		\hline
	\end{tabular}
\end{center}

% This cell is invalid, apparently Polar::rand generates bad cells occasionally
%\subsection{33277}testCase(np, "case 33277", G6(6.486, 46.833, 30.720, -37.984, 27.277, -24.644), true);
%\subsubsection{33277}
%gross failure in both
%Output:    -1.272     0.088     0.357    -0.924    -0.805    -1.515
%Converged: YES
%Is Niggli reduced: NO
%Standard Niggli:        0.088     0.331    -0.479    -1.572     0.748     1.691

\subsection{10130}testCase(np, "case 10130", G6(78.145, 6.409, 1.485, -6.169, -4.536, 8.971), true);
\subsubsection[10130]
testCase(np, "acute", G6(0.825,     1.000,    65.850,     0.611,     0.051,     0.825), true);
converged at 100, but wandered a lot
  Iteration 99:     0.011     0.295    51.361     0.282     0.015     0.009
Output:     0.011     0.295    51.357    -0.273    -0.007    -0.009
Converged: YES
Is Niggli reduced: YES
Standard Niggli:        0.011     0.295    51.357    -0.273    -0.007    -0.009
Max difference: 6.39488e-14
? MATCH

Iteration 100:     0.011     0.295    51.357    -0.273    -0.007    -0.009
Checking all boundaries:
IsNiggli true
g     0.011     0.295    51.357    -0.273    -0.007    -0.009

Checking boundaries:
Checking boundaries: 8 (0.022)
Checking boundaries: D' (0) 7' (0) A' (0) 8' (0.0179722) B' (0.00943398) E' (0.00714143)     CONVERGED

Output:     0.011     0.295    51.357    -0.273    -0.007    -0.009
Converged: YES
Is Niggli reduced: YES
Standard Niggli:        0.011     0.295    51.357    -0.273    -0.007    -0.009

\subsection{acute}
   testCase(np, "acute", G6(0.825,     1.000,    65.850,     0.611,     0.051,     0.825), true);
\subsection{acute}
Output:     0.825     1.000    65.850    -0.560    -0.051    -0.825
Converged: YES
Is Niggli reduced: YES
Standard Niggli:        0.825     1.000    65.850     0.611     0.051     0.825
Max difference: 1.65
?

\subsection{testCase(np, "obtuse", G6(78.145, 6.409, 1.485, -6.169, -4.536, 8.971), true);}
\subsection[obtuse]
======================================================================
Test: obtuse
Input:     78.145     6.409     1.485     0.000     0.000     0.000

======================================================================
Test: Debug test
Input:    78.145     6.409     1.485     0.000     0.000     0.000

Initial Phase 1:
Phase1: Pattern 111, applying M2, M1
Phase1: Pattern 010, applying M2
Phase1 complete after 3 iterations
Phase1 complete after 2 iterations

Iteration 0:     1.485     6.409    78.145     0.000     0.000     0.000
Checking all boundaries:
IsNiggli true
g     1.485     6.409    78.145     0.000     0.000     0.000
3 (6.409) 4 (0) 5 (0)
Checking boundaries: 6/7 (6.409) 9/A (1.485) C/D (1.485)
Checking boundaries: 8 (6.409)
Checking boundaries: D' (0) 7' (0) A' (0) 8' (0) B' (0) E' (0)     CONVERGED

Output:     1.485     6.409    78.145     0.000     0.000     0.000
Converged: YES
Is Niggli reduced: YES
Standard Niggli:        1.485     6.409    78.145     0.000     0.000     0.000
Max difference: 0
? MATCH

\end{document}

### 2025-11-23 P3_uv5038/P3_uv5038.tex
just saving work

--- 2025-08-27 P3/P3.tex
@@
 	\cauthor[a]{Lawrence C.}{Andrews}{larry6640995@gmail.com}{}
-	\author[b]{Herbert J.}{Bernstein}
+	\author[a]{Herbert J.}{Bernstein}
 
 \aff[a]{Ronin Institute for Independent Scholarship 2.0, USA}
-\aff[b]{Ronin Institute for Independent Scholarship 2.0, USA}
 
@@
 	\maketitle                        % DO NOT DELETE THIS LINE
-
-	\begin{synopsis}
-
-	\end{synopsis}
 
@@
 	\begin{abstract}
-		The space \PIII{} is introduced, derived from unit cell axial lengths and interaxial angles. \PIII enables linearization of unit cell parameters using three polar coordinate bases.
+		The space \PIII{} is introduced, derived from unit cell axial lengths and interaxial angles. \PIII{} enables linearization of unit cell parameters using three polar coordinate bases.
 
@@
 
-	For the above reasons, other spaces are often used to describe lattices, often for specialized purposes; see Table  \ref{tbl_spaces}. Figure \ref{fig_spaces} shows relationships between
+[\underline{	Note: Boris Delaunay in his later publications chose to render
+	his surname as ``Delone'' (which is still pronounced the same as Delaunay).}]
+
+	For the above reasons, other spaces are often used to describe lattices, often for specialized purposes; see Table  \ref{tbl_spaces}. Figure \ref{fig_spaces} shows relationships among
 	several of these spaces. Although some of these spaces, such as \HVI{} and	\FIII{} are in almost daily use, they are not often referred to 
@@
 Lengths are unconstrained; angles are geometrically restricted.
-The sum of the angles must be less than $360^\circ$ and obey the triangle inequality \HVI{} is useful for conventional lattice descriptions, but limited in symmetry representation and metric-based classification.
+The angles must obey the spherical triangle inequalities (summing to less than 360 degrees, and the sum of any 2 must not exceed the third).
+\HVI{} is useful for conventional lattice descriptions, but limited in symmetry representation and metric-based classification.
 \\
@@
 	} &
-	Polar coordinate space defined in terms to unit cell axial
-	lengths and their opposing angle. It is a compact and smooth alternative to $\HVI$. Useful for comparing unit cells. (This work) \\
+	Polar coordinate space defined in terms of unit cell axial
+	lengths and their opposing angles (opposing, the angle between the other two unit cell axes). It is a compact and smooth alternative to $\HVI$. Useful for comparing unit cells. (This work) \\
 		\midrule
@@
 The scalar components are dot products between lattice vectors.
-\SVI{} supports Bravsi lattice classification.
+\SVI{} supports Bravais lattice classification.
 24 Bravais types were defined using the scalar components of \SVI{} by Delone (1975), later revised using \SVI{} \cite{andrews2023measuring}.
@@
 	} &
-	\CIII{} associates scalar pairs identified by Delone as “opposites.” Symmetry and Selling/Delone reduction simplify in this space \cite{andrews2023complex}. \\
+	\CIII{} associates scalar pairs identified by Delone as ``opposites.'' Symmetry and Selling/Delone reduction simplify in this space \cite{andrews2023complex}. \\
 		\midrule
@@
 \begin{figure}
-	\centering
-\begin{tikzpicture}[
-	node distance=1.2cm and 2.0cm,
-	coordspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
-		fill=gray!20, rounded corners, font=\small, thick},
-	metricspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
-		fill=blue!10, rounded corners, font=\small, thick},
-	derivedspace/.style={text width=3.2cm, align=center, rectangle, draw=black,
-		fill=green!10, rounded corners, font=\small, thick},
-	arrowlabel/.style={font=\scriptsize, midway, fill=white, inner sep=1pt}
-	]
-
-	% Row 1: Fundamental space
-	\node[coordspace] (F3) {\FIII{} \\ Base vectors};
-
-	% Row 2: Scalar & parameter spaces
-	\node[metricspace] (G6) [below left=of F3] {\GVI{} \\ Metric tensor};
-	\node[metricspace] (S6) [below=of F3] {\SVI{} \\ Selling/Delone};
-	\node[coordspace] (H6) [below right=of F3] {\HVI \\ Cell params};
-
-	% Row 3: Derived/reduced spaces
-	\node[derivedspace] (DC7u) [below=of G6] {DC7u \\ Dirichlet coords};
-	\node[derivedspace] (C3) [below=of S6] {\CIII{} \\ Reduced coords};
-	\node[derivedspace] (P3) [below=of H6] {\PIII{} \\ Polar coords};
-
-	% Arrows from F3 to row 2
-	\draw[->, thick] (F3) -- (G6);
-	\draw[->, thick] (F3) -- (S6);
-	\draw[->, thick] (F3) -- (H6);
-
-	% Arrows from row 2 to row 3
-	\draw[->, thick] (G6) -- (DC7u);
-	\draw[->, thick] (S6) -- (C3);
-	\draw[->, thick] (H6) -- (P3);
-
-\end{tikzpicture}
+\includegraphics[]{P3_fig1.png}
 \label{fig_spaces}
@@
 
+As described in Table \ref{tbl_spaces}, the Selling scalars are a vector:
+[$s_1, s_2, s_3, s_4, s_5, s_6$]. \citeasnoun{Delaunay1932} represents
+their relationships as the edge components of a tetrahedron as 
+shown in Figure \ref{tetrahedron}.
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=0.3\columnwidth]{tetrahedron}
+		\label{tetrahedron}
+		\caption{The relationships of the Selling scalars as
+		described by  \citeasnoun{Delaunay1932}, but labeled using the vector components of \SVI{} \cite{andrews2023measuring}. Visualized as a
+		3 dimensional tetrahedron, scalar $s_1$ is
+		opposite scalar $s_4$, etc. The number of each scalar is the 
+		position in the \SVI{} vector. See Table \ref{tbl_spaces}
+		for the definitions of the scalars in space \SVI{}. }
+
+	\end{center}
+\end{figure}
+
 Extending the observation of \citeasnoun{Delone1975}, the idea occurs
@@
 interpreted as crystallographic axial length/opposite angle pairs. The
-distance of a point has a length equal to the axial length, and the interaxisl
-angle of the two other bases is illustrated on the plots. See Section  \ref{graphical}. Basically, together the three plots of the three coordinates are
+distance of a point from the origin has a length equal to the axial length, and the interaxial
+angle between the 2 other bases (plotted as the angle from the x-axis) is illustrated on the plots. Basically, together the three plots of the three coordinates are
 simple projections of \HVI{}. See Section \ref{graphical}.
@@
 While \PIII{} is derived from the same foundational cell parameters as $\mathbf{H^6}$, $\mathbf{P^3}$ sidesteps issues of angular nonlinearity  that arise in the conventional cell representation.
-Compared with $\mathbf{C^3}$ \cite{andrews2023delone}, which derives coordinates from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic.
+Compared with $\mathbf{C^3}$ \cite{andrews2023complex}, which derives coordinates from $\mathbf{S^6}$, the construction of $\mathbf{P^3}$ is more geometric than algebraic.
 It provides a useful intermediary: more interpretable than metric tensors or other spaces, but less abstract than \SVI{} or \GVI{}.
@@
 allows simple-to-plot representations of collections of unit cells. 
-See Figures \ref{PlotPolar_raw}, \ref{PlotPolar_Niggli}, and  \ref{PlotPolar_Delone} where 100 randomly chosen triclinic cells are plotted
-(Figure \ref{PlotPolar_raw}) and the effects of unit cell reduction
+See Figures \ref{PlotPolar_raw}, \ref{PlotPolar_Niggli}, and  \ref{PlotPolar_Delone}, where 100 randomly chosen triclinic cells are plotted
+(Figure \ref{PlotPolar_raw}), and the effects of unit cell reduction
 are demonstrated (Figure \ref{PlotPolar_Niggli} and Figure \ref{PlotPolar_Delone} ).
@@
 		 primitive triclinic unit cells represented in \PIII{}. Each of the 3 components of \PIII{}
-		 is shown.}
+		 is shown. (Colors reflect the order of input of the unit cells.)}
 	\end{center}
@@
 	\label{PlotPolar_Niggli}
-	\caption{The same 100 random primitive triclinic unit cells shown in Figure \ref{PlotPolar_raw}, Niggli-reduced}
+	\caption{The same 100 randomly selected primitive triclinic unit cells shown in Figure \ref{PlotPolar_raw}, Niggli-reduced}.
 	\end{center}
@@
 	\caption{Scatter plots of 100 randomly selected
-		primitive triclinic unit cells represented in \PIII{} as shown in Figure \ref{PlotPolar_raw}, Selling (Delaunay)-reduced}
+		primitive triclinic unit cells represented in \PIII{} as shown in Figure \ref{PlotPolar_raw}, Selling /Delone-reduced}.
 	\end{center}
@@
 
-In three-space, measuring the Euclidean distance between two points is just computing the square root
+In 3-space, measuring the Euclidean distance between two points is just computing the square root
 of the sum of the squares of the differences in the three coordinates. Similarly in \PIII{},
-we compute distances from the sum of squared difference between the three
-polar coordinates. But polar coordinates have a Euclidean measure from
-their xy representation. So taking the six scalar values of two unit cells
+we compute distances from the square root of the sum of squared differences between the three
+polar coordinates. Polar coordinates have a Euclidean measure from
+their xy representation. So, taking the 6 scalar values of 2 unit cells
 presented in \PIII{}, we calculate a Euclidean distance by the ordinary
@@
 
-%; CmdDists, calculate distances
-%; Enter control variables and input vectors (type 'end' to finish):
-%P 10 10 10  90 90 90
-%p 10 10 10  90 90 91
-%p 10 10 10.1 90 90 90
-%end
-%10.000    10.000    10.000    90.000    90.000    90.000
-%14.142    30.113     8.479   147.466    56.504   144.893
-%
-%10.000    10.000    10.000    90.000    90.000    90.000
-%14.142    29.916     8.564   147.586    56.504   144.621
-%
-%<pre style="font-size: 105%; line-height: 60%; display:inline;" >;
-%; ordinals   P3-Euclid    P3Dist       NCDist    || ordinals     P3-Euclid    P3Dist       NCDist
-%; 1-2        0.175      19191.000      3.490         1-2         0.175      19191.000      3.490
-%; 2-3        0.202      19191.000      4.028         1-3         0.100      19191.000      2.010
-%
-%
-%;.......... input.cells ........................................ ordinals .....
-%P   10.000    10.000    10.000    90.000    90.000    90.000     1
-%P   10.000    10.000    10.000    90.000    90.000    91.000     2
-%P   10.000    10.000    10.100    90.000    90.000    90.000     3
-%;</pre>
-
-%; CmdDists, calculate distances
-%; Enter control variables and input vectors (type 'end' to finish):
-%p 10 10 10   90 90 90
-%p 10 10 10   90 90 91
-%p 10 10 11   90 90 90
-%end
-%<pre style="font-size: 105%; line-height: 60%; display:inline;" >;
-%; ordinals   P3-Euclid    P3Dist       NCDist    || ordinals     P3-Euclid    P3Dist       NCDist
-%; 1-2        3.490        0.203        3.490         1-2         0.175       26.959        3.490
-%; 2-3        21.288       0.872       21.288         1-3         1.000       26.896       21.000
-%
-%
-%;.......... input.cells ........................................ ordinals .....
-%P   10.000    10.000    10.000    90.000    90.000    90.000     1
-%P   10.000    10.000    10.000    90.000    90.000    91.000     2
-%P   10.000    10.000    11.000    90.000    90.000    90.000     3
-%;</pre>
-
-Let us compare two cells derived by modifying a primitive unit cell to form two slightly modified cells, one by changing $\gamma$, and one by changing $c$.
+Let us compare two cells derived by modifying a primitive unit cell to form two slightly modified cells, one by changing $\gamma$, and one by changing $\vec{c}$.
 \begin{enumerate}
@@
 \begin{itemize}
-	\item 1-2: 0.175
-	\item 1-3: 0.100
-	\item 2-3: 0.202
+	\item 1--2: 0.175
+	\item 1--3: 0.100
+	\item 2--3: 0.202
 \end{itemize}
 
-In this example, a 1.0° change in $\gamma$ yields a Euclidean distance comparable to a 0.1 Å change in the c-axis length.
-%
-%\section{Scalar Reduction and Canonical Presentation in Projected Vector Triplets}
-%
-%\subsection{1. Standard Presentation of Projected Vectors}
-%	
-%	Let a set of three projected vectors be given:
-%	
-%	
-%	\[
-%	(p_1, p_2, p_3), \quad p_i \in \mathbb{R}^2
-%	\]
-%	
-%	
-%	Each \( p_i \) is constructed from a lattice vector via magnitude and opposing angle:
-%	
-%	
-%	\[
-%	p_i = |\vec{v}_i| (\cos \theta_i, \sin \theta_i)
-%	\]
-%	
-%	
-%	
-%	The \textbf{Standard Presentation} algorithm assigns a canonical ordering and orientation to the triplet:
-%	
-%	\begin{enumerate}
-%		\item \textbf{Magnitude Sort:} Reorder vectors so that
-%		
-%		
-%		\[
-%		|p_1| \leq |p_2| \leq |p_3|
-%		\]
-%		
-%		
-%		using Euclidean norms \( |p_i| = \sqrt{x_i^2 + y_i^2} \). Ties are preserved up to a numerical tolerance \( \delta \).
-%		
-%\item \textbf{Directional Coherence:} If the \( x \)-components of the triplet are not all of consistent sign, flip the necessary vectors by negating only the \( x_i \) component:
-%
-%
-%\[
-%p_i \to (-x_i,\ y_i)
-%\]
-%
-%
-%This maps quadrant I vectors into quadrant II, preserving their placement in the upper half-plane and maintaining angular conventions typical in crystallographic analysis.
-%
-%
-%
-%\[
-%p_i \to (-x_i,\ y_i)
-%\]
-%
-%
-%to align the triplet into a directionally consistent state. If all \( x_i < 0 \), the configuration is already coherent and requires no change.
-%
-%		
-%		
-%		\[
-%		p_i \to (-x_i, y_i)
-%		\]
-%		
-%		
-%		for any two vectors, preserving chirality.
-%		
-%		\item \textbf{Return Ordered Triplet:} Output the permuted and flipped triplet with consistent magnitude order and cosine directionality.
-%	\end{enumerate}
-%	
-%	\vspace{1em}
-%	
-%	\subsection{2. Scalar Projection Interaction and Cost Function}
-%	
-%	For any pair of projected vectors \( p_i, p_j \in \mathbb{R}^2 \), define the scalar projection interaction (or directional affinity):
-%	
-%	
-%	\[
-%	\mu_{ij} = p_i \cdot p_j = x_i x_j + y_i y_j
-%	\]
-%	
-%	
-%	
-%	The \textbf{scalar coupling cost} is then defined over the triplet as:
-%	
-%	
-%	\[
-%	C' = \frac{|\mu_{12}| + |\mu_{13}| + |\mu_{23}|}{p_1 \cdot p_1 + p_2 \cdot p_2 + p_3 \cdot p_3}
-%	\]
-%	
-%	
-%	
-%	This cost measures the normalized sum of pairwise directional alignment.
-Lower values of \( C' \) indicate that the projected vectors are less directionally coupled — i.e., more orthogonal or dispersed.
-%	
-%	\vspace{1em}
-%	
-%	\subsection{3. Scalar Reduction Algorithm}
-%	
-%	The goal of scalar reduction is to transform the triplet \( (p_1, p_2, p_3) \) into an algebraically simpler form — one with lower scalar coupling cost \( C' \) — while preserving lattice identity (e.g., volume and reconstructability).
-%	
-%	The reduction proceeds iteratively:
-%	
-%	\begin{enumerate}
-%		\item \textbf{Initialization:}
-%		Apply standard presentation. Compute initial cost \( C'_0 \) and physical volume \( V_0 > 0 \).
-%		
-%		\item \textbf{Pairwise Projection Subtractions:}
-%		For each unordered pair \( (p_i, p_j),\ i \neq j \), compute the scalar coefficient:
-%		
-%		
-%		\[
-%		\lambda = \frac{p_i \cdot p_j}{p_j \cdot p_j}
-%		\]
-%		
-%		
-%		Then define the candidate vector:
-%		
-%		
-%		\[
-%		p_i' = p_i - \lambda p_j
-%		\]
-%		
-%		
-%		
-%		\item \textbf{Evaluate Candidate:}
-%		Apply standard presentation to updated triplet. Recompute cost \( C'_{\text{new}} \) and volume \( V_{\text{new}} \).
-%		
-%		\item \textbf{Acceptance Criteria:}
-%		Accept the transformation if:
-%		
-%		
-%		\[
-%		C'_{\text{new}} + \delta < C' \quad \text{and} \quad V_{\text{new}} > 0
-%		\]
-%		
-%		
-%		
-%		\item \textbf{Repeat:}
-%		Continue until no further accepted transformations lower the cost.
-%		
-%		\item \textbf{Output:}
-%		Return the reduced triplet \( (p_1, p_2, p_3)_{\text{red}} \) with minimal scalar cost.
-%	\end{enumerate}
-%	
-%	\vspace{1em}
-%	
-%	\subsection{4. Interpretation of Cost and Reduction}
-%	
-%	The term "reduction" refers to minimizing the scalar coupling among vectors — reducing directional alignment that obscures presentation clarity.
-The cost \( C' \) is a rigorously defined metric for this alignment.
-Transformations reduce \( C' \) while preserving:
-%	\begin{itemize}
-%		\item The vector magnitudes and their reconstructable relationships
-%		\item The unit cell's volume and crystallographic identity
-%		\item The triplet’s algebraic validity in scalar projection space
-%	\end{itemize}
-%	
-%	Reduction leads to a cleaner, more canonical representation of the original structure — just as Niggli reduction provides a minimal cell in metric tensor space, scalar reduction provides a minimal configuration in projected vector space.
-%	
-%
+In this example, a 1.0$^\circ$ change in $\gamma$ yields a Euclidean distance comparable to a 0.1~\AA{} change in the $c$-axis length.
 
@@
 
-	The space \PIII{} is introduced. Less abstract than space based
-	on dot products and metric tensors, it provides a simple to understand
+	The space \PIII{} is introduced. Less abstract than spaces based
+	on dot products and metric tensors, it provides a simple way to understand the
 	method for comparing unit cells and a convenient method of displaying 
@@
 gratefully acknowledged.
-	%	Our thanks to Jean Jakoncic and Alexei Soares for 
-	%	helpful conversations and access to data and facilities at 
-	%	Brookhaven National Laboratory.
-	%	
+
 	\ack{{\bf Funding information}}
@@
 	Funding for this research was provided in part by:  
-	US Department of Energy Offices of Biological and 
+	U.S. Department of Energy Offices of Biological and 
 	Environmental Research and of Basic Energy Sciences 
@@
 	\PIII{} is a linearized description of unit cell parameters,
-	derived from usual lengths and angles of the unit cell. It 
+	created from the usual lengths and angles of the unit cell. It 
 	provides a useful metric for comparing unit cells and a

### 2025-11-23 P3_uv5038/Presentation Isometry in Projected Triplet Reduction.tex
just saving work

\subsection{Presentation Isometry in Projected Triplet Reduction}

In crystallographic applications, it is conventionally preferred that projected vectors occupy the upper half-plane, such that computed angles via \texttt{atan2} lie within $[0^\circ, 180^\circ]$. However, scalar transformations during reduction may produce vectors in the lower half-plane ($y < 0$), leading to angle values in $(-180^\circ, 0^\circ]$ which, while geometrically valid, can be undesirable in presentation.

To enforce a consistent representation, we define a \emph{presentation isometry} that maps each vector $v_i = (x_i, y_i)$ with $y_i < 0$ to its negation $-v_i = (-x_i, -y_i)$, rotating it $180^\circ$ while preserving its magnitude.

The cost function used during reduction is defined as:

\[
\text{Cost}(P_3) = \frac{|\vec{v}_1 \cdot \vec{v}_2| + |\vec{v}_1 \cdot \vec{v}_3| + |\vec{v}_2 \cdot \vec{v}_3|}{|\vec{v}_1|^2 + |\vec{v}_2|^2 + |\vec{v}_3|^2}
\]

Since dot products are wrapped in absolute value, and magnitudes remain unchanged under negation, it follows that flipping vectors into the upper half-plane does not alter the computed cost. Similarly, the triplet's volume, derived from the area spanned by the vectors, remains invariant.

Thus, the quadrant correction process is mathematically sound: it constitutes a direction-preserving transformation that maintains scalar relationships, magnitudes, and reduction metrics, while aligning the final geometry with crystallographic standards.

\textbf{Conclusion:} The upper-half-plane enforcement step introduces no distortion in the scalar reduction process and may be safely applied as a post-processing operation.

### 2025-11-23 SELLA/Sella_OLD.tex
just saving work

--- 2025-07-02 SELLA/Sella.tex
@@
+%Table 3. Illustration of the power of the 'matrix approach to symmetry'
+%The initial cell was originally tested by Himes & Mighell (1987a). BLAF confirmed the reported cubic F-centred cell with a = 15.728 A
+%and, in addition, calculated the estimated error in the cell edge and produced the admitted subgroup sequence. The
+%output for cells 2-5 has been shortened and rearranged. The execution time on the PDI 1/44 computer was 9.8 s.
+%INITIAL UNIT CELL DATA
+%. . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . .
+%. . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . . .
+%Real parameters and errors
+%a = 19.2600(9) alpha = 5.7696(134)
+%b = 63.8825(62) beta = 19.4709(129)
+%c = 27.2394(27) gamma = 17.2952(126)
+%
+%SEL_2025-09-18.11_13_13_2314_0.svg
+%SEL_2025-09-18.11_13_13_2314_0_fit_plots.svg
+%
+
 %------------------------------------------------------------------------------
@@
 \usepackage{booktabs}
+\usepackage{makecell}
 
@@
 
+\begin{table}[htbp]
+	\centering
+	\renewcommand{\arraystretch}{1.3}
+	\begin{tabular}{@{}p{4cm} p{5.5cm} p{5.5cm}@{}}
+		\toprule
+		\textbf{Property} & \textbf{Phase II (Room Temp)} & \textbf{Phase I (Low Temp)} \\
+		\midrule
+		Crystal System & Triclinic & Monoclinic \\
+		Space Group & \textit{P$\bar{1}$} & \textit{C2/c} or \textit{Cc} \\
+		Z (Formula Units) & 2 & 8 \\
+		Unit Cell Parameters &
+		\makecell[l]{
+			$a = 7.485$ Å \\
+			$b = 7.515$ Å \\
+			$c = 7.520$ Å \\
+			$\alpha = 89.98^\circ$ \\
+			$\beta = 90.01^\circ$ \\
+			$\gamma = 90.03^\circ$ \\
+			$V \approx 423$ Å$^3$
+		} &
+		\makecell[l]{
+			$a = 10.60$ Å \\
+			$b = 18.30$ Å \\
+			$c = 7.52$ Å \\
+			$\beta = 92.8^\circ$ \\
+			$V \approx 1457$ Å$^3$
+		} \\
+		Metric Symmetry & Nearly Cubic (pseudo-symmetry) & Monoclinic \\
+		Transformation Proposed &
+		\makecell[l]{
+			Non-unimodular matrix: \\
+			$\begin{bmatrix}
+				2 & 1 & 0 \\
+				0 & 1 & 0 \\
+				0 & 0 & 1
+			\end{bmatrix}$ \\
+			Yields monoclinic cell: \\
+			$a \approx 15.73$ Å, $b \approx 9.07$ Å, \\
+			$c \approx 6.44$ Å, $\beta \approx 92.8^\circ$
+		} & — \\
+		Key Insight &
+		\makecell[l]{
+			Pseudo-symmetry masks true triclinic \\
+			structure; reinterpretation improves \\
+			crystallographic clarity
+		} &
+		\makecell[l]{
+			Cation disorder and complex packing \\
+			in monoclinic symmetry
+		} \\
+		\bottomrule
+	\end{tabular}
+	\caption{Summary of structural phases for [(CH$_3$)$_4$N]CuCl$_3$ from Marsh (1988).}
+\end{table}
+
 	\section{What could go wrong?}
@@
 	\begin{table}
-		\caption{The 21 non-triclinic Delone types with the count of representations.
+		\caption{The 21 non-triclinic Delone types with the count of permutations.
 			The asterisks indicate non-crystallographic types.}
@@
 	\begin{table}
-			\caption{Projectors for the 24 Delone types}
+		\caption{Properly normalized projectors for the 24 Delone types}
 		\label{table:DeloneTypes}
-			\begin{tabular}{lcl}
+		\begin{tabular}{lclc}
 			\toprule
-				C1  & (cI) & [1 1 1 1 1 1/  1 1 1 1 1 1/  1 1 1 1 1 1/  1 1 1 1 1 1/  1 1 1 1 1 1/  1 1 1 1 1 1] \\ 
-				C3  & (cF) & [1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 0/  1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 0] \\ 
-				C5  & (cP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 1 1/  0 0 0 1 1 1/  0 0 0 1 1 1] \\ \\
+			Type & Lattice & Projector & Characteristic \\
+			\midrule
+			C1  & (cI) & [1/6 1/6 1/6 1/6 1/6 1/6/ 1/6 1/6 1/6 1/6 1/6 1/6/ 1/6 1/6 1/6 1/6 1/6 1/6/ 1/6 1/6 1/6 1/6 1/6 1/6/ 1/6 1/6 1/6 1/6 1/6 1/6/ 1/6 1/6 1/6 1/6 1/6 1/6] & (rrr rrr) \\ 
+			C3  & (cF) & [1/4 1/4 0 1/4 1/4 0/ 1/4 1/4 0 1/4 1/4 0/ 0 0 0 0 0 0/ 1/4 1/4 0 1/4 1/4 0/ 1/4 1/4 0 1/4 1/4 0/ 0 0 0 0 0 0] & (rr0 rr0) \\ 
+			C5  & (cP) & [0 0 0 0 0 0/ 0 0 0 0 0 0/ 0 0 0 0 0 0/ 0 0 0 1/3 1/3 1/3/ 0 0 0 1/3 1/3 1/3/ 0 0 0 1/3 1/3 1/3] & (000 rrr) \\ \\
 
-				T1  & (tI) & [1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 1 0 0 1/  1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 1 0 0 1] \\ 
-				T2  & (tI) & [1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 0/  1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 1] \\ 
-				T5  & (tP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 1 0/  0 0 0 1 1 0/  0 0 0 0 0 1] \\ \\
+			T1  & (tI) & [1/4 1/4 0 1/4 1/4 0/ 1/4 1/4 0 1/4 1/4 0/ 0 0 1/2 0 0 1/2/ 1/4 1/4 0 1/4 1/4 0/ 1/4 1/4 0 1/4 1/4 0/ 0 0 1/2 0 0 1/2] & (rrr rrs) \\ 
+			T2  & (tI) & [1/4 1/4 0 1/4 1/4 0/ 1/4 1/4 0 1/4 1/4 0/ 0 0 0 0 0 0/ 1/4 1/4 0 1/4 1/4 0/ 1/4 1/4 0 1/4 1/4 0/ 0 0 0 0 0 1] & (rr0 rrs) \\ 
+			T5  & (tP) & [0 0 0 0 0 0/ 0 0 0 0 0 0/ 0 0 0 0 0 0/ 0 0 0 1/2 1/2 0/ 0 0 0 1/2 1/2 0/ 0 0 0 0 0 1] & (000 rrs) \\ \\
 
-				H1  & (hR) & [1 1 1 0 0 0/  1 1 1 0 0 0/  1 1 1 0 0 0/  0 0 0 1 1 1/  0 0 0 1 1 1/  0 0 0 1 1 1] \\ 
-				H3  & (hR) & [1 1 0 0 1 0/  1 1 0 0 1 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  1 1 0 0 1 0/  0 0 0 0 0 0] \\
-				H4  & (hP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 1 1 1 0/  0 0 1 1 1 0/  0 0 1 1 1 0/  0 0 0 0 0 1] \\ \\
+			H1  & (hR) & [1/3 1/3 1/3 0 0 0/ 1/3 1/3 1/3 0 0 0/ 1/3 1/3 1/3 0 0 0/ 0 0 0 1/3 1/3 1/3/ 0 0 0 1/3 1/3 1/3/ 0 0 0 1/3 1/3 1/3] & (rrr sss) \\ 
+			H3  & (hR) & [1/4 1/4 0 0 1/4 0/ 1/4 1/4 0 0 1/4 0/ 0 0 0 0 0 0/ 0 0 0 1 0 0/ 1/4 1/4 0 0 1/4 0/ 0 0 0 0 0 0] & (rr0 sr0) \\
+			H4  & (hP) & [0 0 0 0 0 0/ 0 0 0 0 0 0/ 0 0 1/3 1/3 1/3 0/ 0 0 1/3 1/3 1/3 0/ 0 0 1/3 1/3 1/3 0/ 0 0 0 0 0 1] & (00r rrs) \\ \\
 
-				O1A & (oF) & [1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 1 0 0 0/  1 1 0 1 1 0/  1 1 0 1 1 0/  0 0 0 0 0 1] \\ 
-				O1B & (oI) & [1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 1 0 0 1/  1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 1 0 0 1] \\ 
-				O2  & (oI) & [1 0 0 0 1 0/  0 1 0 1 0 0/  0 0 0 0 0 0/  0 1 0 1 0 0/  1 0 0 0 1 0/  0 0 0 0 0 1] \\ 
-				O3  & (oI) & [1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0/  1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0] \\ 
-				O4  & (oS) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 1 0 0 0/  0 0 0 1 1 0/  0 0 0 1 1 0/  0 0 0 0 0 1] \\ 
-				O5  & (oP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ \\
+			O1A & (oF) & [1/4 1/4 0 1/4 1/4 0/ 1/4 1/4 0 1/4 1/4 0/ 0 0 1 0 0 0/ 1/4 1/4 0 1/4 1/4 0/ 1/4 1/4 0 1/4 1/4 0/ 0 0 0 0 0 1] & (rrs rrt) \\ 
+			O1B & (oI) & [1/2 0 0 1/2 0 0/ 0 1/2 0 0 1/2 0/ 0 0 1/2 0 0 1/2/ 1/2 0 0 1/2 0 0/ 0 1/2 0 0 1/2 0/ 0 0 1/2 0 0 1/2] & (rst rst) \\ 
+			O2  & (oI) & [1/2 0 0 0 1/2 0/ 0 1/2 0 1/2 0 0/ 0 0 0 0 0 0/ 0 1/2 0 1/2 0 0/ 1/2 0 0 0 1/2 0/ 0 0 0 0 0 1] & (rs0 srt) \\ 
+			O3  & (oI) & [1/2 0 0 1/2 0 0/ 0 1/2 0 0 1/2 0/ 0 0 0 0 0 0/ 1/2 0 0 1/2 0 0/ 0 1/2 0 0 1/2 0/ 0 0 0 0 0 0] & (rs0 rs0) \\ 
+			O4  & (oS) & [0 0 0 0 0 0/ 0 0 0 0 0 0/ 0 0 1 0 0 0/ 0 0 0 1/2 1/2 0/ 0 0 0 1/2 1/2 0/ 0 0 0 0 0 1] & (00r sst) \\ 
+			O5  & (oP) & [0 0 0 0 0 0/ 0 0 0 0 0 0/ 0 0 0 0 0 0/ 0 0 0 1 0 0/ 0 0 0 0 1 0/ 0 0 0 0 0 1] & (000 rst) \\ \\
 
-				M1A & (mC) & [1 1 0 0 0 0/  1 1 0 0 0 0/  0 0 1 0 0 0/  0 0 0 1 1 0/  0 0 0 1 1 0/  0 0 0 0 0 1] \\ 
-				M1B & (mC) & [1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 1 0 0 0/  1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 1] \\ 
-				M2A & (mC) & [1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0/  1 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 1] \\ 
-				M2B & (mC) & [1 0 0 0 0 0/  0 1 0 1 0 0/  0 0 0 0 0 0/  0 1 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ 
-				M3  & (mC) & [1 0 0 0 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  0 1 0 0 1 0/  0 0 0 0 0 0] \\ 
-				M4  & (mP) & [0 0 0 0 0 0/  0 0 0 0 0 0/  0 0 1 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ \\
+			M1A & (mC) & [1/2 1/2 0 0 0 0/ 1/2 1/2 0 0 0 0/ 0 0 1 0 0 0/ 0 0 0 1/2 1/2 0/ 0 0 0 1/2 1/2 0/ 0 0 0 0 0 1] & (rrs ttu) \\ 
+			M1B & (mC) & [1/2 0 0 1/2 0 0/ 0 1/2 0 0 1/2 0/ 0 0 1 0 0 0/ 1/2 0 0 1/2 0 0/ 0 1/2 0 0 1/2 0/ 0 0 0 0 0 1] & (rst rsu) \\ 
+			M2A & (mC) & [1 0 0 0 0 0/ 0 1/2 0 1/2 0 0/ 0 0 0 0 0 0/ 0 1/2 0 1/2 0 0/ 0 0 0 0 1 0/ 0 0 0 0 0 1] & (rs0 stu) \\ 
+			M2B & (mC) & [1/2 0 0 1/2 0 0/ 0 1/2 0 0 1/2 0/ 0 0 0 0 0 0/ 1/2 0 0 1/2 0 0/ 0 1/2 0 0 1/2 0/ 0 0 0 0 0 1] & (rs0 rst) \\ 
+			M3  & (mC) & [1 0 0 0 0 0/ 0 1/2 0 0 1/2 0/ 0 0 0 0 0 0/ 0 0 0 1 0 0/ 0 1/2 0 0 1/2 0/ 0 0 0 0 0 0] & (rs0 ts0) \\ 
+			M4  & (mP) & [0 0 0 0 0 0/ 0 0 0 0 0 0/ 0 0 1 0 0 0/ 0 0 0 1 0 0/ 0 0 0 0 1 0/ 0 0 0 0 0 1] & (00r stu) \\ \\
 
-				A1  & (aP) & [1 0 0 0 0 0/  0 1 0 0 0 0/  0 0 1 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ 
-				A2  & (aP) & [1 0 0 0 0 0/  0 1 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 1] \\ 
-				A3  & (aP) & [1 0 0 0 0 0/  0 1 0 0 0 0/  0 0 0 0 0 0/  0 0 0 1 0 0/  0 0 0 0 1 0/  0 0 0 0 0 0] \\ \\				
+			A1  & (aP) & [1 0 0 0 0 0/ 0 1 0 0 0 0/ 0 0 1 0 0 0/ 0 0 0 1 0 0/ 0 0 0 0 1 0/ 0 0 0 0 0 1] & (rst uvw) \\ 
+			A2  & (aP) & [1 0 0 0 0 0/ 0 1 0 0 0 0/ 0 0 0 0 0 0/ 0 0 0 1 0 0/ 0 0 0 0 1 0/ 0 0 0 0 0 1] & (rs0 tuv) \\ 
+			A3  & (aP) & [1 0 0 0 0 0/ 0 1 0 0 0 0/ 0 0 0 0 0 0/ 0 0 0 1 0 0/ 0 0 0 0 1 0/ 0 0 0 0 0 0] & (rs0 tu0) \\ \\				
 			\bottomrule
@@
 
+	\bibliography{Reduced}
+
+	\bibliographystyle{iucr}
+
 \end{document}                    % DO NOT DELETE THIS LINE

### 2025-11-23 hausdorff/haus_claude.tex
just saving work

\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{cite}

\title{Hausdorff Dimension of Crystallographic Lattices}
\author{}
\date{}

\begin{document}

	\maketitle

	\section{Introduction to Hausdorff Dimension}

	The Hausdorff dimension provides a rigorous mathematical framework for characterizing the geometric complexity of point sets in metric spaces, extending beyond the integer dimensions of classical Euclidean geometry. For a set $S$ in $\mathbb{R}^n$, the Hausdorff dimension $d_{\text{H}}$ quantifies how the ``mass'' (or number of points) scales with the radius of measurement. Unlike topological dimension, which is always an integer, Hausdorff dimension can take non-integer values, making it particularly useful for characterizing fractal-like distributions and irregular point configurations.

	For crystallographic applications, the Hausdorff dimension characterizes how unit cell parameters populate the abstract lattice representation spaces (G6, S6, etc.). While individual crystal structures occupy discrete points, collections of related structures---such as those within a Bravais lattice type, space group, or structural family---form point clouds whose dimensional properties reflect the underlying constraints and symmetries of the crystallographic system.

	\section{Theoretical Foundation}

	The formal definition of Hausdorff dimension relies on $\delta$-covers of a set $S$. For radius $\delta > 0$, a $\delta$-cover is a collection of sets $\{U_i\}$ with $\text{diam}(U_i) \leq \delta$ that covers $S$. The $d$-dimensional Hausdorff measure is defined as:
	%
	\begin{equation}
		\mathcal{H}^d(S) = \lim_{\delta \to 0} \inf \left\{ \sum_i (\text{diam}(U_i))^d : \{U_i\} \text{ is a } \delta\text{-cover of } S \right\}
	\end{equation}
	%
	The Hausdorff dimension is then:
	%
	\begin{equation}
		d_{\text{H}} = \inf \{ d : \mathcal{H}^d(S) = 0 \} = \sup \{ d : \mathcal{H}^d(S) = \infty \}
	\end{equation}

	For practical computation with finite point sets, we employ the box-counting dimension, which converges to the Hausdorff dimension for many well-behaved sets. The box-counting dimension exploits the power-law scaling relationship:
	%
	\begin{equation}
		N(r) \propto r^{-d}
	\end{equation}
	%
	where $N(r)$ represents the number of points (or boxes) needed to cover the set at scale $r$.

	\section{Computational Method}

	Our algorithm estimates the local Hausdorff dimension through an adaptive spherical box-counting method with Monte Carlo sampling. The approach addresses several practical challenges inherent in analyzing finite crystallographic datasets: heterogeneous point distributions, varying local densities, and the need for statistically robust estimates.

	\subsection{Adaptive Radius Selection}

	The algorithm begins by determining an appropriate measurement scale. Given a point cloud of $N$ structures in a representation space with estimated diameter $D$, we compute the mean nearest-neighbor spacing $\mu$ and point density $\rho = N/D$. An initial target radius $r_{\text{target}}$ is selected to capture approximately 4096 points in one dimension:
	%
	\begin{equation}
		r_{\text{target}} = \frac{4096}{\rho}
	\end{equation}
	%
	This radius is constrained to the range $[10\mu, D/1.1]$ to ensure statistically meaningful measurements while avoiding boundary effects.

	To refine $r_{\text{target}}$ for the actual data distribution, we iteratively shrink the radius by factors of 1.1 until identifying a sphere containing 256--266 points around a randomly selected probe structure. This adaptive procedure ensures sufficient local sampling regardless of the global distribution characteristics.

	\subsection{Monte Carlo Dimension Estimation}

	The dimension is estimated through repeated measurements at randomly selected probe points. For each trial, we:
	%
	\begin{enumerate}
		\item Select a random structure $p$ from the dataset as probe center
		\item Count points within radius $r$: $N_{\text{large}} = |\{x : d(x, p) \leq r\}|$
		\item Count points within radius $r/1.1$: $N_{\text{small}} = |\{x : d(x, p) \leq r/1.1\}|$
		\item Compute local dimension estimate: 
		%
		\begin{equation}
			d_{\text{est}} = \frac{\ln(N_{\text{large}}/N_{\text{small}})}{\ln(1.1)}
		\end{equation}
	\end{enumerate}

	The number of trials is set to $\sqrt{N}$ with a minimum of 10, providing a balance between computational efficiency and statistical reliability. Trials where $N_{\text{small}} \geq N_{\text{large}}$ are rejected as non-physical (indicating insufficient local point density).

	\subsection{Statistical Analysis and Convergence}

	The algorithm accumulates the mean dimension $\langle d \rangle$ and variance $\sigma^2$ across successful trials. Early termination occurs when at least half the trials have succeeded and the standard error satisfies:
	%
	\begin{equation}
		\sigma^2 \leq \epsilon^2
	\end{equation}
	%
	where $\epsilon$ is the requested precision. The final estimate is reported as $d_{\text{H}} \pm \sigma$, with physically meaningless results ($d_{\text{H}} + 3\sigma < 0$) rejected.

	\section{Implementation Considerations}

	The method requires minimum 32 structures and non-trivial diameter ($> \epsilon_{\text{machine}}$) for meaningful estimation. For crystallographic datasets, the distance metric $d(x, y)$ is typically the NCDist (Niggli cell distance) or CS6Dist (Selling-reduced cell distance), ensuring the dimension measurement respects crystallographic equivalence and reduction conventions.

	Cached results are returned when previous estimates meet the requested precision, avoiding redundant computation during iterative analyses. The algorithm's robustness derives from its adaptive scaling, Monte Carlo sampling strategy, and explicit validation of statistical convergence.

	\section{Applications to Crystallographic Analysis}

	The Hausdorff dimension provides quantitative insight into the geometric organization of crystal structure families. Low-dimensional manifolds ($d_{\text{H}} \approx 1$--2) within 6-dimensional lattice spaces indicate strong geometric constraints, such as those arising from coordination chemistry or packing efficiency. Higher dimensions suggest greater structural diversity or weaker systematic relationships. Comparison of Hausdorff dimensions across different structural classifications (space groups, composition families, structure types) reveals the relative geometric complexity and constraint hierarchy within crystallographic databases.

\end{document}

### 2025-11-23 hausdorff/haus_copilot.tex
just saving work

\section{Hausdorff Dimension and Its Estimation in Crystallographic Unit Cells}

The \emph{Hausdorff dimension} provides a rigorous mathematical measure of the effective dimensionality of a set, extending the familiar notion of integer dimensions to fractal and irregular structures. For a point cloud, such as a collection of crystallographic unit cells represented in a metric space, the Hausdorff dimension characterizes how the number of points within a ball of radius $r$ scales as $r$ increases. Formally, if $N(r)$ denotes the number of points within distance $r$ of a given probe point, then

\[
N(r) \sim r^d,
\]

where $d$ is the Hausdorff dimension. Unlike the topological dimension, which is always an integer, the Hausdorff dimension may take non‑integer values, reflecting the scaling properties of the distribution.

\subsection{Method of Estimation}

To estimate the Hausdorff dimension of a finite set of unit cells, we employ a \emph{local scaling analysis} around randomly chosen probe points:

\begin{enumerate}
	\item \textbf{Preconditions.} The dataset must contain a sufficient number of unit cells (at least several dozen) and a non‑trivial diameter to ensure meaningful statistics.
	\item \textbf{Radius Selection.} A target radius is chosen based on the overall point density, adjusted to avoid extremes. The radius is scaled so that spheres around probe points contain a reasonable number of neighbors, typically several hundred.
	\item \textbf{Neighbor Counts.} For each probe point, we count the number of unit cells within two concentric spheres:
	\begin{itemize}
		\item A larger sphere of radius $R$
		\item A smaller sphere of radius $R/\lambda$, with $\lambda$ slightly greater than 1 (e.g.\ $\lambda = 1.1$)
	\end{itemize}
	\item \textbf{Local Dimension Estimate.} The ratio of populations in the two spheres provides a local estimate of dimension:

	\[
	d = \frac{\log\!\left(\tfrac{N(R)}{N(R/\lambda)}\right)}{\log(\lambda)}.
	\]

	\item \textbf{Averaging and Error.} Repeating this procedure across many randomly selected probe points yields a distribution of local dimension estimates. The mean provides the global dimension estimate, while the variance supplies an uncertainty measure.
\end{enumerate}

\subsection{Application to Unit Cells}

When applied to sets of crystallographic unit cells expressed in reduced metric spaces (e.g.\ $S_6$ or $G_6$), this method quantifies the effective dimensionality of the ensemble. A dimension close to 6 indicates a nearly uniform distribution in the six‑dimensional parameter space, whereas lower values suggest clustering along lower‑dimensional manifolds, reflecting structural correlations or constraints among the cells.
